### 2020-12-13 manuscript.Rmd
Initial commit

---
title: "manuscript"
author: "Daniel Baker"
date: "13/12/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-12-20 NBDTtemplate.tex
Intermediate update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% A template for NBDT article submissions.
% Based on a template developed by Overleaf for Wiley journal submissions.
% https://www.overleaf.com/latex/templates/template-for-submissions-to-statistics-in-medicine/pmwddvhjgmwj
% relative to that template we added an icon, changed the text, and removed wiley specific functionality
%
% Please note that whilst this template provides a 
% preview of the typeset manuscript for submission, it 
% will not necessarily be the final publication layout.
%
% Usage notes:
% The "blind" option will make anonymous all author, affiliation, correspondence and funding information.
% Use "num-refs" option for numerical citation and references style.
% Use "alpha-refs" option for author-year citation and references style.

\documentclass[num-refs]{nbdt-article}

% Add additional packages here if required
\usepackage{siunitx}

% Update article type if known
\papertype{Original Article}
% Include section in journal if known, otherwise delete
\paperfield{Journal Section}

\title{This is my title}

% List abbreviations here, if any. Please note that it is preferred that abbreviations be defined at the first instance they appear in the text, rather than creating an abbreviations list.
\abbrevs{ABC, a black cat; DEF, doesn't ever fret; GHI, goes home immediately.}

% Include full author names and degrees, when required by the journal.
% Use the \authfn to add symbols for additional footnotes and present addresses, if any. Usually start with 1 for notes about author contributions; then continuing with 2 etc if any author has a different present address.
\author[1\authfn{1}]{Author One PhD}
\author[2\authfn{1}]{Author A.~Two MD}
\author[2\authfn{2}]{Author Three PhD}
\author[2]{Author B.~Four}

\contrib[\authfn{1}]{Equally contributing authors.}

% Include full affiliation details for all authors
\affil[1]{Department, Institution, City, State or Province, Postal Code, Country}
\affil[2]{Department, Institution, City, State or Province, Postal Code, Country}

\corraddress{Author One PhD, Department, Institution, City, State or Province, Postal Code, Country}
\corremail{correspondingauthor@email.com}

\presentadd[\authfn{2}]{Department, Institution, City, State or Province, Postal Code, Country}

\fundinginfo{Funder One, Funder One Department, Grant/Award Number: 123456, 123457 and 123458; Funder Two, Funder Two Department, Grant/Award Number: 123459}

% Include the name of the author that should appear in the running header
\runningauthor{Author One et al.}

\begin{document}

\maketitle

\begin{abstract}
This is a generic template designed for use by NBDT, which includes several options for customization. Please consult the author guidelines for the journal to which you are submitting in order to confirm that your manuscript will comply with the journal's requirements. Please replace this text with your abstract. On top of this style we also recommend using standard writing style for the contents of the paper and recommend the ten simple rules paper: 
\href{http://journals.plos.org/ploscompbiol/article/metrics?id=10.1371/journal.pcbi.1005619}{Standard Structure}

% Please include a maximum of seven keywords
\keywords{keyword 1, \emph{keyword 2}, keyword 3, keyword 4, keyword 5, keyword 6, keyword 7}
\end{abstract}

\section{First Level Heading}
Please lay out your article using the section headings and example objects below, and remember to delete all help text prior to submitting your article to the journal.

\begin{figure}[bt]
\centering
\includegraphics[width=6cm]{example-image-rectangle}
\caption{Although we encourage authors to send us the highest-quality figures possible, for peer-review purposes we are can accept a wide variety of formats, sizes, and resolutions. Legends should be concise but comprehensive – the figure and its legend must be understandable without reference to the text. Include definitions of any symbols used and define/explain all abbreviations and units of measurement.}
\end{figure}

\subsection{Second Level Heading}
If data, scripts or other artifacts used to generate the analyses presented in the article are available via a publicly available data repository, please include a reference to the location of the material within the article.

% Equations should be inserted using standard LaTeX equation and eqnarray environments, not as graphics, and should be set in the main text
This is an equation, numbered
\begin{equation}
\int_0^{+\infty}e^{-x^2}dx=\frac{\sqrt{\pi}}{2}
\end{equation}
And one that is not numbered
\begin{equation*}
e^{i\pi}=-1
\end{equation*}

\subsection{Adding Citations and a References List}

Please use a \verb|.bib| file to store your references. When using Overleaf to prepare your manuscript, you can upload a \verb|.bib| file or import your Mendeley, CiteULike or Zotero library directly as a \verb|.bib| file\footnote{see \url{https://www.overleaf.com/blog/184}}. You can then cite entries from it, like this: \cite{lees2010theoretical}. Just remember to specify a bibliography style, as well as the filename of the \verb|.bib|.

You can find a video tutorial here to learn more about BibTeX: \url{https://www.overleaf.com/help/97-how-to-include-a-bibliography-using-bibtex}.

This template provides two options for the citation and reference list style: 
\begin{description}
\item[Numerical style] Use \verb|\documentclass[...,num-refs]{nbdt-article}|
\item[Author-year style] Use \verb|\documentclass[...,alpha-refs]{nbdt-article}|
\end{description}

\subsubsection{Third Level Heading}
Supporting information will be included with the published article. For submission any supporting information should be supplied as separate files but referred to in the text.

Appendices will be published after the references. For submission they should be supplied as separate files but referred to in the text.

\paragraph{Fourth Level Heading}
% Here are examples of quotes and epigraphs.
\begin{quote}
The significant problems we have cannot be solved at the same level of thinking with which we created them.\endnote{Albert Einstein said this.}
\end{quote}

\begin{epigraph}{Albert Einstein}
Anyone who has never made a mistake has never tried anything new.
\end{epigraph}

\subparagraph{Fifth level heading}
Measurements should be given in SI or SI-derived units.
Chemical substances should be referred to by the generic name only. Trade names should not be used. Drugs should be referred to by their generic names. If proprietary drugs have been used in the study, refer to these by their generic name, mentioning the proprietary name, and the name and location of the manufacturer, in parentheses.

\begin{table}[bt]
\caption{This is a table. Tables should be self-contained and complement, but not duplicate, information contained in the text. They should be not be provided as images. Legends should be concise but comprehensive – the table, legend and footnotes must be understandable without reference to the text. All abbreviations must be defined in footnotes.}
\begin{threeparttable}
\begin{tabular}{lccrr}
\headrow
\thead{Variables} & \thead{JKL ($\boldsymbol{n=30}$)} & \thead{Control ($\boldsymbol{n=40}$)} & \thead{MN} & \thead{$\boldsymbol t$ (68)}\\
Age at testing & 38 & 58 & 504.48 & 58 ms\\
Age at testing & 38 & 58 & 504.48 & 58 ms\\
Age at testing & 38 & 58 & 504.48 & 58 ms\\
Age at testing & 38 & 58 & 504.48 & 58 ms\\
\hiderowcolors
stop alternating row colors from here onwards\\
Age at testing & 38 & 58 & 504.48 & 58 ms\\
Age at testing & 38 & 58 & 504.48 & 58 ms\\
\hline  % Please only put a hline at the end of the table
\end{tabular}

\begin{tablenotes}
\item JKL, just keep laughing; MN, merry noise.
\end{tablenotes}
\end{threeparttable}
\end{table}

\section*{acknowledgements}
Acknowledgements should include contributions from anyone who does not meet the criteria for authorship (for example, to recognize contributions from people who provided technical help, collation of data, writing assistance, acquisition of funding, or a department chairperson who provided general support), as well as any funding or other support information.

\section*{conflict of interest}
You may be asked to provide a conflict of interest statement during the submission process. Please check the journal's author guidelines for details on what to include in this section. Please ensure you liaise with all co-authors to confirm agreement with the final statement.

\printendnotes

% Submissions are not required to reflect the precise reference formatting of the journal (use of italics, bold etc.), however it is important that all key elements of each reference are included.
\bibliography{sample}

\begin{biography}[example-image-1x1]{A.~One}
Please check with the journal's author guidelines whether author biographies are required. They are usually only included for review-type articles, and typically require photos and brief biographies (up to 75 words) for each author.
\bigskip
\bigskip
\end{biography}

\graphicalabstract{example-image-1x1}{Please check the journal's author guildines for whether a graphical abstract, key points, new findings, or other items are required for display in the Table of Contents.}

\end{document}

### 2020-12-20 manuscript.Rmd
Intermediate update

@@
 ---
-title: "manuscript"
-author: "Daniel Baker"
+title: "Statistical analysis of periodic data in neuroscience"
+author: "Daniel H. Baker"
 date: "13/12/2020"
 output: pdf_document
+bibliography: bibliography.bib
+biblio-style: apalike
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+library(pals)
+
+nsims <- 100  # global value for the whole script - increase to 100000 for final run
+
+nbdtpal <- c(rgb(0,0,0),rgb(187/255,40/255,107/255),rgb(56/255,111/255,164/255),'darkgreen','darkorange')
+
+tsq1.test <- function(data,mu=NULL){
+  # function to calculate a one-sample t-squared test after Hotelling (1931)
+  # for two-sample tests, use the Hotelling package
+
+  # compare each DV to some value
+  if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}
+
+  xbar <- colMeans(data) # estimate means for each DV
+  m <- ncol(data)   # record number of DVs
+  N <- nrow(data)   # record sample size
+  C <- cov(data)    # calculate covariance matrix
+  Cinv <- solve(C)  # calculate inverse matrix
+  tsq <- N*t(xbar)%*%Cinv%*%xbar # compute t2 statistic
+  Fratio <- tsq * (N-m)/(m*(N-1)) # convert to F-ratio
+  df1 <- m      # record degrees of freedom
+  df2 <- N-m
+  pval <- 1- pf(Fratio,df1,df2) # estimate p-value
+  p.value <- min(pval,1)
+
+  # store outputs in a data structure
+  output <- data.frame(tsq,Fratio,df1,df2,p.value)
+  return(output)}
+
+tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
+  # function to calculate the t-squared circ statistic from Victor & Mast (1991)
+  # the inputs must be Nx2 or 2xN matrices of numbers
+  # for a one-sample test, mu is an optional vector to which the data are compared
+
+  # transpose if dimensions are the wrong way round
+  if (is.matrix(x)){d <- dim(x)
+  if (d[1]<d[2]){x <- t(x)}}   
+  if (!is.null(y)){if (is.matrix(y)){d <- dim(y)
+  if (d[1]<d[2]){y <- t(y)}}}
+
+  if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}
+
+  # convert data to complex representation
+  x <- complex(real=x[,1],imaginary=x[,2])
+  if (!is.null(y)){y <- complex(real=y[,1],imaginary=y[,2])}
+
+  if (is.null(y)){method <- 'One-sample t-squared circ test'}
+  if (!is.null(y) & paired==TRUE){method <- 'Paired t-squared circ test'}
+  if (!is.null(y) & paired==FALSE){method <- 'Independent samples t-squared circ test'}
+
+  if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}
+
+  # paired or one-sample version of the test
+  if (paired==TRUE | is.null(y)){
+
+    nobs <- length(x)
+    cohmean <- mean(x)
+    absmean <- abs(cohmean)
+
+    displacement <- abs(x - cohmean)
+    diffsumsq <- sum(displacement^2)
+    tsqc <- (nobs-1) * (absmean^2) / diffsumsq
+    Fratio <- nobs*tsqc
+    df2 <- nobs*2-2
+    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
+  }
+
+  # independent samples version of the test
+  if (paired==FALSE & !is.null(y)){  
+    nobs1 <- length(x)
+    cohmean1 <- mean(x)
+    absmean1 <- abs(cohmean1)
+    displacement1 <- abs(x - cohmean1)
+
+    nobs2 <- length(y)
+    cohmean2 <- mean(y)
+    absmean2 <- abs(cohmean2)    
+    displacement2 <- abs(y - cohmean2)
+
+    meandiff <- cohmean1 - cohmean2
+    absdiff <- abs(meandiff)
+    diffsumsq <- sum(displacement1^2) + sum(displacement2^2)
+
+    tsqc <- (nobs1 + nobs2 - 2) * (absdiff^2) / diffsumsq
+    Fratio <- ((nobs1*nobs2)/(nobs1 + nobs2))*tsqc
+    df2 <- (2*nobs1 + 2*nobs2 - 4)
+    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
+  }
+
+  df1 <- 2
+  p.value <- min(pval,1)
+
+  output <- data.frame(tsqc,Fratio,df1,df2,p.value,method)
+  return(output)}
+
+CI.test <- function(data){
+  # function to calculate the condition index of bivariate data
+  # compare for significance with an expected distribution function
+
+  C <- cov(data)  # covariance matrix
+  N <- nrow(data) # sample size
+  eigVal  <- eigen(C)$values  # calculate eigenvalues
+  CI <- sqrt(eigVal[1]/eigVal[2]) # calculate condition index
+
+  maxci <- max(20,round(CI+5))   # make sure our target CI is in range
+  cilist <- seq(1,maxci,0.01)    # list of possible CIs
+
+  # calculate probability density function using modified Edelman equation
+  pdffunction <- ((N-2)*(2^(N-2))) * 
+    ((cilist^2 - 1)/((cilist^2+1)^(N-1))) * (cilist^(N-3))
+  cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
+  criticalCI <- cilist[min(which(cdfinverse<=0.05))]  # find the threshold CI
+
+  indices <- which(cilist>=CI)  # find the CI values larger than our CI
+  pval <- cdfinverse[indices[1]] # estimate the p-value
+
+  output <- data.frame(CI,N,criticalCI,pval)
+  return(output)}
+
+addalpha <- function(col, alpha=1){apply(sapply(col, col2rgb)/255, 2, function(x) rgb(x[1], x[2], x[3], alpha=alpha))}
+
 ```
 
-## R Markdown
+## Abstract
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+## Background
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+A widely used paradigm in many branches of neuroscience is to drive the nervous system using periodic stimuli.
+This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations.
+These periodic responses, often called the _steady-state_ or _frequency following_ response, can be recorded from single neurons [@Enroth-Cugell1966] and local field potentials [@Morrone1987] using invasive methods, or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans [@Norcia2015] and in diverse animal species including insects [@Afsari2014], birds [@Porciatti1990], rodents [@Hwang2019] and primates [@Nakayama1982].
+Steady-state methods are used to measure early sensory responses in vision [@Regan1966], hearing [@Rees1986] and somatosensation [@Snyder1992], and closely related paradigms have been developed to target specific stimulus features such as orientation [@Braddick1986], and facial expression [@Gray2020] and identity [@Liu-Shuang2014].
+In fMRI research, _travelling wave_ methods [@Engel1994; @Sereno1995] are used to map the retinotopic responses of early visual cortex, again using stimuli that change periodically, this time in spatial position.
+Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way [@Spitschan2014].
 
-```{r cars}
-summary(cars)
+A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal.
+The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a convenient, and well-isolated index of the brain's response.
+Fourier spectra comprise both amplitude and phase information, which can be equivalently expressed as complex numbers with real and imaginary components.
+In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data only.
+However an alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary components).
+Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods.
+
+For pointwise and pairwise comparisons, Hotelling's $T^2$ statistic [@Hotelling1931] is a multivariate extension of the t-test. For the one-sample case, the test statistic is defined as:
+
+\begin{equation}
+\label{eq:t2eq}
+T^2 = N(\bar{x} - \mu)' C^{-1} (\bar{x} - \mu),
+\end{equation}
+
+where _N_ is the number of observations, $\bar{x}$ is the multivariate sample mean, $\mu$ is the point of comparison, $C^{-1}$ is the inverse covariance matrix, and $'$ denotes vector transposition.
+Conceptually, the $T^2$ statistic extends the univariate t-statistic by incorporating the covariance between the dependent variables.
+Two-sample and paired variants are also available, and the test can be applied to an arbitrary number of dependent variables (though here we will consider only the bivariate case).
+
+More recently, Victor and Mast [-@Victor1991] proposed a simpler version of $T^2$, called $T^2_{circ}$.
+The $T^2_{circ}$ statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance.
+When these conditions are met, the test statistic is defined as:
+
+\begin{equation}
+\label{eq:t2c}
+T^2_{circ} = (N-1)\frac{|\bar{x}-\mu|^2}{\Sigma|x_j - \bar{x}|^2}
+\end{equation}
+
+where $x_j$ denotes the $j$th observation of the dependent variables, and all other terms retain their previous meanings.
+Notice that no covariance term is present in equation \@ref(eq:t2c), because of the independence assumption.
+This makes the statistic simpler to calculate, but causes problems when the assumption is violated.
+
+In the present paper, we develop best practice guidelines for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms.
+We first demonstrate why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed.
+We then investigate conditions under which we should use either the $T^2$ or $T^2_{circ}$ statistic.
+The range of sample sizes and effect sizes where $T^2_{circ}$ is more sensitive is identified.
+We develop a novel method for testing the assumptions of the $T^2_{circ}$ statistic, based on calculating the condition index of a multivariate data set.
+We then discuss appropriate methods for identifying outliers and calculating error bars.
+Finally, we demonstrate the proposed techniques on two example SSVEP data sets, and recommend some best practice guidelines for analysis decisions.
+All data and scripts used to produce this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats).
+These include _R_ functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index statistic proposed in this paper.
+
+## Fourier amplitudes violate parametric assumptions of univariate statistics
+
+Many studies use parametric t-tests and analysis of variance (ANOVA) to analyse periodic data.
+Specifically, the amplitude component of the Fourier spectrum at the stimulation frequency is used as the dependent variable, discarding the phase information.
+This is problematic, because the amplitude is an absolute quantity, and can never fall below zero.
+Distributions of amplitudes for weak signals are therefore likely to be positively skewed, and in general will violate the assumption of normality.
+
+```{r amphists, fig.cap="Demonstration of skew in absolute Fourier amplitudes for weak signals.
+Signal strength is quantified as Cohen's d, defined as the ratio of the mean to the standard deviation of the sample.
+The upper row shows samples of 50 grey points, and the population mean (coloured points).
+The lower row shows kernel density functions generated from 100,000 amplitude values.
+Note that the mean phase of the signal is irrelevant for these simulations, and is shown in the positive x-direction for consistency.", fig.align="center", echo=FALSE, fig.width=12, fig.height=6}
+
+npoints <- 50
+dlist <- c(0,0.5,1,2,4)
+
+plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,60), ylim=c(0,24))
+for (n in 1:5){lines(c((n-1)*12,(n*12)-2),c(18,18),lwd=2)}
+for (n in 1:5){lines(c(5,5)+(n-1)*12,c(13,23),lwd=2)}
+
+for (n in 1:5){lines(c((n-1)*12,(n*12)-2),c(0,0),lwd=2)}
+for (n in 1:5){lines(c(0,0)+(n-1)*12,c(0,10),lwd=2)}
+
+text(0.5,19,'x',adj=0.5,cex=1.5)
+text(4,13.5,'y',adj=0.5,cex=1.5)
+
+for (d in 1:5){
+simdata <- matrix(rnorm(npoints*2,mean=0,sd=1),nrow=npoints,ncol=2)
+simdata[,1] <- simdata[,1] + dlist[d]
+for (n in 1:npoints){lines(c(0,simdata[n,1])+5+(d-1)*12,c(0,simdata[n,2])+18,col=rgb(0,0,0,alpha=0.2))}
+points(simdata[,1]+5+(d-1)*12,simdata[,2]+18,type='p',pch=16,col=rgb(0,0,0,alpha=0.2))
+points(5+(d-1)*12+dlist[d],18,type='p',pch=16,cex=2,col=nbdtpal[d])
+text(5+(d-1)*12,24,paste('d =',dlist[d]),adj=0.5,cex=1.5)
+}
+
+for (d in 1:5){
+simdata <- matrix(rnorm(nsims*2,mean=0,sd=1),nrow=nsims,ncol=2)
+simdata[,1] <- simdata[,1] + dlist[d]
+absvals <- abs(complex(real=simdata[,1],imaginary=simdata[,2]))
+a <- density(absvals)
+a$x <- a$x-min(a$x)
+a$x <- a$x/max(a$x)
+a$x <- 10*a$x
+a$y <- 10*a$y/max(a$y)
+polygon(a$x+(d-1)*12,a$y,col=addalpha(nbdtpal[d],alpha=0.3),border=NA)
+
+# lines(seq(0,10,0.01)+(d-1)*12,10*pnorm(seq(-5,5,0.01)),lwd=2)
+# q <- quantile(scale(absvals),probs=seq(0,1,0.001))
+# lines(5+q+(d-1)*12,seq(0,10,0.01),lwd=2,col=nbdtpal[d])
+
+}
+
 ```
 
-## Including Plots
+The upper row of Figure \@ref(fig:amphists) shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components.
+The amplitudes are the lengths of the lines joining each grey point to the origin.
+The lower row in Figure \@ref(fig:amphists) shows distributions of amplitudes for the same set of signal strengths.
+These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d > 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]).
+One consequence of this is that t-tests will typically have an inflated Type 1 error (false positive) rate for many signals encountered empirically.
 
-You can also embed plots, for example:
+Typical solutions, such as log-transforming the data, are unlikely to be equally applicable to all conditions.
+For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal.
+Applying a transform to both conditions is therefore problematic.
+Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents.
+Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, as we will demonstrate in the next section.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+## Conditions under which $T^2_{circ}$ is more sensitive than $T^2$
+
+Using the $T^2$ and $T^2_{circ}$ allow us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
+Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$.
+Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and demonstrate that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f).
+However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations.
+This advantage is lost for large sample sizes (N > 32) and large effect sizes (d > 2 when N > 8)
+
+```{r powerfig, fig.cap=".", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+
+layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))
+abc <- c('a','b','c')
+
+vinds <- c(2,5,17)
+dlist <- seq(0,4,0.25)
+slist <- seq(1.5,5,0.5)
+samplesizes <- ceiling(2^slist)
+
+tsqmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
+circmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
+for (n in 1:nsims){
+  for (s in 1:length(samplesizes)){
+    for (s2 in 1:length(dlist)){
+    data <- matrix(rnorm(2*samplesizes[s]),nrow=samplesizes[s],ncol=2)
+    data[,1] <- data[,1] + dlist[s2]
+    a <- tsq1.test(data)
+    if (a[5]<0.05){tsqmatrix[s,s2] <- tsqmatrix[s,s2] + 1}
+    a <- tsqc.test(data)
+    if (a[5]<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
+  }}
+}
+
+tsqpropM <- tsqmatrix/nsims
+circpropM <- circmatrix/nsims
+tsqvscirc <- circpropM - tsqpropM
+
+plotlims <- c(1,5,0,1)
+ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
+ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
+ticklabelsx <- 2^ticklocsx        # set labels for x ticks
+ticklabelsy <- ticklocsy    # set labels for y ticks
+
+for (n in 1:3){
+par(pty="s")
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2.2, cex.lab=1.5)
+title(paste('d =',dlist[vinds[n]]),cex.main=2)
+lines(slist,tsqpropM[,vinds[n]],lwd=3,col=nbdtpal[1])
+lines(slist,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])
+
+points(slist,tsqpropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
+points(slist,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])
+
+text(1.1,0.95,paste('(',abc[n],')',sep=''),cex=2)
+
+if (n==1){legend(3.8,1,c(expression(T^2),expression('T'[circ]^2)),pch=15:16,cex=1.5,col=nbdtpal[1:2],box.lwd=2)}
+}
+
+par(pty="s")
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
+ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
+ticklocsy <- seq(0,4,1)    # locations of tick marks on y axis
+ticklabelsx <- 2^ticklocsx        # set labels for x ticks
+ticklabelsy <- seq(0,4,1)    # set labels for y ticks
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(expression(T^2),cex.main=2)
+image(slist,dlist,tsqpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+text(1.1,3.8,'(d)',cex=2)
+legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white')
+
+par(pty="s")
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+ttext <- expression('T'[circ]^2)
+title(ttext,cex.main=2)
+image(slist,dlist,circpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+text(1.1,3.8,'(e)',cex=2)
+
+par(pty="s")
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+ttext <- expression({'T'[circ]^2}~-~{T^{"2"}})
+title(ttext,cex.main=2)
+image(slist,dlist,tsqvscirc,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+cl <- contourLines(slist,dlist,tsqvscirc,levels=c(0.02,0.05,0.1,0.2,0.4))
+for (n in 1:length(cl)){
+  temp <- cl[n]
+  lines(temp[[1]]$x,temp[[1]]$y,col='black',lwd=3)
+}
+text(1.1,3.8,'(f)',cex=2)
+
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+## Limitations of $T^2_{circ}$ when assumptions are violated
+
+```{r falsealarms, fig.asp=1.4, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
+Estimates are for 100000 simulations per condition, with N=10 observations.
+The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}
+
+par(mfrow=c(2,1))
+
+nobs <- 10
+corvals <- seq(-0.9,0.9,0.1)
+
+plotlims <- c(-1,1,0,0.1) 
+ticklocsx <- seq(-1,1,0.5)    # locations of tick marks on x axis
+ticklocsy <- seq(0,0.1,0.025)    # locations of tick marks on y axis
+ticklabelsx <- ticklocsx        # set labels for x ticks
+ticklabelsy <- ticklocsy*100    # set labels for y ticks
+
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2) 
+title(xlab="Correlation (R)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
+title(ylab="Percent significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+
+lines(c(-1,1),c(0.05,0.05),lty=2,lwd=2)
+
+hsig <- corvals*0
+csig <- corvals*0
+simdata <- matrix(0,nrow=nobs,ncol=2)
+for (cond in 1:length(corvals)){
+  for (simno in 1:nsims){
+
+    xdata <- rnorm(nobs,mean=0,sd=1)
+    ydata <- sign(corvals[cond])*sqrt(abs(corvals[cond]))*xdata + sqrt(1-abs(corvals[cond]))*rnorm(nobs,mean=0,sd=1)
+    simdata[,1] <- xdata
+    simdata[,2] <- ydata
+    output <- tsqc.test(simdata,paired=FALSE)
+    if (output$p.value < 0.05){csig[cond] <- csig[cond] + 1}
+    output <- tsq1.test(simdata)
+    if (output$p.value < 0.05){hsig[cond] <- hsig[cond] + 1}
+  }}
+propc <- csig/nsims
+lines(corvals,propc,lty=1,lwd=4,col=nbdtpal[2])
+proph <- hsig/nsims
+lines(corvals,proph,lty=1,lwd=4,col=nbdtpal[1])
+
+legend(-0.25,0.1, c(expression('T'[circ]^2),expression('T'^2)), cex=1, col=nbdtpal[2:1],lty=1, lwd=3, box.lwd=2)
+
+nobs <- 10000
+corvals <- seq(-0.9,0.9,0.45)
+simdata <- matrix(0,nrow=nobs,ncol=2)
+
+for (cond in 1:length(corvals)){
+  xdata <- rnorm(nobs,mean=0,sd=1)
+  ydata <- sign(corvals[cond])*sqrt(abs(corvals[cond]))*xdata + sqrt(1-abs(corvals[cond]))*rnorm(nobs,mean=0,sd=1)
+  compdata <- data.frame(xdata,ydata)
+
+  A <- cov(compdata)
+  ctr    <- colMeans(compdata) 
+  RR     <- chol(A)                               # Cholesky decomposition
+  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
+  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
+  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
+  eigVal  <- eigen(A)$values
+  eigVec  <- eigen(A)$vectors
+  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
+  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
+  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
+  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
+  ellRot  <- eigVec %*% t(ellBase)
+
+  points(0.05*xdata[1:100]+corvals[cond],0.005*ydata[1:100]+0.02,pch=16,cex=0.5,col=nbdtpal[3])
+  matlines(0.1*xMat+corvals[cond], 0.01*yMat+0.02, lty=1, lwd=3, col="black")
+  lines(0.1*(ellRot+ctr)[1, ]+corvals[cond], 0.01*(ellRot+ctr)[2, ]+0.02, lwd=2)
+}
+text(-0.95,0.095,'(a)',cex=1.5)
+
+nobs <- 10
+corvals <- seq(-0.9,0.9,0.1)
+varratios <- 10^(corvals/2)
+
+plotlims <- c(-1,1,0,0.1) 
+ticklocsx <- seq(-1,1,1)    # locations of tick marks on x axis
+ticklocsy <- seq(0,0.1,0.025)    # locations of tick marks on y axis
+ticklabelsx <- c('1/10','1','10')        # set labels for x ticks
+ticklabelsy <- ticklocsy*100    # set labels for y ticks
+
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2) 
+title(xlab="Variance ratio (y/x)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
+title(ylab="Percent significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+
+lines(c(-1,1),c(0.05,0.05),lty=2,lwd=2)
+
+hsig <- corvals*0
+csig <- corvals*0
+simdata <- matrix(0,nrow=nobs,ncol=2)
+for (cond in 1:length(corvals)){
+  for (simno in 1:nsims){
+
+    xdata <- rnorm(nobs,mean=0,sd=sqrt(1/varratios[cond]))
+    ydata <- rnorm(nobs,mean=0,sd=sqrt(varratios[cond]))
+    simdata[,1] <- xdata
+    simdata[,2] <- ydata
+
+    output <- tsqc.test(simdata,paired=FALSE)
+    if (output$p.value < 0.05){csig[cond] <- csig[cond] + 1}
+    output <- tsq1.test(simdata)
+    if (output$p.value < 0.05){hsig[cond] <- hsig[cond] + 1}
+  }}
+propc <- csig/nsims
+lines(corvals,propc,lty=1,lwd=4,col=nbdtpal[2])
+proph <- hsig/nsims
+lines(corvals,proph,lty=1,lwd=4,col=nbdtpal[1])
+
+legend(-0.25,0.1, c(expression('T'[circ]^2),expression('T'^2)), cex=1, col=nbdtpal[2:1],lty=1, lwd=3, box.lwd=2)
+
+nobs <- 10000
+corvals <- seq(-0.9,0.9,0.45)
+varratios <- 10^(corvals/2)
+simdata <- matrix(0,nrow=nobs,ncol=2)
+
+for (cond in 1:length(corvals)){
+  xdata <- rnorm(nobs,mean=0,sd=sqrt(1/varratios[cond]))
+  ydata <- rnorm(nobs,mean=0,sd=sqrt(varratios[cond]))
+  compdata <- data.frame(xdata,ydata)
+
+  A <- cov(compdata)
+  ctr    <- colMeans(compdata) 
+  RR     <- chol(A)                               # Cholesky decomposition
+  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
+  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
+  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
+  eigVal  <- eigen(A)$values
+  eigVec  <- eigen(A)$vectors
+  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
+  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
+  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
+  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
+  ellRot  <- eigVec %*% t(ellBase)
+
+  points(0.05*xdata[1:100]+corvals[cond],0.005*ydata[1:100]+0.02,pch=16,cex=0.5,col=nbdtpal[3])
+  matlines(0.1*xMat+corvals[cond], 0.01*yMat+0.02, lty=1, lwd=3, col="black")
+  lines(0.1*(ellRot+ctr)[1, ]+corvals[cond], 0.01*(ellRot+ctr)[2, ]+0.02, lwd=2)
+}
+text(-0.95,0.095,'(b)',cex=1.5)
+
+```
+
+## A novel method to test the assumptions of $T^2_{circ}$
+
+Despite the severe consequences of violating the assumptions of the $T^2_{circ}$ statistic, there is currently no accepted test for those assumptions that could be applied to real data.
+Victor and Mast [-@Victor1991] suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single individual, whereas data from multiple participants may be less likely to exhibit independence of the real and imaginary components [@Pei2017].
+
+One convenient way to test the assumptions of $T^2_{circ}$ is to assess the _condition index_ of a data set, which describes the ratio of eigenvalues for a cloud of points.
+The eigenvectors are the major and minor axes of the bounding ellipse (the straight lines in the example icons at the foot of Figure \@ref(fig:falsealarms)a,b).
+Conventionally, the condition index is calculated as the square root of the longest/shortest eigenvector length.
+For uncorrelated random numbers the distribution of condition indices is positively skewed, with a minimum of 1 [@Edelman1988].
+This is because two independent samples of numbers from the same underlying distribution will still by chance have unequal eigenvectors.
+Edelman [-@Edelman1988] provides an equation (Eq 14) for the probability density function of condition indices (x) as a function of sample size (n):
+
+$pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)}$
+
+Attempts to validate this by simulation suggest that for small sample sizes (n<10) a closer approximation is given by:
+
+$pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}$
+
+In Figure \@ref(fig:distcomparison)a we show an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100000 random data sets for n=10 (blue shading).
+The vertical lines show the critical (95%) threshold for the analytic and simulated results.
+Figure \@ref(fig:distcomparison)b shows how these thresholds change as a function of the number of observations, and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
+
+```{r distcomparison, fig.cap='d', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
+
+par(mfrow=c(1,2), las=1)
+
+n <- 10
+x <- seq(1,20,0.01)
+nr <- n-1
+pdffunction <- ((nr-1)*(2^(nr-1))) * ((x^2 - 1)/((x^2+1)^nr)) * (x^(nr-2))
+
+plotlims <- c(1,5,0,1) 
+ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
+ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2) 
+title(xlab="Eigenvalue ratio", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
+title(ylab="Probability", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+
+lines(x,pdffunction/max(pdffunction))
+cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction))
+lines(x,cdfinverse,col=nbdtpal[2])
+lines(c(1,5),c(0.05,0.05),lty=2,col=nbdtpal[1])
+criticalx <- x[min(which(cdfinverse<=0.05))]
+lines(c(criticalx,criticalx),c(0,1),col=nbdtpal[2],lwd=2)
+
+simvect <- NULL
+for (s in 1:nsims){
+  simdata <- matrix(rnorm(2*n,mean=0,sd=1),nrow=n,ncol=2)
+  evs <- eigen(cov(simdata)) 
+  simvect[s] <- sqrt(evs$values[1]/evs$values[2])
+}
+
+a <- hist(simvect,breaks = 200, plot=FALSE)
+axvals <- a$mids
+ayvals <- a$counts/max(a$counts)
+polygon(c(1,axvals,1), c(0,ayvals,0), col=addalpha(nbdtpal[3],alpha=0.2),border=NA)
+simcrit <- quantile(simvect,0.95)
+lines(c(simcrit,simcrit),c(0,1),col=nbdtpal[3],lwd=2,lty=2)
+
+x <- seq(1,50,0.01)
+edelman1 <- NULL
+edelman2 <- NULL
+simratios <- NULL
+for (n in 3:25){
+nr <- n
+exactpdf <- ((nr-1)*(2^(nr-1))) * ((x^2 - 1)/((x^2+1)^nr)) * (x^(nr-2))
+cdfinverse <- 1-(cumsum(exactpdf)/sum(exactpdf))
+edelman1[n-2] <- x[min(which(cdfinverse<=0.05))]
+
+nr <- n - 1
+exactpdf <- ((nr-1)*(2^(nr-1))) * ((x^2 - 1)/((x^2+1)^nr)) * (x^(nr-2))
+cdfinverse <- 1-(cumsum(exactpdf)/sum(exactpdf))
+edelman2[n-2] <- x[min(which(cdfinverse<=0.05))]
+
+simvect <- NULL
+for (s in 1:nsims){
+  simdata <- matrix(rnorm(2*n,mean=0,sd=1),nrow=n,ncol=2)
+  evs <- eigen(cov(simdata))
+  simvect[s] <- sqrt(evs$values[1]/evs$values[2])
+}
+simratios[n-2] <- quantile(simvect,0.95)
+}
+
+plotlims <- c(0,25,0,25) 
+ticklocsx <- seq(0,25,5)    # locations of tick marks on x axis
+ticklocsy <- seq(0,25,5)    # locations of tick marks on y axis
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Threshold", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+
+lines(3:25,edelman1,lwd=2,col=nbdtpal[1])
+lines(3:25,edelman2,lwd=2,col=nbdtpal[2])
+lines(3:25,simratios,lwd=2,col=nbdtpal[3])
+legend(12, 25, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=1, lwd=3, box.lwd=2)
+
+```
+
+The eigenvalue ratio can be used as a test of the assumptions of $T^2_{circ}$.
+If we observe a condition index above the critical threshold for the number of observations then the data set can be said to significantly violate the assumption of equal eigenvalues.
+Because our modified equation provides an inverse density function (red curve in Figure \@ref(fig:distcomparison)a), we can use this to calculate a p-value for the test.
+If the test is non-significant, one can proceed with $T^2_{circ}$; if it is significant, $T^2$ should be used instead.
+
+## Identifying and removing outliers using the Mahalanobis distance
+
+If a data set produces a significant result using the condition index test, this could be due to the presence of one or more outliers.
+The Mahalanobis distance [@Mahalanobis1936] is a useful metric for identifying such multivariate outliers so that they can be excluded.
+It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points.
+This means that any correlations in the data set are taken into account when calculating the distance metric, D.
+
+The effectiveness of this approach to outlier exclusion can be assessed by simulation using the condition index test.
+Figure \@ref(fig:outlierplot) shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves).
+In all cases, the functions depart from the Type 1 error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the Mahalanobis distance exceeds a value around 3.
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean.
+Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the $T^2_{circ}$ test.
+
+```{r outlierplot, fig.cap='d', fig.align="center", echo=FALSE}
+
+dlist <- seq(0,8,0.5)
+slist <- seq(2,6,1)
+samplesizes <- ceiling(2^slist)
+
+nsig <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
+for (s in 1:length(samplesizes)){
+  for (d in 1:length(dlist)){
+    for (n in 1:nsims){
+    data <- matrix(rnorm(2*samplesizes[s]),nrow=samplesizes[s],ncol=2)
+    data[1,] <- c(dlist[d],0)
+    output <- CI.test(data)
+    if(output$pval<0.05){nsig[s,d] <- nsig[s,d] + 1}
+    }
+  }
+}
+
+propsig <- nsig/nsims
+
+plotlims <- c(0,8,0,1) 
+ticklocsx <- seq(0,8,1)    # locations of tick marks on x axis
+ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Outlier Mahalanobis distance", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+
+lines(c(0,8),c(0.05,0.05),lty=2,lwd=2)
+
+for (s in 1:length(samplesizes)){
+  lines(dlist,propsig[s,],lwd=3,col=nbdtpal[s])
+}
+
+legend(0,1,samplesizes,col=nbdtpal,lwd=3,box.lwd=2,title='N observations')
+
+```
+
+## Methods for calculating error bars for Fourier amplitudes
+
+## Applying these methods to real data sets
+
+### Human visual steady-state data
+
+### Mouse auditory steady-state data
+
+https://www.nature.com/articles/s41597-020-00621-z
+https://doi.gin.g-node.org/10.12751/g-node.e5tyek/
+
+## Recommendations for analysing periodic data
+
+Multi/Bivariate statistics should be used for phase-locked Fourier data
+
+Always remove outliers with a Mahalanobis distance > 3
+
+Always test assumptions if using $T^2_{circ}$
+
+No benefit to using tsqcirc for sample sizes>32
+
+With more than two conditions use MANOVA
+
+Use bootstrapping to calculate error bars
+
+## Further considerations
+
+Mention about non phase-locked responses
+
+Greater power with coherent averaging
+
+Could develop ANOVA-circ
+
+## References

### 2020-12-20 CItest.R

@@ -1,0 +1,23 @@
+CI.test <- function(data){
+  # function to calculate the condition index of bivariate data
+  # compare for significance with an expected distribution function
+  
+  C <- cov(data)  # covariance matrix
+  N <- nrow(data) # sample size
+  eigVal  <- eigen(C)$values  # calculate eigenvalues
+  CI <- sqrt(eigVal[1]/eigVal[2]) # calculate condition index
+  
+  maxci <- max(20,round(CI+5))   # make sure our target CI is in range
+  cilist <- seq(1,maxci,0.01)    # list of possible CIs
+  
+  # calculate probability density function using modified Edelman equation
+  pdffunction <- ((N-2)*(2^(N-2))) * 
+    ((cilist^2 - 1)/((cilist^2+1)^(N-1))) * (cilist^(N-3))
+  cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
+  criticalCI <- cilist[min(which(cdfinverse<=0.05))]  # find the threshold CI
+  
+  indices <- which(cilist>=CI)  # find the CI values larger than our CI
+  pval <- cdfinverse[indices[1]] # estimate the p-value
+  
+  output <- data.frame(CI,N,criticalCI,pval)
+  return(output)}

### 2020-12-20 tsq1test.R

@@ -1,0 +1,22 @@
+tsq1.test <- function(data,mu=NULL){
+  # function to calculate a one-sample t-squared test after Hotelling (1931)
+  # for two-sample tests, use the Hotelling package
+  
+  # compare each DV to some value
+  if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}   
+  
+  xbar <- colMeans(data) # estimate means for each DV
+  m <- ncol(data)   # record number of DVs
+  N <- nrow(data)   # record sample size
+  C <- cov(data)    # calculate covariance matrix
+  Cinv <- solve(C)  # calculate inverse matrix
+  tsq <- N*t(xbar)%*%Cinv%*%xbar # compute t2 statistic
+  Fratio <- tsq * (N-m)/(m*(N-1)) # convert to F-ratio
+  df1 <- m      # record degrees of freedom
+  df2 <- N-m
+  pval <- 1- pf(Fratio,df1,df2) # estimate p-value
+  p.value <- min(pval,1)
+  
+  # store outputs in a data structure
+  output <- data.frame(tsq,Fratio,df1,df2,p.value)
+  return(output)}

### 2020-12-20 tsqctest.R

@@ -1,0 +1,65 @@
+tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
+  # function to calculate the t-squared circ statistic from Victor & Mast (1991)
+  # the inputs must be Nx2 or 2xN matrices of numbers
+  # for a one-sample test, mu is an optional vector to which the data are compared
+  
+  # transpose if dimensions are the wrong way round
+  if (is.matrix(x)){d <- dim(x)
+  if (d[1]<d[2]){x <- t(x)}}   
+  if (!is.null(y)){if (is.matrix(y)){d <- dim(y)
+  if (d[1]<d[2]){y <- t(y)}}}
+  
+  if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}
+  
+  # convert data to complex representation
+  x <- complex(real=x[,1],imaginary=x[,2])
+  if (!is.null(y)){y <- complex(real=y[,1],imaginary=y[,2])}
+  
+  if (is.null(y)){method <- 'One-sample t-squared circ test'}
+  if (!is.null(y) & paired==TRUE){method <- 'Paired t-squared circ test'}
+  if (!is.null(y) & paired==FALSE){method <- 'Independent samples t-squared circ test'}
+  
+  if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}
+  
+  # paired or one-sample version of the test
+  if (paired==TRUE | is.null(y)){ 
+    
+    nobs <- length(x)
+    cohmean <- mean(x)
+    absmean <- abs(cohmean)
+    
+    displacement <- abs(x - cohmean)
+    diffsumsq <- sum(displacement^2)
+    tsqc <- (nobs-1) * (absmean^2) / diffsumsq
+    Fratio <- nobs*tsqc
+    df2 <- nobs*2-2
+    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
+  }
+  
+  # independent samples version of the test
+  if (paired==FALSE & !is.null(y)){  
+    nobs1 <- length(x)
+    cohmean1 <- mean(x)
+    absmean1 <- abs(cohmean1)
+    displacement1 <- abs(x - cohmean1)
+    
+    nobs2 <- length(y)
+    cohmean2 <- mean(y)
+    absmean2 <- abs(cohmean2)    
[…]

### 2020-12-23 manuscript.Rmd
Completed first draft of manuscript

@@
 date: "13/12/2020"
-output: pdf_document
+output:
+  bookdown::pdf_document2:
+    fig_caption: yes 
+    toc: false   
+    keep_tex: true
+  html_document:
+    df_print: paged
+  pdf_document: default
 bibliography: bibliography.bib
@@
 
-library(pals)
+# reasonably compact code to check which packages are installed, 
+# install the missing ones, and activate all
+packagelist <- c('bookdown','pals','shape','diagram','HDMD')
+missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
+if (length(missingpackages)>0){install.packages(missingpackages)}
+toinstall <- packagelist[which(!packagelist %in% (.packages()))]
+invisible(lapply(toinstall,library,character.only=TRUE))
 
@@
 
+anovacirc.test <- function(data){
+  # analysis of variance using complex data
+  # an extension of the logic of the t-squared circ statistic
+
+  grouplabels <- data$grouplabels
+  datavals <- data$simdata
+  factorlist <- levels(grouplabels)
+
+  grandmean <- mean(datavals)
+
+  groupmeans <- NULL
+  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
+
+  SSM <- 0
+  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
+
+  dfM <- 2*(nlevels(grouplabels)-1)
+
+  allresiduals <- 1:length(datavals)
+  for (n in 1:nlevels(grouplabels)){allresiduals[which(grouplabels==factorlist[n])] <- abs(datavals[which(grouplabels==factorlist[n])] - groupmeans[n])}
+
+  SSR <- sum(allresiduals^2)
+  dfR <- 2*(length(datavals)-nlevels(grouplabels))
+  MSM <- SSM/dfM
+  MSR <- SSR/dfR
+  Fratio <- MSM/MSR
+  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+
+  output <- NULL
+  output$Fratio <- Fratio
+  output$pvalue <- pvalue
+  output$SSM <- SSM
+  output$SSR <- SSR
+  output$dfM <- dfM
+  output$dfR <- dfR
+  output$MSM <- MSM
+  output$MSR <- MSR
+
+  return(output)
+}
+
+anovacircR.test <- function(data){
+  # analysis of variance using complex data
+  # an extension of the logic of the t-squared circ statistic
+  # this version for repeated measures designs
+
+  grouplabels <- data$grouplabels
+  factorlist <- levels(grouplabels)
+  participantlabels <- data$participant
+  participantlist <- levels(participantlabels)
+  datavals <- data$simdata
+
+  grandmean <- mean(datavals)
+
+  SSW <- 0
+  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
+  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))
+
+  groupmeans <- NULL
+  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
+  SSM <- 0
+  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
+  dfM <- 2*(nlevels(grouplabels)-1)
+
+SSR <- SSW - SSM
+dfR <- dfW - dfM
+
+  MSM <- SSM/dfM
+  MSR <- SSR/dfR
+  Fratio <- MSM/MSR
+  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+
+  output <- NULL
+  output$Fratio <- Fratio
+  output$pvalue <- pvalue
+  output$SSW <- SSW  
+  output$SSM <- SSM
+  output$SSR <- SSR
+  output$dfW <- dfW
+  output$dfM <- dfM
+  output$dfR <- dfR
+  output$MSM <- MSM
+  output$MSR <- MSR
+
+  return(output)
+}
+
+getel <- function(compdata){
+  A <- cov(compdata)
+  ctr    <- colMeans(compdata) 
+  RR     <- chol(A)                               # Cholesky decomposition
+  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
+  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
+  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
+  eigVal  <- eigen(A)$values
+  eigVec  <- eigen(A)$vectors
+  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
+  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
+  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
+  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
+  ellRot  <- eigVec %*% t(ellBase) 
+  return(ellRot)
+  }
+
+v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
+  ## Create a function to perform Matlab's v4 interpolation.
+  ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
+  ## and variables xo and yo, the co-ordinates which will be use to create a grid for interpolation
+  xo <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
+  yo <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
+  xy <- df$x + df$y*sqrt(as.complex(-1))
+  d <- matrix(rep(xy,length(xy)),nrow = length(xy), ncol = length(xy))
+  d <- abs(d - t(d))
+  diag(d) <- 1
+  g <- (d^2) * (log(d)-1)   # Green's function.
+  diag(g) <- 0
+  weights <- qr.solve(g,df$z)
+  xy <- t(xy)
+  outmat <- matrix(nrow = gridRes,ncol = gridRes)
+  for (i in 1:gridRes){
+    for (j in 1:gridRes) {
+      test4 <- abs((xo[i,j] + sqrt(as.complex(-1))*yo[i,j]) - xy)
+      g <- (test4^2) * (log(test4)-1)
+      outmat[i,j] <- g %*% weights}}
+  outDf <- data.frame(x = xo[,1],outmat)
+  names(outDf)[1:length(yo[1,])+1] <- yo[1,]
+  return(outDf)}
+
 addalpha <- function(col, alpha=1){apply(sapply(col, col2rgb)/255, 2, function(x) rgb(x[1], x[2], x[3], alpha=alpha))}
@@
 
-## Abstract
-
-## Background
+# Abstract
+
+Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli.
+Neural signals can then be recorded, and include phase-locked oscillations at the stimulation frequency.
+The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components.
+However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and weak signals might be missed if the phase information is discarded.
+An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components.
+Here we compare the performance of two multivariate extensions of the T-test, Hotelling's $T^2$ and a variant called $T^2_{circ}$.
+We develop a novel test of the assumptions of $T^2_{circ}$ based on the condition index, and a heuristic for excluding outliers using the Mahanalobis distance.
+We then extend the statistic to multi-level designs, and propose a new statistical test termed $ANOVA^2_{circ}$.
+This has identical assumptions to $T^2_{circ}$, and is shown to be more sensitive than MANOVA when these assumptions are met.
+We demonstrate how these tests might be applied to two publicly available empirical data sets, and suggest concrete guidance for choosing which test to run.
+Implementations of these novel tools are provided as _R_ functions, and we hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
+
+# Background
 
@@
 Steady-state methods are used to measure early sensory responses in vision [@Regan1966], hearing [@Rees1986] and somatosensation [@Snyder1992], and closely related paradigms have been developed to target specific stimulus features such as orientation [@Braddick1986], and facial expression [@Gray2020] and identity [@Liu-Shuang2014].
-In fMRI research, _travelling wave_ methods [@Engel1994; @Sereno1995] are used to map the retinotopic responses of early visual cortex, again using stimuli that change periodically, this time in spatial position.
+In fMRI research, _travelling wave_ methods [@Engel1994; @Sereno1995] are used to map the retinotopic responses of early visual cortex using stimuli that change periodically in spatial position.
 Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way [@Spitschan2014].
@@
 
-For pointwise and pairwise comparisons, Hotelling's $T^2$ statistic [@Hotelling1931] is a multivariate extension of the t-test. For the one-sample case, the test statistic is defined as:
+For pointwise and pairwise comparisons, Hotelling's $T^2$ statistic [@Hotelling1931] is a multivariate extension of the T-test. For the one-sample case, the test statistic is defined as:
 
@@
 where _N_ is the number of observations, $\bar{x}$ is the multivariate sample mean, $\mu$ is the point of comparison, $C^{-1}$ is the inverse covariance matrix, and $'$ denotes vector transposition.
-Conceptually, the $T^2$ statistic extends the univariate t-statistic by incorporating the covariance between the dependent variables.
+Conceptually, the $T^2$ statistic extends the univariate T-statistic by incorporating the covariance between the dependent variables.
 Two-sample and paired variants are also available, and the test can be applied to an arbitrary number of dependent variables (though here we will consider only the bivariate case).
@@
 We develop a novel method for testing the assumptions of the $T^2_{circ}$ statistic, based on calculating the condition index of a multivariate data set.
-We then discuss appropriate methods for identifying outliers and calculating error bars.
+Appropriate methods for identifying outliers using the Mahalanobis distance are discussed, and a heuristic proposed.
+Next the logic of $T^2_{circ}$ is extended to situations with more than two levels of the independent variable, and the performance of this novel $ANOVA^2_{circ}$ statistic is compared to MANOVA.
 Finally, we demonstrate the proposed techniques on two example SSVEP data sets, and recommend some best practice guidelines for analysis decisions.
-All data and scripts used to produce this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats).
-These include _R_ functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index statistic proposed in this paper.
-
-## Fourier amplitudes violate parametric assumptions of univariate statistics
-
-Many studies use parametric t-tests and analysis of variance (ANOVA) to analyse periodic data.
+
+All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats).
+These include _R_ functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
+
+# Fourier amplitudes violate parametric assumptions of univariate statistics
+
+Many empirical studies use univariate T-tests or analysis of variance (ANOVA) to analyse periodic data.
 Specifically, the amplitude component of the Fourier spectrum at the stimulation frequency is used as the dependent variable, discarding the phase information.
 This is problematic, because the amplitude is an absolute quantity, and can never fall below zero.
-Distributions of amplitudes for weak signals are therefore likely to be positively skewed, and in general will violate the assumption of normality.
+Distributions of amplitudes for weak signals are therefore positively skewed, and will generally violate the assumption of normality.
 
@@
 These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d > 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]).
-One consequence of this is that t-tests will typically have an inflated Type 1 error (false positive) rate for many signals encountered empirically.
+One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.
 
@@
 Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents.
-Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, as we will demonstrate in the next section.
-
-## Conditions under which $T^2_{circ}$ is more sensitive than $T^2$
-
-Using the $T^2$ and $T^2_{circ}$ allow us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
+Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.
+
+# Conditions under which $T^2_{circ}$ is more sensitive than $T^2$
+
+Using the $T^2$ or $T^2_{circ}$ statistic allows us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
 Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$.
-Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and demonstrate that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f).
+This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant.
+Their simulations show the largest advantage for $T^2_{circ}$ for effect sizes around d = 1 (where the mean signal strength is equal to the standard deviation of the data).
+The advantage appeared to be stronger for smaller sample sizes.
+
+Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and confirm that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f).
 However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations.
-This advantage is lost for large sample sizes (N > 32) and large effect sizes (d > 2 when N > 8)
-
-```{r powerfig, fig.cap=".", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+This advantage is lost for large sample sizes (N > 32) and large effect sizes (d > 2 when N > 8).
+These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for sample sizes, so its use should be restricted to small sample studies.
+
+```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes.
+Panels a-c replicate conditions reported by Victor and Mast (1991).
+Panels (d) and (e) show a wider range of conditions for each statistic.
+Panel (f) shows the difference between the two statistics, with contour lines indicating differences of 0.02, 0.05, 0.1, 0.2 and 0.4.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
 
@@
 text(1.1,3.8,'(d)',cex=2)
-legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white')
+legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white',title='Proportion')
 
@@
 
-## Limitations of $T^2_{circ}$ when assumptions are violated
-
-```{r falsealarms, fig.asp=1.4, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
+# Limitations of $T^2_{circ}$ when assumptions are violated
+
+Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on meeting the test's more stringent assumptions.
+The two variables must be independent, and of equal variance.
+These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not.
+Figure \@ref(fig:falsealarms) shows the results of simulations with randomly generated bivariate data in which no signal is present.
+When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (dashed line).
+However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the variances of the two dependent variables more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2.
+In contrast, the $T^2$ statistic, which explicitly takes account of the covariance matrix (see equation \@ref(eq:t2eq)) shows no increase (black curves).
+
+```{r falsealarms, fig.width=6, fig.height=9, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
 Estimates are for 100000 simulations per condition, with N=10 observations.
@@
 ticklocsx <- seq(-1,1,0.5)    # locations of tick marks on x axis
-ticklocsy <- seq(0,0.1,0.025)    # locations of tick marks on y axis
+ticklocsy <- seq(0,0.1,0.02)    # locations of tick marks on y axis
 ticklabelsx <- ticklocsx        # set labels for x ticks
-ticklabelsy <- ticklocsy*100    # set labels for y ticks
+ticklabelsy <- ticklocsy    # set labels for y ticks
 
@@
 mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
-mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2) 
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
 title(xlab="Correlation (R)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
-title(ylab="Percent significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2, cex.lab=1.5)
 
@@
 ticklocsx <- seq(-1,1,1)    # locations of tick marks on x axis
-ticklocsy <- seq(0,0.1,0.025)    # locations of tick marks on y axis
+ticklocsy <- seq(0,0.1,0.02)    # locations of tick marks on y axis
 ticklabelsx <- c('1/10','1','10')        # set labels for x ticks
-ticklabelsy <- ticklocsy*100    # set labels for y ticks
+ticklabelsy <- ticklocsy    # set labels for y ticks
 
@@
 mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
-mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2) 
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
 title(xlab="Variance ratio (y/x)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
-title(ylab="Percent significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2, cex.lab=1.5)
 
@@
 
-## A novel method to test the assumptions of $T^2_{circ}$
-
-Despite the severe consequences of violating the assumptions of the $T^2_{circ}$ statistic, there is currently no accepted test for those assumptions that could be applied to real data.
-Victor and Mast [-@Victor1991] suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single individual, whereas data from multiple participants may be less likely to exhibit independence of the real and imaginary components [@Pei2017].
+One possible remedy to control the Type I error rate would be to adjust the alpha level to compensate.
+However, this will reduce the statistical power of the $T^2_{circ}$ test, and its advantage over $T^2$ is relatively marginal in most situations to begin with (see Figure \@ref(fig:powerfig)).
+What is required is a method to objectively assess whether the assumptions of $T^2_{circ}$ hold, which we develop in the following section.
+
+# A novel method to test the assumptions of $T^2_{circ}$
+
+Despite the severe consequences of violating the assumptions of the $T^2_{circ}$ statistic (see Figure \@ref(fig:falsealarms)), there is currently no accepted test for those assumptions that could be applied to an empirical data set.
+Victor and Mast [-@Victor1991] suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components [see also @Pei2017].
 
@@
 
-$pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)}$
+\begin{equation}
+\label{eq:edelman1}
+pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)},
+\end{equation}
 
@@
 
-$pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}$
+\begin{equation}
+\label{eq:edelman2}
+pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}
+\end{equation}
 
 In Figure \@ref(fig:distcomparison)a we show an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100000 random data sets for n=10 (blue shading).
-The vertical lines show the critical (95%) threshold for the analytic and simulated results.
+The vertical lines show the critical (95%) threshold for the analytic and simulated results - a ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance.
 Figure \@ref(fig:distcomparison)b shows how these thresholds change as a function of the number of observations, and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
 
-```{r distcomparison, fig.cap='d', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
+```{r distcomparison, fig.cap='Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of 10 observations.
+The vertical blue and red lines show the 95 percent thresholds on the curve (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function.
+Panel (b) shows how 95 percent thresholds change as a functio of the number of observations.', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
 
@@
 mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2) 
-title(xlab="Eigenvalue ratio", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
+title(xlab="Condition index", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
 title(ylab="Probability", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
@@
 lines(c(simcrit,simcrit),c(0,1),col=nbdtpal[3],lwd=2,lty=2)
+text(4.8,0.95,'(a)',cex=1.5)
 
@@
 lines(3:25,simratios,lwd=2,col=nbdtpal[3])
-legend(12, 25, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=1, lwd=3, box.lwd=2)
+legend(12, 15, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=1, lwd=3, box.lwd=2)
+text(23,24,'(b)',cex=1.5)
 
@@
 If the test is non-significant, one can proceed with $T^2_{circ}$; if it is significant, $T^2$ should be used instead.
-
-## Identifying and removing outliers using the Mahalanobis distance
+An _R_ function implementing this test is included in the project code repository (the _CI.test_ function contained in the _CItest.R_ script).
+
+# Identifying and removing outliers using the Mahalanobis distance
 
@@
 In all cases, the functions depart from the Type 1 error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean.
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic return $D^2$, which can be converted to D by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the $T^2_{circ}$ test.
 
-```{r outlierplot, fig.cap='d', fig.align="center", echo=FALSE}
+```{r outlierplot, fig.cap='Simulations illustrating the Mahalanobis distance metric, and showing how a single outlier affects the condition index.
+The upper row shows three example data sets, each with a single outlier shown in red.
+The outliers have Mahalanobis distances of 1, 3 and 5.
+The ellipses are calculated with the outlier included (red) and excluded (black), illustrating how the outlier distorts the aspect ratio of the ellipse.
+The lower plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.', fig.align="center", fig.width=7, fig.height=6, echo=FALSE}
+
+set.seed(2112)
+nobs <- 16
+mdists <- c(1,3,5)
+xlocs <- c(1,3.5,6)
+
+plotlims <- c(0,8,0,2) 
+ticklocsx <- seq(0,8,1)    # locations of tick marks on x axis
+ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Outlier Mahalanobis distance (D)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+mtext(text = "Proportion significant", side = 2, at=0.5, line=1.5, cex=1.5)
+points(0,2,pch=15,col=rgb(0.99,0.99,0.99),cex=0.1)
+
+  xdata <- rnorm(nobs,mean=0,sd=1)
+  ydata <- rnorm(nobs,mean=0,sd=1)
+  compdata <- data.frame(xdata,ydata)
+  ctr <- colMeans(compdata)
+  ellRot <- getel(compdata)
+
+  for (m in 1:3){
+  xdata[1] <- mdists[m]
+  ydata[1] <- 0
+  compdata <- data.frame(xdata,ydata)
+  ctr2 <- colMeans(compdata)
+  ellRot2 <- getel(compdata)
+
+  polygon(0.4*(ellRot2+ctr2)[1, ]+xlocs[m], 0.1*(ellRot2+ctr2)[2, ]+1.5, col=addalpha(nbdtpal[2],0.3),border=NA)
+  lines(0.4*(ellRot+ctr)[1, ]+xlocs[m], 0.1*(ellRot+ctr)[2, ]+1.5, lwd=1)
+  lines(0.4*c(ctr2[1],xdata[1])+xlocs[m],0.1*c(ctr2[2],ydata[1])+1.5,col=nbdtpal[2],lwd=3)
+  points(0.4*xdata+xlocs[m],0.1*ydata+1.5,pch=16,cex=0.5,col=nbdtpal[3])
+  points(0.4*ctr2[1]+xlocs[m],0.1*ctr2[2]+1.5,pch=16,cex=1,col=nbdtpal[1])
+  points(0.4*xdata[1]+xlocs[m],0.1*ydata[1]+1.5,pch=16,cex=1,col=nbdtpal[2])
+}
+
+  text(1,1.8,'D = 1',cex=1.5,adj=0.5)
+  text(3.5,1.8,'D = 3',cex=1.5,adj=0.5)
+  text(6,1.8,'D = 5',cex=1.5,adj=0.5)
 
@@
 
-plotlims <- c(0,8,0,1) 
-ticklocsx <- seq(0,8,1)    # locations of tick marks on x axis
+lines(c(0,8),c(0.05,0.05),lty=2,lwd=2)
+
+for (s in 1:length(samplesizes)){
+  lines(dlist,propsig[s,],lwd=3,col=nbdtpal[s])
+}
+
+legend(0,1,samplesizes,col=nbdtpal,lwd=3,box.lwd=2,title='N observations')
+
+```
+
+A variant of the Mahalanobis distance (the pairwise Mahalanobis distance) can also be used to compute a multivariate measure of effect size, equivalent to Cohen's _d_ statistic [see e.g.
+@Giudice2009].
+This is a valuable statistic to include when reporting the results of multivariate tests, and we include an _R_ function (_mahalES_) that can be used to calculate it.
+
+# Generalising to more than two conditions
+
+Many studies involve more than two conditions that need to be compared.
+Issues with familywise error will quickly become problematic if multiple pairwise $T^2$ or $T^2_{circ}$ statistics are calculated.
+One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's $T^2$, but permits multiple independent variables with more than two levels.
+However, if the assumptions of $T^2_{circ}$ hold for a data set, it might alternatively be possible to extend to logic of the $T^2_{circ}$ test [@Victor1991] to the more general case.
+
+The F statistic for a one-way ANOVA is calculated by taking the ratio of the variance explained by the modelled group means to the residual unexplained variance.
+For multivariate data, the change in group mean would be calculated using the vector distances between the complex means, and the residuals are the vector distances between each data point and its corresponding group mean.
+For an independent one-way design with $k$ groups (or conditions) and $N$ observations per group, the F distribution will have $2(k-1)$ and $2((Nk)-k)$ degrees of freedom.
+The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number.
+A suitable name for such a test might be ${ANOVA}^2_{circ}$, as this reflects the similarity to ANOVA, and the extension of the logic of $T^2_{circ}$ (an alternative name might be $MANOVA_{circ}$, however this feels less appropriate given that many of the key features of MANOVA are absent).
+
+Figure \@ref(fig:powerfig2) shows simulations analogous to those in Figure \@ref(fig:powerfig) for a one-way design with three levels.
+MANOVA is directly compared to the ${ANOVA}^2_{circ}$ statistic across a range of effect sizes and sample sizes.
+Just as for the one-sample statistics, the advantages of ${ANOVA}^2_{circ}$ are particularly apparent for small sample sizes, and larger effect sizes (Figure \@ref(fig:powerfig2)f).
+
+```{r powerfig2, fig.cap="Simulations comparing the sensitivity of MANOVA and ANOVA2circ.
+The format mirrors that of Figure 2.
+In these simulations, there were three conditions, with the signal being added to one condition.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+
+layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))
+abc <- c('a','b','c')
+
+vinds <- c(2,5,17)
+dlist <- seq(0,4,0.25)
+slist <- seq(1.5,5,0.5)
+samplesizes <- ceiling(2^slist)
+
+manovamatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
+circmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
+for (n in 1:nsims){
+  for (s in 1:length(samplesizes)){
+    grouplabels <- as.factor(rep(1:3,each=samplesizes[s]))
+
+    for (s2 in 1:length(dlist)){
+
+  data <- matrix(rnorm(3*2*samplesizes[s]),nrow=3*samplesizes[s],ncol=2)
+  data[1:samplesizes[s],1] <- data[1:samplesizes[s],1] + dlist[s2]
+
+  simdata <- complex(real=data[,1],imaginary=data[,2])
+
+  output <- anovacirc.test(data.frame(simdata,grouplabels))
+  if (output$pvalue<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
+
+dataforManova <- data.frame(grouplabels,data[,1],data[,2])    
+colnames(dataforManova) <- c("Group","Real","Imaginary")
+outcome <- cbind(dataforManova$Real, dataforManova$Imaginary)
+manout <- summary(manova(outcome ~ Group, data = dataforManova))
+  if (manout$stats[1,6]<0.05){manovamatrix[s,s2] <- manovamatrix[s,s2] + 1}
+
+  }}
+}
+
+manovapropM <- manovamatrix/nsims
+circpropM <- circmatrix/nsims
+manvscirc <- circpropM - manovapropM
+
+plotlims <- c(1,5,0,1)
+ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
 ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
+ticklabelsx <- 2^ticklocsx        # set labels for x ticks
+ticklabelsy <- ticklocsy    # set labels for y ticks
+
+for (n in 1:3){
+par(pty="s")
 plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
@@
 axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2.2, cex.lab=1.5)
+title(paste('d =',dlist[vinds[n]]),cex.main=2)
+lines(slist,manovapropM[,vinds[n]],lwd=3,col=nbdtpal[1])
+lines(slist,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])
+
+points(slist,manovapropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
+points(slist,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])
+
+text(1.1,0.95,paste('(',abc[n],')',sep=''),cex=2)
+
+if (n==1){legend(3,1,c('MANOVA',expression('ANOVA'[circ]^2)),pch=15:16,cex=1.5,col=nbdtpal[1:2],box.lwd=2)}
+}
+
+par(pty="s")
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
+ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
+ticklocsy <- seq(0,4,1)    # locations of tick marks on y axis
+ticklabelsx <- 2^ticklocsx        # set labels for x ticks
+ticklabelsy <- seq(0,4,1)    # set labels for y ticks
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(expression('MANOVA'),cex.main=2)
+image(slist,dlist,manovapropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+text(1.1,3.8,'(d)',cex=2)
+legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white',title='Proportion')
+
+par(pty="s")
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+ttext <- expression('ANOVA'[circ]^2)
+title(ttext,cex.main=2)
+image(slist,dlist,circpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+text(1.1,3.8,'(e)',cex=2)
+
+par(pty="s")
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
+mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
+title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
+title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+ttext <- expression({'ANOVA'[circ]^2}~-~{MANOVA})
+title(ttext,cex.main=2)
+image(slist,dlist,manvscirc,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+cl <- contourLines(slist,dlist,manvscirc,levels=c(0.02,0.05,0.1,0.2,0.4))
+for (n in 1:length(cl)){
+  temp <- cl[n]
+  lines(temp[[1]]$x,temp[[1]]$y,col='black',lwd=3)
+}
+text(1.1,3.8,'(f)',cex=2)
+
+```
+
+Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that standard Bonferroni corrections are applied.
+Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary).
+A repeated measures version of ${ANOVA}^2_{circ}$ has also been implemented following the same logic.
+The project repository contains _R_ scripts for both of these functions.
+
+# Deciding which test to run
+
+The flowchart in Figure \@ref(fig:flowchart) contains a proposed decision structure for the analysis of periodic data.
+Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios.
+Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, and the $T^2$ statistic if not.
+Comparisons with more than two conditions should be tested with the $ANOVA^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not.
+Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs.
+In _R_, the _Manova_ function from the _car_ package appears to be capable of the necessary calculations, as does the commercial software package SPSS.
+
+```{r flowchart, fig.cap = 'Flowchart illustrating how one might decide which test to conduct for a given data set, based on the study design and the outcome of the condition index test.', fig.align="center", echo=FALSE}
+
+pos <- coordinates(c(1,2,4))
+plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,1), ylim=c(0,1))
+
+mid <- straightarrow(from = pos[1, ], to = pos[2, ], lty = 1)
+text(mid[1]-0.05, mid[2]+0.05, "N <= 2", cex = 0.8)
+mid <- straightarrow(from = pos[1, ], to = pos[3, ], lty = 1)
+text(mid[1]+0.05, mid[2]+0.05, "N > 2", cex = 0.8)
+mid <- straightarrow(from = pos[2, ], to = pos[4, ], lty = 1)
+text(mid[1]-0.05, mid[2]+0.01, "Yes", cex = 0.8)
+mid <- straightarrow(from = pos[2, ], to = pos[5, ], lty = 1)
+text(mid[1]+0.05, mid[2]+0.01, "No", cex = 0.8)
+mid <- straightarrow(from = pos[3, ], to = pos[6, ], lty = 1)
+text(mid[1]-0.05, mid[2]+0.01, "Yes", cex = 0.8)
+mid <- straightarrow(from = pos[3, ], to = pos[7, ], lty = 1)
+text(mid[1]+0.05, mid[2]+0.01, "No", cex = 0.8)
+
+textellipse(mid = pos[1,], radx = 0.1, rady = 0.1, lab = 'How many levels?', cex = 0.5, shadow.col = "grey")
+textellipse(mid = pos[2,], radx = 0.1, rady = 0.1, lab = 'Eigenvalues equal?', cex = 0.5, shadow.col = "grey")
+textellipse(mid = pos[3,], radx = 0.1, rady = 0.1, lab = 'Eigenvalues equal?', cex = 0.5, shadow.col = "grey")
+textellipse(mid = pos[4,], radx = 0.1, rady = 0.1, lab = expression(T[circ]^2), cex = 0.5, shadow.col = "grey")
+textellipse(mid = pos[5,], radx = 0.1, rady = 0.1, lab = expression("Hotelling's T"^2), cex = 0.5, shadow.col = "grey")
+textellipse(mid = pos[6,], radx = 0.1, rady = 0.1, lab = expression(ANOVA[circ]^2), cex = 0.5, shadow.col = "grey")
+textellipse(mid = pos[7,], radx = 0.1, rady = 0.1, lab = 'MANOVA', cex = 0.5, shadow.col = "grey")
+
+```
+
+# Applying these methods to empirical data sets
+
+Having developed some novel tools for the analysis of periodic data, in the following sections we demonstrate them using two different publicly available empirical data sets.
+The first study recorded responses to auditory and optogenetic stimulation in mice.
+The second study measured visual responses to flickering grating patterns in humans.
+These cases provide examples of how the results of the tests described above might be appropriately reported.
+
+## Mouse auditory and optogenetic steady-state data
+
+Hwang et al.
+[-@Hwang2019; -@Hwang2020] measured steady-state responses using implanted scalp electrodes in 6 mice.
+The mice had previously been given a targeted virus that made parvalbumin neurons in their basal forebrain responsive to specific wavelengths of light, delivered through an optical fiber (a technique called optogenetics).
+SSVEPs were recorded from 36 electrodes for 1 second epochs of 40 Hz auditory stimulation, and various schedules of optogenetic stimulation (including at 40 Hz).
+The data set is described more fully by Hwang et al.
+[-@Hwang2020], and was downloaded from: [https://doi.gin.g-node.org/10.12751/g-node.e5tyek/](https://doi.gin.g-node.org/10.12751/g-node.e5tyek/).
+
+Figure \@ref(fig:mousedata)a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz.
+There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
+Indeed, a paired t-test on the amplitudes reveals no significant difference (t=0.84, df=5, p=0.44).
+However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b).
+The condition index test was non-significant for both conditions (sound: CI = 1.59, _p_ = 0.66; light: CI = 1.69, _p_ = 0.59), so a paired-samples $T^2_{circ}$ test was conducted.
+This revealed a significant difference between conditions ($T^2_{circ}$ = 1.39, $F_{(2,10)}$ = 8.32, _p_ = 0.007) with an effect size of D = 1.02.
+This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
+[-@Hwang2019] went on to explore interactions between these two signals.
+
+```{r mousedata, fig.cap = 'Summary of mouse steady-state responses to 40 Hz stimulation.
+Panel (a) shows the Fourier amplitude spectrum with inset scalp plots for sound (red) and light (blue) stimulation, averaged across repetitions and individuals.
+Panel (b) shows complex (x = real, y = imaginary) Fourier components for 6 individual mice (small points) and their average (large points) for both conditions.', fig.width=10, fig.height=6, fig.align="center", echo=FALSE}
+
+load('Hwangdata.RData')
+
+frequencies <- 1:100
+f1 <- which(frequencies==40)
+
+meandata <- abs(apply(mousedata,c(2,3,4),mean))
+
+temp <- rowMeans(mousedata[,1:2,11,f1])
+xdata <- Re(temp)
+ydata <- Im(temp)
+temp2 <- rowMeans(mousedata[,1:2,12,f1])
+xdata2 <- Re(temp2)
+ydata2 <- Im(temp2)
+
+plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,150), ylim=c(0,2))
+ticklocsx <- seq(0,80,20)    # locations of tick marks on x axis
+ticklocsy <- seq(0,2,0.5)    # locations of tick marks on y axis
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
 mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
-mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2, las=1) 
-title(xlab="Outlier Mahalanobis distance", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
-title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
-
-lines(c(0,8),c(0.05,0.05),lty=2,lwd=2)
-
-for (s in 1:length(samplesizes)){
-  lines(dlist,propsig[s,],lwd=3,col=nbdtpal[s])
-}
-
-legend(0,1,samplesizes,col=nbdtpal,lwd=3,box.lwd=2,title='N observations')
+mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2, las=1)  # the 'line' command moves away from the axis, the 'las' command rotates to vertical
+title(ylab="Amplitude (µV)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+mtext('Frequency (Hz)',side=1,at=40,line=1.2,cex=1.5)
+
+rmax <- 7   #specify a maximum boundary for the grid
+gridRes <- 100 #specify the interpolation grid resolution
+ramp <- colorRamp(c("white", rgb(187/255,40/255,107/255)))  # create a ramp from one colour to another
+colmatrix2 <- rgb(ramp(seq(0, 1, length = 100)), max = 255)   # index the ramp at ten points
+
+datatoplot <- abs(meandata[,11,f1])
+testDat<- data.frame(x = montage[,2],
+                     y = -montage[,3],
+                     z = datatoplot)
+
+xo <- seq(min(-rmax, testDat$x), max(rmax, testDat$x), length = gridRes)
+yo <- seq(max(rmax, testDat$y), min(-rmax, testDat$y), length = gridRes)
+
+interpV4 <- v4Interp(testDat, xo, yo, rmax, gridRes)
+
+zo2 <- as.matrix(interpV4[,2:ncol(interpV4)])
+xo2 <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
+yo2 <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
+zo2[which(zo2>2)] <- 2
+
+image(2.4*xo+20,1.5+xo/14,zo2,zlim=c(0,2),col=colmatrix2,add=TRUE,useRaster=TRUE)
+
+rmax2 <- rmax+0.1
+maskx <- outline[which(outline[,1]>=0),1]
+masky <- outline[which(outline[,1]>=0),2]
+polygon(2.4*c(0,maskx,0,rmax2,rmax2,0)+20,1.5+c(rmax2,masky,-rmax2,-rmax2,rmax2,rmax2)/14,border=NA,col="white")
+maskx <- outline[which(outline[,1]<=0),1]
+masky <- outline[which(outline[,1]<=0),2]
+i <- sort(masky,index.return=TRUE)
+maskx <- maskx[i$ix]
+masky <- masky[i$ix]
+polygon(2.4*c(0,maskx,0,-rmax2,-rmax2,0)+20,1.5+c(-rmax2,masky,rmax2,rmax2,-rmax2,-rmax2)/14,border=NA,col="white")
+
+lines(2.4*outline[,1]+20,1.5+outline[,2]/14,lwd=2)
+points(2.4*montage[3:38,2]+20,1.5+montage[3:38,3]/14,pch=16,col='grey')
+points(2.4*montage[1:2,2]+20,1.5+montage[1:2,3]/14,pch=16)
+
+ramp <- colorRamp(c("white", rgb(56/255,111/255,164/255)))  # create a ramp from one colour to another
+colmatrix2 <- rgb(ramp(seq(0, 1, length = 100)), max = 255)   # index the ramp at ten points
+
+datatoplot <- abs(meandata[,12,f1])
+testDat<- data.frame(x = montage[,2],
+                     y = -montage[,3],
+                     z = datatoplot)
+
+xo <- seq(min(-rmax, testDat$x), max(rmax, testDat$x), length = gridRes)
+yo <- seq(max(rmax, testDat$y), min(-rmax, testDat$y), length = gridRes)
+
+interpV4 <- v4Interp(testDat, xo, yo, rmax, gridRes)
+
+zo2 <- as.matrix(interpV4[,2:ncol(interpV4)])
+xo2 <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
+yo2 <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
+zo2[which(zo2>2)] <- 2
+
+image(2.4*xo+60,1.5+xo/14,zo2,zlim=c(0,2),col=colmatrix2,add=TRUE,useRaster=TRUE)
+
+rmax2 <- rmax+0.1
+maskx <- outline[which(outline[,1]>=0),1]
+masky <- outline[which(outline[,1]>=0),2]
+polygon(2.4*c(0,maskx,0,rmax2,rmax2,0)+60,1.5+c(rmax2,masky,-rmax2,-rmax2,rmax2,rmax2)/14,border=NA,col="white")
+maskx <- outline[which(outline[,1]<=0),1]
+masky <- outline[which(outline[,1]<=0),2]
+i <- sort(masky,index.return=TRUE)
+maskx <- maskx[i$ix]
+masky <- masky[i$ix]
+polygon(2.4*c(0,maskx,0,-rmax2,-rmax2,0)+60,1.5+c(-rmax2,masky,rmax2,rmax2,-rmax2,-rmax2)/14,border=NA,col="white")
+
+lines(2.4*outline[,1]+60,1.5+outline[,2]/14,lwd=2)
+points(2.4*montage[3:38,2]+60,1.5+montage[3:38,3]/14,pch=16,col='grey')
+points(2.4*montage[1:2,2]+60,1.5+montage[1:2,3]/14,pch=16)
+
+lines(frequencies[1:79],meandata[1,11,1:79],col=nbdtpal[2],lwd=3)
+lines(frequencies[1:79],meandata[1,12,1:79],col=nbdtpal[3],lwd=3)
+lines(c(90,150),c(1,1),lwd=2)
+lines(c(120,120),c(0,2),lwd=2)
+
+ellRot1 <- getel(data.frame(xdata,ydata))
+ellRot2 <- getel(data.frame(xdata2,ydata2))
+polygon(10*(ellRot1[1, ]+mean(xdata))+120, 1+(ellRot1[2, ]+mean(ydata))/3, col=addalpha(nbdtpal[2],0.3),border=NA)
+polygon(10*(ellRot2[1, ]+mean(xdata2))+120, 1+(ellRot2[2, ]+mean(ydata2))/3, col=addalpha(nbdtpal[3],0.3),border=NA)
+
+points(10*xdata+120,1+ydata/3,pch=16,col=nbdtpal[2])
+points(10*xdata2+120,1+ydata2/3,pch=16,col=nbdtpal[3])
+points(10*mean(xdata)+120,1+mean(ydata)/3,pch=21,bg=nbdtpal[2],cex=1.5)
+points(10*mean(xdata2)+120,1+mean(ydata2)/3,pch=21,bg=nbdtpal[3],cex=1.5)
+
+legend(130,0.6,c('Sound','Light'),pch=21,pt.cex=1.5,pt.bg=nbdtpal[2:3],box.lwd=2)
+
+text(20,1.93,'Sound',cex=1.5,adj=0.5)
+text(60,1.93,'Light',cex=1.5,adj=0.5)
+
+text(0,1.95,'(a)',cex=2,adj=0.5)
+text(90,1.95,'(b)',cex=2,adj=0.5)
+
+text(149,0.94,'3',cex=1.5,adj=0.5)
+text(122,1.97,'3',cex=1.5,adj=0.5)
+
+# totest <- data.frame(xdata,ydata)
+# totest2 <- data.frame(xdata2,ydata2)
+# CI.test(totest)
+# CI.test(totest2)
+# tsqc.test(totest,y=totest2,paired=TRUE)
+# 
+# abs1 <- abs(rowMeans(mousedata[,1:2,11,f1]))
+# abs2 <- abs(rowMeans(mousedata[,1:2,12,f1]))
+# t.test(abs1,abs2,paired=TRUE)
+
+# colnames(totest2) <- colnames(totest)
+# fores <- rbind(totest,totest2)
+# mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=6))
+# D <- sqrt(mahala_sq$distance[1,2])
 
@@
 
-## Methods for calculating error bars for Fourier amplitudes
-
-## Applying these methods to real data sets
-
-### Human visual steady-state data
-
-### Mouse auditory steady-state data
-
-https://www.nature.com/articles/s41597-020-00621-z
-https://doi.gin.g-node.org/10.12751/g-node.e5tyek/
-
-## Recommendations for analysing periodic data
-
-Multi/Bivariate statistics should be used for phase-locked Fourier data
-
-Always remove outliers with a Mahalanobis distance > 3
-
-Always test assumptions if using $T^2_{circ}$
-
-No benefit to using tsqcirc for sample sizes>32
-
-With more than two conditions use MANOVA
-
-Use bootstrapping to calculate error bars
-
-## Further considerations
-
-Mention about non phase-locked responses
-
-Greater power with coherent averaging
-
-Could develop ANOVA-circ
-
-## References
+## Human visual steady-state data
+
+Vilidaite et al.
+[-@Vilidaite2018] measured visual responses to flickering grating stimuli in a large sample of 100 adults.
+Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker).
+Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \@ref(fig:humanSSVEP)), were well-isolated in the Fourier domain, and increased with stimulus contrast.
+For this analysis, we took responses from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant.
+Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \@ref(fig:humanSSVEP).
+Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.
+
+```{r humanSSVEP, fig.cap = 'Summary of human SSVEP data.
+Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes).
+Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point).
+Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.', fig.width=14, fig.height=5.4, fig.align="center", echo=FALSE}
+
+load("/Users/danbaker/Google Drive/Current work/Research/powercontours/headplotspec.RData")
+
+load('~/Google Drive/Current work/R scripts/danlabtoolbox/data/montagedata.RData')
+
+subjmeans <- apply(alltarget[,31,,],1:2,mean)
+
+rmax <- 0.55   #specify a maximum boundary for the grid
+gridRes <- 100 #specify the interpolation grid resolution
+# colmatrix2 <- kovesi.linear_kry_5_95_c72(101)
+ramp <- colorRamp(c("white", rgb(56/255,111/255,164/255)))
+colmatrix2 <- rgb(ramp(seq(0, 1, length = 100)), max = 255)   # index the ramp at ten
+
+selectrodes <- 'Oz'
+montagelabels <- unlist(montage$labels)
+blackelectrodes <- match(toupper(selectrodes),montagelabels,nomatch=0)
+
+plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,8), ylim=c(0,2))
+nexcluded <- NULL
+isincluded <- 1 + 0*(1:100)
+
+for (cond in 1:7){
+datatoplot <- abs(apply(alltarget[,1:64,cond,], 2, mean))
+datatoplot[c(13,19)] <- mean(datatoplot)   # set the mastoids to 0
+testDat<- data.frame(x = montage$electrodelocs[montage$channelmappings,1],
+                     y = -montage$electrodelocs[montage$channelmappings,2],
+                     z = datatoplot)
+
+xo <- seq(min(-rmax, testDat$x), max(rmax, testDat$x), length = gridRes)
+yo <- seq(max(rmax, testDat$y), min(-rmax, testDat$y), length = gridRes)
+
+interpV4 <- v4Interp(testDat, xo, yo, rmax, gridRes)
+
+zo2 <- as.matrix(interpV4[,2:ncol(interpV4)])
+xo2 <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
+yo2 <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
+outsidecircle <- sqrt(xo2^2 + yo2^2) > 0.5
+zo2[outsidecircle] <- 0
+
+image(cond+xo*0.8,1.5+xo*0.8,zo2,zlim=c(0,0.45),col=colmatrix2,add=TRUE,useRaster=TRUE)
+
+points(cond+montage$electrodelocs[blackelectrodes,1]*0.8,1.5+montage$electrodelocs[blackelectrodes,2]*0.8,pch=16,col="black",cex=1.5)
+
+lines(cond+montage$headoutline[,1]*0.8,1.5+montage$headoutline[,2]*0.8,col="black",lwd=2)
+lines(cond+montage$noseoutline[,1]*0.8,1.5+montage$noseoutline[,2]*0.8,col="black",lwd=2)
+lines(cond+montage$Rearoutline[,1]*0.8,1.5+montage$Rearoutline[,2]*0.8,col="black",lwd=2)
+lines(cond+montage$Learoutline[,1]*0.8,1.5+montage$Learoutline[,2]*0.8,col="black",lwd=2)
+
+lines(c(cond,cond),c(0.05,0.95))
+lines(c(cond-0.45,cond+0.45),c(0.5,0.5))
+
+xvals <- Re(subjmeans[,cond])/5
+yvals <- Im(subjmeans[,cond])/5
+points(xvals+cond,yvals+0.5,pch=16,col=addalpha(nbdtpal[1],alpha=0.2),cex=0.5)
+
+bdata <- data.frame(xvals,yvals)
+D <- sqrt(mahalanobis(bdata, colMeans(bdata), cov(bdata)))
+i <- which(D>3)
+points(xvals[i]+cond,yvals[i]+0.5,pch=16,col=nbdtpal[2],cex=0.5)
+points(mean(xvals[-i])+cond,mean(yvals[-i])+0.5,pch=16,col=nbdtpal[3],cex=1)
+nexcluded[cond] <- length(i)
+isincluded[i] <- 0
+}
+
+cleandata <- subjmeans[which(isincluded==1),]
+
+contrastlevels <- c(0,2,4,8,16,32,64)
+
+for (cond in 1:7){
+  xvals <- Re(cleandata[,cond])/5
+  yvals <- Im(cleandata[,cond])/5
+  ellRot <- getel(data.frame(xvals,yvals))
+  lines((ellRot[1, ]+mean(xvals))+cond, (ellRot[2, ]+mean(yvals))+0.5)
+
+  text(cond,1,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
+# totest <- data.frame(xvals,yvals)
+# print(CI.test(totest))
+}
+
+# grouplabels <- as.factor(rep(1:7,each=89))
+# participant <- as.factor(rep(1:89,times=7))
+# a <- melt(cleandata[,1:7])
+# simdata <- a[,3]
+# withincirc <- anovacircR.test(data.frame(grouplabels,simdata,participant))
+
+# for (cond in 2:7){
+#   set1 <- data.frame(Re(cleandata[,1]),Im(cleandata[,1]))
+#   set2 <- data.frame(Re(cleandata[,cond]),Im(cleandata[,cond]))
+# print(tsqc.test(set1,set2,paired=TRUE))
+#   
+# colnames(set1) <- c('x','y')
+# colnames(set2) <- c('x','y')
+# fores <- rbind(set1,set2)
+# mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=89))
+# D <- sqrt(mahala_sq$distance[1,2])
+# print(D)
+# }
+
+```
+
+With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.2, all _p_ > 0.19).
+A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(12,1056)}$ = 38.9, p < 0.001).
+Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = 0.32, $F_{(2,176)}$ = 28.4, D = 0.13, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = 0.28, $F_{(2,176)}$ = 25.3, D = 0.19, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = 0.10, $F_{(2,176)}$ = 8.5, D = 0.14, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = 0.34, $F_{(2,176)}$ = 35.3, D = 0.33, _p_ < 0.001).
+The study by Vilidaite et al.
+[-@Vilidaite2018] compared SSVEP responses between individuals with and without autism, as well as in a _Drosophila_ genetic model of developmental disorders.
+The raw data are available at: [http://dx.doi.org/10.17605/OSF.IO/Y4N5K](http://dx.doi.org/10.17605/OSF.IO/Y4N5K).
+
+# Further considerations
+
+It is worth pointing out that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations.
+This is the case for most paradigms in which the nervous system is driven by a periodic stimulus.
+However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms [@Berger1929; @Buzsaki2004], which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals [e.g.
+@Canolty2010].
+When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average [@Baker2020].
+This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.
+
+The present paper has focussed on the Frequentist statistical tradition.
+However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses.
+Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests [@Rouder2009].
+However this is a non-trivial undertaking, and is beyond the scope of the current paper.
+
+Another possibility is to use machine learning techniques such as multivariate pattern analysis (MVPA) to analyse periodic data.
+This involvess training a classifier algorithm to distinguish between two (or more) experimental conditions or states, and then assessing classifier accuracy on fresh data.
+If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance.
+Such methods have been hugely influential in the fMRI literature, and for analysing event-related potential data collected using EEG or MEG.
+However, they have not been widely applied to steady-state data (though see West et al.
+[-@West2015] for an example).
+In principle, the real and imaginary Fourier components can be treated as separate dependent variables, along with different recording locations and/or frequencies.
+This approach has the potential to offer sensitive, high-powered statistical tests that circumvent many of the shortcomings associated with parametric statistics.
+
+Even when statistics are conducted using both the real and imaginary Fourier components, it is still typical to visualise the mean amplitudes.
+Several approaches to calculating appropriate error bars have been proposed.
+For example, Pei et al.
+[-@Pei2017] suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
+This approach is somewhat computationally demanding, though we have implemented an _R_ function to calculate error bars using this method.
+However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95% confidence intervals on amplitudes.
+This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean.
+The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled amplitudes.
+Upper and lower 95% confidence intervals on the amplitude can then be taken at appropriate quantiles of this population.
+
+# General recommendations for analysing periodic data
+
+The simulations reported here allow us to make several recommendations for how periodic data should be analysed.
+Multivariate statistics should be used for phase-locked Fourier data instead of univariate statistics such as T-tests and ANOVA.
+This avoids problems from distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information.
+Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahanalobis distance reported as a measure of effect size.
+For sample sizes of N < 32, the $T^2_{circ}$ and $ANOVA^2_{circ}$ statistics can be used if the condition index test is non-significant for all conditions.
+Alternatively, the $T^2$ or MANOVA statistics should be used when these conditions are not met.
+The greater power afforded by these tests should in general lead to more accurate statistical inferences when analysing periodic data.
+
+# References

### 2020-12-23 manuscript.tex
Completed first draft of manuscript

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Statistical analysis of periodic data in neuroscience},
            pdfauthor={Daniel H. Baker},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{13/12/2020}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals can then be recorded, and include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components. However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here we compare the performance of two multivariate extensions of the T-test, Hotelling's \(T^2\) and a variant called \(T^2_{circ}\). We develop a novel test of the assumptions of \(T^2_{circ}\) based on the condition index, and a heuristic for excluding outliers using the Mahanalobis distance. We then extend the statistic to multi-level designs, and propose a new statistical test termed \(ANOVA^2_{circ}\). This has identical assumptions to \(T^2_{circ}\), and is shown to be more sensitive than MANOVA when these assumptions are met. We demonstrate how these tests might be applied to two publicly available empirical data sets, and suggest concrete guidance for choosing which test to run. Implementations of these novel tools are provided as \emph{R} functions, and we hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

\hypertarget{background}{%
\section{Background}\label{background}}

A widely used paradigm in many branches of neuroscience is to drive the nervous system using periodic stimuli. This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations. These periodic responses, often called the \emph{steady-state} or \emph{frequency following} response, can be recorded from single neurons (Enroth-Cugell and Robson 1966) and local field potentials (Morrone, Burr, and Speed 1987) using invasive methods, or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans (Norcia et al. 2015) and in diverse animal species including insects (Afsari et al. 2014), birds (Porciatti et al. 1990), rodents (Hwang et al. 2019) and primates (Nakayama and Mackeben 1982). Steady-state methods are used to measure early sensory responses in vision (Regan 1966), hearing (Rees, Green, and Kay 1986) and somatosensation (Snyder 1992), and closely related paradigms have been developed to target specific stimulus features such as orientation (Braddick, Wattam-Bell, and Atkinson 1986), and facial expression (Gray et al. 2020) and identity (Liu-Shuang, Norcia, and Rossion 2014). In fMRI research, \emph{travelling wave} methods (Engel et al. 1994; Sereno et al. 1995) are used to map the retinotopic responses of early visual cortex using stimuli that change periodically in spatial position. Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way (Spitschan et al. 2014).

A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal. The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a convenient, and well-isolated index of the brain's response. Fourier spectra comprise both amplitude and phase information, which can be equivalently expressed as complex numbers with real and imaginary components. In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data only. However an alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary components). Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods.

For pointwise and pairwise comparisons, Hotelling's \(T^2\) statistic (Hotelling 1931) is a multivariate extension of the T-test. For the one-sample case, the test statistic is defined as:

\begin{equation}
\label{eq:t2eq}
T^2 = N(\bar{x} - \mu)' C^{-1} (\bar{x} - \mu),
\end{equation}

where \emph{N} is the number of observations, \(\bar{x}\) is the multivariate sample mean, \(\mu\) is the point of comparison, \(C^{-1}\) is the inverse covariance matrix, and \('\) denotes vector transposition. Conceptually, the \(T^2\) statistic extends the univariate T-statistic by incorporating the covariance between the dependent variables. Two-sample and paired variants are also available, and the test can be applied to an arbitrary number of dependent variables (though here we will consider only the bivariate case).

More recently, Victor and Mast (1991) proposed a simpler version of \(T^2\), called \(T^2_{circ}\). The \(T^2_{circ}\) statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance. When these conditions are met, the test statistic is defined as:

\begin{equation}
\label{eq:t2c}
T^2_{circ} = (N-1)\frac{|\bar{x}-\mu|^2}{\Sigma|x_j - \bar{x}|^2}
\end{equation}

where \(x_j\) denotes the \(j\)th observation of the dependent variables, and all other terms retain their previous meanings. Notice that no covariance term is present in equation \eqref{eq:t2c}, because of the independence assumption. This makes the statistic simpler to calculate, but causes problems when the assumption is violated.

In the present paper, we develop best practice guidelines for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms. We first demonstrate why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed. We then investigate conditions under which we should use either the \(T^2\) or \(T^2_{circ}\) statistic. The range of sample sizes and effect sizes where \(T^2_{circ}\) is more sensitive is identified. We develop a novel method for testing the assumptions of the \(T^2_{circ}\) statistic, based on calculating the condition index of a multivariate data set. Appropriate methods for identifying outliers using the Mahalanobis distance are discussed, and a heuristic proposed. Next the logic of \(T^2_{circ}\) is extended to situations with more than two levels of the independent variable, and the performance of this novel \(ANOVA^2_{circ}\) statistic is compared to MANOVA. Finally, we demonstrate the proposed techniques on two example SSVEP data sets, and recommend some best practice guidelines for analysis decisions.

All scripts used to generate this manuscript are available at: \url{https://github.com/bakerdh/FourierStats}. These include \emph{R} functions to implement one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.

\hypertarget{fourier-amplitudes-violate-parametric-assumptions-of-univariate-statistics}{%
\section{Fourier amplitudes violate parametric assumptions of univariate statistics}\label{fourier-amplitudes-violate-parametric-assumptions-of-univariate-statistics}}

Many empirical studies use univariate T-tests or analysis of variance (ANOVA) to analyse periodic data. Specifically, the amplitude component of the Fourier spectrum at the stimulation frequency is used as the dependent variable, discarding the phase information. This is problematic, because the amplitude is an absolute quantity, and can never fall below zero. Distributions of amplitudes for weak signals are therefore positively skewed, and will generally violate the assumption of normality.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/amphists-1}

}

\caption{Demonstration of skew in absolute Fourier amplitudes for weak signals. Signal strength is quantified as Cohen's d, defined as the ratio of the mean to the standard deviation of the sample. The upper row shows samples of 50 grey points, and the population mean (coloured points). The lower row shows kernel density functions generated from 100,000 amplitude values. Note that the mean phase of the signal is irrelevant for these simulations, and is shown in the positive x-direction for consistency.}\label{fig:amphists}
\end{figure}

The upper row of Figure \ref{fig:amphists} shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \ref{fig:amphists} shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d \textgreater{} 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen (1988)). One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.

Typical solutions, such as log-transforming the data, are unlikely to be equally applicable to all conditions. For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal. Applying a transform to both conditions is therefore problematic. Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents. Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.

\hypertarget{conditions-under-which-t2_circ-is-more-sensitive-than-t2}{%
\section{\texorpdfstring{Conditions under which \(T^2_{circ}\) is more sensitive than \(T^2\)}{Conditions under which T\^{}2\_\{circ\} is more sensitive than T\^{}2}}\label{conditions-under-which-t2_circ-is-more-sensitive-than-t2}}

Using the \(T^2\) or \(T^2_{circ}\) statistic allows us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values. Victor \& Mast (1991) report simulations showing situations where \(T^2_{circ}\) has greater power than \(T^2\). This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant. Their simulations show the largest advantage for \(T^2_{circ}\) for effect sizes around d = 1 (where the mean signal strength is equal to the standard deviation of the data). The advantage appeared to be stronger for smaller sample sizes.

Here we replicate and extend these simulations (see Figure \ref{fig:powerfig}), and confirm that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f). However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (d \textgreater{} 2 when N \textgreater{} 8). These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for sample sizes, so its use should be restricted to small sample studies.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/powerfig-1}

}

\caption{Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes. Panels a-c replicate conditions reported by Victor and Mast (1991). Panels (d) and (e) show a wider range of conditions for each statistic. Panel (f) shows the difference between the two statistics, with contour lines indicating differences of 0.02, 0.05, 0.1, 0.2 and 0.4.}\label{fig:powerfig}
\end{figure}

\hypertarget{limitations-of-t2_circ-when-assumptions-are-violated}{%
\section{\texorpdfstring{Limitations of \(T^2_{circ}\) when assumptions are violated}{Limitations of T\^{}2\_\{circ\} when assumptions are violated}}\label{limitations-of-t2_circ-when-assumptions-are-violated}}

Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on meeting the test's more stringent assumptions. The two variables must be independent, and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \ref{fig:falsealarms} shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (dashed line). However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the variances of the two dependent variables more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2. In contrast, the \(T^2\) statistic, which explicitly takes account of the covariance matrix (see equation \eqref{eq:t2eq}) shows no increase (black curves).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/falsealarms-1.pdf}
\caption{\label{fig:falsealarms}Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b). Estimates are for 100000 simulations per condition, with N=10 observations. The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.}
\end{figure}

One possible remedy to control the Type I error rate would be to adjust the alpha level to compensate. However, this will reduce the statistical power of the \(T^2_{circ}\) test, and its advantage over \(T^2\) is relatively marginal in most situations to begin with (see Figure \ref{fig:powerfig}). What is required is a method to objectively assess whether the assumptions of \(T^2_{circ}\) hold, which we develop in the following section.

\hypertarget{a-novel-method-to-test-the-assumptions-of-t2_circ}{%
\section{\texorpdfstring{A novel method to test the assumptions of \(T^2_{circ}\)}{A novel method to test the assumptions of T\^{}2\_\{circ\}}}\label{a-novel-method-to-test-the-assumptions-of-t2_circ}}

Despite the severe consequences of violating the assumptions of the \(T^2_{circ}\) statistic (see Figure \ref{fig:falsealarms}), there is currently no accepted test for those assumptions that could be applied to an empirical data set. Victor and Mast (1991) suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components (see also Pei et al. 2017).

One convenient way to test the assumptions of \(T^2_{circ}\) is to assess the \emph{condition index} of a data set, which describes the ratio of eigenvalues for a cloud of points. The eigenvectors are the major and minor axes of the bounding ellipse (the straight lines in the example icons at the foot of Figure \ref{fig:falsealarms}a,b). Conventionally, the condition index is calculated as the square root of the longest/shortest eigenvector length. For uncorrelated random numbers the distribution of condition indices is positively skewed, with a minimum of 1 (Edelman 1988). This is because two independent samples of numbers from the same underlying distribution will still by chance have unequal eigenvectors. Edelman (1988) provides an equation (Eq 14) for the probability density function of condition indices (x) as a function of sample size (n):

\begin{equation}
\label{eq:edelman1}
pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)},
\end{equation}

Attempts to validate this by simulation suggest that for small sample sizes (n\textless{}10) a closer approximation is given by:

\begin{equation}
\label{eq:edelman2}
pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}
\end{equation}

In Figure \ref{fig:distcomparison}a we show an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100000 random data sets for n=10 (blue shading). The vertical lines show the critical (95\%) threshold for the analytic and simulated results - a ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance. Figure \ref{fig:distcomparison}b shows how these thresholds change as a function of the number of observations, and it is clear that the modified expression (red) most closely approximates the simulation results (blue).

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/distcomparison-1}

}

\caption{Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of 10 observations. The vertical blue and red lines show the 95 percent thresholds on the curve (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function. Panel (b) shows how 95 percent thresholds change as a functio of the number of observations.}\label{fig:distcomparison}
\end{figure}

The eigenvalue ratio can be used as a test of the assumptions of \(T^2_{circ}\). If we observe a condition index above the critical threshold for the number of observations then the data set can be said to significantly violate the assumption of equal eigenvalues. Because our modified equation provides an inverse density function (red curve in Figure \ref{fig:distcomparison}a), we can use this to calculate a p-value for the test. If the test is non-significant, one can proceed with \(T^2_{circ}\); if it is significant, \(T^2\) should be used instead. An \emph{R} function implementing this test is included in the project code repository (the \emph{CI.test} function contained in the \emph{CItest.R} script).

\hypertarget{identifying-and-removing-outliers-using-the-mahalanobis-distance}{%
\section{Identifying and removing outliers using the Mahalanobis distance}\label{identifying-and-removing-outliers-using-the-mahalanobis-distance}}

If a data set produces a significant result using the condition index test, this could be due to the presence of one or more outliers. The Mahalanobis distance (Mahalanobis 1936) is a useful metric for identifying such multivariate outliers so that they can be excluded. It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points. This means that any correlations in the data set are taken into account when calculating the distance metric, D.

The effectiveness of this approach to outlier exclusion can be assessed by simulation using the condition index test. Figure \ref{fig:outlierplot} shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves). In all cases, the functions depart from the Type 1 error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the Mahalanobis distance exceeds a value around 3. This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic return \(D^2\), which can be converted to D by taking the square root). Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the \(T^2_{circ}\) test.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/outlierplot-1}

}

\caption{Simulations illustrating the Mahalanobis distance metric, and showing how a single outlier affects the condition index. The upper row shows three example data sets, each with a single outlier shown in red. The outliers have Mahalanobis distances of 1, 3 and 5. The ellipses are calculated with the outlier included (red) and excluded (black), illustrating how the outlier distorts the aspect ratio of the ellipse. The lower plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.}\label{fig:outlierplot}
\end{figure}

A variant of the Mahalanobis distance (the pairwise Mahalanobis distance) can also be used to compute a multivariate measure of effect size, equivalent to Cohen's \emph{d} statistic (see e.g. Del Giudice 2009). This is a valuable statistic to include when reporting the results of multivariate tests, and we include an \emph{R} function (\emph{mahalES}) that can be used to calculate it.

\hypertarget{generalising-to-more-than-two-conditions}{%
\section{Generalising to more than two conditions}\label{generalising-to-more-than-two-conditions}}

Many studies involve more than two conditions that need to be compared. Issues with familywise error will quickly become problematic if multiple pairwise \(T^2\) or \(T^2_{circ}\) statistics are calculated. One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's \(T^2\), but permits multiple independent variables with more than two levels. However, if the assumptions of \(T^2_{circ}\) hold for a data set, it might alternatively be possible to extend to logic of the \(T^2_{circ}\) test (Victor and Mast 1991) to the more general case.

The F statistic for a one-way ANOVA is calculated by taking the ratio of the variance explained by the modelled group means to the residual unexplained variance. For multivariate data, the change in group mean would be calculated using the vector distances between the complex means, and the residuals are the vector distances between each data point and its corresponding group mean. For an independent one-way design with \(k\) groups (or conditions) and \(N\) observations per group, the F distribution will have \(2(k-1)\) and \(2((Nk)-k)\) degrees of freedom. The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number. A suitable name for such a test might be \({ANOVA}^2_{circ}\), as this reflects the similarity to ANOVA, and the extension of the logic of \(T^2_{circ}\) (an alternative name might be \(MANOVA_{circ}\), however this feels less appropriate given that many of the key features of MANOVA are absent).

Figure \ref{fig:powerfig2} shows simulations analogous to those in Figure \ref{fig:powerfig} for a one-way design with three levels. MANOVA is directly compared to the \({ANOVA}^2_{circ}\) statistic across a range of effect sizes and sample sizes. Just as for the one-sample statistics, the advantages of \({ANOVA}^2_{circ}\) are particularly apparent for small sample sizes, and larger effect sizes (Figure \ref{fig:powerfig2}f).

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/powerfig2-1}

}

\caption{Simulations comparing the sensitivity of MANOVA and ANOVA2circ. The format mirrors that of Figure 2. In these simulations, there were three conditions, with the signal being added to one condition.}\label{fig:powerfig2}
\end{figure}

Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that standard Bonferroni corrections are applied. Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary). A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic. The project repository contains \emph{R} scripts for both of these functions.

\hypertarget{deciding-which-test-to-run}{%
\section{Deciding which test to run}\label{deciding-which-test-to-run}}

The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios. Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if not. Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs. In \emph{R}, the \emph{Manova} function from the \emph{car} package appears to be capable of the necessary calculations, as does the commercial software package SPSS.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/flowchart-1}

}

\caption{Flowchart illustrating how one might decide which test to conduct for a given data set, based on the study design and the outcome of the condition index test.}\label{fig:flowchart}
\end{figure}

\hypertarget{applying-these-methods-to-empirical-data-sets}{%
\section{Applying these methods to empirical data sets}\label{applying-these-methods-to-empirical-data-sets}}

Having developed some novel tools for the analysis of periodic data, in the following sections we demonstrate them using two different publicly available empirical data sets. The first study recorded responses to auditory and optogenetic stimulation in mice. The second study measured visual responses to flickering grating patterns in humans. These cases provide examples of how the results of the tests described above might be appropriately reported.

\hypertarget{mouse-auditory-and-optogenetic-steady-state-data}{%
\subsection{Mouse auditory and optogenetic steady-state data}\label{mouse-auditory-and-optogenetic-steady-state-data}}

Hwang et al. (2019, 2020) measured steady-state responses using implanted scalp electrodes in 6 mice. The mice had previously been given a targeted virus that made parvalbumin neurons in their basal forebrain responsive to specific wavelengths of light, delivered through an optical fiber (a technique called optogenetics). SSVEPs were recorded from 36 electrodes for 1 second epochs of 40 Hz auditory stimulation, and various schedules of optogenetic stimulation (including at 40 Hz). The data set is described more fully by Hwang et al. (2020), and was downloaded from: \url{https://doi.gin.g-node.org/10.12751/g-node.e5tyek/}.

Figure \ref{fig:mousedata}a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz. There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality. Indeed, a paired t-test on the amplitudes reveals no significant difference (t=0.84, df=5, p=0.44). However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \ref{fig:mousedata}b). The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted. This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of D = 1.02. This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al. (2019) went on to explore interactions between these two signals.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/mousedata-1}

}

\caption{Summary of mouse steady-state responses to 40 Hz stimulation. Panel (a) shows the Fourier amplitude spectrum with inset scalp plots for sound (red) and light (blue) stimulation, averaged across repetitions and individuals. Panel (b) shows complex (x = real, y = imaginary) Fourier components for 6 individual mice (small points) and their average (large points) for both conditions.}\label{fig:mousedata}
\end{figure}

\hypertarget{human-visual-steady-state-data}{%
\subsection{Human visual steady-state data}\label{human-visual-steady-state-data}}

Vilidaite et al. (2018) measured visual responses to flickering grating stimuli in a large sample of 100 adults. Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker). Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast. For this analysis, we took responses from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant. Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}. Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/humanSSVEP-1}

}

\caption{Summary of human SSVEP data. Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes). Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point). Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.}\label{fig:humanSSVEP}
\end{figure}

With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.2, all \emph{p} \textgreater{} 0.19). A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, p \textless{} 0.001). Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.4, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.3, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.5, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.34, \(F_{(2,176)}\) = 35.3, D = 0.33, \emph{p} \textless{} 0.001). The study by Vilidaite et al. (2018) compared SSVEP responses between individuals with and without autism, as well as in a \emph{Drosophila} genetic model of developmental disorders. The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}.

\hypertarget{further-considerations}{%
\section{Further considerations}\label{further-considerations}}

It is worth pointing out that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations. This is the case for most paradigms in which the nervous system is driven by a periodic stimulus. However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger 1929; Buzsáki and Draguhn 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g. Canolty and Knight 2010). When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al. 2020). This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.

The present paper has focussed on the Frequentist statistical tradition. However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses. Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al. 2009). However this is a non-trivial undertaking, and is beyond the scope of the current paper.

Another possibility is to use machine learning techniques such as multivariate pattern analysis (MVPA) to analyse periodic data. This involvess training a classifier algorithm to distinguish between two (or more) experimental conditions or states, and then assessing classifier accuracy on fresh data. If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance. Such methods have been hugely influential in the fMRI literature, and for analysing event-related potential data collected using EEG or MEG. However, they have not been widely applied to steady-state data (though see West et al. (2015) for an example). In principle, the real and imaginary Fourier components can be treated as separate dependent variables, along with different recording locations and/or frequencies. This approach has the potential to offer sensitive, high-powered statistical tests that circumvent many of the shortcomings associated with parametric statistics.

Even when statistics are conducted using both the real and imaginary Fourier components, it is still typical to visualise the mean amplitudes. Several approaches to calculating appropriate error bars have been proposed. For example, Pei et al. (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude. This approach is somewhat computationally demanding, though we have implemented an \emph{R} function to calculate error bars using this method. However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95\% confidence intervals on amplitudes. This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean. The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled amplitudes. Upper and lower 95\% confidence intervals on the amplitude can then be taken at appropriate quantiles of this population.

\hypertarget{general-recommendations-for-analysing-periodic-data}{%
\section{General recommendations for analysing periodic data}\label{general-recommendations-for-analysing-periodic-data}}

The simulations reported here allow us to make several recommendations for how periodic data should be analysed. Multivariate statistics should be used for phase-locked Fourier data instead of univariate statistics such as T-tests and ANOVA. This avoids problems from distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information. Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahanalobis distance reported as a measure of effect size. For sample sizes of N \textless{} 32, the \(T^2_{circ}\) and \(ANOVA^2_{circ}\) statistics can be used if the condition index test is non-significant for all conditions. Alternatively, the \(T^2\) or MANOVA statistics should be used when these conditions are not met. The greater power afforded by these tests should in general lead to more accurate statistical inferences when analysing periodic data.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Afsari2014}{}%
Afsari, Farinaz, Kenneth V Christensen, Garrick Paul Smith, Morten Hentzer, Olivia M Nippe, Christopher J H Elliott, and Alex R Wade. 2014. ``Abnormal Visual Gain Control in a Parkinson's Disease Model.'' \emph{Hum Mol Genet} 23 (17): 4465--78. \url{https://doi.org/10.1093/hmg/ddu159}.

\leavevmode\hypertarget{ref-Baker2020}{}%
Baker, D.H., G. Vilidaite, F.A. Lygo, A.K. Smith, T.R. Flack, A.D. Gouws, and T.J. Andrews. 2020. ``Power Contours: Optimising Sample Size and Precision in Experimental Psychology and Human Neuroscience.'' \emph{Psychological Methods}. \url{https://doi.org/10.1037/met0000337}.

\leavevmode\hypertarget{ref-Berger1929}{}%
Berger, H. 1929. ``Über Das Elektrenkephalogramm Des Menschen.'' \emph{Archiv F. Psychiatrie} 87: 527--70. \url{https://doi.org/10.1007/BF01797193}.

\leavevmode\hypertarget{ref-Braddick1986}{}%
Braddick, O J, J Wattam-Bell, and J Atkinson. 1986. ``Orientation-Specific Cortical Responses Develop in Early Infancy.'' \emph{Nature} 320 (6063): 617--9. \url{https://doi.org/10.1038/320617a0}.

\leavevmode\hypertarget{ref-Buzsaki2004}{}%
Buzsáki, György, and Andreas Draguhn. 2004. ``Neuronal Oscillations in Cortical Networks.'' \emph{Science} 304 (5679): 1926--9. \url{https://doi.org/10.1126/science.1099745}.

\leavevmode\hypertarget{ref-Canolty2010}{}%
Canolty, Ryan T, and Robert T Knight. 2010. ``The Functional Role of Cross-Frequency Coupling.'' \emph{Trends Cogn Sci} 14 (11): 506--15. \url{https://doi.org/10.1016/j.tics.2010.09.001}.

\leavevmode\hypertarget{ref-Cohen1988}{}%
Cohen, J. 1988. \emph{Statistical Power Analysis for the Behavioral Sciences}. Lawrence Erlbaum, Hillsdale, New Jersey.

\leavevmode\hypertarget{ref-Giudice2009}{}%
Del Giudice, M. 2009. ``On the Real Magnitude of Psychological Sex Differences.'' \emph{Evolutionary Psychology} 7: 264--79. \url{https://doi.org/10.1177/147470490900700209}.

\leavevmode\hypertarget{ref-Edelman1988}{}%
Edelman, Alan. 1988. ``Eigenvalues and Condition Numbers of Random Matrices.'' \emph{SIAM Journal on Matrix Analysis and Applications} 9 (4): 543--60. \url{https://doi.org/10.1137/0609045}.

\leavevmode\hypertarget{ref-Engel1994}{}%
Engel, S A, D E Rumelhart, B A Wandell, A T Lee, G H Glover, E J Chichilnisky, and M N Shadlen. 1994. ``FMRI of Human Visual Cortex.'' \emph{Nature} 369 (6481): 525. \url{https://doi.org/10.1038/369525a0}.

\leavevmode\hypertarget{ref-Enroth-Cugell1966}{}%
Enroth-Cugell, C, and J G Robson. 1966. ``The Contrast Sensitivity of Retinal Ganglion Cells of the Cat.'' \emph{J Physiol} 187 (3): 517--52. \url{https://doi.org/10.1113/jphysiol.1966.sp008107}.

\leavevmode\hypertarget{ref-Gray2020}{}%
Gray, Katie L H, Tessa R Flack, Miaomiao Yu, Freya A Lygo, and Daniel H Baker. 2020. ``Nonlinear Transduction of Emotional Facial Expression.'' \emph{Vision Res} 170 (May): 1--11. \url{https://doi.org/10.1016/j.visres.2020.03.004}.

\leavevmode\hypertarget{ref-Hotelling1931}{}%
Hotelling, H. 1931. ``The Generalization of Student's Ratio.'' \emph{The Annals of Mathematical Statistics} 2 (3). Institute of Mathematical Statistics: 360--78. \url{http://www.jstor.org/stable/2957535}.

\leavevmode\hypertarget{ref-Hwang2019}{}%
Hwang, Eunjin, Ritchie E Brown, Bernat Kocsis, Tae Kim, James T McKenna, James M McNally, Hio-Been Han, and Jee Hyun Choi. 2019. ``Optogenetic Stimulation of Basal Forebrain Parvalbumin Neurons Modulates the Cortical Topography of Auditory Steady-State Responses.'' \emph{Brain Struct Funct} 224 (4): 1505--18. \url{https://doi.org/10.1007/s00429-019-01845-5}.

\leavevmode\hypertarget{ref-Hwang2020}{}%
Hwang, Eunjin, Hio-Been Han, Jung Young Kim, and Jee Hyun Choi. 2020. ``High-Density Eeg of Auditory Steady-State Responses During Stimulation of Basal Forebrain Parvalbumin Neurons.'' \emph{Sci Data} 7 (1): 288. \url{https://doi.org/10.1038/s41597-020-00621-z}.

\leavevmode\hypertarget{ref-Liu-Shuang2014}{}%
Liu-Shuang, Joan, Anthony M Norcia, and Bruno Rossion. 2014. ``An Objective Index of Individual Face Discrimination in the Right Occipito-Temporal Cortex by Means of Fast Periodic Oddball Stimulation.'' \emph{Neuropsychologia} 52 (January): 57--72. \url{https://doi.org/10.1016/j.neuropsychologia.2013.10.022}.

\leavevmode\hypertarget{ref-Mahalanobis1936}{}%
Mahalanobis, P.C. 1936. ``On the Generalised Distance in Statistics.'' \emph{Proceedings of the National Academy of Sciences of India} 2: 49--55.

\leavevmode\hypertarget{ref-Morrone1987}{}%
Morrone, M C, D C Burr, and H D Speed. 1987. ``Cross-Orientation Inhibition in Cat Is Gaba Mediated.'' \emph{Exp Brain Res} 67 (3): 635--44. \url{https://doi.org/10.1007/BF00247294}.

\leavevmode\hypertarget{ref-Nakayama1982}{}%
Nakayama, K, and M Mackeben. 1982. ``Steady State Visual Evoked Potentials in the Alert Primate.'' \emph{Vision Res} 22 (10): 1261--71. \url{https://doi.org/10.1016/0042-6989(82)90138-9}.

\leavevmode\hypertarget{ref-Norcia2015}{}%
Norcia, Anthony M, L Gregory Appelbaum, Justin M Ales, Benoit R Cottereau, and Bruno Rossion. 2015. ``The Steady-State Visual Evoked Potential in Vision Research: A Review.'' \emph{J Vis} 15 (6): 4. \url{https://doi.org/10.1167/15.6.4}.

\leavevmode\hypertarget{ref-Pei2017}{}%
Pei, Francesca, Stefano Baldassi, Jeffrey J Tsai, Holly E Gerhard, and Anthony M Norcia. 2017. ``Development of Contrast Normalization Mechanisms During Childhood and Adolescence.'' \emph{Vision Res} 133 (April): 12--20. \url{https://doi.org/10.1016/j.visres.2016.03.010}.

\leavevmode\hypertarget{ref-Porciatti1990}{}%
Porciatti, V, G Fontanesi, A Raffaelli, and P Bagnoli. 1990. ``Binocularity in the Little Owl, Athene Noctua. II. Properties of Visually Evoked Potentials from the Wulst in Response to Monocular and Binocular Stimulation with Sine Wave Gratings.'' \emph{Brain Behav Evol} 35 (1): 40--48. \url{https://doi.org/10.1159/000115855}.

\leavevmode\hypertarget{ref-Rees1986}{}%
Rees, A, G G Green, and R H Kay. 1986. ``Steady-State Evoked Responses to Sinusoidally Amplitude-Modulated Sounds Recorded in Man.'' \emph{Hear Res} 23 (2): 123--33. \url{https://doi.org/10.1016/0378-5955(86)90009-2}.

\leavevmode\hypertarget{ref-Regan1966}{}%
Regan, D. 1966. ``Some Characteristics of Average Steady-State and Transient Responses Evoked by Modulated Light.'' \emph{Electroencephalogr Clin Neurophysiol} 20 (3): 238--48. \url{https://doi.org/10.1016/0013-4694(66)90088-5}.

\leavevmode\hypertarget{ref-Rouder2009}{}%
Rouder, Jeffrey N, Paul L Speckman, Dongchu Sun, Richard D Morey, and Geoffrey Iverson. 2009. ``Bayesian T Tests for Accepting and Rejecting the Null Hypothesis.'' \emph{Psychon Bull Rev} 16 (2): 225--37. \url{https://doi.org/10.3758/PBR.16.2.225}.

\leavevmode\hypertarget{ref-Sereno1995}{}%
Sereno, M I, A M Dale, J B Reppas, K K Kwong, J W Belliveau, T J Brady, B R Rosen, and R B Tootell. 1995. ``Borders of Multiple Visual Areas in Humans Revealed by Functional Magnetic Resonance Imaging.'' \emph{Science} 268 (5212): 889--93. \url{https://doi.org/10.1126/science.7754376}.

\leavevmode\hypertarget{ref-Snyder1992}{}%
Snyder, A Z. 1992. ``Steady-State Vibration Evoked Potentials: Descriptions of Technique and Characterization of Responses.'' \emph{Electroencephalogr Clin Neurophysiol} 84 (3): 257--68. \url{https://doi.org/10.1016/0168-5597(92)90007-x}.

\leavevmode\hypertarget{ref-Spitschan2014}{}%
Spitschan, Manuel, Sandeep Jain, David H Brainard, and Geoffrey K Aguirre. 2014. ``Opponent Melanopsin and S-Cone Signals in the Human Pupillary Light Response.'' \emph{Proc Natl Acad Sci U S A} 111 (43): 15568--72. \url{https://doi.org/10.1073/pnas.1400942111}.

\leavevmode\hypertarget{ref-Victor1991}{}%
Victor, J D, and J Mast. 1991. ``A New Statistic for Steady-State Evoked Potentials.'' \emph{Electroencephalogr Clin Neurophysiol} 78 (5): 378--88. \url{https://doi.org/10.1016/0013-4694(91)90099-p}.

\leavevmode\hypertarget{ref-Vilidaite2018}{}%
Vilidaite, Greta, Anthony M Norcia, Ryan J H West, Christopher J H Elliott, Francesca Pei, Alex R Wade, and Daniel H Baker. 2018. ``Autism Sensory Dysfunction in an Evolutionarily Conserved System.'' \emph{Proc Biol Sci} 285 (1893): 20182255. \url{https://doi.org/10.1098/rspb.2018.2255}.

\leavevmode\hypertarget{ref-West2015}{}%
West, Ryan J H, Christopher J H Elliott, and Alex R Wade. 2015. ``Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.'' \emph{Sci Rep} 5 (November): 16933. \url{https://doi.org/10.1038/srep16933}.

\end{document}

### 2020-12-23 anovacirc.R

@@ -1,0 +1,40 @@
+anovacirc.test <- function(data){
+  # analysis of variance using complex data
+  # an extension of the logic of the t-squared circ statistic
+  
+  grouplabels <- data$grouplabels
+  datavals <- data$simdata
+  factorlist <- levels(grouplabels)
+  
+  grandmean <- mean(datavals)
+  
+  groupmeans <- NULL
+  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
+  
+  SSM <- 0
+  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
+  
+  dfM <- 2*(nlevels(grouplabels)-1)
+  
+  allresiduals <- 1:length(datavals)
+  for (n in 1:nlevels(grouplabels)){allresiduals[which(grouplabels==factorlist[n])] <- abs(datavals[which(grouplabels==factorlist[n])] - groupmeans[n])}
+  
+  SSR <- sum(allresiduals^2)
+  dfR <- 2*(length(datavals)-nlevels(grouplabels))
+  MSM <- SSM/dfM
+  MSR <- SSR/dfR
+  Fratio <- MSM/MSR
+  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+  
+  output <- NULL
+  output$Fratio <- Fratio
+  output$pvalue <- pvalue
+  output$SSM <- SSM
+  output$SSR <- SSR
+  output$dfM <- dfM
+  output$dfR <- dfR
+  output$MSM <- MSM
+  output$MSR <- MSR
+  
+  return(output)
+}

### 2020-12-23 anovacircR.R

@@ -1,0 +1,45 @@
+anovacircR.test <- function(data){
+  # analysis of variance using complex data
+  # an extension of the logic of the t-squared circ statistic
+  # this version for repeated measures designs
+  
+  grouplabels <- data$grouplabels
+  factorlist <- levels(grouplabels)
+  participantlabels <- data$participant
+  participantlist <- levels(participantlabels)
+  datavals <- data$simdata
+  
+  grandmean <- mean(datavals)
+  
+  SSW <- 0
+  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
+  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))
+  
+  groupmeans <- NULL
+  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
+  SSM <- 0
+  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
+  dfM <- 2*(nlevels(grouplabels)-1)
+  
+  SSR <- SSW - SSM
+  dfR <- dfW - dfM
+  
+  MSM <- SSM/dfM
+  MSR <- SSR/dfR
+  Fratio <- MSM/MSR
+  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+  
+  output <- NULL
+  output$Fratio <- Fratio
+  output$pvalue <- pvalue
+  output$SSW <- SSW  
+  output$SSM <- SSM
+  output$SSR <- SSR
+  output$dfW <- dfW
+  output$dfM <- dfM
+  output$dfR <- dfR
+  output$MSM <- MSM
+  output$MSR <- MSR
+  
+  return(output)
+}

### 2020-12-23 importmousedata.m

@@ -1,0 +1,45 @@
+function importmousedata
+
+% function to import mouse SSVEP data from the Hwang et al (2019) study
+% data are stored in EEGlab format and have already been epoched
+% the Fourier transform is taken, and complex spectra at the signal frequencies are stored for analysis in R
+% DHB 21/12/20
+
+close all;
+
+datadir = '/Users/danbaker/Desktop/10.12751_g-node.e5tyek/';
+nconds = [5 7];
+startpoint = 1602;
+endpoint = 3601;    % this is 1 second of data during which the stimulus was presented
+
+mousedata = zeros(6,38,12,100);
+for subj = 1:6
+    for expt = 1:2
+        
+        EEG = pop_loadset(strcat('epochs_animal',num2str(subj),'.set'),strcat(datadir,'dataset_',num2str(expt),'/'));
+        duration = EEG.pnts/EEG.srate;
+        for ch = 1:EEG.nbchan
+            elspec = [];
+            for cond = 1:nconds(expt)
+                
+                condcount = 0;
+                for n = 1:EEG.trials
+                    thiscond = str2num(EEG.event(n).type(1));
+                    if thiscond==cond
+                        condcount = condcount + 1;
+                        temp = fft(squeeze(EEG.data(ch,startpoint:endpoint,n)))/2000;
+                        elspec(condcount,:) = temp(2:101);
+                    end
+                    
+                end
+                mousedata(subj,ch,(cond+(expt-1)*5),:) = mean(elspec,'omitnan');
+            end
+            
+        end
+    end
+end
+
+
+save('Hwangdata.mat','mousedata');
+
+end

### 2020-12-23 plotmousedata.R

@@ -1,0 +1,182 @@
+# library(R.matlab)
+# library(reshape2)
+# data <- readMat('Hwangdata.mat')
+# mousedata <- data$mousedata
+# montage <- read.csv('~/Desktop/10.12751_g-node.e5tyek/montage.csv')
+# outline <- read.csv('~/Desktop/10.12751_g-node.e5tyek/outline.csv',header=FALSE)
+# outline <- t(outline[,1:16])
+# outline2 <- melt(outline)
+# outline <- matrix(outline2[1:268,3],nrow=134,ncol=2,byrow=TRUE)
+# save(file='Hwangdata.RData',list=c('mousedata','montage','outline'))
+
+library(HDMD)
+
+v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
+  ## Create a function to perform Matlab's v4 interpolation.
+  ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
+  ## and variables xo and yo, the co-ordinates which will be use to create a grid for interpolation
+  xo <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
+  yo <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
+  xy <- df$x + df$y*sqrt(as.complex(-1))
+  d <- matrix(rep(xy,length(xy)),nrow = length(xy), ncol = length(xy))
+  d <- abs(d - t(d))
+  diag(d) <- 1
+  g <- (d^2) * (log(d)-1)   # Green's function.
+  diag(g) <- 0
+  weights <- qr.solve(g,df$z)
+  xy <- t(xy)
+  outmat <- matrix(nrow = gridRes,ncol = gridRes)
+  for (i in 1:gridRes){
+    for (j in 1:gridRes) {
+      test4 <- abs((xo[i,j] + sqrt(as.complex(-1))*yo[i,j]) - xy)
+      g <- (test4^2) * (log(test4)-1)
+      outmat[i,j] <- g %*% weights}}
+  outDf <- data.frame(x = xo[,1],outmat)
+  names(outDf)[1:length(yo[1,])+1] <- yo[1,]
[…]

### 2020-12-26 manuscript/manuscript.Rmd
Restructured into an R package

--- 2020-12-23 manuscript.Rmd
@@
 
-nsims <- 100  # global value for the whole script - increase to 100000 for final run
-
-nbdtpal <- c(rgb(0,0,0),rgb(187/255,40/255,107/255),rgb(56/255,111/255,164/255),'darkgreen','darkorange')
+nsims <- 10000  # global value for the whole script - increase to 100000 for final run
+
+# palette using the main colours from the NBDT journal logo
+nbdtpal <- c(rgb(0,0,0),rgb(0.73,0.16,0.42),rgb(0.22,0.44,0.64),'darkgreen','darkorange')
 
@@
 Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli.
-Neural signals can then be recorded, and include phase-locked oscillations at the stimulation frequency.
+Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency.
 The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components.
-However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and weak signals might be missed if the phase information is discarded.
+However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded.
 An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components.
-Here we compare the performance of two multivariate extensions of the T-test, Hotelling's $T^2$ and a variant called $T^2_{circ}$.
-We develop a novel test of the assumptions of $T^2_{circ}$ based on the condition index, and a heuristic for excluding outliers using the Mahanalobis distance.
+Here we compare the performance of two multivariate extensions of the T-test: Hotelling's $T^2$ and a variant called $T^2_{circ}$.
+We develop a novel test of the assumptions of $T^2_{circ}$ based on the condition index, and suggest a heuristic for excluding outliers using the Mahanalobis distance.
 We then extend the statistic to multi-level designs, and propose a new statistical test termed $ANOVA^2_{circ}$.
@@
 tsqvscirc <- circpropM - tsqpropM
+tsqvscirc <- pmax(tsqvscirc,0)  # sets any negative values to 0
 
@@
 
-lines(x,pdffunction/max(pdffunction))
+lines(x,pdffunction/max(pdffunction),lwd=3)
 cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction))
-lines(x,cdfinverse,col=nbdtpal[2])
+lines(x,cdfinverse,col=nbdtpal[2],lwd=3)
 lines(c(1,5),c(0.05,0.05),lty=2,col=nbdtpal[1])
 criticalx <- x[min(which(cdfinverse<=0.05))]
-lines(c(criticalx,criticalx),c(0,1),col=nbdtpal[2],lwd=2)
+lines(c(criticalx,criticalx),c(0,1),col=nbdtpal[2],lwd=3)
 
@@
 simcrit <- quantile(simvect,0.95)
-lines(c(simcrit,simcrit),c(0,1),col=nbdtpal[3],lwd=2,lty=2)
+lines(c(simcrit,simcrit),c(0,1),col=nbdtpal[3],lwd=3,lty=2)
 text(4.8,0.95,'(a)',cex=1.5)
@@
 
-lines(3:25,edelman1,lwd=2,col=nbdtpal[1])
-lines(3:25,edelman2,lwd=2,col=nbdtpal[2])
-lines(3:25,simratios,lwd=2,col=nbdtpal[3])
+lines(3:25,edelman1,lwd=3,col=nbdtpal[1])
+lines(3:25,edelman2,lwd=3,col=nbdtpal[2])
+lines(3:25,simratios,lwd=3,col=nbdtpal[3])
 legend(12, 15, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=1, lwd=3, box.lwd=2)
@@
 manvscirc <- circpropM - manovapropM
+manvscirc <- pmax(manvscirc,0)  # sets any negative values to 0
 
@@
 
-The flowchart in Figure \@ref(fig:flowchart) contains a proposed decision structure for the analysis of periodic data.
+The flowchart in Figure \@ref(fig:flowchart) contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
 Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios.
@@
 There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
-Indeed, a paired t-test on the amplitudes reveals no significant difference (t=0.84, df=5, p=0.44).
+Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = 0.84, _df_ = 5, _p_ = 0.44).
 However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b).
@@
 @Canolty2010].
-When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average [@Baker2020].
+When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average [@Baker2021].
 This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.
@@
 The present paper has focussed on the Frequentist statistical tradition.
-However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses.
-Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests [@Rouder2009].
+However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses [@Jeffreys1961].
+Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests [@Rouder2009] and ANOVA [@Rouder2017].
 However this is a non-trivial undertaking, and is beyond the scope of the current paper.
@@
 If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance.
-Such methods have been hugely influential in the fMRI literature, and for analysing event-related potential data collected using EEG or MEG.
+Such methods have been hugely influential in the fMRI literature [@Schwarzkopf2011], and for analysing event-related potential data collected using EEG or MEG [@Grootswagers2017].
 However, they have not been widely applied to steady-state data (though see West et al.
@@
 In principle, the real and imaginary Fourier components can be treated as separate dependent variables, along with different recording locations and/or frequencies.
-This approach has the potential to offer sensitive, high-powered statistical tests that circumvent many of the shortcomings associated with parametric statistics.
+This approach has the potential to offer sensitive, high-powered statistical tests that circumvent many of the shortcomings associated with traditional statistics.
 
@@
 [-@Pei2017] suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat computationally demanding, though we have implemented an _R_ function to calculate error bars using this method.
+This approach is somewhat demanding, though we have implemented an _R_ function to calculate error bars using this method (available through the _getampCI_ function).
 However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95% confidence intervals on amplitudes.

### 2020-12-26 manuscript/manuscript.tex
Restructured into an R package

--- 2020-12-23 manuscript.tex
@@
 Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli.
-Neural signals can then be recorded, and include phase-locked oscillations at the stimulation frequency.
+Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency.
 The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components.
-However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and weak signals might be missed if the phase information is discarded.
+However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded.
 An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components.

### 2020-12-26 R/amperrors.R

@@ -1,0 +1,1 @@
+# placeholder for errorbar function amperrors

### 2020-12-26 R/fftshift.R

@@ -1,0 +1,1 @@
+fftshift <- function(im) {im * (-1)^(row(im) + col(im))}  

### 2020-12-26 manuscript/manuscript.Rmd
Adding skeleton documentation and package structure

@@
 # install the missing ones, and activate all
-packagelist <- c('bookdown','pals','shape','diagram','HDMD')
+
+# devtools::install_github("bakerdh/FourierStats")
+# devtools::install('FourierStats')
+
+packagelist <- c('bookdown','pals','shape','diagram','HDMD','FourierStats')
 missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
@@
 nbdtpal <- c(rgb(0,0,0),rgb(0.73,0.16,0.42),rgb(0.22,0.44,0.64),'darkgreen','darkorange')
-
-tsq1.test <- function(data,mu=NULL){
-  # function to calculate a one-sample t-squared test after Hotelling (1931)
-  # for two-sample tests, use the Hotelling package
-
-  # compare each DV to some value
-  if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}
-
-  xbar <- colMeans(data) # estimate means for each DV
-  m <- ncol(data)   # record number of DVs
-  N <- nrow(data)   # record sample size
-  C <- cov(data)    # calculate covariance matrix
-  Cinv <- solve(C)  # calculate inverse matrix
-  tsq <- N*t(xbar)%*%Cinv%*%xbar # compute t2 statistic
-  Fratio <- tsq * (N-m)/(m*(N-1)) # convert to F-ratio
-  df1 <- m      # record degrees of freedom
-  df2 <- N-m
-  pval <- 1- pf(Fratio,df1,df2) # estimate p-value
-  p.value <- min(pval,1)
-
-  # store outputs in a data structure
-  output <- data.frame(tsq,Fratio,df1,df2,p.value)
-  return(output)}
-
-tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
-  # function to calculate the t-squared circ statistic from Victor & Mast (1991)
-  # the inputs must be Nx2 or 2xN matrices of numbers
-  # for a one-sample test, mu is an optional vector to which the data are compared
-
-  # transpose if dimensions are the wrong way round
-  if (is.matrix(x)){d <- dim(x)
-  if (d[1]<d[2]){x <- t(x)}}   
-  if (!is.null(y)){if (is.matrix(y)){d <- dim(y)
-  if (d[1]<d[2]){y <- t(y)}}}
-
-  if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}
-
-  # convert data to complex representation
-  x <- complex(real=x[,1],imaginary=x[,2])
-  if (!is.null(y)){y <- complex(real=y[,1],imaginary=y[,2])}
-
-  if (is.null(y)){method <- 'One-sample t-squared circ test'}
-  if (!is.null(y) & paired==TRUE){method <- 'Paired t-squared circ test'}
-  if (!is.null(y) & paired==FALSE){method <- 'Independent samples t-squared circ test'}
-
-  if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}
-
-  # paired or one-sample version of the test
-  if (paired==TRUE | is.null(y)){
-
-    nobs <- length(x)
-    cohmean <- mean(x)
-    absmean <- abs(cohmean)
-
-    displacement <- abs(x - cohmean)
-    diffsumsq <- sum(displacement^2)
-    tsqc <- (nobs-1) * (absmean^2) / diffsumsq
-    Fratio <- nobs*tsqc
-    df2 <- nobs*2-2
-    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
-  }
-
-  # independent samples version of the test
-  if (paired==FALSE & !is.null(y)){  
-    nobs1 <- length(x)
-    cohmean1 <- mean(x)
-    absmean1 <- abs(cohmean1)
-    displacement1 <- abs(x - cohmean1)
-
-    nobs2 <- length(y)
-    cohmean2 <- mean(y)
-    absmean2 <- abs(cohmean2)    
-    displacement2 <- abs(y - cohmean2)
-
-    meandiff <- cohmean1 - cohmean2
-    absdiff <- abs(meandiff)
-    diffsumsq <- sum(displacement1^2) + sum(displacement2^2)
-
-    tsqc <- (nobs1 + nobs2 - 2) * (absdiff^2) / diffsumsq
-    Fratio <- ((nobs1*nobs2)/(nobs1 + nobs2))*tsqc
-    df2 <- (2*nobs1 + 2*nobs2 - 4)
-    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
-  }
-
-  df1 <- 2
-  p.value <- min(pval,1)
-
-  output <- data.frame(tsqc,Fratio,df1,df2,p.value,method)
-  return(output)}
-
-CI.test <- function(data){
-  # function to calculate the condition index of bivariate data
-  # compare for significance with an expected distribution function
-
-  C <- cov(data)  # covariance matrix
-  N <- nrow(data) # sample size
-  eigVal  <- eigen(C)$values  # calculate eigenvalues
-  CI <- sqrt(eigVal[1]/eigVal[2]) # calculate condition index
-
-  maxci <- max(20,round(CI+5))   # make sure our target CI is in range
-  cilist <- seq(1,maxci,0.01)    # list of possible CIs
-
-  # calculate probability density function using modified Edelman equation
-  pdffunction <- ((N-2)*(2^(N-2))) * 
-    ((cilist^2 - 1)/((cilist^2+1)^(N-1))) * (cilist^(N-3))
-  cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
-  criticalCI <- cilist[min(which(cdfinverse<=0.05))]  # find the threshold CI
-
-  indices <- which(cilist>=CI)  # find the CI values larger than our CI
-  pval <- cdfinverse[indices[1]] # estimate the p-value
-
-  output <- data.frame(CI,N,criticalCI,pval)
-  return(output)}
-
-anovacirc.test <- function(data){
-  # analysis of variance using complex data
-  # an extension of the logic of the t-squared circ statistic
-
-  grouplabels <- data$grouplabels
-  datavals <- data$simdata
-  factorlist <- levels(grouplabels)
-
-  grandmean <- mean(datavals)
-
-  groupmeans <- NULL
-  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
-
-  SSM <- 0
-  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
-
-  dfM <- 2*(nlevels(grouplabels)-1)
-
-  allresiduals <- 1:length(datavals)
-  for (n in 1:nlevels(grouplabels)){allresiduals[which(grouplabels==factorlist[n])] <- abs(datavals[which(grouplabels==factorlist[n])] - groupmeans[n])}
-
-  SSR <- sum(allresiduals^2)
-  dfR <- 2*(length(datavals)-nlevels(grouplabels))
-  MSM <- SSM/dfM
-  MSR <- SSR/dfR
-  Fratio <- MSM/MSR
-  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
-
-  output <- NULL
-  output$Fratio <- Fratio
-  output$pvalue <- pvalue
-  output$SSM <- SSM
-  output$SSR <- SSR
-  output$dfM <- dfM
-  output$dfR <- dfR
-  output$MSM <- MSM
-  output$MSR <- MSR
-
-  return(output)
-}
-
-anovacircR.test <- function(data){
-  # analysis of variance using complex data
-  # an extension of the logic of the t-squared circ statistic
-  # this version for repeated measures designs
-
-  grouplabels <- data$grouplabels
-  factorlist <- levels(grouplabels)
-  participantlabels <- data$participant
-  participantlist <- levels(participantlabels)
-  datavals <- data$simdata
-
-  grandmean <- mean(datavals)
-
-  SSW <- 0
-  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
-  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))
-
-  groupmeans <- NULL
-  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
-  SSM <- 0
-  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
-  dfM <- 2*(nlevels(grouplabels)-1)
-
-SSR <- SSW - SSM
-dfR <- dfW - dfM
-
-  MSM <- SSM/dfM
-  MSR <- SSR/dfR
-  Fratio <- MSM/MSR
-  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
-
-  output <- NULL
-  output$Fratio <- Fratio
-  output$pvalue <- pvalue
-  output$SSW <- SSW  
-  output$SSM <- SSM
-  output$SSR <- SSR
-  output$dfW <- dfW
-  output$dfM <- dfM
-  output$dfR <- dfR
-  output$MSM <- MSM
-  output$MSR <- MSR
-
-  return(output)
-}
-
-getel <- function(compdata){
-  A <- cov(compdata)
-  ctr    <- colMeans(compdata) 
-  RR     <- chol(A)                               # Cholesky decomposition
-  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
-  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
-  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
-  eigVal  <- eigen(A)$values
-  eigVec  <- eigen(A)$vectors
-  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
-  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
-  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
-  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
-  ellRot  <- eigVec %*% t(ellBase) 
-  return(ellRot)
-  }
 

### 2020-12-26 R/CI.test.R

@@ -1,23 +1,24 @@
+#' function to calculate the condition index of bivariate data
+#' compare for significance with an expected distribution function
+#' @export
 CI.test <- function(data){
-  # function to calculate the condition index of bivariate data
-  # compare for significance with an expected distribution function
-  
+
   C <- cov(data)  # covariance matrix
   N <- nrow(data) # sample size
   eigVal  <- eigen(C)$values  # calculate eigenvalues
   CI <- sqrt(eigVal[1]/eigVal[2]) # calculate condition index
-  
+
   maxci <- max(20,round(CI+5))   # make sure our target CI is in range
   cilist <- seq(1,maxci,0.01)    # list of possible CIs
-  
+
   # calculate probability density function using modified Edelman equation
-  pdffunction <- ((N-2)*(2^(N-2))) * 
+  pdffunction <- ((N-2)*(2^(N-2))) *
     ((cilist^2 - 1)/((cilist^2+1)^(N-1))) * (cilist^(N-3))
   cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
   criticalCI <- cilist[min(which(cdfinverse<=0.05))]  # find the threshold CI
-  
+
   indices <- which(cilist>=CI)  # find the CI values larger than our CI
   pval <- cdfinverse[indices[1]] # estimate the p-value
-  
+
   output <- data.frame(CI,N,criticalCI,pval)
   return(output)}

### 2020-12-26 R/FourierStats-package.r

@@ -1,0 +1,1 @@
+#' Documentation for FourierStats R package

### 2020-12-26 R/amperrors.R

@@ -1,1 +1,5 @@
-# placeholder for errorbar function amperrors
+#' placeholder for errorbar function amperrors
+#' @export
+amperrors <- function(input){
+
+return(input)}

### 2020-12-26 R/anovacirc.test.R

@@ -1,45 +1,41 @@
-anovacircR.test <- function(data){
-  # analysis of variance using complex data
-  # an extension of the logic of the t-squared circ statistic
-  # this version for repeated measures designs
-  
+#' analysis of variance using complex data
+#' an extension of the logic of the t-squared circ statistic
+#' @export
+anovacirc.test <- function(data){
+
   grouplabels <- data$grouplabels
-  factorlist <- levels(grouplabels)
-  participantlabels <- data$participant
-  participantlist <- levels(participantlabels)
   datavals <- data$simdata
-  
+  factorlist <- levels(grouplabels)
+
   grandmean <- mean(datavals)
-  
-  SSW <- 0
-  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
-  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))
-  
+
   groupmeans <- NULL
   for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
+
   SSM <- 0
   for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
+
   dfM <- 2*(nlevels(grouplabels)-1)
-  
-  SSR <- SSW - SSM
-  dfR <- dfW - dfM
-  
+
+  allresiduals <- 1:length(datavals)
+  for (n in 1:nlevels(grouplabels)){allresiduals[which(grouplabels==factorlist[n])] <- abs(datavals[which(grouplabels==factorlist[n])] - groupmeans[n])}
+
+  SSR <- sum(allresiduals^2)
+  dfR <- 2*(length(datavals)-nlevels(grouplabels))
   MSM <- SSM/dfM
   MSR <- SSR/dfR
   Fratio <- MSM/MSR
   pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
-  
+
   output <- NULL
   output$Fratio <- Fratio
   output$pvalue <- pvalue
-  output$SSW <- SSW  
   output$SSM <- SSM
   output$SSR <- SSR
-  output$dfW <- dfW
   output$dfM <- dfM
   output$dfR <- dfR
   output$MSM <- MSM
   output$MSR <- MSR
-  
+
   return(output)
 }

### 2020-12-26 R/anovacircR.test.R

@@ -1,40 +1,46 @@
-anovacirc.test <- function(data){
-  # analysis of variance using complex data
-  # an extension of the logic of the t-squared circ statistic
-  
+#' analysis of variance using complex data
+#' an extension of the logic of the t-squared circ statistic
+#' this version for repeated measures designs
+#' @export
+anovacircR.test <- function(data){
+
   grouplabels <- data$grouplabels
-  datavals <- data$simdata
   factorlist <- levels(grouplabels)
-  
+  participantlabels <- data$participant
+  participantlist <- levels(participantlabels)
+  datavals <- data$simdata
+
   grandmean <- mean(datavals)
-  
+
+  SSW <- 0
+  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
+  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))
+
   groupmeans <- NULL
   for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
-  
   SSM <- 0
   for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
-  
   dfM <- 2*(nlevels(grouplabels)-1)
-  
-  allresiduals <- 1:length(datavals)
-  for (n in 1:nlevels(grouplabels)){allresiduals[which(grouplabels==factorlist[n])] <- abs(datavals[which(grouplabels==factorlist[n])] - groupmeans[n])}
-  
-  SSR <- sum(allresiduals^2)
-  dfR <- 2*(length(datavals)-nlevels(grouplabels))
+
+  SSR <- SSW - SSM
+  dfR <- dfW - dfM
+
   MSM <- SSM/dfM
   MSR <- SSR/dfR
   Fratio <- MSM/MSR
   pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
-  
+
   output <- NULL
   output$Fratio <- Fratio
   output$pvalue <- pvalue
+  output$SSW <- SSW
   output$SSM <- SSM
   output$SSR <- SSR
+  output$dfW <- dfW
   output$dfM <- dfM
   output$dfR <- dfR
   output$MSM <- MSM
   output$MSR <- MSR
-  
+
   return(output)
 }

### 2020-12-26 R/fftshift.R

@@ -1,1 +1,3 @@
-fftshift <- function(im) {im * (-1)^(row(im) + col(im))}  
+#' one line helper function to do quadrant flip
+#' @export
+fftshift <- function(im) {im * (-1)^(row(im) + col(im))}

### 2020-12-26 R/getel.R

@@ -1,0 +1,19 @@
+#' helper function that calculates an ellipse bounding some data points
+#' the outline of the ellipse is returned
+#' @export
+getel <- function(compdata){
+  A <- cov(compdata)
+  ctr    <- colMeans(compdata)
+  RR     <- chol(A)                               # Cholesky decomposition
+  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
+  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
+  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
+  eigVal  <- eigen(A)$values
+  eigVec  <- eigen(A)$vectors
+  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
+  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
+  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
+  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles))
+  ellRot  <- eigVec %*% t(ellBase)
+  return(ellRot)
+}

### 2020-12-26 R/tsq1.test.R

@@ -1,10 +1,11 @@
+#' function to calculate a one-sample t-squared test after Hotelling (1931)
+#' for two-sample tests, use the Hotelling package
+#' @export
 tsq1.test <- function(data,mu=NULL){
-  # function to calculate a one-sample t-squared test after Hotelling (1931)
-  # for two-sample tests, use the Hotelling package
-  
+
   # compare each DV to some value
-  if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}   
-  
+  if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}
+
   xbar <- colMeans(data) # estimate means for each DV
   m <- ncol(data)   # record number of DVs
   N <- nrow(data)   # record sample size
@@ -16,7 +17,7 @@ tsq1.test <- function(data,mu=NULL){
   df2 <- N-m
   pval <- 1- pf(Fratio,df1,df2) # estimate p-value
   p.value <- min(pval,1)
-  
+
   # store outputs in a data structure
   output <- data.frame(tsq,Fratio,df1,df2,p.value)
   return(output)}

### 2020-12-26 R/tsqc.test.R

@@ -1,33 +1,34 @@
+#' function to calculate the t-squared circ statistic from Victor & Mast (1991)
+#' the inputs must be Nx2 or 2xN matrices of numbers
+#' for a one-sample test, mu is an optional vector to which the data are compared
+#' @export
 tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
-  # function to calculate the t-squared circ statistic from Victor & Mast (1991)
-  # the inputs must be Nx2 or 2xN matrices of numbers
-  # for a one-sample test, mu is an optional vector to which the data are compared
-  
+
   # transpose if dimensions are the wrong way round
   if (is.matrix(x)){d <- dim(x)
-  if (d[1]<d[2]){x <- t(x)}}   
+  if (d[1]<d[2]){x <- t(x)}}
   if (!is.null(y)){if (is.matrix(y)){d <- dim(y)
   if (d[1]<d[2]){y <- t(y)}}}
-  
+
   if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}
-  
+
   # convert data to complex representation
   x <- complex(real=x[,1],imaginary=x[,2])
   if (!is.null(y)){y <- complex(real=y[,1],imaginary=y[,2])}
-  
+
   if (is.null(y)){method <- 'One-sample t-squared circ test'}
   if (!is.null(y) & paired==TRUE){method <- 'Paired t-squared circ test'}
   if (!is.null(y) & paired==FALSE){method <- 'Independent samples t-squared circ test'}
-  
+
   if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}
-  
+
   # paired or one-sample version of the test
-  if (paired==TRUE | is.null(y)){ 
-    
+  if (paired==TRUE | is.null(y)){
+
     nobs <- length(x)
     cohmean <- mean(x)
     absmean <- abs(cohmean)
-    
+
     displacement <- abs(x - cohmean)
     diffsumsq <- sum(displacement^2)
     tsqc <- (nobs-1) * (absmean^2) / diffsumsq
@@ -35,31 +36,31 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
     df2 <- nobs*2-2
     pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
   }
-  
+
[…]

### 2020-12-31 manuscript/manuscript.Rmd
Fixed cluster correction function

@@
 author: "Daniel H. Baker"
-date: "13/12/2020"
+date: "`r Sys.Date()`"
 output:
@@
 
-packagelist <- c('bookdown','pals','shape','diagram','HDMD','FourierStats')
+packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','HDMD','FourierStats')
 missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
@@
 v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
-  ## Create a function to perform Matlab's v4 interpolation.
+  ## helper function to perform Matlab's v4 interpolation.
   ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
@@
 The $T^2_{circ}$ statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance.
-When these conditions are met, the test statistic is defined as:
+When these conditions are met, the test statistic for the one-sample case is defined as:
 
@@
 Notice that no covariance term is present in equation \@ref(eq:t2c), because of the independence assumption.
-This makes the statistic simpler to calculate, but causes problems when the assumption is violated.
+This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as we will demonstrate below).
+Conceptually, this statistic takes the vector difference between the centroid ($\bar{x}$) and a comparison point ($\mu$), and scales by the lengths of the residual vector lines joining each data point to the centroid.
+Two-sample and repeated measures versions of the $T^2_{circ}$ statistic are also possible.
 
@@
 All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats).
-These include _R_ functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
+This includes an _R_ package, featuring functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
 
@@
 
-The upper row of Figure \@ref(fig:amphists) shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components.
+The upper row of Figure \@ref(fig:amphists) shows scatterplots of simulated Fourier components, expressed using real (_x_) and imaginary (_y_) components.
 The amplitudes are the lengths of the lines joining each grey point to the origin.
 The lower row in Figure \@ref(fig:amphists) shows distributions of amplitudes for the same set of signal strengths.
-These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d > 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]).
+These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's _d_ > 2; Cohen's _d_ is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]).
 One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.
@@
 However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations.
-This advantage is lost for large sample sizes (N > 32) and large effect sizes (d > 2 when N > 8).
-These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for sample sizes, so its use should be restricted to small sample studies.
+This advantage is lost for large sample sizes (N > 32) and large effect sizes (_d_ > 2 when N > 8).
+These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use should be restricted to small sample studies.
 
@@
 Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on meeting the test's more stringent assumptions.
-The two variables must be independent, and of equal variance.
+The two variables must be independent (i.e.
+uncorrelated), and of equal variance.
 These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not.
@@
 ```{r falsealarms, fig.width=6, fig.height=9, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
-Estimates are for 100000 simulations per condition, with N=10 observations.
+Estimates are for 100000 simulations per condition, with N = 10 observations.
 The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}
@@
   ellRot  <- eigVec %*% t(ellBase) 
+  ellRot[1,] <- ellRot[1,] + ctr[1]
+  ellRot[2,] <- ellRot[2,] + ctr[2]
 
@@
   matlines(0.1*xMat+corvals[cond], 0.01*yMat+0.02, lty=1, lwd=3, col="black")
-  lines(0.1*(ellRot+ctr)[1, ]+corvals[cond], 0.01*(ellRot+ctr)[2, ]+0.02, lwd=2)
+  lines(0.1*ellRot[1, ]+corvals[cond], 0.01*ellRot[2, ]+0.02, lwd=2)
 }
@@
 mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2, las=1) 
-title(xlab="Outlier Mahalanobis distance (D)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
-mtext(text = "Proportion significant", side = 2, at=0.5, line=1.5, cex=1.5)
+title(xlab="Outlier Mahalanobis distance (D)", col.lab=rgb(0,0,0), line=1.3, cex.lab=1.2)
+mtext(text = "Proportion significant", side = 2, at=0.5, line=1.5, cex=1.3)
 points(0,2,pch=15,col=rgb(0.99,0.99,0.99),cex=0.1)
@@
 
-  polygon(0.4*(ellRot2+ctr2)[1, ]+xlocs[m], 0.1*(ellRot2+ctr2)[2, ]+1.5, col=addalpha(nbdtpal[2],0.3),border=NA)
-  lines(0.4*(ellRot+ctr)[1, ]+xlocs[m], 0.1*(ellRot+ctr)[2, ]+1.5, lwd=1)
+  polygon(0.4*ellRot2[1, ]+xlocs[m], 0.1*ellRot2[2, ]+1.5, col=addalpha(nbdtpal[2],0.3),border=NA)
+  lines(0.4*ellRot[1, ]+xlocs[m], 0.1*ellRot[2, ]+1.5, lwd=1)
   lines(0.4*c(ctr2[1],xdata[1])+xlocs[m],0.1*c(ctr2[2],ydata[1])+1.5,col=nbdtpal[2],lwd=3)
@@
 
-  text(1,1.8,'D = 1',cex=1.5,adj=0.5)
-  text(3.5,1.8,'D = 3',cex=1.5,adj=0.5)
-  text(6,1.8,'D = 5',cex=1.5,adj=0.5)
+  text(1.25,1.8,'D = 1',cex=1.5,adj=0.5)
+  text(3.75,1.8,'D = 3',cex=1.5,adj=0.5)
+  text(6.25,1.8,'D = 5',cex=1.5,adj=0.5)
 
@@
 This is a valuable statistic to include when reporting the results of multivariate tests, and we include an _R_ function (_mahalES_) that can be used to calculate it.
+
+# Controlling for multiple comparisons across location and time
+
+In many studies, it is important to compare responses over space and/or time.
+However with large numbers of sensors, voxels or temporal epochs, the familywise error rate quickly becomes problematic, inflating the Type I error (false positive) rate.
+Solutions such as Bonferroni correction, which adjust the alpha level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power.
+An alternative approach is to use cluster correction methods to control the Type I error rate, using mass univariate tests [e.g.
+@Maris2007].
+These typically involve summing test statistics such as T-values across adjacent significant locations and/or moments in time.
+The summed test statistic is compared to a null distribution generated from the same data by randomly permuting condition labels (or the sign of the data for one-sample tests).
+Such methods control the Type I error rate without substantially reducing statistical power.
+The same approach can be straightforwardly applied to the $T^2$ and $T^2_{circ}$ statistics.
+This allows a principled approach to identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli.
+The _FourierStats_ package includes an implementation of this method with options for multivariate statistics (the _clustercorrect_ function).
 
@@
 
-Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that standard Bonferroni corrections are applied.
+Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that standard multiple comparison corrections are applied (e.g.
+Bonferroni correction).
 Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary).
@@
 
-Figure \@ref(fig:mousedata)a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz.
-There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
-Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = 0.84, _df_ = 5, _p_ = 0.44).
-However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b).
-The condition index test was non-significant for both conditions (sound: CI = 1.59, _p_ = 0.66; light: CI = 1.69, _p_ = 0.59), so a paired-samples $T^2_{circ}$ test was conducted.
-This revealed a significant difference between conditions ($T^2_{circ}$ = 1.39, $F_{(2,10)}$ = 8.32, _p_ = 0.007) with an effect size of D = 1.02.
-This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
-[-@Hwang2019] went on to explore interactions between these two signals.
-
 ```{r mousedata, fig.cap = 'Summary of mouse steady-state responses to 40 Hz stimulation.
@@
 
-load('Hwangdata.RData')
+load('misc/Hwangdata.RData')
 
@@
 ellRot2 <- getel(data.frame(xdata2,ydata2))
-polygon(10*(ellRot1[1, ]+mean(xdata))+120, 1+(ellRot1[2, ]+mean(ydata))/3, col=addalpha(nbdtpal[2],0.3),border=NA)
-polygon(10*(ellRot2[1, ]+mean(xdata2))+120, 1+(ellRot2[2, ]+mean(ydata2))/3, col=addalpha(nbdtpal[3],0.3),border=NA)
+polygon(10*ellRot1[1, ]+120, 1+ellRot1[2, ]/3, col=addalpha(nbdtpal[2],0.3),border=NA)
+polygon(10*ellRot2[1, ]+120, 1+ellRot2[2, ]/3, col=addalpha(nbdtpal[3],0.3),border=NA)
 
@@
 
-# totest <- data.frame(xdata,ydata)
-# totest2 <- data.frame(xdata2,ydata2)
-# CI.test(totest)
-# CI.test(totest2)
-# tsqc.test(totest,y=totest2,paired=TRUE)
-# 
-# abs1 <- abs(rowMeans(mousedata[,1:2,11,f1]))
-# abs2 <- abs(rowMeans(mousedata[,1:2,12,f1]))
-# t.test(abs1,abs2,paired=TRUE)
-
-# colnames(totest2) <- colnames(totest)
-# fores <- rbind(totest,totest2)
-# mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=6))
-# D <- sqrt(mahala_sq$distance[1,2])
+totest <- data.frame(xdata,ydata)
+totest2 <- data.frame(xdata2,ydata2)
+cisound <- CI.test(totest)
+cilight <- CI.test(totest2)
+tsqcresult <- tsqc.test(totest,y=totest2,paired=TRUE)
+
+abs1 <- abs(rowMeans(mousedata[,1:2,11,f1]))
+abs2 <- abs(rowMeans(mousedata[,1:2,12,f1]))
+tresult <- t.test(abs1,abs2,paired=TRUE)
+
+colnames(totest2) <- colnames(totest)
+fores <- rbind(totest,totest2)
+mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=6))
+D <- sqrt(mahala_sq$distance[1,2])
 
 ```
+
+Figure \@ref(fig:mousedata)a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz.
+There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
+Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = `r round(tresult$statistic,digits=2)`, _df_ = `r tresult$parameter`, _p_ = `r round(tresult$p.value,digits=2)`).
+However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b).
+The condition index test was non-significant for both conditions (sound: CI = `r round(cisound$CI,digits=2)`, _p_ = `r round(cisound$pval,digits=2)`; light: CI = `r round(cilight$CI,digits=2)`, _p_ = `r round(cilight$pval,digits=2)`), so a paired-samples $T^2_{circ}$ test was conducted.
+This revealed a significant difference between conditions ($T^2_{circ}$ = `r round(tsqcresult$tsqc,digits=2)`, $F_{(`r tsqcresult$df1`,`r tsqcresult$df2`)}$ = `r round(tsqcresult$Fratio,digits=2)`, _p_ = `r round(tsqcresult$p.value,digits=3)`) with an effect size of D = `r round(D,digits=2)`.
+This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
+[-@Hwang2019] went on to explore interactions between these two signals.
 
@@
 Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \@ref(fig:humanSSVEP)), were well-isolated in the Fourier domain, and increased with stimulus contrast.
-For this analysis, we took responses from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant.
+Significant activity was evident at 4% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \@ref(fig:humanSSVEP).
+For the main analysis, we took responses from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant.
 Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \@ref(fig:humanSSVEP).
@@
 Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes).
+Electrodes marked in red indicate cluster-corrected significance.
 Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point).
@@
 subjmeans <- apply(alltarget[,31,,],1:2,mean)
+
+nexcluded <- NULL
+isincluded <- 1 + 0*(1:100)
+for (cond in 1:7){
+xvals <- Re(subjmeans[,cond])
+yvals <- Im(subjmeans[,cond])
+bdata <- data.frame(xvals,yvals)
+D <- sqrt(mahalanobis(bdata, colMeans(bdata), cov(bdata)))
+i <- which(D>3)
+nexcluded[cond] <- length(i)
+isincluded[i] <- 0}
+
+distmatrix <- matrix(0,nrow=64,ncol=64)
+adjacencymatrix <- matrix(0,nrow=64,ncol=64)
+for (n in 1:64){
+  for (m in 1:64){
+    xy1 <- montage$electrodelocs[montage$channelmappings[n],1:2]
+    xy2 <- montage$electrodelocs[montage$channelmappings[m],1:2]
+    distmatrix[n,m] <- as.numeric(sqrt((xy1[1]-xy2[1])^2 + (xy1[2]-xy2[2])^2))
+  }
+}
+threshold <- 0.18
+i1 <- which(distmatrix<=threshold)
+i2 <- which(distmatrix==0)
+adjacencymatrix[i1] <- 1
+adjacencymatrix[i2] <- 0
 
@@
 plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,8), ylim=c(0,2))
-nexcluded <- NULL
-isincluded <- 1 + 0*(1:100)
 
 for (cond in 1:7){
-datatoplot <- abs(apply(alltarget[,1:64,cond,], 2, mean))
+
+dataforcc <- apply(alltarget[which(isincluded==1),1:64,cond,],c(1,2),mean)
+clusts <- clustercorrect(dataforcc,adjacencymatrix=adjacencymatrix,testtype=3,clustformthresh=0.0001,clustthresh=0.01,nresamples=1000)
+
+datatoplot <- abs(apply(alltarget[which(isincluded==1),1:64,cond,], 2, mean))
 datatoplot[c(13,19)] <- mean(datatoplot)   # set the mastoids to 0
@@
 
+for (n in 1:64){points(cond+montage$electrodelocs[n,1]*0.8,1.5+montage$electrodelocs[n,2]*0.8,pch=16,col="grey",cex=0.8)}
 points(cond+montage$electrodelocs[blackelectrodes,1]*0.8,1.5+montage$electrodelocs[blackelectrodes,2]*0.8,pch=16,col="black",cex=1.5)
+
+if (clusts$nclusters>0){
+for (n in 1:clusts$nclusters){
+  toplot <- clusts$clusterpoints[[n]]
+  points(cond+montage$electrodelocs[montage$channelmappings[toplot],1]*0.8,1.5+montage$electrodelocs[montage$channelmappings[toplot],2]*0.8,pch=16,col=nbdtpal[2],cex=0.8)}}
 
@@
 i <- which(D>3)
+for (i2 in 1:length(i)){
+lines(c(xvals[i[i2]],mean(xvals[-i]))+cond,c(yvals[i[i2]],mean(yvals[-i]))+0.5,col=nbdtpal[2])}
 points(xvals[i]+cond,yvals[i]+0.5,pch=16,col=nbdtpal[2],cex=0.5)
 points(mean(xvals[-i])+cond,mean(yvals[-i])+0.5,pch=16,col=nbdtpal[3],cex=1)
-nexcluded[cond] <- length(i)
-isincluded[i] <- 0
 }
@@
 
+allCI <- NULL
+allCIp <- NULL
 for (cond in 1:7){
@@
   ellRot <- getel(data.frame(xvals,yvals))
-  lines((ellRot[1, ]+mean(xvals))+cond, (ellRot[2, ]+mean(yvals))+0.5)
+  lines(ellRot[1, ]+cond, ellRot[2, ]+0.5)
 
   text(cond,1,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
-# totest <- data.frame(xvals,yvals)
-# print(CI.test(totest))
-}
-
-# grouplabels <- as.factor(rep(1:7,each=89))
-# participant <- as.factor(rep(1:89,times=7))
-# a <- melt(cleandata[,1:7])
-# simdata <- a[,3]
-# withincirc <- anovacircR.test(data.frame(grouplabels,simdata,participant))
-
-# for (cond in 2:7){
-#   set1 <- data.frame(Re(cleandata[,1]),Im(cleandata[,1]))
-#   set2 <- data.frame(Re(cleandata[,cond]),Im(cleandata[,cond]))
-# print(tsqc.test(set1,set2,paired=TRUE))
-#   
-# colnames(set1) <- c('x','y')
-# colnames(set2) <- c('x','y')
-# fores <- rbind(set1,set2)
-# mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=89))
-# D <- sqrt(mahala_sq$distance[1,2])
-# print(D)
-# }
+totest <- data.frame(xvals,yvals)
+temp <- CI.test(totest)
+allCI[cond] <- temp$CI
+allCIp[cond] <- temp$pval
+}
+
+grouplabels <- as.factor(rep(1:7,each=89))
+participant <- as.factor(rep(1:89,times=7))
+a <- reshape2::melt(cleandata[,1:7])
+simdata <- a[,3]
+withincirc <- anovacircR.test(data.frame(grouplabels,simdata,participant))
+
+allD <- NULL
+alltsq <- data.frame(matrix(0,nrow=6,ncol=6))
+for (cond in 2:7){
+  set1 <- data.frame(Re(cleandata[,1]),Im(cleandata[,1]))
+  set2 <- data.frame(Re(cleandata[,cond]),Im(cleandata[,cond]))
+alltsq[cond-1,1:6] <- tsqc.test(set1,set2,paired=TRUE)
+
+colnames(set1) <- c('x','y')
+colnames(set2) <- c('x','y')
+fores <- rbind(set1,set2)
+mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=89))
+allD[cond-1] <- sqrt(mahala_sq$distance[1,2])
+
+}
+colnames(alltsq) <- colnames(tsqc.test(set1,set2,paired=TRUE))
 
@@
 
-With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.2, all _p_ > 0.19).
-A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(12,1056)}$ = 38.9, p < 0.001).
-Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = 0.32, $F_{(2,176)}$ = 28.4, D = 0.13, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = 0.28, $F_{(2,176)}$ = 25.3, D = 0.19, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = 0.10, $F_{(2,176)}$ = 8.5, D = 0.14, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = 0.34, $F_{(2,176)}$ = 35.3, D = 0.33, _p_ < 0.001).
+With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = `r format(round(max(allCI),digits=2),nsmall=2)`, all _p_ > `r round(min(allCIp),digits=2)`).
+A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(`r withincirc$dfM`,`r withincirc$dfR`)}$ = `r round(withincirc$Fratio,digits=1)`, _p_ < `r max(round(withincirc$pvalue,digits=3),0.001)`).
+Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[3],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[3],digits=2)`, D = `r round(allD[3],digits=2)`, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[4],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[4],digits=2)`, D = `r round(allD[4],digits=2)`, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[5],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[5],digits=2)`, D = `r round(allD[5],digits=2)`, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[6],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[6],digits=2)`, D = `r round(allD[6],digits=2)`, _p_ < 0.001).
 The study by Vilidaite et al.
@@
 [-@Pei2017] suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat demanding, though we have implemented an _R_ function to calculate error bars using this method (available through the _getampCI_ function).
+This approach is somewhat demanding, though we have implemented an _R_ function to calculate error bars using this method (available through the _amperrors_ function).
 However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95% confidence intervals on amplitudes.

### 2020-12-31 manuscript/manuscript.tex
Fixed cluster correction function

@@
 \author{Daniel H. Baker}
-\date{13/12/2020}
+\date{2020-12-30}
 
@@
 An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components.
-Here we compare the performance of two multivariate extensions of the T-test, Hotelling's \(T^2\) and a variant called \(T^2_{circ}\).
-We develop a novel test of the assumptions of \(T^2_{circ}\) based on the condition index, and a heuristic for excluding outliers using the Mahanalobis distance.
+Here we compare the performance of two multivariate extensions of the T-test: Hotelling's \(T^2\) and a variant called \(T^2_{circ}\).
+We develop a novel test of the assumptions of \(T^2_{circ}\) based on the condition index, and suggest a heuristic for excluding outliers using the Mahanalobis distance.
 We then extend the statistic to multi-level designs, and propose a new statistical test termed \(ANOVA^2_{circ}\).
@@
 The \(T^2_{circ}\) statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance.
-When these conditions are met, the test statistic is defined as:
+When these conditions are met, the test statistic for the one-sample case is defined as:
 
@@
 Notice that no covariance term is present in equation \eqref{eq:t2c}, because of the independence assumption.
-This makes the statistic simpler to calculate, but causes problems when the assumption is violated.
+This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as we will demonstrate below).
+Conceptually, this statistic takes the vector difference between the centroid (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the lengths of the residual vector lines joining each data point to the centroid.
+Two-sample and repeated measures versions of the \(T^2_{circ}\) statistic are also possible.
 
@@
 All scripts used to generate this manuscript are available at: \url{https://github.com/bakerdh/FourierStats}.
-These include \emph{R} functions to implement one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.
+This includes an \emph{R} package, featuring functions to implement one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.
 
@@
 
-The upper row of Figure \ref{fig:amphists} shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components.
+The upper row of Figure \ref{fig:amphists} shows scatterplots of simulated Fourier components, expressed using real (\emph{x}) and imaginary (\emph{y}) components.
 The amplitudes are the lengths of the lines joining each grey point to the origin.
 The lower row in Figure \ref{fig:amphists} shows distributions of amplitudes for the same set of signal strengths.
-These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d \textgreater{} 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen (1988)).
+These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's \emph{d} \textgreater{} 2; Cohen's \emph{d} is the mean difference scaled by the standard deviation, see Cohen (1988)).
 One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.
@@
 However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations.
-This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (d \textgreater{} 2 when N \textgreater{} 8).
-These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for sample sizes, so its use should be restricted to small sample studies.
+This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (\emph{d} \textgreater{} 2 when N \textgreater{} 8).
+These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use should be restricted to small sample studies.
 
@@
 Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on meeting the test's more stringent assumptions.
-The two variables must be independent, and of equal variance.
+The two variables must be independent (i.e.~uncorrelated), and of equal variance.
 These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not.
@@
 \caption{\label{fig:falsealarms}Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
-Estimates are for 100000 simulations per condition, with N=10 observations.
+Estimates are for 100000 simulations per condition, with N = 10 observations.
 The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.}
@@
 
+\hypertarget{controlling-for-multiple-comparisons-across-location-and-time}{%
+\section{Controlling for multiple comparisons across location and time}\label{controlling-for-multiple-comparisons-across-location-and-time}}
+
+In many studies, it is important to compare responses over space and/or time.
+However with large numbers of sensors, voxels or temporal epochs, the familywise error rate quickly becomes problematic, inflating the Type I error (false positive) rate.
+Solutions such as Bonferroni correction, which adjust the alpha level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power.
+An alternative approach is to use cluster correction methods to control the Type I error rate, using mass univariate tests (e.g.
+Maris and Oostenveld 2007).
+These typically involve summing test statistics such as T-values across adjacent significant locations and/or moments in time.
+The summed test statistic is compared to a null distribution generated from the same data by randomly permuting condition labels (or the sign of the data for one-sample tests).
+Such methods control the Type I error rate without substantially reducing statistical power.
+The same approach can be straightforwardly applied to the \(T^2\) and \(T^2_{circ}\) statistics.
+This allows a principled approach to identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli.
+The \emph{FourierStats} package includes an implementation of this method with options for multivariate statistics (the \emph{clustercorrect} function).
+
 \hypertarget{generalising-to-more-than-two-conditions}{%
@@
 
-Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that standard Bonferroni corrections are applied.
+Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that standard multiple comparison corrections are applied (e.g.~Bonferroni correction).
 Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary).
@@
 
-The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data.
+The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
 Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios.
@@
 
-Figure \ref{fig:mousedata}a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz.
-There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
-Indeed, a paired t-test on the amplitudes reveals no significant difference (t=0.84, df=5, p=0.44).
-However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \ref{fig:mousedata}b).
-The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted.
-This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of D = 1.02.
-This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
-(2019) went on to explore interactions between these two signals.
-
 \begin{figure}
@@
 
+Figure \ref{fig:mousedata}a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz.
+There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
+Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (\emph{t} = 0.84, \emph{df} = 5, \emph{p} = 0.44).
+However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \ref{fig:mousedata}b).
+The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted.
+This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of D = 1.02.
+This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
+(2019) went on to explore interactions between these two signals.
+
 \hypertarget{human-visual-steady-state-data}{%
@@
 Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast.
-For this analysis, we took responses from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant.
+Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \ref{fig:humanSSVEP}.
+For the main analysis, we took responses from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant.
 Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}.
@@
 
-With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.2, all \emph{p} \textgreater{} 0.19).
-A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, p \textless{} 0.001).
-Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.4, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.3, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.5, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.34, \(F_{(2,176)}\) = 35.3, D = 0.33, \emph{p} \textless{} 0.001).
+With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.20, all \emph{p} \textgreater{} 0.19).
+A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, \emph{p} \textless{} 0.001).
+Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, D = 0.33, \emph{p} \textless{} 0.001).
 The study by Vilidaite et al.
@@
 When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al.
-2020).
+2021).
 This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.
@@
 The present paper has focussed on the Frequentist statistical tradition.
-However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses.
+However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses (Jeffreys 1961).
 Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al.
-2009).
+2009) and ANOVA (Rouder et al.
+2017).
 However this is a non-trivial undertaking, and is beyond the scope of the current paper.
@@
 If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance.
-Such methods have been hugely influential in the fMRI literature, and for analysing event-related potential data collected using EEG or MEG.
+Such methods have been hugely influential in the fMRI literature (Schwarzkopf and Rees 2011), and for analysing event-related potential data collected using EEG or MEG (Grootswagers, Wardle, and Carlson 2017).
 However, they have not been widely applied to steady-state data (though see West et al.
@@
 (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat computationally demanding, though we have implemented an \emph{R} function to calculate error bars using this method.
+This approach is somewhat demanding, though we have implemented an \emph{R} function to calculate error bars using this method (available through the \emph{getampCI} function).
 However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95\% confidence intervals on amplitudes.
@@
 
-\leavevmode\hypertarget{ref-Baker2020}{}%
+\leavevmode\hypertarget{ref-Baker2021}{}%
 Baker, D.H., G.
@@
 Andrews.
-2020.
+2021.
 ``Power Contours: Optimising Sample Size and Precision in Experimental Psychology and Human Neuroscience.'' \emph{Psychological Methods}.
@@
 2020.
-``Nonlinear Transduction of Emotional Facial Expression.'' \emph{Vision Res} 170 (May): 1--11.
+``Nonlinear Transduction of Emotional Facial Expression.'' \emph{Vision Res} 170: 1--11.
 \url{https://doi.org/10.1016/j.visres.2020.03.004}.
+
+\leavevmode\hypertarget{ref-Grootswagers2017}{}%
+Grootswagers, Tijl, Susan G Wardle, and Thomas A Carlson.
+2017.
+``Decoding Dynamic Brain Patterns from Evoked Responses: A Tutorial on Multivariate Pattern Analysis Applied to Time Series Neuroimaging Data.'' \emph{J Cogn Neurosci} 29 (4): 677--97.
+\url{https://doi.org/10.1162/jocn_a_01068}.
 
@@
 
+\leavevmode\hypertarget{ref-Jeffreys1961}{}%
+Jeffreys, H. 1961. \emph{Theory of Probability}. 3rd ed. Oxford University Press.
+
 \leavevmode\hypertarget{ref-Liu-Shuang2014}{}%
@@
 2014.
-``An Objective Index of Individual Face Discrimination in the Right Occipito-Temporal Cortex by Means of Fast Periodic Oddball Stimulation.'' \emph{Neuropsychologia} 52 (January): 57--72.
+``An Objective Index of Individual Face Discrimination in the Right Occipito-Temporal Cortex by Means of Fast Periodic Oddball Stimulation.'' \emph{Neuropsychologia} 52: 57--72.
 \url{https://doi.org/10.1016/j.neuropsychologia.2013.10.022}.
@@
 Mahalanobis, P.C. 1936. ``On the Generalised Distance in Statistics.'' \emph{Proceedings of the National Academy of Sciences of India} 2: 49--55.
+
+\leavevmode\hypertarget{ref-Maris2007}{}%
+Maris, Eric, and Robert Oostenveld. 2007. ``Nonparametric Statistical Testing of EEG- and MEG-Data.'' \emph{J Neurosci Methods} 164 (1): 177--90. \url{https://doi.org/10.1016/j.jneumeth.2007.03.024}.
 
@@
 2017.
-``Development of Contrast Normalization Mechanisms During Childhood and Adolescence.'' \emph{Vision Res} 133 (April): 12--20.
+``Development of Contrast Normalization Mechanisms During Childhood and Adolescence.'' \emph{Vision Res} 133: 12--20.
 \url{https://doi.org/10.1016/j.visres.2016.03.010}.
@@
 
+\leavevmode\hypertarget{ref-Rouder2017}{}%
+Rouder, Jeffrey N, Richard D Morey, Josine Verhagen, April R Swagman, and Eric-Jan Wagenmakers.
+2017.
+``Bayesian Analysis of Factorial Designs.'' \emph{Psychol Methods} 22 (2): 304--21.
+\url{https://doi.org/10.1037/met0000057}.
+
 \leavevmode\hypertarget{ref-Rouder2009}{}%
@@
 \url{https://doi.org/10.3758/PBR.16.2.225}.
+
+\leavevmode\hypertarget{ref-Schwarzkopf2011}{}%
+Schwarzkopf, D S, and G Rees. 2011. ``Pattern Classification Using Functional Magnetic Resonance Imaging.'' \emph{Wiley Interdiscip Rev Cogn Sci} 2 (5): 568--79. \url{https://doi.org/10.1002/wcs.141}.
 
@@
 2015.
-``Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.'' \emph{Sci Rep} 5 (November): 16933.
+``Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.'' \emph{Sci Rep} 5: 16933.
 \url{https://doi.org/10.1038/srep16933}.

### 2020-12-31 R/CI.test.R

@@ -1,7 +1,12 @@
-#' function to calculate the condition index of bivariate data
-#' compare for significance with an expected distribution function
+#' CI.test: function to calculate the condition index of bivariate data
+#' the condition index is the ratio of eigenvalues (eigenvector lengths), calculated longest/shortest
+#' the index is then compared for significance with an expected distribution function
+#' significant tests (p<0.05) violate the assumptions of the T-squared-circ and ANOVA-squared-circ tests
+#' see Baker (2021) for further details
 #' @export
 CI.test <- function(data){
+
+  if (is.complex(data)){data <- data.frame(Re(data),Im(data))}
 
   C <- cov(data)  # covariance matrix
   N <- nrow(data) # sample size

### 2020-12-31 R/FourierStats-package.r

@@ -1,1 +1,14 @@
 #' Documentation for FourierStats R package
+#' The FourierStats package implements several statistical tests for use with bivariate data such as the complex (real and imaginary) components of the Fourier spectrum. Included functions implement the T-squared-circ test of Victor & Mast (1991), and the condition index and ANOVA-squared-circ described by Baker (2021).
+#' Functions include:
+#'  - analysecplx: a function to analyse complex Fourier data, following the guidelines from the Baker (2021) paper
+#'  - tsqc.test: implementation of the T-squared-circ test of Victor & Mast (1991)
+#'  - tsq1.test: one-sample Hotelling's T-squared test
+#'  - CI.test: condition index test, to test the assumptions of the T-squared-circ test
+#'  - anovacirc.test: implementation of one-way between subjects ANOVA-squared-circ test
+#'  - anovacircR.test: implementation of one-way repeated measures ANOVA-squared-circ test
+#'  - amperrors: function to calculate error bars for amplitudes incorporating coherent averaging
+#'  - getel: helper function that calculates the bounding ellipse for a cloud of points
+#'  - fftshift: helper function that performs the quadrant shift of a 2D Fourier spectrum
+#'
+#'  DHB 26/12/20

### 2020-12-31 R/amperrors.R

@@ -1,5 +1,79 @@
-#' placeholder for errorbar function amperrors
+#' amperrors: calculate error bars on the amplitude component of coherently averaged data
+#' the input is a vector of complex numbers, or an Nx2 matrix of real and imaginary components
+#' the optional 'method' flag allows the following values:
+#'  - 'boot': (default) bootstraps the confidence intervals (10000 iterations)
+#'  - 'circ': calculates confidence intervals based on a circular bounding region (used when the condition index test is non-significant)
+#'  - 'ellipse': calculates confidence intervals based on an elliptical bounding region (based on Pei et al. 2017, doi: 10.1016/j.visres.2016.03.010)
+#'  - 'abs': calculates the standard error (and the mean) using the absolute amplitudes
+#'  the optional 'quantiles' flag allows the quantile to be set for the confidence intervals
+#'  typical values are 95 (for 95% confidence intervals), and 68 (for standard errors)
+#'  the function returns the mean amplitude and the upper and lower error bars
 #' @export
-amperrors <- function(input){
+amperrors <- function(input, method='boot',quantiles=95){
+  d <- dim(input)
+  compdata <- input
+  if (!is.complex(input)){compdata <- complex(real=input[,1],imaginary=input[,2])}
+  xydata <- data.frame(Re(compdata),Im(compdata))
+  # compdata is now expressed in complex values, xy data is an Nx2 data frame of real and imaginary values
+  output <- NULL
+  output$meanamp <- abs(mean(compdata)) # calculate the mean amplitude
 
-return(input)}
+  if (method=='boot'){
+    # bootstrap the error bars on the amplitude using 10000 resamples
+    bspop <- NULL
+    for (n in 1:10000){bspop[n] <- abs(mean(sample(compdata,length(compdata),replace=TRUE)))}
+    output$upperCI <- quantile(bspop,probs=1-(1-(quantiles/100))/2)
+    output$lowerCI <- quantile(bspop,probs=(1-(quantiles/100))/2)
+  }
+
+  if (method=='circ'){
+    # pools the variance in the x (real) and y (imaginary) directions
+    # calculates the standard error based on this estimate
[…]

### 2020-12-31 R/analysecplx.R

@@ -1,0 +1,143 @@
+#' analysecplx: wrapper function that decides which statistical test to run for a given data set
+#' the expected input is a data frame in long format:
+#' one column should contain complex valued data, or two columns headed x and y contain the real and imaginary parts
+#' if a column is headed 'group' (or any word beginning with a g) is included, this will be interpreted as level data for the independent variable
+#' if a column is headed 'participant' (or any word beginning with a p) is included, this will be interpreted as a random factor
+#'
+#' the function first runs the condition index test for each level of the IV
+#' if the condition index test is non-significant for all levels, a T-squared-circ test is run for one- and two-sample designs, or an ANOVA-squared-circ test otherwise
+#' if the condition index test is significant for any level, a T-squared or MANOVA test is run instead
+#' the paired-samples Mahalanobis effect size statistic is also calculated
+#' @export
+analysecplx <- function(input){
+
+output <- NULL
+
+# first check inputs and get into sensible format
+quiterror <- 0
+colheadings <- tolower(colnames(input))
+# search for columns starting with x, y, g and p
+xcol <- 0
+ycol <- 0
+groupcol <- 0
+randcol <- 0
+for (n in 1:ncol(input)){
+  temp <- substr(colheadings[n],start=1,stop=1)
+  if (pmatch('x',temp,nomatch=0)==1){xcol <- n}
+  if (pmatch('y',temp,nomatch=0)==1){ycol <- n}
+  if (pmatch('g',temp,nomatch=0)==1){groupcol <- n}
+  if (pmatch('p',temp,nomatch=0)==1){randcol <- n}
+}
+
+# look for a column containing complex data
+findcplx <- 0*ncol(input)
+for (n in 1:ncol(input)){if (is.complex(input[1,n])){findcplx[n] <- 1}}
+if (sum(findcplx)>0){cplxdata <- input[,which(findcplx==1)[1]]}
+
+# if no complex data found, look for columns headed x and y
+if (sum(findcplx)==0){
+  if (xcol>0 && ycol>0){cplxdata <- complex(real=input[,xcol], imaginary=input[,ycol])}
+  if (xcol==0 && ycol==0){quiterror <- 1
[…]

### 2020-12-31 R/anovacirc.test.R

@@ -1,5 +1,10 @@
-#' analysis of variance using complex data
+#' anovacirc.test: two-dimensional analysis of variance using complex data
 #' an extension of the logic of the t-squared circ statistic
+#' this is the one-way, between subjects implementation of the test
+#' the expected input is a long-format data frame with the following columns:
+#' simdata: the data stored as complex values
+#' grouplabels: condition labels indicating the level of the independent variable each data point corresponds to
+#' see Baker (2021) for further details
 #' @export
 anovacirc.test <- function(data){
 

### 2020-12-31 R/anovacircR.test.R

@@ -1,6 +1,11 @@
-#' analysis of variance using complex data
+#' anovacircR.test: two-dimensional analysis of variance using complex data
 #' an extension of the logic of the t-squared circ statistic
-#' this version for repeated measures designs
+#' this is the one-way, repeated measures implementation of the test
+#' the expected input is a long-format data frame with the following columns:
+#' simdata: the data stored as complex values
+#' grouplabels: condition labels indicating the level of the independent variable each data point corresponds to
+#' participant: participant ID number
+#' see Baker (2021) for further details
 #' @export
 anovacircR.test <- function(data){
 

### 2020-12-31 R/clustercorrect.R

@@ -1,0 +1,199 @@
+#' clustercorrect: implements non-parametric cluster correction method described by Maris & Oostenveld (2007) J Neurosci Methods, 164: 177-190, doi: 10.1016/j.jneumeth.2007.03.024
+#' this version can deal with complex data using the T-squared or T-squared-circ statistics
+#'
+#' The input datax should be an N (participants) x m (sensors/timepoints/locations) matrix
+#' The optional input datay can either be a matrix of the same dimensions, or N2 x m (for unpaired designs with unbalanced samples), or a single value to compare to datax
+#'
+#' Legal values of the testtype input are 1: t-test, 2: Hotelling's T-squared, 3: Victor & Mast's T-squared-circ
+#' Complex values are required for options 2 & 3. If complex values are passed in for test 1 (t-test), they will be converted to amplitudes
+#'
+#' if the measures are from adjacent time points, an adjacency matrix will be automatically constructed
+#' if the measures are from different spatial locations (e.g. an electrode montage) you should calculate an adjacency matrix yourself
+#' this should be an m x m matrix (where m is the number of sensors), where values are zero for non-adjacent sensor pairs, and 1 for adjacent pairs
+#'
+#' outputs an object containing the indices of all significant clusters
+#' @export
+clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,paired=TRUE,clustformthresh=0.05,clustthresh=0.05,nresamples=10000) {
+  clustout <- NULL
+  sigclustercount <- 0
+  clusterpout <- NULL
+  clusterpoints <- list()
+  nulldist <- NULL
+  N <- nrow(datax)  # number of observations (e.g. participants)
+  m <- ncol(datax)  # number of measurements (e.g. sensors, voxels, timepoints)
+
+  # if complex data have been passed in and t-test selected, convert to amplitudes
+  if (is.complex(datax) && testtype==1){datax <- abs(datax)}
+  if (is.complex(datay) && testtype==1){datay <- abs(datay)}
+
+  # if two data sets have been passed in, note the size of the second matrix
+  if (!pracma::isempty(datay)){N2 <- nrow(datay)}
+
+  # if it's a paired design, subtract the two data sets and delete the second one
[…]

### 2020-12-31 R/fftshift.R

@@ -1,3 +1,5 @@
-#' one line helper function to do quadrant flip
+#' one line helper function to do quadrant flip for 2D Fourier transforms
+#' shifts the low spatial frequencies to the centre of the spectrum
+#' this is included in most programming languages, but is absent from the base R Fourier transform tools
 #' @export
 fftshift <- function(im) {im * (-1)^(row(im) + col(im))}

### 2020-12-31 R/getel.R

@@ -1,5 +1,6 @@
-#' helper function that calculates an ellipse bounding some data points
-#' the outline of the ellipse is returned
+#' getel: helper function that calculates an ellipse bounding some data points
+#' the outline of the ellipse is returned, sampled at 200 points
+#' the input should be an Nx2 matrix of observations
 #' @export
 getel <- function(compdata){
   A <- cov(compdata)
@@ -15,5 +16,7 @@ getel <- function(compdata){
   yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
   ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles))
   ellRot  <- eigVec %*% t(ellBase)
+  ellRot[1,] <- ellRot[1,] + ctr[1]
+  ellRot[2,] <- ellRot[2,] + ctr[2]
   return(ellRot)
 }

### 2020-12-31 R/tsq1.test.R

@@ -1,8 +1,14 @@
-#' function to calculate a one-sample t-squared test after Hotelling (1931)
-#' for two-sample tests, use the Hotelling package
+#' tsq1.test: function to calculate a one-sample T-squared test of Hotelling (1931)
+#' the input is an N (observations) by m (dependent variables) matrix
+#' optional input mu is a vector with m entries, to which the data are compared (otherwise compared to 0)
+#' a repeated measures version can be calculated by subtracting the two conditions and then running the one-sample test
+#' for two-sample tests, use the hotelling.test function from the Hotelling package
 #' @export
 tsq1.test <- function(data,mu=NULL){
 
+  # if data are complex, convert to real and absolute values
+  if (is.complex(data)){data <- data.frame(Re(data),Im(data))}
+  if (is.complex(mu)){mu <- c(Re(mu),Im(mu))}
   # compare each DV to some value
   if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}
 

### 2020-12-31 R/tsqc.test.R

@@ -1,6 +1,8 @@
-#' function to calculate the t-squared circ statistic from Victor & Mast (1991)
+#' tsqc.test: function to calculate the T-squared circ statistic from Victor & Mast (1991)
 #' the inputs must be Nx2 or 2xN matrices of numbers
 #' for a one-sample test, mu is an optional vector to which the data are compared
+#' for a two-sample test, x and y are the data from the two conditions
+#' if paired=TRUE, the matrices (x and y) must be the same size
 #' @export
 tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
 
@@ -13,8 +15,8 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
   if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}
 
   # convert data to complex representation
-  x <- complex(real=x[,1],imaginary=x[,2])
-  if (!is.null(y)){y <- complex(real=y[,1],imaginary=y[,2])}
+  if (!is.complex(x)){x <- complex(real=x[,1],imaginary=x[,2])}
+  if (!is.null(y)){if (!is.complex(y)){y <- complex(real=y[,1],imaginary=y[,2])}}
 
   if (is.null(y)){method <- 'One-sample t-squared circ test'}
   if (!is.null(y) & paired==TRUE){method <- 'Paired t-squared circ test'}

### 2020-12-31 manuscript/misc/testamperrors.R

@@ -1,0 +1,40 @@
+tic()
+N <- 10
+simdata <- matrix(rnorm(2*N,mean=0.5,sd=0.2),nrow=N,ncol=2)
+simdata <- complex(real=simdata[,1],imaginary=simdata[,2])
+qval <- 95
+bootstrapped <- amperrors(simdata, method='boot',quantiles=qval)
+circle <- amperrors(simdata, method='circ',quantiles=qval)
+ellipse <- amperrors(simdata, method='ellipse',quantiles=qval)
+absamp <- amperrors(simdata, method='abs',quantiles=qval)
+toc()
+
+plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,5), ylim=c(0,1))
+ticklocsx <- 0:5   # locations of tick marks on x axis
+ticklocsy <- (0:5)/5    # locations of tick marks on y axis
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+
+trueCI <- 1.96*(0.2/sqrt(N))
+polygon(c(0,5,5,0), 0.5*sqrt(2)+c(-1,-1,1,1)*trueCI, col=rgb(0,0,0,alpha=0.2),border=NA)
+lines(c(0,5),0.5*sqrt(c(2,2)),lty=2)
+
+
+arrows(1,absamp$meanamp,x1=1, y1=absamp$upperCI, length=0.015, angle=90, lwd=2, col='black')
+arrows(1,absamp$meanamp,x1=1, y1=absamp$lowerCI, length=0.015, angle=90, lwd=2, col='black')
+arrows(2,circle$meanamp,x1=2, y1=circle$upperCI, length=0.015, angle=90, lwd=2, col='black')
+arrows(2,circle$meanamp,x1=2, y1=circle$lowerCI, length=0.015, angle=90, lwd=2, col='black')
+arrows(3,ellipse$meanamp,x1=3, y1=ellipse$upperCI, length=0.015, angle=90, lwd=2, col='black')
+arrows(3,ellipse$meanamp,x1=3, y1=ellipse$lowerCI, length=0.015, angle=90, lwd=2, col='black')
+arrows(4,bootstrapped$meanamp,x1=4, y1=bootstrapped$upperCI, length=0.015, angle=90, lwd=2, col='black')
[…]

### 2020-12-31 manuscript/misc/testanalysecplx.R

@@ -1,0 +1,61 @@
+N <- 10
+
+x <- rnorm(2*N)
+y <- rnorm(2*N)
+x[(N+1):(2*N)] <- x[(N+1):(2*N)] + 0.1
+xy <- complex(real=x,imaginary=y)
+group <- rep(1:2,each=N)
+participant <- rep(1:N,times=2)
+
+input <- data.frame(xy,group)
+analysecplx(input)
+
+input <- data.frame(xy,group,participant)
+analysecplx(input)
+
+
+x <- rnorm(2*N)
+y <- rnorm(2*N)
+x <- x + 0.5*y
+xy <- complex(real=x,imaginary=y)
+group <- rep(1:2,each=N)
+participant <- rep(1:N,times=2)
+
+input <- data.frame(xy,group)
+analysecplx(input)
+
+input <- data.frame(xy,group,participant)
+analysecplx(input)
+
+
+N <- 20
+x <- rnorm(3*N)
+y <- rnorm(3*N)
+x[(N+1):(2*N)] <- x[(N+1):(2*N)] + 0.1
+xy <- complex(real=x,imaginary=y)
+group <- rep(1:3,each=N)
+participant <- rep(1:N,times=3)
+
+input <- data.frame(xy,group)
+analysecplx(input)
+
+input <- data.frame(xy,group,participant)
+analysecplx(input)
+
+
+
+N <- 20
+x <- rnorm(3*N)
+y <- rnorm(3*N) + 0.5*x
+x[(N+1):(2*N)] <- x[(N+1):(2*N)] + 0.5
+xy <- complex(real=x,imaginary=y)
+group <- rep(1:3,each=N)
+participant <- rep(1:N,times=3)
+
+input <- data.frame(xy,group)
+analysecplx(input)
+
+input <- data.frame(xy,group,participant)
+analysecplx(input)
+
+

### 2020-12-31 manuscript/misc/testcluster.R

@@ -1,0 +1,237 @@
+
+N <- 10
+m <- 12
+
+data <- matrix(rnorm(N*m),nrow=N,ncol=12)
+data[,3:6] <- data[,3:6] + 1
+
+allp <- NULL
+for (i in 1:m){
+  output <- t.test(data[,i])
+  allp[i] <- output$p.value
+}
+
+clusts <- clustercorrect(data,testtype=1,clustformthresh=0.05,nresamples=1000)
+
+
+
+load("/Users/danbaker/Google Drive/Current work/Research/powercontours/headplotspec.RData")
+
+load('~/Google Drive/Current work/R scripts/danlabtoolbox/data/montagedata.RData')
+
+distmatrix <- matrix(0,nrow=64,ncol=64)
+adjacencymatrix <- matrix(0,nrow=64,ncol=64)
+for (n in 1:64){
+  for (m in 1:64){
+    xy1 <- montage$electrodelocs[montage$channelmappings[n],1:2]
+    xy2 <- montage$electrodelocs[montage$channelmappings[m],1:2]
+    distmatrix[n,m] <- as.numeric(sqrt((xy1[1]-xy2[1])^2 + (xy1[2]-xy2[2])^2))
+  }
+}
+threshold <- 0.18
+i1 <- which(distmatrix<=threshold)
+i2 <- which(distmatrix==0)
+adjacencymatrix[i1] <- 1
+adjacencymatrix[i2] <- 0
+
+par(pty="s")  # make axis square
+plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(-0.6,0.6), ylim=c(-0.6,0.6))
+
+for (n in 1:64){
+  for (m in 1:64){
+    if (adjacencymatrix[n,m]){
+      lines(montage$electrodelocs[montage$channelmappings[c(n,m)],1],montage$electrodelocs[montage$channelmappings[c(n,m)],2])
+    }
+}}
+points(montage$electrodelocs[1:64,1],montage$electrodelocs[1:64,2],pch=16,col='grey')
+
+lines(montage$headoutline[,1],montage$headoutline[,2],col="black",lwd=2)
+lines(montage$noseoutline[,1],montage$noseoutline[,2],col="black",lwd=2)
+lines(montage$Rearoutline[,1],montage$Rearoutline[,2],col="black",lwd=2)
+lines(montage$Learoutline[,1],montage$Learoutline[,2],col="black",lwd=2)
+
+
+
[…]

### 2020-12-31 manuscript/manuscript.Rmd
Incorporated example data into package. Improvements to anova.circ (merged independent and repeated measures versions into a single function).

@@
 bibliography: bibliography.bib
-biblio-style: apalike
+csl: elife.csl
+# biblio-style: apalike
 ---
@@
 
-# devtools::install_github("bakerdh/FourierStats")
-# devtools::install('FourierStats')
-
-packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','HDMD','FourierStats')
+packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','HDMD','devtools')
 missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
@@
 
-nsims <- 10000  # global value for the whole script - increase to 100000 for final run
+# install the FourierStats package from github using the devtools package
+packagelist <- c('FourierStats')
+missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
+if (length(missingpackages)>0){devtools::install_github("bakerdh/FourierStats")}
+toinstall <- packagelist[which(!packagelist %in% (.packages()))]
+invisible(lapply(toinstall,library,character.only=TRUE))
+
+nsims <- 10000  # global value for the whole script - increase to 100000 for final run, but note this takes around 15 hours to complete
 
@@
 v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
-  ## helper function to perform Matlab's v4 interpolation.
+  ## helper function to perform interpolation.
   ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
@@
 
+# helper function to add transparency to a colour
 addalpha <- function(col, alpha=1){apply(sapply(col, col2rgb)/255, 2, function(x) rgb(x[1], x[2], x[3], alpha=alpha))}
@@
 Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency.
-The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components.
+The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase).
 However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded.
 An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components.
-Here we compare the performance of two multivariate extensions of the T-test: Hotelling's $T^2$ and a variant called $T^2_{circ}$.
-We develop a novel test of the assumptions of $T^2_{circ}$ based on the condition index, and suggest a heuristic for excluding outliers using the Mahanalobis distance.
-We then extend the statistic to multi-level designs, and propose a new statistical test termed $ANOVA^2_{circ}$.
+Here the performance of two multivariate extensions of the T-test are compared: Hotelling's $T^2$ and a variant called $T^2_{circ}$.
+A novel test of the assumptions of $T^2_{circ}$ is developed, based on the condition index of the data (the square root of the ratio of eigenvalues of a bounding ellipse), and a heuristic for excluding outliers using the Mahalanobis distance is proposed.
+The $T^2_{circ}$ statistic is then extended to multi-level designs, resulting in a new statistical test termed $ANOVA^2_{circ}$.
 This has identical assumptions to $T^2_{circ}$, and is shown to be more sensitive than MANOVA when these assumptions are met.
-We demonstrate how these tests might be applied to two publicly available empirical data sets, and suggest concrete guidance for choosing which test to run.
-Implementations of these novel tools are provided as _R_ functions, and we hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
+The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run.
+Implementations of these novel tools are provided as an _R_ package, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
+
+_Keywords:_ multivariate statistics, Fourier analysis, steady-state, condition index, Mahalanobis distance.
 
@@
 This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations.
-These periodic responses, often called the _steady-state_ or _frequency following_ response, can be recorded from single neurons [@Enroth-Cugell1966] and local field potentials [@Morrone1987] using invasive methods, or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans [@Norcia2015] and in diverse animal species including insects [@Afsari2014], birds [@Porciatti1990], rodents [@Hwang2019] and primates [@Nakayama1982].
+These periodic responses, often called the _steady-state_ or _frequency following_ response, can be recorded using invasive methods from single neurons [@Enroth-Cugell1966] and local field potentials [@Morrone1987], or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans [@Norcia2015] and in diverse animal species including insects [@Afsari2014], birds [@Porciatti1990], rodents [@Hwang2019] and primates [@Nakayama1982].
 Steady-state methods are used to measure early sensory responses in vision [@Regan1966], hearing [@Rees1986] and somatosensation [@Snyder1992], and closely related paradigms have been developed to target specific stimulus features such as orientation [@Braddick1986], and facial expression [@Gray2020] and identity [@Liu-Shuang2014].
@@
 A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal.
-The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a convenient, and well-isolated index of the brain's response.
+The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response.
 Fourier spectra comprise both amplitude and phase information, which can be equivalently expressed as complex numbers with real and imaginary components.
@@
 Conceptually, the $T^2$ statistic extends the univariate T-statistic by incorporating the covariance between the dependent variables.
-Two-sample and paired variants are also available, and the test can be applied to an arbitrary number of dependent variables (though here we will consider only the bivariate case).
+Two-sample and paired variants are also available, and the test can be applied with an arbitrary number of dependent variables (though here only the bivariate case will be considered).
 
 More recently, Victor and Mast [-@Victor1991] proposed a simpler version of $T^2$, called $T^2_{circ}$.
-The $T^2_{circ}$ statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance.
+The $T^2_{circ}$ statistic makes the strong assumption that the dependent variables (real and imaginary Fourier components) are uncorrelated and have equal variance.
 When these conditions are met, the test statistic for the one-sample case is defined as:
@@
 Notice that no covariance term is present in equation \@ref(eq:t2c), because of the independence assumption.
-This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as we will demonstrate below).
-Conceptually, this statistic takes the vector difference between the centroid ($\bar{x}$) and a comparison point ($\mu$), and scales by the lengths of the residual vector lines joining each data point to the centroid.
+This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as will be demonstrated below).
+Conceptually, this statistic takes the vector difference between the centroid ($\bar{x}$) and a comparison point ($\mu$), and scales by the mean length of the residual vector lines joining each data point to the centroid.
 Two-sample and repeated measures versions of the $T^2_{circ}$ statistic are also possible.
 
-In the present paper, we develop best practice guidelines for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms.
-We first demonstrate why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed.
-We then investigate conditions under which we should use either the $T^2$ or $T^2_{circ}$ statistic.
+In the present paper, best practice guidelines are developed for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms.
+It is first demonstrated why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed.
+Then conditions are investigated under which either the $T^2$ or $T^2_{circ}$ statistic should be used.
 The range of sample sizes and effect sizes where $T^2_{circ}$ is more sensitive is identified.
-We develop a novel method for testing the assumptions of the $T^2_{circ}$ statistic, based on calculating the condition index of a multivariate data set.
+A novel method for testing the assumptions of the $T^2_{circ}$ statistic is developed, based on calculating the condition index of a multivariate data set.
 Appropriate methods for identifying outliers using the Mahalanobis distance are discussed, and a heuristic proposed.
 Next the logic of $T^2_{circ}$ is extended to situations with more than two levels of the independent variable, and the performance of this novel $ANOVA^2_{circ}$ statistic is compared to MANOVA.
-Finally, we demonstrate the proposed techniques on two example SSVEP data sets, and recommend some best practice guidelines for analysis decisions.
+Finally, the proposed techniques are demonstrated on two example SSVEP data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.
 
 All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats).
-This includes an _R_ package, featuring functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
+This includes an _R_ package called _FourierStats_, featuring functions to implement a one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
 
@@
 
-```{r amphists, fig.cap="Demonstration of skew in absolute Fourier amplitudes for weak signals.
+```{r amphists, fig.cap="Demonstration of skew in absolute Fourier amplitudes for signals of different strengths.
 Signal strength is quantified as Cohen's d, defined as the ratio of the mean to the standard deviation of the sample.
@@
 These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's _d_ > 2; Cohen's _d_ is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]).
-One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.
+One consequence of this is that T-tests will potentially have an inflated Type I error (false positive) rate for many signals encountered empirically, especially if used to make pointwise comparisons to an amplitude of 0.
 
@@
 
-Using the $T^2$ or $T^2_{circ}$ statistic allows us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
+Using the $T^2$ or $T^2_{circ}$ statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
 Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$.
 This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant.
-Their simulations show the largest advantage for $T^2_{circ}$ for effect sizes around d = 1 (where the mean signal strength is equal to the standard deviation of the data).
+Their simulations show the largest advantage for $T^2_{circ}$ for effect sizes around _d_ = 1 (where the mean signal strength is equal to the standard deviation of the data).
 The advantage appeared to be stronger for smaller sample sizes.
 
-Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and confirm that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f).
+Here these simulations are replicated and extended (see Figure \@ref(fig:powerfig)), and it is shown that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f).
 However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations.
-This advantage is lost for large sample sizes (N > 32) and large effect sizes (_d_ > 2 when N > 8).
+This advantage is lost for large sample sizes (N > 32) and large effect sizes (when _d_ > 2 and N > 8).
 These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use should be restricted to small sample studies.
 
-```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes.
+```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes (100,000 simulated data sets per condition).
 Panels a-c replicate conditions reported by Victor and Mast (1991).
@@
 
-Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on meeting the test's more stringent assumptions.
+Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on satisfying the $T^2_{circ}$ test's more stringent assumptions.
 The two variables must be independent (i.e.
@@
 Figure \@ref(fig:falsealarms) shows the results of simulations with randomly generated bivariate data in which no signal is present.
-When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (dashed line).
+When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (horizontal dashed lines).
 However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the variances of the two dependent variables more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2.
@@
 
-```{r falsealarms, fig.width=6, fig.height=9, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
-Estimates are for 100000 simulations per condition, with N = 10 observations.
+```{r falsealarms, fig.width=6, fig.height=7, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
+Estimates are for 100,000 simulated data sets per condition, with N = 10 observations.
 The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}
 
-par(mfrow=c(2,1))
+par(mfrow=c(2,1), mar=c(3,4,1,2))
 
@@
 However, this will reduce the statistical power of the $T^2_{circ}$ test, and its advantage over $T^2$ is relatively marginal in most situations to begin with (see Figure \@ref(fig:powerfig)).
-What is required is a method to objectively assess whether the assumptions of $T^2_{circ}$ hold, which we develop in the following section.
+What is required is a method to objectively assess whether the assumptions of $T^2_{circ}$ hold; this is developed in the following section.
 
@@
 
-Despite the severe consequences of violating the assumptions of the $T^2_{circ}$ statistic (see Figure \@ref(fig:falsealarms)), there is currently no accepted test for those assumptions that could be applied to an empirical data set.
+Despite the severe consequences of violating the assumptions of the $T^2_{circ}$ statistic (see Figure \@ref(fig:falsealarms)), there is currently no accepted test of those assumptions that could be applied to an empirical data set.
 Victor and Mast [-@Victor1991] suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components [see also @Pei2017].
@@
 For uncorrelated random numbers the distribution of condition indices is positively skewed, with a minimum of 1 [@Edelman1988].
-This is because two independent samples of numbers from the same underlying distribution will still by chance have unequal eigenvectors.
-Edelman [-@Edelman1988] provides an equation (Eq 14) for the probability density function of condition indices (x) as a function of sample size (n):
+This is because two independent samples of numbers from the same underlying distribution will generally by chance have unequal eigenvectors, and the definition of the condition index ($\sqrt{longest/shortest}$) prevents its values dropping below 1 (values < 1 would imply that the shortest eigenvector is longer than the longest one).
+Edelman [-@Edelman1988] provides an equation (his Eq.
+14) for the probability density function of condition indices (_x_) as a function of sample size (_N_):
 
@@
 \label{eq:edelman1}
-pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)},
+pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)},
 \end{equation}
 
-Attempts to validate this by simulation suggest that for small sample sizes (n<10) a closer approximation is given by:
+Attempts to validate this by simulation suggest that for small sample sizes (_N_ < 10) a closer approximation is given by:
 
@@
 \label{eq:edelman2}
-pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}
+pdf = (N-2)2^{N-2}\frac{x^2 - 1}{(x^2 + 1)^{(N-1)}}x^{(N-3)}.
 \end{equation}
 
-In Figure \@ref(fig:distcomparison)a we show an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100000 random data sets for n=10 (blue shading).
-The vertical lines show the critical (95%) threshold for the analytic and simulated results - a ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance.
-Figure \@ref(fig:distcomparison)b shows how these thresholds change as a function of the number of observations, and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
-
-```{r distcomparison, fig.cap='Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of 10 observations.
-The vertical blue and red lines show the 95 percent thresholds on the curve (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function.
-Panel (b) shows how 95 percent thresholds change as a functio of the number of observations.', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
+Figure \@ref(fig:distcomparison)a shows an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100,000 random data sets with _N_ = 10 (blue shading).
+The vertical lines show the critical (95%) threshold for the analytic and simulated results.
+A ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance, because it has a condition index larger than expected by chance (assuming $\alpha$ = 0.05).
+Ratios below the threshold imply that the eigenvalues are equal (in a statistical sense).
+Figure \@ref(fig:distcomparison)b shows how these critical thresholds change as a function of the number of observations (N), and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
+
+```{r distcomparison, fig.cap='Logic of the condition index test.
+Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of N = 10 observations.
+The vertical blue and red lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function.
+Ellipse icons above panel (a) illustrate different condition indices between 1 and 5 (note that rotation does not affect the condition index).
+Panel (b) shows how 95 percent thresholds change as a function of the number of observations.', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
 
@@
 
-plotlims <- c(1,5,0,1) 
+plotlims <- c(1,5,0,1.2) 
 ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
@@
 plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
-axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)
 axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
-mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklocsx, side = 1, at=ticklocsx)
 mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2) 
 title(xlab="Condition index", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
-title(ylab="Probability", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(ylab="Probability density      ", line=1.5, cex.lab=1.5)
 
@@
 text(4.8,0.95,'(a)',cex=1.5)
+
+xpos <- c(1.2,2.1,3,3.9,4.8)
+angles <- seq(0, 2*pi, length.out=200) 
+RR <- matrix(c(1,0,0,1),nrow=2,ncol=2)
+for (n in 1:5){
+  RR[2,2] <- 1/sqrt(n)
+  ell <- 1 * cbind(cos(angles), sin(angles)) %*% RR
+  polygon((ell[,1]*0.4*(5/6))+xpos[n],(ell[,2]/10)+1.1,col=rgb(0.9,0.9,0.9),border=NA)
+}
 
@@
 lines(3:25,edelman2,lwd=3,col=nbdtpal[2])
-lines(3:25,simratios,lwd=3,col=nbdtpal[3])
-legend(12, 15, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=1, lwd=3, box.lwd=2)
+lines(3:25,simratios,lwd=3,lty=2,col=nbdtpal[3])
+legend(12, 15, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=c(1,1,2), lwd=3, box.lwd=2)
 text(23,24,'(b)',cex=1.5)
@@
 The eigenvalue ratio can be used as a test of the assumptions of $T^2_{circ}$.
-If we observe a condition index above the critical threshold for the number of observations then the data set can be said to significantly violate the assumption of equal eigenvalues.
-Because our modified equation provides an inverse density function (red curve in Figure \@ref(fig:distcomparison)a), we can use this to calculate a p-value for the test.
+If a condition index is observed that is above the critical threshold for the number of observations, then the data set can be said to significantly violate the assumption of equal eigenvalues.
+Because the modified equation provides an inverse density function (red curve in Figure \@ref(fig:distcomparison)a), this can be used to calculate a _p_-value for the test.
 If the test is non-significant, one can proceed with $T^2_{circ}$; if it is significant, $T^2$ should be used instead.
-An _R_ function implementing this test is included in the project code repository (the _CI.test_ function contained in the _CItest.R_ script).
+An _R_ function implementing this test is included in the _FourierStats_ package (the _CI.test_ function).
 
@@
 Figure \@ref(fig:outlierplot) shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves).
-In all cases, the functions depart from the Type 1 error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic return $D^2$, which can be converted to D by taking the square root).
+In all cases, the functions depart from the Type I error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the Mahalanobis distance exceeds a value around 3.
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core _mahalanobis_ function in _R_, return $D^2$, which can be converted to D by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the $T^2_{circ}$ test.
@@
 The upper row shows three example data sets, each with a single outlier shown in red.
-The outliers have Mahalanobis distances of 1, 3 and 5.
+The outliers have approximate Mahalanobis distances of 1, 3 and 5.
 The ellipses are calculated with the outlier included (red) and excluded (black), illustrating how the outlier distorts the aspect ratio of the ellipse.
@@
 @Giudice2009].
-This is a valuable statistic to include when reporting the results of multivariate tests, and we include an _R_ function (_mahalES_) that can be used to calculate it.
+This is a valuable statistic to include when reporting the results of multivariate tests, and an _R_ function (_pairwise.mahalanobis_) to calculate it is available as part of the _HDMD_ package ([https://CRAN.R-project.org/package=HDMD](https://CRAN.R-project.org/package=HDMD)).
 
@@
 
-In many studies, it is important to compare responses over space and/or time.
-However with large numbers of sensors, voxels or temporal epochs, the familywise error rate quickly becomes problematic, inflating the Type I error (false positive) rate.
-Solutions such as Bonferroni correction, which adjust the alpha level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power.
+In some studies, it is important to compare responses over space and/or time.
+However with large numbers of sensors, voxels or temporal epochs, the familywise error quickly becomes problematic, inflating the Type I error (false positive) rate.
+Solutions such as Bonferroni correction, which adjust the $\alpha$ level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power.
 An alternative approach is to use cluster correction methods to control the Type I error rate, using mass univariate tests [e.g.
@@
 Such methods control the Type I error rate without substantially reducing statistical power.
-The same approach can be straightforwardly applied to the $T^2$ and $T^2_{circ}$ statistics.
-This allows a principled approach to identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli.
+The same approach can be applied to the $T^2$ and $T^2_{circ}$ statistics.
+This allows a principled method for identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli.
 The _FourierStats_ package includes an implementation of this method with options for multivariate statistics (the _clustercorrect_ function).
@@
 Many studies involve more than two conditions that need to be compared.
-Issues with familywise error will quickly become problematic if multiple pairwise $T^2$ or $T^2_{circ}$ statistics are calculated.
+Again, issues with familywise error will quickly become problematic if multiple pairwise $T^2$ or $T^2_{circ}$ statistics are calculated.
 One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's $T^2$, but permits multiple independent variables with more than two levels.
-However, if the assumptions of $T^2_{circ}$ hold for a data set, it might alternatively be possible to extend to logic of the $T^2_{circ}$ test [@Victor1991] to the more general case.
-
-The F statistic for a one-way ANOVA is calculated by taking the ratio of the variance explained by the modelled group means to the residual unexplained variance.
+However, if the assumptions of $T^2_{circ}$ hold for a data set, it might alternatively be possible to extend the logic of the $T^2_{circ}$ test [@Victor1991] to the more general case.
+
+The F-statistic for a one-way ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance.
 For multivariate data, the change in group mean would be calculated using the vector distances between the complex means, and the residuals are the vector distances between each data point and its corresponding group mean.
-For an independent one-way design with $k$ groups (or conditions) and $N$ observations per group, the F distribution will have $2(k-1)$ and $2((Nk)-k)$ degrees of freedom.
+For an independent one-way design with $k$ groups (or conditions) and $N$ observations per group, the F-distribution will have $2(k-1)$ and $2((Nk)-k)$ degrees of freedom.
 The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number.
+A _p_-value can be derived from the F-distribution in the usual way.
+
 A suitable name for such a test might be ${ANOVA}^2_{circ}$, as this reflects the similarity to ANOVA, and the extension of the logic of $T^2_{circ}$ (an alternative name might be $MANOVA_{circ}$, however this feels less appropriate given that many of the key features of MANOVA are absent).
-
-Figure \@ref(fig:powerfig2) shows simulations analogous to those in Figure \@ref(fig:powerfig) for a one-way design with three levels.
+Figure \@ref(fig:powerfig2) shows simulations analogous to those in Figure \@ref(fig:powerfig) for a one-way between-subjects design with three levels.
 MANOVA is directly compared to the ${ANOVA}^2_{circ}$ statistic across a range of effect sizes and sample sizes.
@@
 The format mirrors that of Figure 2.
-In these simulations, there were three conditions, with the signal being added to one condition.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+In these simulations, there were three conditions, with the signal being added to one condition only.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
 
@@
   output <- anovacirc.test(data.frame(simdata,grouplabels))
-  if (output$pvalue<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
+  if (output$p.value<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
 
@@
 
-Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that standard multiple comparison corrections are applied (e.g.
+Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.
 Bonferroni correction).
@@
 A repeated measures version of ${ANOVA}^2_{circ}$ has also been implemented following the same logic.
-The project repository contains _R_ scripts for both of these functions.
+The project repository contains an _R_ function (_anovacirc.test_) to run both of these tests.
 
@@
 The flowchart in Figure \@ref(fig:flowchart) contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
-Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios.
+Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test.
 Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, and the $T^2$ statistic if not.
@@
 Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs.
-In _R_, the _Manova_ function from the _car_ package appears to be capable of the necessary calculations, as does the commercial software package SPSS.
+In _R_, the _Manova_ function from the _car_ package appears to be capable of the necessary calculations, as does the commercial statistics package _SPSS_.
 
@@
 
-# Applying these methods to empirical data sets
-
-Having developed some novel tools for the analysis of periodic data, in the following sections we demonstrate them using two different publicly available empirical data sets.
+# Applying multivariate methods to empirical data sets
+
+Having developed some novel tools for the analysis of periodic data, in the following sections their use is demonstrated for two different publicly available empirical data sets.
 The first study recorded responses to auditory and optogenetic stimulation in mice.
 The second study measured visual responses to flickering grating patterns in humans.
-These cases provide examples of how the results of the tests described above might be appropriately reported.
+These examples also provide an illustration of how the results of the tests described above might be appropriately reported.
 
@@
 [-@Hwang2020], and was downloaded from: [https://doi.gin.g-node.org/10.12751/g-node.e5tyek/](https://doi.gin.g-node.org/10.12751/g-node.e5tyek/).
+A processed data file is included with permission from the authors in the _FourierStats_ package.
 
@@
 Panel (a) shows the Fourier amplitude spectrum with inset scalp plots for sound (red) and light (blue) stimulation, averaged across repetitions and individuals.
+Grey and black points in the insets indicate electrode locations.
 Panel (b) shows complex (x = real, y = imaginary) Fourier components for 6 individual mice (small points) and their average (large points) for both conditions.', fig.width=10, fig.height=6, fig.align="center", echo=FALSE}
 
-load('misc/Hwangdata.RData')
+# example data are bundled with the package
+load(system.file("extdata", "Hwangdata.RData", package = "FourierStats"))
 
@@
 datatoplot <- abs(meandata[,11,f1])
-testDat<- data.frame(x = montage[,2],
-                     y = -montage[,3],
+testDat<- data.frame(x = mousemontage[,2],
+                     y = -mousemontage[,3],
                      z = datatoplot)
@@
 rmax2 <- rmax+0.1
-maskx <- outline[which(outline[,1]>=0),1]
-masky <- outline[which(outline[,1]>=0),2]
+maskx <- mouseoutline[which(mouseoutline[,1]>=0),1]
+masky <- mouseoutline[which(mouseoutline[,1]>=0),2]
 polygon(2.4*c(0,maskx,0,rmax2,rmax2,0)+20,1.5+c(rmax2,masky,-rmax2,-rmax2,rmax2,rmax2)/14,border=NA,col="white")
-maskx <- outline[which(outline[,1]<=0),1]
-masky <- outline[which(outline[,1]<=0),2]
+maskx <- mouseoutline[which(mouseoutline[,1]<=0),1]
+masky <- mouseoutline[which(mouseoutline[,1]<=0),2]
 i <- sort(masky,index.return=TRUE)
@@
 
-lines(2.4*outline[,1]+20,1.5+outline[,2]/14,lwd=2)
-points(2.4*montage[3:38,2]+20,1.5+montage[3:38,3]/14,pch=16,col='grey')
-points(2.4*montage[1:2,2]+20,1.5+montage[1:2,3]/14,pch=16)
+lines(2.4*mouseoutline[,1]+20,1.5+mouseoutline[,2]/14,lwd=2)
+points(2.4*mousemontage[3:38,2]+20,1.5+mousemontage[3:38,3]/14,pch=16,col='grey')
+points(2.4*mousemontage[1:2,2]+20,1.5+mousemontage[1:2,3]/14,pch=16)
 
@@
 datatoplot <- abs(meandata[,12,f1])
-testDat<- data.frame(x = montage[,2],
-                     y = -montage[,3],
+testDat<- data.frame(x = mousemontage[,2],
+                     y = -mousemontage[,3],
                      z = datatoplot)
@@
 rmax2 <- rmax+0.1
-maskx <- outline[which(outline[,1]>=0),1]
-masky <- outline[which(outline[,1]>=0),2]
+maskx <- mouseoutline[which(mouseoutline[,1]>=0),1]
+masky <- mouseoutline[which(mouseoutline[,1]>=0),2]
 polygon(2.4*c(0,maskx,0,rmax2,rmax2,0)+60,1.5+c(rmax2,masky,-rmax2,-rmax2,rmax2,rmax2)/14,border=NA,col="white")
-maskx <- outline[which(outline[,1]<=0),1]
-masky <- outline[which(outline[,1]<=0),2]
+maskx <- mouseoutline[which(mouseoutline[,1]<=0),1]
+masky <- mouseoutline[which(mouseoutline[,1]<=0),2]
 i <- sort(masky,index.return=TRUE)
@@
 
-lines(2.4*outline[,1]+60,1.5+outline[,2]/14,lwd=2)
-points(2.4*montage[3:38,2]+60,1.5+montage[3:38,3]/14,pch=16,col='grey')
-points(2.4*montage[1:2,2]+60,1.5+montage[1:2,3]/14,pch=16)
+lines(2.4*mouseoutline[,1]+60,1.5+mouseoutline[,2]/14,lwd=2)
+points(2.4*mousemontage[3:38,2]+60,1.5+mousemontage[3:38,3]/14,pch=16,col='grey')
+points(2.4*mousemontage[1:2,2]+60,1.5+mousemontage[1:2,3]/14,pch=16)
 
@@
 text(122,1.97,'3',cex=1.5,adj=0.5)
+text(95,1.1,'Real',cex=1.5,adj=0.5)
+text(117,0.28,'Imaginary',cex=1.5,adj=0.5,srt=90)
 
@@
 Significant activity was evident at 4% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \@ref(fig:humanSSVEP).
-For the main analysis, we took responses from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant.
+For the main analysis, responses were taken from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant.
 Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \@ref(fig:humanSSVEP).
@@
 
-load("/Users/danbaker/Google Drive/Current work/Research/powercontours/headplotspec.RData")
-
-load('~/Google Drive/Current work/R scripts/danlabtoolbox/data/montagedata.RData')
-
-subjmeans <- apply(alltarget[,31,,],1:2,mean)
+# example data are bundled with the package
+load(system.file("extdata", "humanSSVEPdata.RData", package = "FourierStats"))
+
+subjmeans <- alltarget[,31,]
 
@@
 
-dataforcc <- apply(alltarget[which(isincluded==1),1:64,cond,],c(1,2),mean)
+dataforcc <- alltarget[which(isincluded==1),1:64,cond]
 clusts <- clustercorrect(dataforcc,adjacencymatrix=adjacencymatrix,testtype=3,clustformthresh=0.0001,clustthresh=0.01,nresamples=1000)
 
-datatoplot <- abs(apply(alltarget[which(isincluded==1),1:64,cond,], 2, mean))
+datatoplot <- abs(apply(alltarget[which(isincluded==1),1:64,cond], 2, mean))
 datatoplot[c(13,19)] <- mean(datatoplot)   # set the mastoids to 0
@@
 
-for (n in 1:64){points(cond+montage$electrodelocs[n,1]*0.8,1.5+montage$electrodelocs[n,2]*0.8,pch=16,col="grey",cex=0.8)}
+for (n in 1:64){points(cond+montage$electrodelocs[n,1]*0.8,1.5+montage$electrodelocs[n,2]*0.8,pch=16,col="grey",cex=0.5)}
 points(cond+montage$electrodelocs[blackelectrodes,1]*0.8,1.5+montage$electrodelocs[blackelectrodes,2]*0.8,pch=16,col="black",cex=1.5)
@@
   toplot <- clusts$clusterpoints[[n]]
-  points(cond+montage$electrodelocs[montage$channelmappings[toplot],1]*0.8,1.5+montage$electrodelocs[montage$channelmappings[toplot],2]*0.8,pch=16,col=nbdtpal[2],cex=0.8)}}
+  points(cond+montage$electrodelocs[montage$channelmappings[toplot],1]*0.8,1.5+montage$electrodelocs[montage$channelmappings[toplot],2]*0.8,pch=16,col=nbdtpal[2],cex=0.5)}}
 
@@
 
-  text(cond,1,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
+  text(cond,1.02,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
 totest <- data.frame(xvals,yvals)
@@
 simdata <- a[,3]
-withincirc <- anovacircR.test(data.frame(grouplabels,simdata,participant))
+withincirc <- anovacirc.test(data.frame(simdata,grouplabels,participant))
 
@@
 With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = `r format(round(max(allCI),digits=2),nsmall=2)`, all _p_ > `r round(min(allCIp),digits=2)`).
-A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(`r withincirc$dfM`,`r withincirc$dfR`)}$ = `r round(withincirc$Fratio,digits=1)`, _p_ < `r max(round(withincirc$pvalue,digits=3),0.001)`).
+A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(`r withincirc$dfM`,`r withincirc$dfR`)}$ = `r round(withincirc$Fratio,digits=1)`, _p_ < `r max(round(withincirc$p.value,digits=3),0.001)`).
 Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[3],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[3],digits=2)`, D = `r round(allD[3],digits=2)`, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[4],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[4],digits=2)`, D = `r round(allD[4],digits=2)`, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[5],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[5],digits=2)`, D = `r round(allD[5],digits=2)`, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[6],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[6],digits=2)`, D = `r round(allD[6],digits=2)`, _p_ < 0.001).
@@
 [-@Vilidaite2018] compared SSVEP responses between individuals with and without autism, as well as in a _Drosophila_ genetic model of developmental disorders.
-The raw data are available at: [http://dx.doi.org/10.17605/OSF.IO/Y4N5K](http://dx.doi.org/10.17605/OSF.IO/Y4N5K).
+The raw data are available at: [http://dx.doi.org/10.17605/OSF.IO/Y4N5K](http://dx.doi.org/10.17605/OSF.IO/Y4N5K), and a processed version is included with the _FourierStats_ package.
 
@@
 However, they have not been widely applied to steady-state data (though see West et al.
-[-@West2015] for an example).
+[-@West2015] for one example).
 In principle, the real and imaginary Fourier components can be treated as separate dependent variables, along with different recording locations and/or frequencies.
@@
 [-@Pei2017] suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat demanding, though we have implemented an _R_ function to calculate error bars using this method (available through the _amperrors_ function).
-However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95% confidence intervals on amplitudes.
+This approach is somewhat demanding, though an _R_ function is provided to calculate error bars using this method (available through the _amperrors_ function).
+However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes.
 This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean.
-The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled amplitudes.
-Upper and lower 95% confidence intervals on the amplitude can then be taken at appropriate quantiles of this population.
+The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled mean amplitudes.
+Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68% and 95% are typical, with the 68% interval corresponding to the standard error).
 
@@
 
-The simulations reported here allow us to make several recommendations for how periodic data should be analysed.
+The simulations reported here allow several recommendations to be made for how periodic data should be analysed.
 Multivariate statistics should be used for phase-locked Fourier data instead of univariate statistics such as T-tests and ANOVA.
-This avoids problems from distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information.
-Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahanalobis distance reported as a measure of effect size.
+This avoids problems from non-normal distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information.
+Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahalanobis distance reported as a measure of effect size.
 For sample sizes of N < 32, the $T^2_{circ}$ and $ANOVA^2_{circ}$ statistics can be used if the condition index test is non-significant for all conditions.

### 2020-12-31 manuscript/manuscript.tex
Incorporated example data into package. Improvements to anova.circ (merged independent and repeated measures versions into a single function).

@@
 \author{Daniel H. Baker}
-\date{2020-12-30}
+\date{2020-12-31}
 
@@
 Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency.
-The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components.
+The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase).
 However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded.
 An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components.
-Here we compare the performance of two multivariate extensions of the T-test: Hotelling's \(T^2\) and a variant called \(T^2_{circ}\).
-We develop a novel test of the assumptions of \(T^2_{circ}\) based on the condition index, and suggest a heuristic for excluding outliers using the Mahanalobis distance.
-We then extend the statistic to multi-level designs, and propose a new statistical test termed \(ANOVA^2_{circ}\).
+Here the performance of two multivariate extensions of the T-test are compared: Hotelling's \(T^2\) and a variant called \(T^2_{circ}\).
+A novel test of the assumptions of \(T^2_{circ}\) is developed, based on the condition index of the data (the square root of the ratio of eigenvalues of a bounding ellipse), and a heuristic for excluding outliers using the Mahalanobis distance is proposed.
+The \(T^2_{circ}\) statistic is then extended to multi-level designs, resulting in a new statistical test termed \(ANOVA^2_{circ}\).
 This has identical assumptions to \(T^2_{circ}\), and is shown to be more sensitive than MANOVA when these assumptions are met.
-We demonstrate how these tests might be applied to two publicly available empirical data sets, and suggest concrete guidance for choosing which test to run.
-Implementations of these novel tools are provided as \emph{R} functions, and we hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
+The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run.
+Implementations of these novel tools are provided as an \emph{R} package, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
+
+\emph{Keywords:} multivariate statistics, Fourier analysis, steady-state, condition index, Mahalanobis distance.
 
@@
 This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations.
-These periodic responses, often called the \emph{steady-state} or \emph{frequency following} response, can be recorded from single neurons (Enroth-Cugell and Robson 1966) and local field potentials (Morrone, Burr, and Speed 1987) using invasive methods, or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans (Norcia et al.
-2015) and in diverse animal species including insects (Afsari et al.
-2014), birds (Porciatti et al.
-1990), rodents (Hwang et al.
-2019) and primates (Nakayama and Mackeben 1982).
-Steady-state methods are used to measure early sensory responses in vision (Regan 1966), hearing (Rees, Green, and Kay 1986) and somatosensation (Snyder 1992), and closely related paradigms have been developed to target specific stimulus features such as orientation (Braddick, Wattam-Bell, and Atkinson 1986), and facial expression (Gray et al.
-2020) and identity (Liu-Shuang, Norcia, and Rossion 2014).
-In fMRI research, \emph{travelling wave} methods (Engel et al.
-1994; Sereno et al.
-1995) are used to map the retinotopic responses of early visual cortex using stimuli that change periodically in spatial position.
-Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way (Spitschan et al.
-2014).
+These periodic responses, often called the \emph{steady-state} or \emph{frequency following} response, can be recorded using invasive methods from single neurons (Enroth-Cugell and Robson, 1966) and local field potentials (Morrone et al., 1987), or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans (Norcia et al., 2015) and in diverse animal species including insects (Afsari et al., 2014), birds (Porciatti et al., 1990), rodents (Hwang et al., 2019) and primates (Nakayama and Mackeben, 1982).
+Steady-state methods are used to measure early sensory responses in vision (Regan, 1966), hearing (Rees et al., 1986) and somatosensation (Snyder, 1992), and closely related paradigms have been developed to target specific stimulus features such as orientation (Braddick et al., 1986), and facial expression (Gray et al., 2020) and identity (Liu-Shuang et al., 2014).
+In fMRI research, \emph{travelling wave} methods (Engel et al., 1994; Sereno et al., 1995) are used to map the retinotopic responses of early visual cortex using stimuli that change periodically in spatial position.
+Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way (Spitschan et al., 2014).
 
 A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal.
-The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a convenient, and well-isolated index of the brain's response.
+The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response.
 Fourier spectra comprise both amplitude and phase information, which can be equivalently expressed as complex numbers with real and imaginary components.
@@
 
-For pointwise and pairwise comparisons, Hotelling's \(T^2\) statistic (Hotelling 1931) is a multivariate extension of the T-test. For the one-sample case, the test statistic is defined as:
+For pointwise and pairwise comparisons, Hotelling's \(T^2\) statistic (Hotelling, 1931) is a multivariate extension of the T-test. For the one-sample case, the test statistic is defined as:
 
@@
 Conceptually, the \(T^2\) statistic extends the univariate T-statistic by incorporating the covariance between the dependent variables.
-Two-sample and paired variants are also available, and the test can be applied to an arbitrary number of dependent variables (though here we will consider only the bivariate case).
+Two-sample and paired variants are also available, and the test can be applied with an arbitrary number of dependent variables (though here only the bivariate case will be considered).
 
 More recently, Victor and Mast (1991) proposed a simpler version of \(T^2\), called \(T^2_{circ}\).
-The \(T^2_{circ}\) statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance.
+The \(T^2_{circ}\) statistic makes the strong assumption that the dependent variables (real and imaginary Fourier components) are uncorrelated and have equal variance.
 When these conditions are met, the test statistic for the one-sample case is defined as:
@@
 Notice that no covariance term is present in equation \eqref{eq:t2c}, because of the independence assumption.
-This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as we will demonstrate below).
-Conceptually, this statistic takes the vector difference between the centroid (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the lengths of the residual vector lines joining each data point to the centroid.
+This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as will be demonstrated below).
+Conceptually, this statistic takes the vector difference between the centroid (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the mean length of the residual vector lines joining each data point to the centroid.
 Two-sample and repeated measures versions of the \(T^2_{circ}\) statistic are also possible.
 
-In the present paper, we develop best practice guidelines for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms.
-We first demonstrate why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed.
-We then investigate conditions under which we should use either the \(T^2\) or \(T^2_{circ}\) statistic.
+In the present paper, best practice guidelines are developed for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms.
+It is first demonstrated why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed.
+Then conditions are investigated under which either the \(T^2\) or \(T^2_{circ}\) statistic should be used.
 The range of sample sizes and effect sizes where \(T^2_{circ}\) is more sensitive is identified.
-We develop a novel method for testing the assumptions of the \(T^2_{circ}\) statistic, based on calculating the condition index of a multivariate data set.
+A novel method for testing the assumptions of the \(T^2_{circ}\) statistic is developed, based on calculating the condition index of a multivariate data set.
 Appropriate methods for identifying outliers using the Mahalanobis distance are discussed, and a heuristic proposed.
 Next the logic of \(T^2_{circ}\) is extended to situations with more than two levels of the independent variable, and the performance of this novel \(ANOVA^2_{circ}\) statistic is compared to MANOVA.
-Finally, we demonstrate the proposed techniques on two example SSVEP data sets, and recommend some best practice guidelines for analysis decisions.
+Finally, the proposed techniques are demonstrated on two example SSVEP data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.
 
 All scripts used to generate this manuscript are available at: \url{https://github.com/bakerdh/FourierStats}.
-This includes an \emph{R} package, featuring functions to implement one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.
+This includes an \emph{R} package called \emph{FourierStats}, featuring functions to implement a one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.
 
@@
 
-\caption{Demonstration of skew in absolute Fourier amplitudes for weak signals.
+\caption{Demonstration of skew in absolute Fourier amplitudes for signals of different strengths.
 Signal strength is quantified as Cohen's d, defined as the ratio of the mean to the standard deviation of the sample.
@@
 These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's \emph{d} \textgreater{} 2; Cohen's \emph{d} is the mean difference scaled by the standard deviation, see Cohen (1988)).
-One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.
+One consequence of this is that T-tests will potentially have an inflated Type I error (false positive) rate for many signals encountered empirically, especially if used to make pointwise comparisons to an amplitude of 0.
 
@@
 
-Using the \(T^2\) or \(T^2_{circ}\) statistic allows us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
+Using the \(T^2\) or \(T^2_{circ}\) statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
 Victor \& Mast (1991) report simulations showing situations where \(T^2_{circ}\) has greater power than \(T^2\).
 This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant.
-Their simulations show the largest advantage for \(T^2_{circ}\) for effect sizes around d = 1 (where the mean signal strength is equal to the standard deviation of the data).
+Their simulations show the largest advantage for \(T^2_{circ}\) for effect sizes around \emph{d} = 1 (where the mean signal strength is equal to the standard deviation of the data).
 The advantage appeared to be stronger for smaller sample sizes.
 
-Here we replicate and extend these simulations (see Figure \ref{fig:powerfig}), and confirm that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f).
+Here these simulations are replicated and extended (see Figure \ref{fig:powerfig}), and it is shown that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f).
 However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations.
-This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (\emph{d} \textgreater{} 2 when N \textgreater{} 8).
+This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (when \emph{d} \textgreater{} 2 and N \textgreater{} 8).
 These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use should be restricted to small sample studies.
@@
 
-\caption{Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes.
+\caption{Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes (100,000 simulated data sets per condition).
 Panels a-c replicate conditions reported by Victor and Mast (1991).
@@
 
-Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on meeting the test's more stringent assumptions.
+Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on satisfying the \(T^2_{circ}\) test's more stringent assumptions.
 The two variables must be independent (i.e.~uncorrelated), and of equal variance.
@@
 Figure \ref{fig:falsealarms} shows the results of simulations with randomly generated bivariate data in which no signal is present.
-When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (dashed line).
+When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (horizontal dashed lines).
 However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the variances of the two dependent variables more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2.
@@
 \caption{\label{fig:falsealarms}Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
-Estimates are for 100000 simulations per condition, with N = 10 observations.
+Estimates are for 100,000 simulated data sets per condition, with N = 10 observations.
 The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.}
@@
 However, this will reduce the statistical power of the \(T^2_{circ}\) test, and its advantage over \(T^2\) is relatively marginal in most situations to begin with (see Figure \ref{fig:powerfig}).
-What is required is a method to objectively assess whether the assumptions of \(T^2_{circ}\) hold, which we develop in the following section.
+What is required is a method to objectively assess whether the assumptions of \(T^2_{circ}\) hold; this is developed in the following section.
 
@@
 
-Despite the severe consequences of violating the assumptions of the \(T^2_{circ}\) statistic (see Figure \ref{fig:falsealarms}), there is currently no accepted test for those assumptions that could be applied to an empirical data set.
-Victor and Mast (1991) suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components (see also Pei et al.
-2017).
+Despite the severe consequences of violating the assumptions of the \(T^2_{circ}\) statistic (see Figure \ref{fig:falsealarms}), there is currently no accepted test of those assumptions that could be applied to an empirical data set.
+Victor and Mast (1991) suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components (see also Pei et al., 2017).
 
@@
 Conventionally, the condition index is calculated as the square root of the longest/shortest eigenvector length.
-For uncorrelated random numbers the distribution of condition indices is positively skewed, with a minimum of 1 (Edelman 1988).
-This is because two independent samples of numbers from the same underlying distribution will still by chance have unequal eigenvectors.
-Edelman (1988) provides an equation (Eq 14) for the probability density function of condition indices (x) as a function of sample size (n):
+For uncorrelated random numbers the distribution of condition indices is positively skewed, with a minimum of 1 (Edelman, 1988).
+This is because two independent samples of numbers from the same underlying distribution will generally by chance have unequal eigenvectors, and the definition of the condition index (\(\sqrt{longest/shortest}\)) prevents its values dropping below 1 (values \textless{} 1 would imply that the shortest eigenvector is longer than the longest one).
+Edelman (1988) provides an equation (his Eq.
+14) for the probability density function of condition indices (\emph{x}) as a function of sample size (\emph{N}):
 
@@
 \label{eq:edelman1}
-pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)},
+pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)},
 \end{equation}
 
-Attempts to validate this by simulation suggest that for small sample sizes (n\textless{}10) a closer approximation is given by:
+Attempts to validate this by simulation suggest that for small sample sizes (\emph{N} \textless{} 10) a closer approximation is given by:
 
@@
 \label{eq:edelman2}
-pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}
+pdf = (N-2)2^{N-2}\frac{x^2 - 1}{(x^2 + 1)^{(N-1)}}x^{(N-3)}.
 \end{equation}
 
-In Figure \ref{fig:distcomparison}a we show an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100000 random data sets for n=10 (blue shading).
-The vertical lines show the critical (95\%) threshold for the analytic and simulated results - a ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance.
-Figure \ref{fig:distcomparison}b shows how these thresholds change as a function of the number of observations, and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
+Figure \ref{fig:distcomparison}a shows an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100,000 random data sets with \emph{N} = 10 (blue shading).
+The vertical lines show the critical (95\%) threshold for the analytic and simulated results.
+A ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance, because it has a condition index larger than expected by chance (assuming \(\alpha\) = 0.05).
+Ratios below the threshold imply that the eigenvalues are equal (in a statistical sense).
+Figure \ref{fig:distcomparison}b shows how these critical thresholds change as a function of the number of observations (N), and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
 
@@
 
-\caption{Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of 10 observations.
-The vertical blue and red lines show the 95 percent thresholds on the curve (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function.
-Panel (b) shows how 95 percent thresholds change as a functio of the number of observations.}\label{fig:distcomparison}
+\caption{Logic of the condition index test.
+Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of N = 10 observations.
+The vertical blue and red lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function.
+Ellipse icons above panel (a) illustrate different condition indices between 1 and 5 (note that rotation does not affect the condition index).
+Panel (b) shows how 95 percent thresholds change as a function of the number of observations.}\label{fig:distcomparison}
 \end{figure}
@@
 The eigenvalue ratio can be used as a test of the assumptions of \(T^2_{circ}\).
-If we observe a condition index above the critical threshold for the number of observations then the data set can be said to significantly violate the assumption of equal eigenvalues.
-Because our modified equation provides an inverse density function (red curve in Figure \ref{fig:distcomparison}a), we can use this to calculate a p-value for the test.
+If a condition index is observed that is above the critical threshold for the number of observations, then the data set can be said to significantly violate the assumption of equal eigenvalues.
+Because the modified equation provides an inverse density function (red curve in Figure \ref{fig:distcomparison}a), this can be used to calculate a \emph{p}-value for the test.
 If the test is non-significant, one can proceed with \(T^2_{circ}\); if it is significant, \(T^2\) should be used instead.
-An \emph{R} function implementing this test is included in the project code repository (the \emph{CI.test} function contained in the \emph{CItest.R} script).
+An \emph{R} function implementing this test is included in the \emph{FourierStats} package (the \emph{CI.test} function).
 
@@
 If a data set produces a significant result using the condition index test, this could be due to the presence of one or more outliers.
-The Mahalanobis distance (Mahalanobis 1936) is a useful metric for identifying such multivariate outliers so that they can be excluded.
+The Mahalanobis distance (Mahalanobis, 1936) is a useful metric for identifying such multivariate outliers so that they can be excluded.
 It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points.
@@
 Figure \ref{fig:outlierplot} shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves).
-In all cases, the functions depart from the Type 1 error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic return \(D^2\), which can be converted to D by taking the square root).
+In all cases, the functions depart from the Type I error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the Mahalanobis distance exceeds a value around 3.
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R}, return \(D^2\), which can be converted to D by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the \(T^2_{circ}\) test.
@@
 A variant of the Mahalanobis distance (the pairwise Mahalanobis distance) can also be used to compute a multivariate measure of effect size, equivalent to Cohen's \emph{d} statistic (see e.g.
-Del Giudice 2009).
-This is a valuable statistic to include when reporting the results of multivariate tests, and we include an \emph{R} function (\emph{mahalES}) that can be used to calculate it.
+Del Giudice, 2009).
+This is a valuable statistic to include when reporting the results of multivariate tests, and an \emph{R} function (\emph{pairwise.mahalanobis}) to calculate it is available as part of the \emph{HDMD} package \url{https://CRAN.R-project.org/package=HDMD}.
 
@@
 
-In many studies, it is important to compare responses over space and/or time.
-However with large numbers of sensors, voxels or temporal epochs, the familywise error rate quickly becomes problematic, inflating the Type I error (false positive) rate.
-Solutions such as Bonferroni correction, which adjust the alpha level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power.
+In some studies, it is important to compare responses over space and/or time.
+However with large numbers of sensors, voxels or temporal epochs, the familywise error quickly becomes problematic, inflating the Type I error (false positive) rate.
+Solutions such as Bonferroni correction, which adjust the \(\alpha\) level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power.
 An alternative approach is to use cluster correction methods to control the Type I error rate, using mass univariate tests (e.g.
-Maris and Oostenveld 2007).
+Maris and Oostenveld, 2007).
 These typically involve summing test statistics such as T-values across adjacent significant locations and/or moments in time.
@@
 Such methods control the Type I error rate without substantially reducing statistical power.
-The same approach can be straightforwardly applied to the \(T^2\) and \(T^2_{circ}\) statistics.
-This allows a principled approach to identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli.
+The same approach can be applied to the \(T^2\) and \(T^2_{circ}\) statistics.
+This allows a principled method for identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli.
 The \emph{FourierStats} package includes an implementation of this method with options for multivariate statistics (the \emph{clustercorrect} function).
@@
 Many studies involve more than two conditions that need to be compared.
-Issues with familywise error will quickly become problematic if multiple pairwise \(T^2\) or \(T^2_{circ}\) statistics are calculated.
+Again, issues with familywise error will quickly become problematic if multiple pairwise \(T^2\) or \(T^2_{circ}\) statistics are calculated.
 One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's \(T^2\), but permits multiple independent variables with more than two levels.
-However, if the assumptions of \(T^2_{circ}\) hold for a data set, it might alternatively be possible to extend to logic of the \(T^2_{circ}\) test (Victor and Mast 1991) to the more general case.
+However, if the assumptions of \(T^2_{circ}\) hold for a data set, it might alternatively be possible to extend the logic of the \(T^2_{circ}\) test (Victor and Mast, 1991) to the more general case.
 
-The F statistic for a one-way ANOVA is calculated by taking the ratio of the variance explained by the modelled group means to the residual unexplained variance.
+The F-statistic for a one-way ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance.
 For multivariate data, the change in group mean would be calculated using the vector distances between the complex means, and the residuals are the vector distances between each data point and its corresponding group mean.
-For an independent one-way design with \(k\) groups (or conditions) and \(N\) observations per group, the F distribution will have \(2(k-1)\) and \(2((Nk)-k)\) degrees of freedom.
+For an independent one-way design with \(k\) groups (or conditions) and \(N\) observations per group, the F-distribution will have \(2(k-1)\) and \(2((Nk)-k)\) degrees of freedom.
 The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number.
+A \emph{p}-value can be derived from the F-distribution in the usual way.
+
 A suitable name for such a test might be \({ANOVA}^2_{circ}\), as this reflects the similarity to ANOVA, and the extension of the logic of \(T^2_{circ}\) (an alternative name might be \(MANOVA_{circ}\), however this feels less appropriate given that many of the key features of MANOVA are absent).
-
-Figure \ref{fig:powerfig2} shows simulations analogous to those in Figure \ref{fig:powerfig} for a one-way design with three levels.
+Figure \ref{fig:powerfig2} shows simulations analogous to those in Figure \ref{fig:powerfig} for a one-way between-subjects design with three levels.
 MANOVA is directly compared to the \({ANOVA}^2_{circ}\) statistic across a range of effect sizes and sample sizes.
@@
 The format mirrors that of Figure 2.
-In these simulations, there were three conditions, with the signal being added to one condition.}\label{fig:powerfig2}
+In these simulations, there were three conditions, with the signal being added to one condition only.}\label{fig:powerfig2}
 \end{figure}
 
-Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that standard multiple comparison corrections are applied (e.g.~Bonferroni correction).
+Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction).
 Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary).
 A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic.
-The project repository contains \emph{R} scripts for both of these functions.
+The project repository contains \emph{R} functions for both of these tests (\emph{anovacirc.test} and \emph{anovacircR.test}).
 
@@
 The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
-Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios.
+Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test.
 Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if not.
@@
 Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs.
-In \emph{R}, the \emph{Manova} function from the \emph{car} package appears to be capable of the necessary calculations, as does the commercial software package SPSS.
+In \emph{R}, the \emph{Manova} function from the \emph{car} package appears to be capable of the necessary calculations, as does the commercial statistics package \emph{SPSS}.
 
@@
 
-\hypertarget{applying-these-methods-to-empirical-data-sets}{%
-\section{Applying these methods to empirical data sets}\label{applying-these-methods-to-empirical-data-sets}}
+\hypertarget{applying-multivariate-methods-to-empirical-data-sets}{%
+\section{Applying multivariate methods to empirical data sets}\label{applying-multivariate-methods-to-empirical-data-sets}}
 
-Having developed some novel tools for the analysis of periodic data, in the following sections we demonstrate them using two different publicly available empirical data sets.
+Having developed some novel tools for the analysis of periodic data, in the following sections their use is demonstrated for two different publicly available empirical data sets.
 The first study recorded responses to auditory and optogenetic stimulation in mice.
 The second study measured visual responses to flickering grating patterns in humans.
-These cases provide examples of how the results of the tests described above might be appropriately reported.
+These examples also provide an illustration of how the results of the tests described above might be appropriately reported.
 
@@
 (2020), and was downloaded from: \url{https://doi.gin.g-node.org/10.12751/g-node.e5tyek/}.
+A processed data file is included with permission from the authors in the \emph{FourierStats} package.
 
@@
 Panel (a) shows the Fourier amplitude spectrum with inset scalp plots for sound (red) and light (blue) stimulation, averaged across repetitions and individuals.
+Grey and black points in the insets indicate electrode locations.
 Panel (b) shows complex (x = real, y = imaginary) Fourier components for 6 individual mice (small points) and their average (large points) for both conditions.}\label{fig:mousedata}
@@
 Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \ref{fig:humanSSVEP}.
-For the main analysis, we took responses from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant.
+For the main analysis, responses were taken from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant.
 Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}.
@@
 (2018) compared SSVEP responses between individuals with and without autism, as well as in a \emph{Drosophila} genetic model of developmental disorders.
-The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}.
+The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}, and a processed version is included with the \emph{FourierStats} package.
 
@@
 This is the case for most paradigms in which the nervous system is driven by a periodic stimulus.
-However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger 1929; Buzsáki and Draguhn 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g.
-Canolty and Knight 2010).
-When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al.
-2021).
+However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger, 1929; Buzsáki and Draguhn, 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g.
+Canolty and Knight, 2010).
+When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al., 2021).
 This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.
@@
 The present paper has focussed on the Frequentist statistical tradition.
-However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses (Jeffreys 1961).
-Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al.
-2009) and ANOVA (Rouder et al.
-2017).
+However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses (Jeffreys, 1961).
+Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al., 2009) and ANOVA (Rouder et al., 2017).
 However this is a non-trivial undertaking, and is beyond the scope of the current paper.
@@
 If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance.
-Such methods have been hugely influential in the fMRI literature (Schwarzkopf and Rees 2011), and for analysing event-related potential data collected using EEG or MEG (Grootswagers, Wardle, and Carlson 2017).
+Such methods have been hugely influential in the fMRI literature (Schwarzkopf and Rees, 2011), and for analysing event-related potential data collected using EEG or MEG (Grootswagers et al., 2017).
 However, they have not been widely applied to steady-state data (though see West et al.
-(2015) for an example).
+(2015) for one example).
 In principle, the real and imaginary Fourier components can be treated as separate dependent variables, along with different recording locations and/or frequencies.
@@
 (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat demanding, though we have implemented an \emph{R} function to calculate error bars using this method (available through the \emph{getampCI} function).
-However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95\% confidence intervals on amplitudes.
+This approach is somewhat demanding, though an \emph{R} function is provided to calculate error bars using this method (available through the \emph{amperrors} function).
+However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes.
 This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean.
-The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled amplitudes.
-Upper and lower 95\% confidence intervals on the amplitude can then be taken at appropriate quantiles of this population.
+The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled mean amplitudes.
+Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68\% and 95\% are typical, with the 68\% interval corresponding to the standard error).
 
@@
 
-The simulations reported here allow us to make several recommendations for how periodic data should be analysed.
+The simulations reported here allow several recommendations to be made for how periodic data should be analysed.
 Multivariate statistics should be used for phase-locked Fourier data instead of univariate statistics such as T-tests and ANOVA.
-This avoids problems from distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information.
-Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahanalobis distance reported as a measure of effect size.
+This avoids problems from non-normal distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information.
+Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahalanobis distance reported as a measure of effect size.
 For sample sizes of N \textless{} 32, the \(T^2_{circ}\) and \(ANOVA^2_{circ}\) statistics can be used if the condition index test is non-significant for all conditions.
@@
 \leavevmode\hypertarget{ref-Afsari2014}{}%
-Afsari, Farinaz, Kenneth V Christensen, Garrick Paul Smith, Morten Hentzer, Olivia M Nippe, Christopher J H Elliott, and Alex R Wade.
+Afsari F, Christensen KV, Smith GP, Hentzer M, Nippe OM, Elliott CJH, Wade AR.
 2014.
-``Abnormal Visual Gain Control in a Parkinson's Disease Model.'' \emph{Hum Mol Genet} 23 (17): 4465--78.
-\url{https://doi.org/10.1093/hmg/ddu159}.
+Abnormal visual gain control in a parkinson's disease model.
+\emph{Hum Mol Genet} \textbf{23}:4465--78.
+doi:\href{https://doi.org/10.1093/hmg/ddu159}{10.1093/hmg/ddu159}
 
 \leavevmode\hypertarget{ref-Baker2021}{}%
-Baker, D.H., G.
-Vilidaite, F.A.
-Lygo, A.K.
-Smith, T.R.
-Flack, A.D.
-Gouws, and T.J.
-Andrews.
+Baker DH, Vilidaite G, Lygo FA, Smith AK, Flack TR, Gouws AD, Andrews TJ.
 2021.
-``Power Contours: Optimising Sample Size and Precision in Experimental Psychology and Human Neuroscience.'' \emph{Psychological Methods}.
-\url{https://doi.org/10.1037/met0000337}.
+Power contours: Optimising sample size and precision in experimental psychology and human neuroscience.
+\emph{Psychological Methods}.
+doi:\href{https://doi.org/10.1037/met0000337}{10.1037/met0000337}
 
 \leavevmode\hypertarget{ref-Berger1929}{}%
-Berger, H. 1929. ``Über Das Elektrenkephalogramm Des Menschen.'' \emph{Archiv F. Psychiatrie} 87: 527--70. \url{https://doi.org/10.1007/BF01797193}.
+Berger H. 1929. Über das elektrenkephalogramm des menschen. \emph{Archiv f Psychiatrie} \textbf{87}:527--570. doi:\href{https://doi.org/10.1007/BF01797193}{10.1007/BF01797193}
 
 \leavevmode\hypertarget{ref-Braddick1986}{}%
-Braddick, O J, J Wattam-Bell, and J Atkinson. 1986. ``Orientation-Specific Cortical Responses Develop in Early Infancy.'' \emph{Nature} 320 (6063): 617--9. \url{https://doi.org/10.1038/320617a0}.
+Braddick OJ, Wattam-Bell J, Atkinson J.
+1986.
+Orientation-specific cortical responses develop in early infancy.
+\emph{Nature} \textbf{320}:617--9.
+doi:\href{https://doi.org/10.1038/320617a0}{10.1038/320617a0}
 
 \leavevmode\hypertarget{ref-Buzsaki2004}{}%
-Buzsáki, György, and Andreas Draguhn. 2004. ``Neuronal Oscillations in Cortical Networks.'' \emph{Science} 304 (5679): 1926--9. \url{https://doi.org/10.1126/science.1099745}.
+Buzsáki G, Draguhn A. 2004. Neuronal oscillations in cortical networks. \emph{Science} \textbf{304}:1926--9. doi:\href{https://doi.org/10.1126/science.1099745}{10.1126/science.1099745}
 
 \leavevmode\hypertarget{ref-Canolty2010}{}%
-Canolty, Ryan T, and Robert T Knight. 2010. ``The Functional Role of Cross-Frequency Coupling.'' \emph{Trends Cogn Sci} 14 (11): 506--15. \url{https://doi.org/10.1016/j.tics.2010.09.001}.
+Canolty RT, Knight RT.
+2010.
+The functional role of cross-frequency coupling.
+\emph{Trends Cogn Sci} \textbf{14}:506--15.
+doi:\href{https://doi.org/10.1016/j.tics.2010.09.001}{10.1016/j.tics.2010.09.001}
 
 \leavevmode\hypertarget{ref-Cohen1988}{}%
-Cohen, J. 1988. \emph{Statistical Power Analysis for the Behavioral Sciences}. Lawrence Erlbaum, Hillsdale, New Jersey.
+Cohen J. 1988. Statistical power analysis for the behavioral sciences. Lawrence Erlbaum, Hillsdale, New Jersey.
 
 \leavevmode\hypertarget{ref-Giudice2009}{}%
-Del Giudice, M. 2009. ``On the Real Magnitude of Psychological Sex Differences.'' \emph{Evolutionary Psychology} 7: 264--79. \url{https://doi.org/10.1177/147470490900700209}.
+Del Giudice M.
+2009.
+On the real magnitude of psychological sex differences.
+\emph{Evolutionary Psychology} \textbf{7}:264--279.
+doi:\href{https://doi.org/10.1177/147470490900700209}{10.1177/147470490900700209}
 
 \leavevmode\hypertarget{ref-Edelman1988}{}%
-Edelman, Alan. 1988. ``Eigenvalues and Condition Numbers of Random Matrices.'' \emph{SIAM Journal on Matrix Analysis and Applications} 9 (4): 543--60. \url{https://doi.org/10.1137/0609045}.
+Edelman A.
+1988.
+Eigenvalues and condition numbers of random matrices.
+\emph{SIAM Journal on Matrix Analysis and Applications} \textbf{9}:543--560.
+doi:\href{https://doi.org/10.1137/0609045}{10.1137/0609045}
 
 \leavevmode\hypertarget{ref-Engel1994}{}%
-Engel, S A, D E Rumelhart, B A Wandell, A T Lee, G H Glover, E J Chichilnisky, and M N Shadlen.
+Engel SA, Rumelhart DE, Wandell BA, Lee AT, Glover GH, Chichilnisky EJ, Shadlen MN.
 1994.
-``FMRI of Human Visual Cortex.'' \emph{Nature} 369 (6481): 525.
-\url{https://doi.org/10.1038/369525a0}.
+FMRI of human visual cortex.
+\emph{Nature} \textbf{369}:525.
+doi:\href{https://doi.org/10.1038/369525a0}{10.1038/369525a0}
 
 \leavevmode\hypertarget{ref-Enroth-Cugell1966}{}%
-Enroth-Cugell, C, and J G Robson. 1966. ``The Contrast Sensitivity of Retinal Ganglion Cells of the Cat.'' \emph{J Physiol} 187 (3): 517--52. \url{https://doi.org/10.1113/jphysiol.1966.sp008107}.
+Enroth-Cugell C, Robson JG.
+1966.
+The contrast sensitivity of retinal ganglion cells of the cat.
+\emph{J Physiol} \textbf{187}:517--52.
+doi:\href{https://doi.org/10.1113/jphysiol.1966.sp008107}{10.1113/jphysiol.1966.sp008107}
 
 \leavevmode\hypertarget{ref-Gray2020}{}%
-Gray, Katie L H, Tessa R Flack, Miaomiao Yu, Freya A Lygo, and Daniel H Baker.
+Gray KLH, Flack TR, Yu M, Lygo FA, Baker DH.
 2020.
-``Nonlinear Transduction of Emotional Facial Expression.'' \emph{Vision Res} 170: 1--11.
-\url{https://doi.org/10.1016/j.visres.2020.03.004}.
+Nonlinear transduction of emotional facial expression.
+\emph{Vision Res} \textbf{170}:1--11.
+doi:\href{https://doi.org/10.1016/j.visres.2020.03.004}{10.1016/j.visres.2020.03.004}
 
 \leavevmode\hypertarget{ref-Grootswagers2017}{}%
-Grootswagers, Tijl, Susan G Wardle, and Thomas A Carlson.
+Grootswagers T, Wardle SG, Carlson TA.
 2017.
-``Decoding Dynamic Brain Patterns from Evoked Responses: A Tutorial on Multivariate Pattern Analysis Applied to Time Series Neuroimaging Data.'' \emph{J Cogn Neurosci} 29 (4): 677--97.
-\url{https://doi.org/10.1162/jocn_a_01068}.
+Decoding dynamic brain patterns from evoked responses: A tutorial on multivariate pattern analysis applied to time series neuroimaging data.
+\emph{J Cogn Neurosci} \textbf{29}:677--697.
+doi:\href{https://doi.org/10.1162/jocn_a_01068}{10.1162/jocn\_a\_01068}
 
 \leavevmode\hypertarget{ref-Hotelling1931}{}%
-Hotelling, H.
-1931.
-``The Generalization of Student's Ratio.'' \emph{The Annals of Mathematical Statistics} 2 (3).
-Institute of Mathematical Statistics: 360--78.
-\url{http://www.jstor.org/stable/2957535}.
+Hotelling H. 1931. The generalization of student's ratio. \emph{The Annals of Mathematical Statistics} \textbf{2}:360--378.
 
 \leavevmode\hypertarget{ref-Hwang2019}{}%
-Hwang, Eunjin, Ritchie E Brown, Bernat Kocsis, Tae Kim, James T McKenna, James M McNally, Hio-Been Han, and Jee Hyun Choi.
+Hwang E, Brown RE, Kocsis B, Kim T, McKenna JT, McNally JM, Han H-B, Choi JH.
 2019.
-``Optogenetic Stimulation of Basal Forebrain Parvalbumin Neurons Modulates the Cortical Topography of Auditory Steady-State Responses.'' \emph{Brain Struct Funct} 224 (4): 1505--18.
-\url{https://doi.org/10.1007/s00429-019-01845-5}.
+Optogenetic stimulation of basal forebrain parvalbumin neurons modulates the cortical topography of auditory steady-state responses.
+\emph{Brain Struct Funct} \textbf{224}:1505--1518.
+doi:\href{https://doi.org/10.1007/s00429-019-01845-5}{10.1007/s00429-019-01845-5}
 
 \leavevmode\hypertarget{ref-Hwang2020}{}%
-Hwang, Eunjin, Hio-Been Han, Jung Young Kim, and Jee Hyun Choi.
+Hwang E, Han H-B, Kim JY, Choi JH.
 2020.
-``High-Density Eeg of Auditory Steady-State Responses During Stimulation of Basal Forebrain Parvalbumin Neurons.'' \emph{Sci Data} 7 (1): 288.
-\url{https://doi.org/10.1038/s41597-020-00621-z}.
+High-density eeg of auditory steady-state responses during stimulation of basal forebrain parvalbumin neurons.
+\emph{Sci Data} \textbf{7}:288.
+doi:\href{https://doi.org/10.1038/s41597-020-00621-z}{10.1038/s41597-020-00621-z}
 
 \leavevmode\hypertarget{ref-Jeffreys1961}{}%
-Jeffreys, H. 1961. \emph{Theory of Probability}. 3rd ed. Oxford University Press.
+Jeffreys H. 1961. Theory of probability, 3rd ed. Oxford University Press.
 
 \leavevmode\hypertarget{ref-Liu-Shuang2014}{}%
-Liu-Shuang, Joan, Anthony M Norcia, and Bruno Rossion.
+Liu-Shuang J, Norcia AM, Rossion B.
 2014.
-``An Objective Index of Individual Face Discrimination in the Right Occipito-Temporal Cortex by Means of Fast Periodic Oddball Stimulation.'' \emph{Neuropsychologia} 52: 57--72.
-\url{https://doi.org/10.1016/j.neuropsychologia.2013.10.022}.
+An objective index of individual face discrimination in the right occipito-temporal cortex by means of fast periodic oddball stimulation.
+\emph{Neuropsychologia} \textbf{52}:57--72.
+doi:\href{https://doi.org/10.1016/j.neuropsychologia.2013.10.022}{10.1016/j.neuropsychologia.2013.10.022}
 
 \leavevmode\hypertarget{ref-Mahalanobis1936}{}%
-Mahalanobis, P.C. 1936. ``On the Generalised Distance in Statistics.'' \emph{Proceedings of the National Academy of Sciences of India} 2: 49--55.
+Mahalanobis P. 1936. On the generalised distance in statistics. \emph{Proceedings of the National Academy of Sciences of India} \textbf{2}:49--55.
 
 \leavevmode\hypertarget{ref-Maris2007}{}%
-Maris, Eric, and Robert Oostenveld. 2007. ``Nonparametric Statistical Testing of EEG- and MEG-Data.'' \emph{J Neurosci Methods} 164 (1): 177--90. \url{https://doi.org/10.1016/j.jneumeth.2007.03.024}.
+Maris E, Oostenveld R.
+2007.
+Nonparametric statistical testing of EEG- and MEG-data.
+\emph{J Neurosci Methods} \textbf{164}:177--90.
+doi:\href{https://doi.org/10.1016/j.jneumeth.2007.03.024}{10.1016/j.jneumeth.2007.03.024}
 
 \leavevmode\hypertarget{ref-Morrone1987}{}%
-Morrone, M C, D C Burr, and H D Speed. 1987. ``Cross-Orientation Inhibition in Cat Is Gaba Mediated.'' \emph{Exp Brain Res} 67 (3): 635--44. \url{https://doi.org/10.1007/BF00247294}.
+Morrone MC, Burr DC, Speed HD. 1987. Cross-orientation inhibition in cat is gaba mediated. \emph{Exp Brain Res} \textbf{67}:635--44. doi:\href{https://doi.org/10.1007/BF00247294}{10.1007/BF00247294}
 
 \leavevmode\hypertarget{ref-Nakayama1982}{}%
-Nakayama, K, and M Mackeben. 1982. ``Steady State Visual Evoked Potentials in the Alert Primate.'' \emph{Vision Res} 22 (10): 1261--71. \url{https://doi.org/10.1016/0042-6989(82)90138-9}.
+Nakayama K, Mackeben M.
+1982.
+Steady state visual evoked potentials in the alert primate.
+\emph{Vision Res} \textbf{22}:1261--71.
+doi:\href{https://doi.org/10.1016/0042-6989(82)90138-9}{10.1016/0042-6989(82)90138-9}
 
 \leavevmode\hypertarget{ref-Norcia2015}{}%
-Norcia, Anthony M, L Gregory Appelbaum, Justin M Ales, Benoit R Cottereau, and Bruno Rossion.
+Norcia AM, Appelbaum LG, Ales JM, Cottereau BR, Rossion B.
 2015.
-``The Steady-State Visual Evoked Potential in Vision Research: A Review.'' \emph{J Vis} 15 (6): 4.
-\url{https://doi.org/10.1167/15.6.4}.
+The steady-state visual evoked potential in vision research: A review.
+\emph{J Vis} \textbf{15}:4.
+doi:\href{https://doi.org/10.1167/15.6.4}{10.1167/15.6.4}
 
 \leavevmode\hypertarget{ref-Pei2017}{}%
-Pei, Francesca, Stefano Baldassi, Jeffrey J Tsai, Holly E Gerhard, and Anthony M Norcia.
+Pei F, Baldassi S, Tsai JJ, Gerhard HE, Norcia AM.
 2017.
-``Development of Contrast Normalization Mechanisms During Childhood and Adolescence.'' \emph{Vision Res} 133: 12--20.
-\url{https://doi.org/10.1016/j.visres.2016.03.010}.
+Development of contrast normalization mechanisms during childhood and adolescence.
+\emph{Vision Res} \textbf{133}:12--20.
+doi:\href{https://doi.org/10.1016/j.visres.2016.03.010}{10.1016/j.visres.2016.03.010}
 
 \leavevmode\hypertarget{ref-Porciatti1990}{}%
-Porciatti, V, G Fontanesi, A Raffaelli, and P Bagnoli.
+Porciatti V, Fontanesi G, Raffaelli A, Bagnoli P.
 1990.
-``Binocularity in the Little Owl, Athene Noctua.
+Binocularity in the little owl, athene noctua.
 II.
-Properties of Visually Evoked Potentials from the Wulst in Response to Monocular and Binocular Stimulation with Sine Wave Gratings.'' \emph{Brain Behav Evol} 35 (1): 40--48.
-\url{https://doi.org/10.1159/000115855}.
+Properties of visually evoked potentials from the wulst in response to monocular and binocular stimulation with sine wave gratings.
+\emph{Brain Behav Evol} \textbf{35}:40--8.
+doi:\href{https://doi.org/10.1159/000115855}{10.1159/000115855}
 
 \leavevmode\hypertarget{ref-Rees1986}{}%
-Rees, A, G G Green, and R H Kay.
+Rees A, Green GG, Kay RH.
 1986.
-``Steady-State Evoked Responses to Sinusoidally Amplitude-Modulated Sounds Recorded in Man.'' \emph{Hear Res} 23 (2): 123--33.
-\url{https://doi.org/10.1016/0378-5955(86)90009-2}.
+Steady-state evoked responses to sinusoidally amplitude-modulated sounds recorded in man.
+\emph{Hear Res} \textbf{23}:123--33.
+doi:\href{https://doi.org/10.1016/0378-5955(86)90009-2}{10.1016/0378-5955(86)90009-2}
 
 \leavevmode\hypertarget{ref-Regan1966}{}%
-Regan, D.
+Regan D.
 1966.
-``Some Characteristics of Average Steady-State and Transient Responses Evoked by Modulated Light.'' \emph{Electroencephalogr Clin Neurophysiol} 20 (3): 238--48.
-\url{https://doi.org/10.1016/0013-4694(66)90088-5}.
+Some characteristics of average steady-state and transient responses evoked by modulated light.
+\emph{Electroencephalogr Clin Neurophysiol} \textbf{20}:238--48.
+doi:\href{https://doi.org/10.1016/0013-4694(66)90088-5}{10.1016/0013-4694(66)90088-5}
 
 \leavevmode\hypertarget{ref-Rouder2017}{}%
-Rouder, Jeffrey N, Richard D Morey, Josine Verhagen, April R Swagman, and Eric-Jan Wagenmakers.
+Rouder JN, Morey RD, Verhagen J, Swagman AR, Wagenmakers E-J.
 2017.
-``Bayesian Analysis of Factorial Designs.'' \emph{Psychol Methods} 22 (2): 304--21.
-\url{https://doi.org/10.1037/met0000057}.
+Bayesian analysis of factorial designs.
+\emph{Psychol Methods} \textbf{22}:304--321.
+doi:\href{https://doi.org/10.1037/met0000057}{10.1037/met0000057}
 
 \leavevmode\hypertarget{ref-Rouder2009}{}%
-Rouder, Jeffrey N, Paul L Speckman, Dongchu Sun, Richard D Morey, and Geoffrey Iverson.
+Rouder JN, Speckman PL, Sun D, Morey RD, Iverson G.
 2009.
-``Bayesian T Tests for Accepting and Rejecting the Null Hypothesis.'' \emph{Psychon Bull Rev} 16 (2): 225--37.
-\url{https://doi.org/10.3758/PBR.16.2.225}.
+Bayesian t tests for accepting and rejecting the null hypothesis.
+\emph{Psychon Bull Rev} \textbf{16}:225--37.
+doi:\href{https://doi.org/10.3758/PBR.16.2.225}{10.3758/PBR.16.2.225}
 
 \leavevmode\hypertarget{ref-Schwarzkopf2011}{}%
-Schwarzkopf, D S, and G Rees. 2011. ``Pattern Classification Using Functional Magnetic Resonance Imaging.'' \emph{Wiley Interdiscip Rev Cogn Sci} 2 (5): 568--79. \url{https://doi.org/10.1002/wcs.141}.
+Schwarzkopf DS, Rees G.
+2011.
+Pattern classification using functional magnetic resonance imaging.
+\emph{Wiley Interdiscip Rev Cogn Sci} \textbf{2}:568--579.
+doi:\href{https://doi.org/10.1002/wcs.141}{10.1002/wcs.141}
 
 \leavevmode\hypertarget{ref-Sereno1995}{}%
-Sereno, M I, A M Dale, J B Reppas, K K Kwong, J W Belliveau, T J Brady, B R Rosen, and R B Tootell.
+Sereno MI, Dale AM, Reppas JB, Kwong KK, Belliveau JW, Brady TJ, Rosen BR, Tootell RB.
 1995.
-``Borders of Multiple Visual Areas in Humans Revealed by Functional Magnetic Resonance Imaging.'' \emph{Science} 268 (5212): 889--93.
-\url{https://doi.org/10.1126/science.7754376}.
+Borders of multiple visual areas in humans revealed by functional magnetic resonance imaging.
+\emph{Science} \textbf{268}:889--93.
+doi:\href{https://doi.org/10.1126/science.7754376}{10.1126/science.7754376}
 
 \leavevmode\hypertarget{ref-Snyder1992}{}%
-Snyder, A Z.
+Snyder AZ.
 1992.
-``Steady-State Vibration Evoked Potentials: Descriptions of Technique and Characterization of Responses.'' \emph{Electroencephalogr Clin Neurophysiol} 84 (3): 257--68.
-\url{https://doi.org/10.1016/0168-5597(92)90007-x}.
+Steady-state vibration evoked potentials: Descriptions of technique and characterization of responses.
+\emph{Electroencephalogr Clin Neurophysiol} \textbf{84}:257--68.
+doi:\href{https://doi.org/10.1016/0168-5597(92)90007-x}{10.1016/0168-5597(92)90007-x}
 
 \leavevmode\hypertarget{ref-Spitschan2014}{}%
-Spitschan, Manuel, Sandeep Jain, David H Brainard, and Geoffrey K Aguirre.
+Spitschan M, Jain S, Brainard DH, Aguirre GK.
 2014.
-``Opponent Melanopsin and S-Cone Signals in the Human Pupillary Light Response.'' \emph{Proc Natl Acad Sci U S A} 111 (43): 15568--72.
-\url{https://doi.org/10.1073/pnas.1400942111}.
+Opponent melanopsin and s-cone signals in the human pupillary light response.
+\emph{Proc Natl Acad Sci U S A} \textbf{111}:15568--72.
+doi:\href{https://doi.org/10.1073/pnas.1400942111}{10.1073/pnas.1400942111}
 
 \leavevmode\hypertarget{ref-Victor1991}{}%
-Victor, J D, and J Mast. 1991. ``A New Statistic for Steady-State Evoked Potentials.'' \emph{Electroencephalogr Clin Neurophysiol} 78 (5): 378--88. \url{https://doi.org/10.1016/0013-4694(91)90099-p}.
+Victor JD, Mast J.
+1991.
+A new statistic for steady-state evoked potentials.
+\emph{Electroencephalogr Clin Neurophysiol} \textbf{78}:378--88.
+doi:\href{https://doi.org/10.1016/0013-4694(91)90099-p}{10.1016/0013-4694(91)90099-p}
 
 \leavevmode\hypertarget{ref-Vilidaite2018}{}%
-Vilidaite, Greta, Anthony M Norcia, Ryan J H West, Christopher J H Elliott, Francesca Pei, Alex R Wade, and Daniel H Baker.
+Vilidaite G, Norcia AM, West RJH, Elliott CJH, Pei F, Wade AR, Baker DH.
 2018.
-``Autism Sensory Dysfunction in an Evolutionarily Conserved System.'' \emph{Proc Biol Sci} 285 (1893): 20182255.
-\url{https://doi.org/10.1098/rspb.2018.2255}.
+Autism sensory dysfunction in an evolutionarily conserved system.
+\emph{Proc Biol Sci} \textbf{285}:20182255.
+doi:\href{https://doi.org/10.1098/rspb.2018.2255}{10.1098/rspb.2018.2255}
 
 \leavevmode\hypertarget{ref-West2015}{}%
-West, Ryan J H, Christopher J H Elliott, and Alex R Wade.
+West RJH, Elliott CJH, Wade AR.
 2015.
-``Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.'' \emph{Sci Rep} 5: 16933.
-\url{https://doi.org/10.1038/srep16933}.
+Classification of parkinson's disease genotypes in drosophila using spatiotemporal profiling of vision.
+\emph{Sci Rep} \textbf{5}:16933.
+doi:\href{https://doi.org/10.1038/srep16933}{10.1038/srep16933}
 

### 2020-12-31 R/CI.test.R

@@ -1,10 +1,13 @@
 #' CI.test: function to calculate the condition index of bivariate data
-#' the condition index is the ratio of eigenvalues (eigenvector lengths), calculated longest/shortest
+#' Inputs:
+#'   data is an N x 2 matrix of bivariate (x,y) observations, or a vector of N complex values
+#'   alpha is the criterion for statistical significance, set to a default of 0.05
+#' the condition index is square root of the the ratio of eigenvalues (eigenvector lengths), calculated longest/shortest
 #' the index is then compared for significance with an expected distribution function
-#' significant tests (p<0.05) violate the assumptions of the T-squared-circ and ANOVA-squared-circ tests
+#' significant tests (p < alpha) violate the assumptions of the T-squared-circ and ANOVA-squared-circ tests
 #' see Baker (2021) for further details
 #' @export
-CI.test <- function(data){
+CI.test <- function(data, alpha=0.05){
 
   if (is.complex(data)){data <- data.frame(Re(data),Im(data))}
 
@@ -20,7 +23,7 @@ CI.test <- function(data){
   pdffunction <- ((N-2)*(2^(N-2))) *
     ((cilist^2 - 1)/((cilist^2+1)^(N-1))) * (cilist^(N-3))
   cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
-  criticalCI <- cilist[min(which(cdfinverse<=0.05))]  # find the threshold CI
+  criticalCI <- cilist[min(which(cdfinverse<=alpha))]  # find the threshold CI
 
   indices <- which(cilist>=CI)  # find the CI values larger than our CI
   pval <- cdfinverse[indices[1]] # estimate the p-value

### 2020-12-31 R/analysecplx.R

@@ -1,108 +1,88 @@
-#' analysecplx: wrapper function that decides which statistical test to run for a given data set
-#' the expected input is a data frame in long format:
-#' one column should contain complex valued data, or two columns headed x and y contain the real and imaginary parts
-#' if a column is headed 'group' (or any word beginning with a g) is included, this will be interpreted as level data for the independent variable
-#' if a column is headed 'participant' (or any word beginning with a p) is included, this will be interpreted as a random factor
+#' analysecplx: wrapper function that automatically decides which statistical test to run for a given data set
+#'
+#' Inputs--
+#' data:  this can either be a vector of complex numbers, or a matrix
+#'        if it is a matrix, either the first column contains complex values, or the first two columns are the x and y (real and imaginary) values of the DV
+#'        if there are further columns, these are treated as the group labels and participant IDs
+#' group: condition labels indicating the level of the independent variable that each data point corresponds to
+#'        this is an optional input, but if it is supplied it supercedes values from the matrix
+#' participant: variable storing participant/subject IDs (i.e. the random factor) for repeated measures analysis
+#'        this is an optional input, but if it is supplied it supercedes values from the matrix
 #'
 #' the function first runs the condition index test for each level of the IV
 #' if the condition index test is non-significant for all levels, a T-squared-circ test is run for one- and two-sample designs, or an ANOVA-squared-circ test otherwise
-#' if the condition index test is significant for any level, a T-squared or MANOVA test is run instead
+#' if the condition index test is significant for any level, a T-squared test or MANOVA is run instead
+#' if participant labels are included, a repeated measures test is conducted
[…]

### 2020-12-31 R/anovacirc.test.R

@@ -1,19 +1,43 @@
 #' anovacirc.test: two-dimensional analysis of variance using complex data
-#' an extension of the logic of the t-squared circ statistic
-#' this is the one-way, between subjects implementation of the test
-#' the expected input is a long-format data frame with the following columns:
-#' simdata: the data stored as complex values
-#' grouplabels: condition labels indicating the level of the independent variable each data point corresponds to
+#' an extension of the logic of the T-squared circ statistic of Victor & Mast (1991)
+#' this is a one-way implementation of the test - if participant IDs are supplied a repeated measures version is conducted
+#'
+#' Inputs--
+#' data:  this can either be a vector of complex numbers, or a matrix
+#'        if it is a matrix, either the first column contains complex values, or the first two columns are the x and y (real and imaginary) values of the DV
+#'        if there are further columns, these are treated as the group labels and participant IDs
+#' group: condition labels indicating the level of the independent variable that each data point corresponds to
+#'        this is an optional input, but if it is supplied it supercedes values from the matrix
+#' participant: variable storing participant/subject IDs (i.e. the random factor) for repeated measures analysis
+#'        this is an optional input, but if it is supplied it supercedes values from the matrix
+#'
 #' see Baker (2021) for further details
 #' @export
-anovacirc.test <- function(data){
+anovacirc.test <- function(data, group=NULL, participant=NULL){
 
-  grouplabels <- data$grouplabels
-  datavals <- data$simdata
+  if (!is.matrix(data)){datavals <- data}
+
+  if (is.matrix(data)){d <- dim(data)
+  if (is.complex(data[,1])){datavals <- data[,1]
+  if (d[2]>1){grouplabels <- data[,2]}
+  if (d[2]>2){participantlabels <- data[,3]}}
+
+  if (!is.complex(data[,1])){datavals <- complex(real=data[,1],imaginary=data[,2])
+  if (d[2]>2){grouplabels <- data[,3]}
+  if (d[2]>3){participantlabels <- data[,4]}}
+  }
+
+  if (!is.null(group)){grouplabels <- group}
[…]

### 2020-12-31 R/anovacircR.test.R

@@ -1,51 +1,0 @@
-#' anovacircR.test: two-dimensional analysis of variance using complex data
-#' an extension of the logic of the t-squared circ statistic
-#' this is the one-way, repeated measures implementation of the test
-#' the expected input is a long-format data frame with the following columns:
-#' simdata: the data stored as complex values
-#' grouplabels: condition labels indicating the level of the independent variable each data point corresponds to
-#' participant: participant ID number
-#' see Baker (2021) for further details
-#' @export
-anovacircR.test <- function(data){
-
-  grouplabels <- data$grouplabels
-  factorlist <- levels(grouplabels)
-  participantlabels <- data$participant
-  participantlist <- levels(participantlabels)
-  datavals <- data$simdata
-
-  grandmean <- mean(datavals)
-
-  SSW <- 0
-  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
-  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))
-
-  groupmeans <- NULL
-  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
-  SSM <- 0
-  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
-  dfM <- 2*(nlevels(grouplabels)-1)
-
-  SSR <- SSW - SSM
-  dfR <- dfW - dfM
-
-  MSM <- SSM/dfM
-  MSR <- SSR/dfR
-  Fratio <- MSM/MSR
-  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
-
-  output <- NULL
-  output$Fratio <- Fratio
-  output$pvalue <- pvalue
-  output$SSW <- SSW
-  output$SSM <- SSM
-  output$SSR <- SSR
-  output$dfW <- dfW
-  output$dfM <- dfM
-  output$dfR <- dfR
-  output$MSM <- MSM
-  output$MSR <- MSR
-
-  return(output)
-}

### 2020-12-31 R/fftshift2D.R

@@ -2,4 +2,4 @@
 #' shifts the low spatial frequencies to the centre of the spectrum
 #' this is included in most programming languages, but is absent from the base R Fourier transform tools
 #' @export
-fftshift <- function(im) {im * (-1)^(row(im) + col(im))}
+fftshift2D <- function(im) {im * (-1)^(row(im) + col(im))}

### 2020-12-31 R/getel.R

@@ -1,8 +1,11 @@
 #' getel: helper function that calculates an ellipse bounding some data points
 #' the outline of the ellipse is returned, sampled at 200 points
-#' the input should be an Nx2 matrix of observations
+#' the input should be an Nx2 matrix of observations or an N-length vector of complex values
 #' @export
 getel <- function(compdata){
+
+  if (is.complex(compdata)){compdata <- data.frame(Re(compdata),Im(compdata))}
+
   A <- cov(compdata)
   ctr    <- colMeans(compdata)
   RR     <- chol(A)                               # Cholesky decomposition

### 2021-01-01 manuscript/manuscript.Rmd
Several minor updates and got testanalysecplx working with the MANOVA.RM functions

@@
 @Giudice2009].
-This is a valuable statistic to include when reporting the results of multivariate tests, and an _R_ function (_pairwise.mahalanobis_) to calculate it is available as part of the _HDMD_ package ([https://CRAN.R-project.org/package=HDMD](https://CRAN.R-project.org/package=HDMD)).
+This is a valuable statistic to include when reporting the results of multivariate tests, and an _R_ function (_pairwise.mahalanobis_) to calculate it is available as part of the [_HDMD_](https://CRAN.R-project.org/package=HDMD) package.
 
@@
 Comparisons with more than two conditions should be tested with the $ANOVA^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not.
-Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs.
-In _R_, the _Manova_ function from the _car_ package appears to be capable of the necessary calculations, as does the commercial statistics package _SPSS_.
+Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs.
+However the _multRM_ function in the [_MANOVA.RM_]{https://CRAN.R-project.org/package=MANOVA.RM} package [@Friedrich2019] is able to handle such designs appropriately using a bootstrapping approach.
 
@@
 [-@Vilidaite2018] compared SSVEP responses between individuals with and without autism, as well as in a _Drosophila_ genetic model of developmental disorders.
-The raw data are available at: [http://dx.doi.org/10.17605/OSF.IO/Y4N5K](http://dx.doi.org/10.17605/OSF.IO/Y4N5K), and a processed version is included with the _FourierStats_ package.
+The raw data are available at: [http://dx.doi.org/10.17605/OSF.IO/Y4N5K](http://dx.doi.org/10.17605/OSF.IO/Y4N5K), and a processed version is included with the [_FourierStats_](https://github.com/bakerdh/FourierStats) package.
 

### 2021-01-01 R/analysecplx.R

@@ -13,7 +13,7 @@
 #' if the condition index test is non-significant for all levels, a T-squared-circ test is run for one- and two-sample designs, or an ANOVA-squared-circ test otherwise
 #' if the condition index test is significant for any level, a T-squared test or MANOVA is run instead
 #' if participant labels are included, a repeated measures test is conducted
-#' the paired-samples Mahalanobis effect size statistic is also calculated
+#' the Mahalanobis distance effect size statistic (D) is also calculated
 #' @export
 analysecplx <- function(data, group=NULL, participant=NULL){
 
@@ -22,9 +22,11 @@ complexdata <- NULL
 grouplabels <- NULL
 participantlabels <- NULL
 
-if (!is.matrix(data)){complexdata <- data}
+# if the input data is a vector, it must be complex Fourier components
+if (is.null(ncol(data))){complexdata <- data}
 
-if (is.matrix(data)){d <- dim(data)
+# if the input is not a vector, then work out what it contains based on its dimensions
+if (!is.null(ncol(data))){d <- dim(data)
 
 if (is.complex(data[,1])){complexdata <- data[,1]
 if (d[2]>1){grouplabels <- data[,2]}
@@ -32,11 +34,11 @@ if (d[2]>2){participantlabels <- data[,3]}}
 
 if (!is.complex(data[,1])){complexdata <- complex(real=data[,1],imaginary=data[,2])
 if (d[2]>2){grouplabels <- data[,3]}
-if (d[2]>3){participantlabels <- data[,4]}}
+if (d[2]>3){participantlabels <- as.factor(data[,4])}}
 }
 
 if (!is.null(group)){grouplabels <- group}
-if (!is.null(participant)){participantlabels <- participant}
+if (!is.null(participant)){participantlabels <- as.factor(participant)}
 
 if (is.null(grouplabels)){grouplabels <- rep(1,length(complexdata))}
 isRM <- 1
@@ -45,6 +47,10 @@ if (is.null(participantlabels)){isRM <- 0}
 grouplabels <- as.factor(grouplabels)
 factorlist <- levels(grouplabels)
 ngroups <- nlevels(grouplabels)
+nobservations <- length(complexdata)
+
+message(paste('Design has',ngroups,'levels, with',nobservations/ngroups,'observations per level'))
+if (!is.null(participantlabels)){message('Design is repeated measures')}
 
[…]

### 2021-01-01 R/anovacirc.test.R

@@ -15,9 +15,12 @@
 #' @export
 anovacirc.test <- function(data, group=NULL, participant=NULL){
 
-  if (!is.matrix(data)){datavals <- data}
+  # if the input data is a vector, it must be complex Fourier components
+  if (is.null(ncol(data))){datavals <- data}
 
-  if (is.matrix(data)){d <- dim(data)
+  # if the input is not a vector, then work out what it contains based on its dimensions
+  if (!is.null(ncol(data))){d <- dim(data)
+
   if (is.complex(data[,1])){datavals <- data[,1]
   if (d[2]>1){grouplabels <- data[,2]}
   if (d[2]>2){participantlabels <- data[,3]}}
@@ -54,17 +57,9 @@ anovacirc.test <- function(data, group=NULL, participant=NULL){
   MSM <- SSM/dfM
   MSR <- SSR/dfR
   Fratio <- MSM/MSR
-  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+  p.value <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
 
-  output <- NULL
-  output$Fratio <- Fratio
-  output$p.value <- pvalue
-  output$SSM <- SSM
-  output$SSR <- SSR
-  output$dfM <- dfM
-  output$dfR <- dfR
-  output$MSM <- MSM
-  output$MSR <- MSR
+  output <- data.frame(Fratio,p.value,SSM,SSR,dfM,dfR,MSM,MSR)
 }
 
   # if participant labels have been supplied, run a repeated measures ANOVA-circ
@@ -88,19 +83,9 @@ anovacirc.test <- function(data, group=NULL, participant=NULL){
   MSM <- SSM/dfM
   MSR <- SSR/dfR
   Fratio <- MSM/MSR
-  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+  p.value <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
 
-  output <- NULL
-  output$Fratio <- Fratio
-  output$p.value <- pvalue
-  output$SSW <- SSW
-  output$SSM <- SSM
-  output$SSR <- SSR
-  output$dfW <- dfW
-  output$dfM <- dfM
-  output$dfR <- dfR
-  output$MSM <- MSM
-  output$MSR <- MSR
[…]

### 2021-01-01 R/tsqc.test.R

@@ -7,9 +7,9 @@
 tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
 
   # transpose if dimensions are the wrong way round
-  if (is.matrix(x)){d <- dim(x)
+  if (!is.null(ncol(x))){d <- dim(x)
   if (d[1]<d[2]){x <- t(x)}}
-  if (!is.null(y)){if (is.matrix(y)){d <- dim(y)
+  if (!is.null(y)){if (!is.null(ncol(y))){d <- dim(y)
   if (d[1]<d[2]){y <- t(y)}}}
 
   if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}

### 2021-01-01 manuscript/misc/testanalysecplx.R

@@ -7,11 +7,13 @@ xy <- complex(real=x,imaginary=y)
 group <- rep(1:2,each=N)
 participant <- rep(1:N,times=2)
 
-input <- data.frame(xy,group)
-analysecplx(input)
+data <- data.frame(xy,group)
+group=NULL
+participant=NULL
+results <- analysecplx(data)
 
-input <- data.frame(xy,group,participant)
-analysecplx(input)
+data <- data.frame(xy,group,participant)
+analysecplx(data)
 
 
 x <- rnorm(2*N)
@@ -36,7 +38,7 @@ xy <- complex(real=x,imaginary=y)
 group <- rep(1:3,each=N)
 participant <- rep(1:N,times=3)
 
-input <- data.frame(xy,group)
+data <- data.frame(xy,group)
 analysecplx(input)
 
 input <- data.frame(xy,group,participant)

### 2021-01-01 manuscript/manuscript.Rmd
Minor improvements

@@
 
-nsims <- 10000  # global value for the whole script - increase to 100000 for final run, but note this takes around 15 hours to complete
+nsims <- 100  # global value for the whole script - increase to 100000 for final run, but note this takes around 15 hours to complete
 
@@
 Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs.
-However the _multRM_ function in the [_MANOVA.RM_]{https://CRAN.R-project.org/package=MANOVA.RM} package [@Friedrich2019] is able to handle such designs appropriately using a bootstrapping approach.
+However the _multRM_ function in the [_MANOVA.RM_](https://CRAN.R-project.org/package=MANOVA.RM) package [@Friedrich2019] is able to handle such designs appropriately using a bootstrapping approach.
 

### 2021-01-01 manuscript/manuscript.tex
Minor improvements

@@
 \author{Daniel H. Baker}
-\date{2020-12-31}
+\date{2021-01-01}
 
@@
 Del Giudice, 2009).
-This is a valuable statistic to include when reporting the results of multivariate tests, and an \emph{R} function (\emph{pairwise.mahalanobis}) to calculate it is available as part of the \emph{HDMD} package \url{https://CRAN.R-project.org/package=HDMD}.
+This is a valuable statistic to include when reporting the results of multivariate tests, and an \emph{R} function (\emph{pairwise.mahalanobis}) to calculate it is available as part of the \href{https://CRAN.R-project.org/package=HDMD}{\emph{HDMD}} package.
 
@@
 A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic.
-The project repository contains \emph{R} functions for both of these tests (\emph{anovacirc.test} and \emph{anovacircR.test}).
+The project repository contains an \emph{R} function (\emph{anovacirc.test}) to run both of these tests.
 
@@
 Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not.
-Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs.
-In \emph{R}, the \emph{Manova} function from the \emph{car} package appears to be capable of the necessary calculations, as does the commercial statistics package \emph{SPSS}.
+Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs.
+However the \emph{multRM} function in the {[}\emph{MANOVA.RM}{]}\{\url{https://CRAN.R-project.org/package=MANOVA.RM}\} package (Friedrich et al., 2019) is able to handle such designs appropriately using a bootstrapping approach.
 
@@
 (2018) compared SSVEP responses between individuals with and without autism, as well as in a \emph{Drosophila} genetic model of developmental disorders.
-The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}, and a processed version is included with the \emph{FourierStats} package.
+The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}, and a processed version is included with the \href{https://github.com/bakerdh/FourierStats}{\emph{FourierStats}} package.
 
@@
 doi:\href{https://doi.org/10.1113/jphysiol.1966.sp008107}{10.1113/jphysiol.1966.sp008107}
+
+\leavevmode\hypertarget{ref-Friedrich2019}{}%
+Friedrich S, Konietschke F, Pauly M.
+2019.
+Resampling-Based Analysis of Multivariate Data and Repeated Measures Designs with the R Package MANOVA.RM.
+\emph{The R Journal} \textbf{11}:380--400.
+doi:\href{https://doi.org/10.32614/RJ-2019-051}{10.32614/RJ-2019-051}
 

### 2021-01-01 R/anovacirc.test.R

@@ -15,6 +15,9 @@
 #' @export
 anovacirc.test <- function(data, group=NULL, participant=NULL){
 
+  grouplabels <- NULL
+  participantlabels <- NULL
+
   # if the input data is a vector, it must be complex Fourier components
   if (is.null(ncol(data))){datavals <- data}
 

### 2021-01-02 manuscript/manuscript.Rmd
Added new figure

@@
 
-nsims <- 100  # global value for the whole script - increase to 100000 for final run, but note this takes around 15 hours to complete
+nsims <- 100  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
 
@@
 
+getbanana <- function(p){
+  thetaRange <- seq(0.01, pi, 0.01)
+  Tval <- atan((p[3]*tan(thetaRange))/p[4])
+  x1 <- p[1] + (p[3] * cos(Tval))
+  x2 <- p[1] - (p[3] * cos(Tval))
+  y <- p[2] + (p[4] * sin(Tval))
+
+  XXX <- c(x1[1:(length(x1)/2)], x2[(length(x2)/2):1])
+  Y <- c(y[1:(length(y)/2)], y[(length(y)/2):1], y[length(y):(1+length(y)/2)], y[(1+length(y)/2):length(y)])
+  X <- c(XXX, XXX[length(XXX):1])
+  errormatrix <- pol2cart(matrix(data=c(Y,X),ncol=2))
+  return(errormatrix)
+}
+
 # helper function to add transparency to a colour
@@
 A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal.
-The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response.
-Fourier spectra comprise both amplitude and phase information, which can be equivalently expressed as complex numbers with real and imaginary components.
+The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-defined index of the brain's response (see Figure \@ref(fig:fourierexplain)a,b).
+Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \@ref(fig:fourierexplain)c), or equivalently as complex numbers with real and imaginary components (Figure \@ref(fig:fourierexplain)d).
 In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data only.
@@
 Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods.
+
+```{r fourierexplain, fig.cap="Illustration of the principles of Fourier analysis for time-varying signals.
+Panel (a) shows a sinusoidal signal waveform, and simulated neural responses for successive observations (repetitions in the same individual, or recorded from multiple individuals).
+Panel (b) shows the Fourier amplitude spectra of the waveforms in (a), with clear peaks at the signal frequency (5Hz) in each example.
+The Fourier spectrum also includes a phase term, which can be represented in polar coordinates (amplitude and phase); panel (c) shows this for example responses at the signal frequency (grey points) and their average (blue point), which is computed separately for amplitude and phase terms.
+Alternatively, the same information is contained in a Cartesian representation of the real and imaginary parts of the complex spectrum (panel (d)).
+Notice that the individual observations (grey points) are the same in panels (c,d), but the different representations have implications for how the average (blue points) and measures of spread (shaded regions) are calculated.", fig.align="center", echo=FALSE, fig.width=9, fig.height=8}
+
+set.seed(170382)
+
+par(mfrow=c(2,1), mar=c(3,4,1,2))
+
+plotlims <- c(0,2.4,0,6) 
+ticklocsx <- seq(0,1,0.5)    # locations of tick marks on x axis
+ticklocsy <- seq(0,0.1,0.02)    # locations of tick marks on y axis
+ticklabelsx <- ticklocsx        # set labels for x ticks
+ticklabelsy <- ticklocsy    # set labels for y ticks
+
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=seq(0,1,0.2), tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(1, at=seq(1.4,2.4,0.1), tck=0.01, lab=F, lwd=2) 
+mtext(text = seq(0,1,0.2), side = 1, at=seq(0,1,0.2), line=0.2)
+mtext(text = seq(0,20,2), side = 1, at=seq(1.4,2.4,0.1), line=0.2)
+mtext(text = 'Time (s)', side = 1, cex=1.5, line=1.7, at=0.5) 
+mtext(text = 'Frequency (Hz)', side = 1, cex=1.5, line=1.7, at=1.9)
+
+text(1.2,5.5,'Signal',cex=1.2,adj=0.5)
+text(1.2,4.5,'Observation',cex=1,adj=0.5)
+text(1.2,3.75,'1',cex=1.2,adj=0.5)
+text(1.2,2.75,'2',cex=1.2,adj=0.5)
+text(1.2,1.75,'3',cex=1.2,adj=0.5)
+text(1.2,1.1,'.',cex=1.2,adj=0.5)
+text(1.2,1,'.',cex=1.2,adj=0.5)
+text(1.2,0.9,'.',cex=1.2,adj=0.5)
+text(1.2,0.25,'N',cex=1.2,adj=0.5)
+
+text(0,5.8,'(a)',cex=1.5,adj=0.5)
+text(1.45,5.8,'(b)',cex=1.5,adj=0.5)
+
+tvals <- seq(0,10-0.001,0.001)
+fvals <- c(0:(length(tvals)/2-1),(length(tvals)/2-1):0)/10
+signal <- 0.8*cos(5*tvals * 2*pi + pi)
+
+lines(tvals[1:1000],5+(signal[1:1000]+1)/2,lwd=2)
+
+fsignal <- abs(fft(signal))/length(signal)
+lines(1.4+fvals[10:200]/20,5+(2*fsignal[10:200]),lwd=2)
+
+signal <- 0.8*cos(5*tvals * 2*pi + 3*pi/2)
+ypos <- c(0,1.5,2.5,3.5)
+for (n in 1:4){
+nspec <- ((fft(signal)/length(signal)) + 0.2*rnorm(length(fvals))/fvals)
+nspec[c(1:10,9991:10000)] <- 0
+lines(1.4+fvals[10:200]/20,ypos[n]+(2*abs(nspec[10:200])),lwd=2)
+
+nsignal <- Re(fft(nspec, inverse=TRUE))
+nsignal <- nsignal[1:1000]/max(abs(nsignal[1:1000]))
+nsignal <- nsignal - mean(nsignal)
+lines(tvals[1:1000],ypos[n]-0.5+(nsignal[1:1000]+1),lwd=2)
+}
+
+signal <- 0.4*cos(5*tvals * 2*pi + pi/4)
+allsimsignals <- NULL
+nspec <- fft(signal)/length(signal)
+for (n in 1:20){
+  respec <- Re(nspec) + 1*rnorm(length(fvals))/fvals
+  imspec <- Im(nspec) + 1*rnorm(length(fvals))/fvals
+
+  allsimsignals[n] <- complex(real=respec[51],imaginary=imspec[51])
+}
+
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(-1,3.8), ylim=c(-1,1))
+
+rangles <- seq(0,360)*(pi/180)
+r <- seq(0.2,1,0.2)
+for (n in 1:length(r)){
+  lines(r[n]*cos(rangles),r[n]*sin(rangles),col='gray')
+}
+lines(c(0,0),c(-1,1),lwd=2)
+lines(cos(c(30,210)*pi/180),sin(c(30,210)*pi/180),lwd=2)
+lines(cos(c(150,330)*pi/180),sin(c(150,330)*pi/180),lwd=2)
+
+text(0.5,-0.2,'Amplitude',cex=1.2,adj=0.5,srt=-30)
+text(1.1,0,'Phase',cex=1.2,adj=0.5,srt=-90)
+
+stdresps <- sd(Mod(allsimsignals))
+stdangles <- sd(Arg(allsimsignals))
+pvals <- getbanana(c(mean(Mod(allsimsignals)),Arg(mean(allsimsignals)),stdresps,stdangles))
+polygon(pvals[,1], pvals[,2], col=addalpha(nbdtpal[3],0.2), border=NA)
+
+for (n in 1:length(allsimsignals)){lines(c(0,Re(allsimsignals[n])),c(0,Im(allsimsignals[n])),col=addalpha(nbdtpal[1],0.5))}
+points(Re(allsimsignals),Im(allsimsignals),pch=16,col=addalpha(nbdtpal[1],0.5),cex=0.8)
+
+meanangle <- mean(((Arg(allsimsignals)-1) %% 2*pi)+1)
+meanangle <- Arg(mean(allsimsignals))
+meanamp <- mean(Mod(allsimsignals))
+y <- meanamp*sin(meanangle)
+x <- meanamp*cos(meanangle)
+lines(c(0,x),c(0,y),lwd=3,col=nbdtpal[3])
+points(x,y,pch=21,bg=nbdtpal[3],cex=1.5)
+
+arrows(2.8,-1,2.8,1,length=0.1,lwd=2)
+arrows(1.8,0,3.8,0,length=0.1,lwd=2)
+text(1.93,0.1,'Real',cex=1.2,adj=0.5)
+text(2.7,-0.7,'Imaginary',cex=1.2,adj=0.5,srt=90)
+
+points(Re(allsimsignals)+2.8,Im(allsimsignals),pch=16,col=addalpha(nbdtpal[1],0.5),cex=0.8)
+points(Re(mean(allsimsignals))+2.8,Im(mean(allsimsignals)),pch=21,bg=nbdtpal[3],cex=1.5)
+eldata <- getel(allsimsignals)
+polygon(eldata[1,]+2.8,eldata[2,], col=addalpha(nbdtpal[3],0.2), border=NA)
+
+text(-1,1,'(c)',cex=1.5,adj=0.5)
+text(1.8,1,'(d)',cex=1.5,adj=0.5)
+
+```
 
@@
 
-text(0.5,19,'x',adj=0.5,cex=1.5)
-text(4,13.5,'y',adj=0.5,cex=1.5)
+text(0.6,19,'Re',adj=0.5,cex=1.5)
+text(4,13.6,'Im',adj=0.5,cex=1.5,srt=90)
 
@@
 polygon(a$x+(d-1)*12,a$y,col=addalpha(nbdtpal[d],alpha=0.3),border=NA)
-
-# lines(seq(0,10,0.01)+(d-1)*12,10*pnorm(seq(-5,5,0.01)),lwd=2)
-# q <- quantile(scale(absvals),probs=seq(0,1,0.001))
-# lines(5+q+(d-1)*12,seq(0,10,0.01),lwd=2,col=nbdtpal[d])
 
@@
 text(1.1,3.8,'(d)',cex=2)
-legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white',title='Proportion')
+legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white',title='Power')
 
@@
 ```{r powerfig2, fig.cap="Simulations comparing the sensitivity of MANOVA and ANOVA2circ.
-The format mirrors that of Figure 2.
+The format mirrors that of Figure 3.
 In these simulations, there were three conditions, with the signal being added to one condition only.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}

### 2021-01-02 manuscript/manuscript.tex
Added new figure

@@
 \author{Daniel H. Baker}
-\date{2021-01-01}
+\date{2021-01-02}
 
@@
 A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal.
-The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response.
-Fourier spectra comprise both amplitude and phase information, which can be equivalently expressed as complex numbers with real and imaginary components.
+The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-defined index of the brain's response (see Figure \ref{fig:fourierexplain}a,b).
+Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \ref{fig:fourierexplain}c), or equivalently as complex numbers with real and imaginary components (Figure \ref{fig:fourierexplain}d).
 In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data only.
@@
 Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/fourierexplain-1}
+
+}
+
+\caption{Illustration of the principles of Fourier analysis for time-varying signals.
+Panel (a) shows a sinusoidal signal waveform, and simulated neural responses for successive observations (repetitions in the same individual, or recorded from multiple individuals).
+Panel (b) shows the Fourier amplitude spectra of the waveforms in (a), with clear peaks at the signal frequency (5Hz) in each example.
+The Fourier spectrum also includes a phase term, which can be represented in polar coordinates (amplitude and phase); panel (c) shows this for example responses at the signal frequency (grey points) and their average (blue point), which is computed separately for amplitude and phase terms.
+Alternatively, the same information is contained in a Cartesian representation of the real and imaginary parts of the complex spectrum (panel (d)).
+Notice that the individual observations (grey points) are the same in panels (c,d), but the different representations have implications for how the average (blue points) and measures of spread (shaded regions) are calculated.}\label{fig:fourierexplain}
+\end{figure}
 
@@
 \caption{Simulations comparing the sensitivity of MANOVA and ANOVA2circ.
-The format mirrors that of Figure 2.
+The format mirrors that of Figure 3.
 In these simulations, there were three conditions, with the signal being added to one condition only.}\label{fig:powerfig2}
@@
 Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs.
-However the \emph{multRM} function in the {[}\emph{MANOVA.RM}{]}\{\url{https://CRAN.R-project.org/package=MANOVA.RM}\} package (Friedrich et al., 2019) is able to handle such designs appropriately using a bootstrapping approach.
+However the \emph{multRM} function in the \href{https://CRAN.R-project.org/package=MANOVA.RM}{\emph{MANOVA.RM}} package (Friedrich et al., 2019) is able to handle such designs appropriately using a bootstrapping approach.
 
@@
 2014.
-Abnormal visual gain control in a parkinson's disease model.
+Abnormal visual gain control in a Parkinson's disease model.
 \emph{Hum Mol Genet} \textbf{23}:4465--78.
@@
 1994.
-FMRI of human visual cortex.
+fMRI of human visual cortex.
 \emph{Nature} \textbf{369}:525.
@@
 \leavevmode\hypertarget{ref-Hotelling1931}{}%
-Hotelling H. 1931. The generalization of student's ratio. \emph{The Annals of Mathematical Statistics} \textbf{2}:360--378.
+Hotelling H. 1931. The generalization of Student's ratio. \emph{The Annals of Mathematical Statistics} \textbf{2}:360--378.
 
@@
 2020.
-High-density eeg of auditory steady-state responses during stimulation of basal forebrain parvalbumin neurons.
+High-density EEG of auditory steady-state responses during stimulation of basal forebrain parvalbumin neurons.
 \emph{Sci Data} \textbf{7}:288.
@@
 \leavevmode\hypertarget{ref-Morrone1987}{}%
-Morrone MC, Burr DC, Speed HD. 1987. Cross-orientation inhibition in cat is gaba mediated. \emph{Exp Brain Res} \textbf{67}:635--44. doi:\href{https://doi.org/10.1007/BF00247294}{10.1007/BF00247294}
+Morrone MC, Burr DC, Speed HD. 1987. Cross-orientation inhibition in cat is GABA mediated. \emph{Exp Brain Res} \textbf{67}:635--44. doi:\href{https://doi.org/10.1007/BF00247294}{10.1007/BF00247294}
 
@@
 2015.
-Classification of parkinson's disease genotypes in drosophila using spatiotemporal profiling of vision.
+Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.
 \emph{Sci Rep} \textbf{5}:16933.

### 2021-01-02 manuscript/misc/fourierfig.R

@@ -1,0 +1,127 @@
+nbdtpal <- c(rgb(0,0,0),rgb(0.73,0.16,0.42),rgb(0.22,0.44,0.64),'darkgreen','darkorange')
+addalpha <- function(col, alpha=1){apply(sapply(col, col2rgb)/255, 2, function(x) rgb(x[1], x[2], x[3], alpha=alpha))}
+set.seed(170382)
+
+getbanana <- function(p){
+  thetaRange <- seq(0.01, pi, 0.01)
+  Tval <- atan((p[3]*tan(thetaRange))/p[4])
+  x1 <- p[1] + (p[3] * cos(Tval))
+  x2 <- p[1] - (p[3] * cos(Tval))
+  y <- p[2] + (p[4] * sin(Tval))
+  
+  XXX <- c(x1[1:(length(x1)/2)], x2[(length(x2)/2):1])
+  Y <- c(y[1:(length(y)/2)], y[(length(y)/2):1], y[length(y):(1+length(y)/2)], y[(1+length(y)/2):length(y)])
+  X <- c(XXX, XXX[length(XXX):1])
+  errormatrix <- pol2cart(matrix(data=c(Y,X),ncol=2))
+  return(errormatrix)
+}
+
+par(mfrow=c(2,1), mar=c(3,4,1,2))
+
+plotlims <- c(0,2.4,0,6) 
+ticklocsx <- seq(0,1,0.5)    # locations of tick marks on x axis
+ticklocsy <- seq(0,0.1,0.02)    # locations of tick marks on y axis
+ticklabelsx <- ticklocsx        # set labels for x ticks
+ticklabelsy <- ticklocsy    # set labels for y ticks
+
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
+axis(1, at=c(0,1), tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+# axis(2, at=c(0,6), tck=0.01, lab=F, lwd=2)
+axis(1, at=c(1.4,2.4), tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+# axis(4, at=c(0,6), tck=0.01, lab=F, lwd=2)
[…]

### 2021-01-03 manuscript/manuscript.Rmd
Minor updates to some functions

@@
 
-nsims <- 100  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
+nsims <- 1000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
 
@@
 Next the logic of $T^2_{circ}$ is extended to situations with more than two levels of the independent variable, and the performance of this novel $ANOVA^2_{circ}$ statistic is compared to MANOVA.
-Finally, the proposed techniques are demonstrated on two example SSVEP data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.
+Finally, the proposed techniques are demonstrated on two example data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.
 
@@
 
-One possible remedy to control the Type I error rate would be to adjust the alpha level to compensate.
+One possible remedy to control the Type I error rate would be to adjust either the alpha level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated).
 However, this will reduce the statistical power of the $T^2_{circ}$ test, and its advantage over $T^2$ is relatively marginal in most situations to begin with (see Figure \@ref(fig:powerfig)).
@@
 Victor and Mast [-@Victor1991] suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components [see also @Pei2017].
+However it would be useful to develop a method that can tell us whether the assumptions hold for a given data set.
 
@@
 Conventionally, the condition index is calculated as the square root of the longest/shortest eigenvector length.
-For uncorrelated random numbers the distribution of condition indices is positively skewed, with a minimum of 1 [@Edelman1988].
+For uncorrelated random numbers the expected distribution of condition indices is positively skewed, with a minimum of 1 [@Edelman1988].
 This is because two independent samples of numbers from the same underlying distribution will generally by chance have unequal eigenvectors, and the definition of the condition index ($\sqrt{longest/shortest}$) prevents its values dropping below 1 (values < 1 would imply that the shortest eigenvector is longer than the longest one).
-Edelman [-@Edelman1988] provides an equation (his Eq.
+For bivariate data, Edelman [-@Edelman1988] provides an equation (his Eq.
 14) for the probability density function of condition indices (_x_) as a function of sample size (_N_):
@@
 
-Figure \@ref(fig:distcomparison)a shows an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100,000 random data sets with _N_ = 10 (blue shading).
+Figure \@ref(fig:distcomparison)a shows example distributions derived from both expressions (black and red curves), and simulations from 100,000 random data sets with _N_ = 4 (blue shading).
 The vertical lines show the critical (95%) threshold for the analytic and simulated results.
 A ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance, because it has a condition index larger than expected by chance (assuming $\alpha$ = 0.05).
-Ratios below the threshold imply that the eigenvalues are equal (in a statistical sense).
-Figure \@ref(fig:distcomparison)b shows how these critical thresholds change as a function of the number of observations (N), and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
+Ratios below the threshold imply that the eigenvalues can be considered equal (in a statistical sense).
+Figure \@ref(fig:distcomparison)b shows how these critical thresholds change as a function of the number of observations (N), and it appears that the modified expression (red) most closely approximates the simulation results (blue).
 
 ```{r distcomparison, fig.cap='Logic of the condition index test.
-Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of N = 10 observations.
-The vertical blue and red lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function.
-Ellipse icons above panel (a) illustrate different condition indices between 1 and 5 (note that rotation does not affect the condition index).
+Panel (a) shows the distribution of condition indices derived from Equations 3 (black curve) and 4 (red curve), and by stochastic simulation (blue shading), for a sample size of N = 4 observations.
+The vertical lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line).
+Ellipse icons above panel (a) illustrate different condition indices between 1 and 19 (note that rotation does not affect the condition index).
 Panel (b) shows how 95 percent thresholds change as a function of the number of observations.', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
@@
 
-n <- 10
-x <- seq(1,20,0.01)
+n <- 4
+x <- seq(1,19,0.01)
+nr <- n
+pdfED <- ((nr-1)*(2^(nr-1))) * ((x^2 - 1)/((x^2+1)^nr)) * (x^(nr-2))
 nr <- n-1
@@
 
-plotlims <- c(1,5,0,1.2) 
-ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
+plotlims <- c(1,19,0,1.2) 
+ticklocsx <- seq(1,19,2)    # locations of tick marks on x axis
 ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
@@
 title(xlab="Condition index", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
-title(ylab="Probability density      ", line=1.5, cex.lab=1.5)
-
-lines(x,pdffunction/max(pdffunction),lwd=3)
+title(ylab="Probability density", line=1.5, cex.lab=1.5)
+
+lines(x,pdfED/max(pdfED),lwd=3)
+lines(x,pdffunction/max(pdffunction),lwd=3,col=nbdtpal[2])
 cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction))
-lines(x,cdfinverse,col=nbdtpal[2],lwd=3)
-lines(c(1,5),c(0.05,0.05),lty=2,col=nbdtpal[1])
+cdfinverseE <- 1-(cumsum(pdfED)/sum(pdfED))
+# lines(x,cdfinverse,col=nbdtpal[2],lwd=3)
+lines(c(1,19),c(0.05,0.05),lty=2,col=nbdtpal[1])
 criticalx <- x[min(which(cdfinverse<=0.05))]
 lines(c(criticalx,criticalx),c(0,1),col=nbdtpal[2],lwd=3)
+criticalxED <- x[min(which(cdfinverseE<=0.05))]
+lines(c(criticalxED,criticalxED),c(0,1),lwd=3)
 
@@
 }
-
-a <- hist(simvect,breaks = 200, plot=FALSE)
+simvect[which(simvect>200)] <- 200
+a <- hist(simvect,breaks = 2000, plot=FALSE)
 axvals <- a$mids
@@
 lines(c(simcrit,simcrit),c(0,1),col=nbdtpal[3],lwd=3,lty=2)
-text(4.8,0.95,'(a)',cex=1.5)
-
-xpos <- c(1.2,2.1,3,3.9,4.8)
-angles <- seq(0, 2*pi, length.out=200) 
-RR <- matrix(c(1,0,0,1),nrow=2,ncol=2)
-for (n in 1:5){
-  RR[2,2] <- 1/sqrt(n)
-  ell <- 1 * cbind(cos(angles), sin(angles)) %*% RR
-  polygon((ell[,1]*0.4*(5/6))+xpos[n],(ell[,2]/10)+1.1,col=rgb(0.9,0.9,0.9),border=NA)
-}
-
-x <- seq(1,50,0.01)
+
+text(11,0.95,'N = 4',pos=4,cex=1.5)
+text(18,0.95,'(a)',cex=1.5)
+
+x <- seq(1,1000,0.001)
 edelman1 <- NULL
@@
 simratios[n-2] <- quantile(simvect,0.95)
+}
+
+xpos <- c(2,4.75,7.5,10,12.5,15.25,18)
+cis <- c(1,4,7,10,13,16,19)
+angles <- seq(0, 2*pi, length.out=200) 
+RR <- matrix(c(1,0,0,1),nrow=2,ncol=2)
+for (n in 1:length(cis)){
+  RR[2,2] <- 1/sqrt(cis[n])
+  ell <- 1 * cbind(cos(angles), sin(angles)) %*% RR
+  polygon((ell[,1]*0.4*4*(5/6))+xpos[n],(ell[,2]/10)+1.1,col=rgb(0,0,0,alpha=0.2),border=NA)
 }
@@
 If a condition index is observed that is above the critical threshold for the number of observations, then the data set can be said to significantly violate the assumption of equal eigenvalues.
-Because the modified equation provides an inverse density function (red curve in Figure \@ref(fig:distcomparison)a), this can be used to calculate a _p_-value for the test.
+Because the modified equation permits estimation of an inverse density function, this can be used to calculate a _p_-value for the test.
 If the test is non-significant, one can proceed with $T^2_{circ}$; if it is significant, $T^2$ should be used instead.
@@
 It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points.
-This means that any correlations in the data set are taken into account when calculating the distance metric, D.
+This means that any correlations in the data set are taken into account when calculating the distance metric, _D_.
 
@@
 Figure \@ref(fig:outlierplot) shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves).
-In all cases, the functions depart from the Type I error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core _mahalanobis_ function in _R_, return $D^2$, which can be converted to D by taking the square root).
+In all cases, the functions depart from the Type I error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the outlier's Mahalanobis distance exceeds a value around 3.
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core _mahalanobis_ function in _R_, return $D^2$, which can be converted to _D_ by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the $T^2_{circ}$ test.
@@
 The ellipses are calculated with the outlier included (red) and excluded (black), illustrating how the outlier distorts the aspect ratio of the ellipse.
-The lower plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.', fig.align="center", fig.width=7, fig.height=6, echo=FALSE}
+The main plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.', fig.align="center", fig.width=7, fig.height=6, echo=FALSE}
 
@@
 
-Many studies involve more than two conditions that need to be compared.
+Many studies involve more than two experimental conditions that need to be compared.
 Again, issues with familywise error will quickly become problematic if multiple pairwise $T^2$ or $T^2_{circ}$ statistics are calculated.
-One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's $T^2$, but permits multiple independent variables with more than two levels.
-However, if the assumptions of $T^2_{circ}$ hold for a data set, it might alternatively be possible to extend the logic of the $T^2_{circ}$ test [@Victor1991] to the more general case.
+One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's $T^2$, but permits independent variables with more than two levels, as well as factorial designs.
+However, if the assumptions of $T^2_{circ}$ hold for a data set, it might alternatively be possible to extend the logic of the $T^2_{circ}$ test [@Victor1991] to the more general case, and obtain a sensitivity benefit similar to that shown in Figure \@ref(fig:powerfig).
 
@@
 The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number.
-A _p_-value can be derived from the F-distribution in the usual way.
+A _p_-value can then be derived from the F-distribution in the usual way.
 
@@
 
-Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.
+Following a significant ${ANOVA}^2_{circ}$ test, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.
 Bonferroni correction).
-Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary).
-A repeated measures version of ${ANOVA}^2_{circ}$ has also been implemented following the same logic.
-The project repository contains an _R_ function (_anovacirc.test_) to run both of these tests.
+Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary; see Figure \@ref(fig:fourierexplain)d).
+A repeated measures version of ${ANOVA}^2_{circ}$ can also be implemented following the same logic.
+The project repository contains an _R_ function (_anovacirc.test_) to run both of these tests for one-way designs.
+In principle factorial versions might also be derived.
 
@@
 Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test.
-Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, and the $T^2$ statistic if not.
+Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, and the $T^2$ statistic if they are not.
 Comparisons with more than two conditions should be tested with the $ANOVA^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not.
-Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs.
+Many MANOVA implementations cannot deal correctly with random factors (repeated measures), particularly in complex factorial designs.
 However the _multRM_ function in the [_MANOVA.RM_](https://CRAN.R-project.org/package=MANOVA.RM) package [@Friedrich2019] is able to handle such designs appropriately using a bootstrapping approach.
@@
 The second study measured visual responses to flickering grating patterns in humans.
-These examples also provide an illustration of how the results of the tests described above might be appropriately reported.
+These examples also provide a demonstration of how the results of the tests described above might be appropriately reported.
 
@@
 The mice had previously been given a targeted virus that made parvalbumin neurons in their basal forebrain responsive to specific wavelengths of light, delivered through an optical fiber (a technique called optogenetics).
-SSVEPs were recorded from 36 electrodes for 1 second epochs of 40 Hz auditory stimulation, and various schedules of optogenetic stimulation (including at 40 Hz).
+Steady-state evoked potentials were recorded from 36 electrodes for 1 second epochs of 40 Hz auditory stimulation, and various schedules of optogenetic stimulation (including at 40 Hz).
 The data set is described more fully by Hwang et al.
@@
 The condition index test was non-significant for both conditions (sound: CI = `r round(cisound$CI,digits=2)`, _p_ = `r round(cisound$pval,digits=2)`; light: CI = `r round(cilight$CI,digits=2)`, _p_ = `r round(cilight$pval,digits=2)`), so a paired-samples $T^2_{circ}$ test was conducted.
-This revealed a significant difference between conditions ($T^2_{circ}$ = `r round(tsqcresult$tsqc,digits=2)`, $F_{(`r tsqcresult$df1`,`r tsqcresult$df2`)}$ = `r round(tsqcresult$Fratio,digits=2)`, _p_ = `r round(tsqcresult$p.value,digits=3)`) with an effect size of D = `r round(D,digits=2)`.
+This revealed a significant difference between conditions ($T^2_{circ}$ = `r round(tsqcresult$tsqc,digits=2)`, $F_{(`r tsqcresult$df1`,`r tsqcresult$df2`)}$ = `r round(tsqcresult$Fratio,digits=2)`, _p_ = `r round(tsqcresult$p.value,digits=3)`) with an effect size of _D_ = `r round(D,digits=2)`.
 This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
@@
 Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \@ref(fig:humanSSVEP)), were well-isolated in the Fourier domain, and increased with stimulus contrast.
-Significant activity was evident at 4% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \@ref(fig:humanSSVEP).
-For the main analysis, responses were taken from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant.
+Significant activity was evident at 4% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \@ref(fig:humanSSVEP).
+For the main analysis, responses were taken from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), and averaged across repetition for each participant.
 Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \@ref(fig:humanSSVEP).
@@
 A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(`r withincirc$dfM`,`r withincirc$dfR`)}$ = `r round(withincirc$Fratio,digits=1)`, _p_ < `r max(round(withincirc$p.value,digits=3),0.001)`).
-Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[3],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[3],digits=2)`, D = `r round(allD[3],digits=2)`, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[4],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[4],digits=2)`, D = `r round(allD[4],digits=2)`, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[5],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[5],digits=2)`, D = `r round(allD[5],digits=2)`, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[6],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[6],digits=2)`, D = `r round(allD[6],digits=2)`, _p_ < 0.001).
+Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[3],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[3],digits=2)`, _D_ = `r round(allD[3],digits=2)`, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[4],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[4],digits=2)`, _D_ = `r round(allD[4],digits=2)`, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[5],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[5],digits=2)`, _D_ = `r round(allD[5],digits=2)`, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[6],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[6],digits=2)`, _D_ = `r round(allD[6],digits=2)`, _p_ < 0.001).
 The study by Vilidaite et al.
@@
 
-It is worth pointing out that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations.
+It is worth stating explicitly that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations.
 This is the case for most paradigms in which the nervous system is driven by a periodic stimulus.
@@
 Another possibility is to use machine learning techniques such as multivariate pattern analysis (MVPA) to analyse periodic data.
-This involvess training a classifier algorithm to distinguish between two (or more) experimental conditions or states, and then assessing classifier accuracy on fresh data.
+This involves training a classifier algorithm to distinguish between two (or more) experimental conditions or states, and then assessing classifier accuracy for predicting the group labels of fresh data.
 If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance.

### 2021-01-03 R/CI.test.R

@@ -16,8 +16,7 @@ CI.test <- function(data, alpha=0.05){
   eigVal  <- eigen(C)$values  # calculate eigenvalues
   CI <- sqrt(eigVal[1]/eigVal[2]) # calculate condition index
 
-  maxci <- max(20,round(CI+5))   # make sure our target CI is in range
-  cilist <- seq(1,maxci,0.01)    # list of possible CIs
+  cilist <- seq(1,100,0.001)    # list of possible CIs
 
   # calculate probability density function using modified Edelman equation
   pdffunction <- ((N-2)*(2^(N-2))) *
@@ -25,8 +24,9 @@ CI.test <- function(data, alpha=0.05){
   cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
   criticalCI <- cilist[min(which(cdfinverse<=alpha))]  # find the threshold CI
 
+  pvals <- 0
   indices <- which(cilist>=CI)  # find the CI values larger than our CI
-  pval <- cdfinverse[indices[1]] # estimate the p-value
+  if (length(indices)>0){pval <- cdfinverse[indices[1]]} # estimate the p-value
 
   output <- data.frame(CI,N,criticalCI,pval)
   return(output)}

### 2021-01-03 manuscript/manuscript.tex
Minor updates to some functions

@@
 \author{Daniel H. Baker}
-\date{2021-01-02}
+\date{2021-01-03}
 
@@
 Next the logic of \(T^2_{circ}\) is extended to situations with more than two levels of the independent variable, and the performance of this novel \(ANOVA^2_{circ}\) statistic is compared to MANOVA.
-Finally, the proposed techniques are demonstrated on two example SSVEP data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.
+Finally, the proposed techniques are demonstrated on two example data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.
 
@@
 
-One possible remedy to control the Type I error rate would be to adjust the alpha level to compensate.
+One possible remedy to control the Type I error rate would be to adjust either the alpha level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated).
 However, this will reduce the statistical power of the \(T^2_{circ}\) test, and its advantage over \(T^2\) is relatively marginal in most situations to begin with (see Figure \ref{fig:powerfig}).
@@
 Victor and Mast (1991) suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components (see also Pei et al., 2017).
+However it would be useful to develop a method that can tell us whether the assumptions hold for a given data set.
 
@@
 Conventionally, the condition index is calculated as the square root of the longest/shortest eigenvector length.
-For uncorrelated random numbers the distribution of condition indices is positively skewed, with a minimum of 1 (Edelman, 1988).
+For uncorrelated random numbers the expected distribution of condition indices is positively skewed, with a minimum of 1 (Edelman, 1988).
 This is because two independent samples of numbers from the same underlying distribution will generally by chance have unequal eigenvectors, and the definition of the condition index (\(\sqrt{longest/shortest}\)) prevents its values dropping below 1 (values \textless{} 1 would imply that the shortest eigenvector is longer than the longest one).
-Edelman (1988) provides an equation (his Eq.
+For bivariate data, Edelman (1988) provides an equation (his Eq.
 14) for the probability density function of condition indices (\emph{x}) as a function of sample size (\emph{N}):
@@
 
-Figure \ref{fig:distcomparison}a shows an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100,000 random data sets with \emph{N} = 10 (blue shading).
+Figure \ref{fig:distcomparison}a shows example distributions derived from both expressions (black and red curves), and simulations from 100,000 random data sets with \emph{N} = 4 (blue shading).
 The vertical lines show the critical (95\%) threshold for the analytic and simulated results.
 A ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance, because it has a condition index larger than expected by chance (assuming \(\alpha\) = 0.05).
-Ratios below the threshold imply that the eigenvalues are equal (in a statistical sense).
-Figure \ref{fig:distcomparison}b shows how these critical thresholds change as a function of the number of observations (N), and it is clear that the modified expression (red) most closely approximates the simulation results (blue).
+Ratios below the threshold imply that the eigenvalues can be considered equal (in a statistical sense).
+Figure \ref{fig:distcomparison}b shows how these critical thresholds change as a function of the number of observations (N), and it appears that the modified expression (red) most closely approximates the simulation results (blue).
 
@@
 \caption{Logic of the condition index test.
-Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of N = 10 observations.
-The vertical blue and red lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function.
-Ellipse icons above panel (a) illustrate different condition indices between 1 and 5 (note that rotation does not affect the condition index).
+Panel (a) shows the distribution of condition indices derived from Equations 3 (black curve) and 4 (red curve), and by stochastic simulation (blue shading), for a sample size of N = 4 observations.
+The vertical lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line).
+Ellipse icons above panel (a) illustrate different condition indices between 1 and 19 (note that rotation does not affect the condition index).
 Panel (b) shows how 95 percent thresholds change as a function of the number of observations.}\label{fig:distcomparison}
@@
 If a condition index is observed that is above the critical threshold for the number of observations, then the data set can be said to significantly violate the assumption of equal eigenvalues.
-Because the modified equation provides an inverse density function (red curve in Figure \ref{fig:distcomparison}a), this can be used to calculate a \emph{p}-value for the test.
+Because the modified equation permits estimation of an inverse density function, this can be used to calculate a \emph{p}-value for the test.
 If the test is non-significant, one can proceed with \(T^2_{circ}\); if it is significant, \(T^2\) should be used instead.
@@
 It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points.
-This means that any correlations in the data set are taken into account when calculating the distance metric, D.
+This means that any correlations in the data set are taken into account when calculating the distance metric, \emph{D}.
 
@@
 Figure \ref{fig:outlierplot} shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves).
-In all cases, the functions depart from the Type I error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R}, return \(D^2\), which can be converted to D by taking the square root).
+In all cases, the functions depart from the Type I error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the outlier's Mahalanobis distance exceeds a value around 3.
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R}, return \(D^2\), which can be converted to \emph{D} by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the \(T^2_{circ}\) test.
@@
 The ellipses are calculated with the outlier included (red) and excluded (black), illustrating how the outlier distorts the aspect ratio of the ellipse.
-The lower plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.}\label{fig:outlierplot}
+The main plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.}\label{fig:outlierplot}
 \end{figure}
@@
 
-Many studies involve more than two conditions that need to be compared.
+Many studies involve more than two experimental conditions that need to be compared.
 Again, issues with familywise error will quickly become problematic if multiple pairwise \(T^2\) or \(T^2_{circ}\) statistics are calculated.
-One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's \(T^2\), but permits multiple independent variables with more than two levels.
-However, if the assumptions of \(T^2_{circ}\) hold for a data set, it might alternatively be possible to extend the logic of the \(T^2_{circ}\) test (Victor and Mast, 1991) to the more general case.
+One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's \(T^2\), but permits independent variables with more than two levels, as well as factorial designs.
+However, if the assumptions of \(T^2_{circ}\) hold for a data set, it might alternatively be possible to extend the logic of the \(T^2_{circ}\) test (Victor and Mast, 1991) to the more general case, and obtain a sensitivity benefit similar to that shown in Figure \ref{fig:powerfig}.
 
@@
 The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number.
-A \emph{p}-value can be derived from the F-distribution in the usual way.
+A \emph{p}-value can then be derived from the F-distribution in the usual way.
 
@@
 
-Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction).
-Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary).
-A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic.
-The project repository contains an \emph{R} function (\emph{anovacirc.test}) to run both of these tests.
+Following a significant \({ANOVA}^2_{circ}\) test, one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction).
+Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary; see Figure \ref{fig:fourierexplain}d).
+A repeated measures version of \({ANOVA}^2_{circ}\) can also be implemented following the same logic.
+The project repository contains an \emph{R} function (\emph{anovacirc.test}) to run both of these tests for one-way designs.
+In principle factorial versions might also be derived.
 
@@
 Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test.
-Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if not.
+Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if they are not.
 Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not.
-Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs.
+Many MANOVA implementations cannot deal correctly with random factors (repeated measures), particularly in complex factorial designs.
 However the \emph{multRM} function in the \href{https://CRAN.R-project.org/package=MANOVA.RM}{\emph{MANOVA.RM}} package (Friedrich et al., 2019) is able to handle such designs appropriately using a bootstrapping approach.
@@
 The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted.
-This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of D = 1.02.
+This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of \emph{D} = 1.02.
 This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
@@
 Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast.
-Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \ref{fig:humanSSVEP}.
-For the main analysis, responses were taken from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant.
+Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \ref{fig:humanSSVEP}.
+For the main analysis, responses were taken from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), and averaged across repetition for each participant.
 Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}.
@@
 A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, \emph{p} \textless{} 0.001).
-Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, D = 0.33, \emph{p} \textless{} 0.001).
+Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, \emph{D} = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, \emph{D} = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, \emph{D} = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, \emph{D} = 0.33, \emph{p} \textless{} 0.001).
 The study by Vilidaite et al.
@@
 2015.
-Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.
+Classification of Parkinson's disease genotypes in Drosophila using spatiotemporal profiling of vision.
 \emph{Sci Rep} \textbf{5}:16933.

### 2021-01-03 R/amperrors.R

@@ -1,15 +1,16 @@
 #' amperrors: calculate error bars on the amplitude component of coherently averaged data
 #' the input is a vector of complex numbers, or an Nx2 matrix of real and imaginary components
 #' the optional 'method' flag allows the following values:
-#'  - 'boot': (default) bootstraps the confidence intervals (10000 iterations)
+#'  - 'boot': (default) bootstraps the confidence intervals
 #'  - 'circ': calculates confidence intervals based on a circular bounding region (used when the condition index test is non-significant)
 #'  - 'ellipse': calculates confidence intervals based on an elliptical bounding region (based on Pei et al. 2017, doi: 10.1016/j.visres.2016.03.010)
 #'  - 'abs': calculates the standard error (and the mean) using the absolute amplitudes
 #'  the optional 'quantiles' flag allows the quantile to be set for the confidence intervals
+#'  the optional 'nresamples' variable allows the number of iterations to be determined for the bootstrapping option (ignored for other methods)
 #'  typical values are 95 (for 95% confidence intervals), and 68 (for standard errors)
 #'  the function returns the mean amplitude and the upper and lower error bars
 #' @export
-amperrors <- function(input, method='boot',quantiles=95){
+amperrors <- function(input, method='boot',quantiles=95,nresamples=10000){
   d <- dim(input)
   compdata <- input
   if (!is.complex(input)){compdata <- complex(real=input[,1],imaginary=input[,2])}
@@ -21,7 +22,7 @@ amperrors <- function(input, method='boot',quantiles=95){
   if (method=='boot'){
     # bootstrap the error bars on the amplitude using 10000 resamples
     bspop <- NULL
-    for (n in 1:10000){bspop[n] <- abs(mean(sample(compdata,length(compdata),replace=TRUE)))}
+    for (n in 1:nresamples){bspop[n] <- abs(mean(sample(compdata,length(compdata),replace=TRUE)))}
     output$upperCI <- quantile(bspop,probs=1-(1-(quantiles/100))/2)
     output$lowerCI <- quantile(bspop,probs=(1-(quantiles/100))/2)
   }

### 2021-01-04 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 
-nsims <- 1000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
+nsims <- 100000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
 

### 2021-01-06 FourierStatsMatlab/CI_test.m
Started implementing Matlab version of toolbox

@@ -1,0 +1,44 @@
+function output = CI_test(data,alpha)
+% CI.test: function to calculate the condition index of bivariate data
+% Inputs:
+%   data is an N x 2 matrix of bivariate (x,y) observations, or a vector of N complex values
+%   alpha is the criterion for statistical significance, set to a default of 0.05
+% Output:
+%   a structure with fields: CI (the condition index), N (sample size),
+%   criticalCI (the critical index, given alpha), pval (the p-value)
+% the condition index is square root of the the ratio of eigenvalues (eigenvector lengths), calculated longest/shortest
+% the index is then compared for significance with an expected distribution function
+% significant tests (p < alpha) violate the assumptions of the T-squared-circ and ANOVA-squared-circ tests
+% see Baker (2021) for further details
+
+if (isempty(alpha))
+    alpha = 0.05;
+end
+
+if (~isreal(data))
+    data = [real(data); imag(data)]';
+end
+
+C = cov(data);
+N = length(data);
+eigVal = eig(C);
+CI = sqrt(max(eigVal)/min(eigVal));
+
+cilist = 1:0.001:100;
+
+pdffunction = ((N-2).*(2.^(N-2))) .* ((cilist.^2 - 1)./((cilist.^2+1).^(N-1))) .* (cilist.^(N-3));
+cdfinverse = 1 - (cumsum(pdffunction)./sum(pdffunction));
+criticalCI = min(cilist(find(cdfinverse<alpha)));
+
+pval = 0;
+indices = find(cilist>=CI);
+if (~isempty(indices))
+    pval = cdfinverse(indices(1));
+end
+
+output.CI = CI;
+output.N = N;
+output.criticalCI = criticalCI;
+output.pval = pval;
+    
+end

### 2021-01-06 FourierStatsMatlab/anovacirc_test.m

@@ -1,0 +1,117 @@
+function output = anovacirc_test(data,group,participant)
+
+% anovacirc.test: two-dimensional analysis of variance using complex data
+% an extension of the logic of the T-squared circ statistic of Victor & Mast (1991)
+% this is a one-way implementation of the test - if participant IDs are supplied a repeated measures version is conducted
+%
+% Inputs--
+% data:  this can either be a vector of complex numbers, or a matrix
+%        if it is a matrix, either the first two columns are the x and y (real and imaginary) values of the DV
+%        if there are further columns, these are treated as the group labels and participant IDs
+% group: condition labels indicating the level of the independent variable that each data point corresponds to
+%        this is an optional input, but if it is supplied it supercedes values from the matrix
+% participant: variable storing participant/subject IDs (i.e. the random factor) for repeated measures analysis
+%        this is an optional input, but if it is supplied it supercedes values from the matrix
+%
+% see Baker (2021) for further details
+
+grouplabels = [];
+participantlabels = [];
+
+s = size(data);
+
+if (~isreal(data(1)))
+    datavals = data;
+else
+    datavals = complex(data(:,1),data(:,2));
+    
+    if (s(2)>2)
+        grouplabels = data(:,3);
+    end
+    if (s(2)>3)
+        participantlabels = data(:,4);
+    end
+end
+
+if (~isempty(group))
+    grouplabels = group;
+end
+if (~isempty(participant))
+    participantlabels = participant;
+end
+
+factorlist = unique(grouplabels);
+
+grandmean = mean(datavals);
+
+if (isempty(participantlabels))
+    % if no participant labels have been supplied run a between-subject ANOVA-circ test
+    
+    SSM = 0;
+    allresiduals = zeros(1,length(datavals));
+    for n = 1:length(factorlist)
+        groupmeans(n) = mean(datavals(find(grouplabels==factorlist(n))));
+        SSM = SSM + length(find(grouplabels==factorlist(n))) .* abs(groupmeans(n)-grandmean).^2;
[…]

### 2021-01-06 FourierStatsMatlab/tsqc_test.m

@@ -1,0 +1,98 @@
+function output = tsqc_test(x,y,paired,mu)
+
+% tsqc_test: function to calculate the T-squared circ statistic from Victor & Mast (1991)
+% the inputs must be Nx2 or 2xN matrices of numbers
+% for a one-sample test, mu is an optional vector to which the data are compared
+% for a two-sample test, x and y are the data from the two conditions
+% if paired=1, the matrices (x and y) must be the same size
+% the output is a structure containing the test results
+
+d = size(x);
+if (d(1)<d(2))
+    x = x';
+end
+d = size(y);
+if (d(1)<d(2))
+    y = y';
+end
+
+if (~isempty(mu))
+    for n = 1:length(mu)
+        x(:,n) = x(:,n) - mu(n);
+    end
+end
+
+if isreal(x)
+    x = complex(x(:,1),x(:,2));
+end
+if (~isempty(y))
+    if isreal(y)
+        y = complex(y(:,1),y(:,2));
+    end
+end
+
+if (isempty(y))
+    method = 'One-sample t-squared circ test';
+end
+if (~isempty(y) && paired==1)
+    method = 'Paired t-squared circ test';
+end
+if (~isempty(y) && paired==0)
+    method = 'Independent samples t-squared circ test';
+end
+
+if (paired==1)
+    if (~isempty(y))
+        x = x - y;
+    end
+end
+
+df1 = 2;
+if (paired==1 || isempty(y))
+    % paired or one-sample version of the test
+    
+    nobs = length(x);
+    cohmean = mean(x);
+    absmean = abs(cohmean);
+    
+    displacement = abs(x - cohmean);
+    diffsumsq = sum(displacement.^2);
+    tsqc = (nobs-1) .* (absmean.^2) ./ diffsumsq;
+    Fratio = nobs*tsqc;
+    df2 = nobs*2 - 2;
+    pval = fcdf(Fratio,df1,df2,'upper');
+    
+else
+    % independent samples version of the test
+    
+    nobs1 = length(x);
+    cohmean1 = mean(x);
+    absmean1 = abs(cohmean1);
+    displacement1 = abs(x - cohmean1);
+    
+    nobs2 = length(y);
+    cohmean2 = mean(y);
+    absmean2 = abs(cohmean2);
+    displacement2 = abs(y - cohmean2);
+    
[…]

### 2021-01-06 R/tsqc.test.R

@@ -24,6 +24,7 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
 
   if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}
 
+  df1 <- 2
   # paired or one-sample version of the test
   if (paired==TRUE | is.null(y)){
 
@@ -36,7 +37,7 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
     tsqc <- (nobs-1) * (absmean^2) / diffsumsq
     Fratio <- nobs*tsqc
     df2 <- nobs*2-2
-    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
+    pval <- pf(Fratio,df1=df1,df2=df2,lower.tail=FALSE)
   }
 
   # independent samples version of the test
@@ -58,10 +59,9 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
     tsqc <- (nobs1 + nobs2 - 2) * (absdiff^2) / diffsumsq
     Fratio <- ((nobs1*nobs2)/(nobs1 + nobs2))*tsqc
     df2 <- (2*nobs1 + 2*nobs2 - 4)
-    pval <- pf(Fratio,df1=2,df2=df2,lower.tail=FALSE)
+    pval <- pf(Fratio,df1=df1,df2=df2,lower.tail=FALSE)
   }
 
-  df1 <- 2
   p.value <- min(pval,1)
 
   output <- data.frame(tsqc,Fratio,df1,df2,p.value,method)

### 2021-01-06 FourierStatsMatlab/tsq1_test.m
Started implementing Matlab version of toolbox

@@ -1,0 +1,46 @@
+function output = tsq1_test(x,mu)
+
+% tsq1_test: function to calculate a one-sample Hotelling's T-squared statistic 
+% the input (x) must be Nx2 or 2xN matrices of numbers
+% or can be stored as a single vector of complex numbers
+% mu is an optional vector to which the data are compared
+% the output is a structure containing the test results
+
+d = size(x);
+if (d(1)<d(2))
+    x = x';
+end
+
+if (~isempty(mu))
+    for n = 1:length(mu)
+        x(:,n) = x(:,n) - mu(n);
+    end
+end
+
+if ~isreal(x)
+    x = [real(x); imag(x)]';
+end
+
+    method = 'One-sample Hotellings t-squared test';
+
+    xbar = mean(x);
+    s = size(x);
+    m = s(2);
+    N = s(1);
+    C = cov(x);
+    Cinv = inv(C);
+    tsq = N * xbar * Cinv * xbar';
+    Fratio = tsq .* (N - m)./(m.*(N-1));
+    df1 = m;
+    df2 = N-m;
+    pval = fcdf(Fratio,df1,df2,'upper');
+    pval = min(pval,1);
+    
+output.tsq = tsq;
+output.Fratio = Fratio;
+output.df1 = df1;
+output.df2 = df2;
+output.pval = pval;
+output.method = method;
+
+end

### 2021-01-06 R/CI.test.R

@@ -24,7 +24,7 @@ CI.test <- function(data, alpha=0.05){
   cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
   criticalCI <- cilist[min(which(cdfinverse<=alpha))]  # find the threshold CI
 
-  pvals <- 0
+  pval <- 0
   indices <- which(cilist>=CI)  # find the CI values larger than our CI
   if (length(indices)>0){pval <- cdfinverse[indices[1]]} # estimate the p-value
 

### 2021-01-06 R/tsq1.test.R

@@ -21,7 +21,7 @@ tsq1.test <- function(data,mu=NULL){
   Fratio <- tsq * (N-m)/(m*(N-1)) # convert to F-ratio
   df1 <- m      # record degrees of freedom
   df2 <- N-m
-  pval <- 1- pf(Fratio,df1,df2) # estimate p-value
+  pval <- pf(Fratio,df1,df2,lower.tail=FALSE) # estimate p-value
   p.value <- min(pval,1)
 
   # store outputs in a data structure

### 2021-01-06 manuscript/manuscript.Rmd

@@
 All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats).
-This includes an _R_ package called _FourierStats_, featuring functions to implement a one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
+This includes a _Matlab_ toolbox and an _R_ package called _FourierStats_, featuring functions to implement a one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
 

### 2021-01-07 manuscript/manuscript.Rmd
Implemented independent T-squared test. Replaced the tsq1.test function with tsqh.test (which can do both one and two sample tests).

--- 2021-01-04 manuscript/manuscript.Rmd
@@
 All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats).
-This includes an _R_ package called _FourierStats_, featuring functions to implement a one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
+This includes a _Matlab_ toolbox and an _R_ package called _FourierStats_, featuring functions to implement Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.
 
@@
     data[,1] <- data[,1] + dlist[s2]
-    a <- tsq1.test(data)
+    a <- tsqh.test(data)
     if (a[5]<0.05){tsqmatrix[s,s2] <- tsqmatrix[s,s2] + 1}
@@
     if (output$p.value < 0.05){csig[cond] <- csig[cond] + 1}
-    output <- tsq1.test(simdata)
+    output <- tsqh.test(simdata)
     if (output$p.value < 0.05){hsig[cond] <- hsig[cond] + 1}

### 2021-01-07 FourierStatsMatlab/analysecplx.m

@@ -1,0 +1,208 @@
+function output = analysecplx(data, group, participant)
+
+% analysecplx: wrapper function that automatically decides which statistical test to run for a given data set
+%
+% Inputs--
+% data:  this can either be a vector of complex numbers, or a matrix
+%        if it is a matrix, the first two columns are the x and y (real and imaginary) values of the DV
+%        if there are further columns, these are treated as the group labels and participant IDs
+% group: condition labels indicating the level of the independent variable that each data point corresponds to
+%        this is an optional input, but if it is supplied it supercedes values from the matrix
+% participant: variable storing participant/subject IDs (i.e. the random factor) for repeated measures analysis
+%        this is an optional input, but if it is supplied it supercedes values from the matrix
+%
+% the function first runs the condition index test for each level of the IV
+% if the condition index test is non-significant for all levels, a T-squared-circ test is run for one- and two-sample designs, or an ANOVA-squared-circ test otherwise
+% if the condition index test is significant for any level, a T-squared test or MANOVA is run instead
+% if participant labels are included, a repeated measures test is conducted
+% the Mahalanobis distance effect size statistic (D) is also calculated
+
+grouplabels = [];
+participantlabels = [];
+
+s = size(data);
+
+if (~isreal(data(1)))
+    datavals = data;
+else
+    datavals = complex(data(:,1),data(:,2));
+    
+    if (s(2)>2)
+        grouplabels = data(:,3);
+    end
+    if (s(2)>3)
+        participantlabels = data(:,4);
+    end
+end
+
+if (~isempty(group))
+    grouplabels = group;
+end
+if (~isempty(participant))
+    participantlabels = participant;
+end
+
+factorlist = unique(grouplabels);
+
+if (isempty(grouplabels))
+    grouplabels = ones(size(complexdata));
+end
+isRM = 1;
+if (isempty(participantlabels))
+    isRM = 0;
+end
+
+factorlist = unique(grouplabels);
+ngroups = length(factorlist);
+nobservations = length(complexdata);
+
+disp(strcat('Design has ',num2str(ngroups),' levels, with ',numstr(nobservations/ngroups),' observations per level'));
+if (isRM==1)
+    disp('Design is repeated measures');
+end
+
+
[…]

### 2021-01-07 FourierStatsMatlab/tsqh_test.m

@@ -1,0 +1,103 @@
+function output = tsqh_test(x,y,paired,mu)
+
+% tsqh.test: function to calculate a different variants of the T-squared test of Hotelling (1931)
+% the inputs must be Nx2 or 2xN matrices of numbers
+% for a one-sample test, mu is an optional vector to which the data are compared
+% for a two-sample test, x and y are the data from the two conditions
+% if paired=TRUE, the matrices (x and y) must be the same size
+
+d = size(x);
+if (d(1)<d(2))
+    x = x';
+end
+d = size(y);
+if (d(1)<d(2))
+    y = y';
+end
+
+if (~isempty(mu))
+    for n = 1:length(mu)
+        x(:,n) = x(:,n) - mu(n);
+    end
+end
+
+if ~isreal(x)
+    x = [real(x); imag(x)]';
+end
+if (~isempty(y))
+    if ~isreal(y)
+        y = [real(y); imag(y)]';
+    end
+end
+
+if (isempty(y))
+    method = 'One-sample T-squared test';
+end
+if (~isempty(y) && paired==1)
+    method = 'Paired samples T-squared test';
+end
+if (~isempty(y) && paired==0)
+    method = 'Independent samples T-squared test';
+end
+
+if (paired==1)
+    if (~isempty(y))
+        x = x - y;
+    end
+end
+
+
+if (paired==1 || isempty(y))
+    % paired or one-sample version of the test
+
+    xbar = mean(x);
+    s = size(x);
+    m = s(2);
+    N = s(1);
+    C = cov(x);
+    Cinv = inv(C);
+    tsq = N * xbar * Cinv * xbar';
+    Fratio = tsq .* (N - m)./(m.*(N-1));
+    df1 = m;
+    df2 = N-m;
+    pval = fcdf(Fratio,df1,df2,'upper');
+ 
+else
+    % independent samples version of the test
+    
+    xbarA = mean(x);
+    s = size(x);
+    m = s(2);
+    N1 = s(1);
+    
+    xbarB = mean(y);
+    s = size(y);
+    N2 = s(1); 
+    
+    CA = cov(x);
+    CB = cov(y);
+    
+    meandiff = xbarA - xbarB;
+    
+    C = ((N1-1)*CA + (N2-1)*CB)/(N1+N2-m);
+    Cinv = inv(C);
+    
[…]

### 2021-01-07 R/FourierStats-package.r

@@ -3,10 +3,9 @@
 #' Functions include:
 #'  - analysecplx: a function to analyse complex Fourier data, following the guidelines from the Baker (2021) paper
 #'  - tsqc.test: implementation of the T-squared-circ test of Victor & Mast (1991)
-#'  - tsq1.test: one-sample Hotelling's T-squared test
+#'  - tsqh.test: Hotelling's T-squared test (one and two sample versions)
 #'  - CI.test: condition index test, to test the assumptions of the T-squared-circ test
-#'  - anovacirc.test: implementation of one-way between subjects ANOVA-squared-circ test
-#'  - anovacircR.test: implementation of one-way repeated measures ANOVA-squared-circ test
+#'  - anovacirc.test: implementation of one-way between subjects and repeated measures ANOVA-squared-circ test
 #'  - amperrors: function to calculate error bars for amplitudes incorporating coherent averaging
 #'  - getel: helper function that calculates the bounding ellipse for a cloud of points
 #'  - fftshift: helper function that performs the quadrant shift of a 2D Fourier spectrum

### 2021-01-07 R/analysecplx.R

@@ -95,7 +95,7 @@ if (assumptionsmet==0 && ngroups<3){
   if (ngroups==1){
     message('Running a one-sample T-squared test')
     output$testtype <- 'One-sample T-squared test'
-    results <- tsq1.test(complexdata)
+    results <- tsqh.test(complexdata)
     message(paste('The test statistic T^2 =',round(results$tsq,digits=2)))
     message(paste('The equivalent F-ratio with',results$df1,'and',results$df2,'degrees of freedom is F =',round(results$Fratio,digits=2)))
     output$teststat <- results$tsq
@@ -110,19 +110,19 @@ if (assumptionsmet==0 && ngroups<3){
     if (isRM==0){
       message('Running an independent T-squared test')
       output$testtype <- 'Independent T-squared test'
-      results <- Hotelling::hotelling.test(x=data.frame(Re(dataA),Im(dataA)),y=data.frame(Re(dataB),Im(dataB)))
-      message(paste('The test statistic T^2 =',round(results$stats$statistic,digits=2)))
-      output$teststat <- results$stats$statistic
-      output$Fratio <- results$stats$statistic * results$stats$m
-      output$df1 <- results$stats$df[1]
-      output$df2 <- results$stats$df[2]
-      output$p.value <- results$pval
-      message(paste('The equivalent F-ratio with',output$df1,'and',output$df2,'degrees of freedom is F =',round(output$Fratio,digits=2)))
+      results <- tsqh.test(dataA,dataB)
+      output$teststat <- results$tsq
+      output$Fratio <- results$Fratio
+      output$df1 <- results$df1
+      output$df2 <- results$df2
+      output$p.value <- results$p.value
+      message(paste('The test statistic T^2 =',round(results$tsq,digits=2)))
+      message(paste('The equivalent F-ratio with',results$df1,'and',results$df2,'degrees of freedom is F =',round(results$Fratio,digits=2)))
     }
     if (isRM==1){
       message('Running a repeated measures T-squared test')
       output$testtype <- 'Repeated measures T-squared test'
-      results <- tsq1.test(dataA-dataB)   # take the difference between groups and run a 1 sample test
+      results <- tsqh.test(dataA-dataB)   # take the difference between groups and run a 1 sample test
       message(paste('The test statistic T^2 =',round(results$tsq,digits=2)))
[…]

### 2021-01-07 R/tsq1.test.R

@@ -1,29 +1,0 @@
-#' tsq1.test: function to calculate a one-sample T-squared test of Hotelling (1931)
-#' the input is an N (observations) by m (dependent variables) matrix
-#' optional input mu is a vector with m entries, to which the data are compared (otherwise compared to 0)
-#' a repeated measures version can be calculated by subtracting the two conditions and then running the one-sample test
-#' for two-sample tests, use the hotelling.test function from the Hotelling package
-#' @export
-tsq1.test <- function(data,mu=NULL){
-
-  # if data are complex, convert to real and absolute values
-  if (is.complex(data)){data <- data.frame(Re(data),Im(data))}
-  if (is.complex(mu)){mu <- c(Re(mu),Im(mu))}
-  # compare each DV to some value
-  if (!is.null(mu)){for (m in 1:length(mu)){data[,m] <- data[,m]-mu[m]}}
-
-  xbar <- colMeans(data) # estimate means for each DV
-  m <- ncol(data)   # record number of DVs
-  N <- nrow(data)   # record sample size
-  C <- cov(data)    # calculate covariance matrix
-  Cinv <- solve(C)  # calculate inverse matrix
-  tsq <- N*t(xbar)%*%Cinv%*%xbar # compute t2 statistic
-  Fratio <- tsq * (N-m)/(m*(N-1)) # convert to F-ratio
-  df1 <- m      # record degrees of freedom
-  df2 <- N-m
-  pval <- pf(Fratio,df1,df2,lower.tail=FALSE) # estimate p-value
-  p.value <- min(pval,1)
-
-  # store outputs in a data structure
-  output <- data.frame(tsq,Fratio,df1,df2,p.value)
-  return(output)}

### 2021-01-07 R/tsqh.test.R

@@ -1,0 +1,75 @@
+#' tsqh.test: function to calculate a different variants of the T-squared test of Hotelling (1931)
+#' the inputs must be Nx2 or 2xN matrices of numbers, or vectors of complex numbers
+#' for a one-sample test, mu is an optional vector to which the data are compared
+#' for a two-sample test, x and y are the data from the two conditions
+#' if paired=TRUE, the inputs (x and y) must be the same size
+#' @export
+tsqh.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
+
+  # transpose if dimensions are the wrong way round
+  if (!is.null(ncol(x))){d <- dim(x)
+  if (d[1]<d[2]){x <- t(x)}}
+  if (!is.null(y)){if (!is.null(ncol(y))){d <- dim(y)
+  if (d[1]<d[2]){y <- t(y)}}}
+
+  if (!is.null(mu)){for (n in 1:length(mu)){x[,n] <- x[,n] - mu[n]}}
+
+  # convert data to bivariate representation if complex
+  if (is.complex(x)){x <- data.frame(Re(x),Im(x))}
+  if (!is.null(y)){if (is.complex(y)){y <- data.frame(Re(y),Im(y))}}
+
+  if (is.null(y)){method <- 'One-sample T-squared test'}
+  if (!is.null(y) & paired==TRUE){method <- 'Paired T-squared test'}
+  if (!is.null(y) & paired==FALSE){method <- 'Independent samples T-squared test'}
+
+  if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}
+
+  # paired or one-sample version of the test
+  if (paired==TRUE | is.null(y)){
+
+    xbar <- colMeans(x) # estimate means for each DV
+    m <- ncol(x)   # record number of DVs
+    N <- nrow(x)   # record sample size
+    C <- cov(x)    # calculate covariance matrix
+    Cinv <- solve(C)  # calculate inverse matrix
+    tsq <- N*t(xbar)%*%Cinv%*%xbar # compute t2 statistic
+    Fratio <- tsq * (N-m)/(m*(N-1)) # convert to F-ratio
+    df1 <- m      # record degrees of freedom
+    df2 <- N-m
+    pval <- pf(Fratio,df1,df2,lower.tail=FALSE) # estimate p-value
+
+  }
+
+  # independent samples version of the test
+  if (paired==FALSE & !is.null(y)){
+
[…]

### 2021-01-07 FourierStatsMatlab/analysecplx.m
Implemented independent T-squared test. Replaced the tsq1.test function with tsqh.test (which can do both one and two sample tests).

@@ -1,0 +1,203 @@
+function output = analysecplx(data, group, participant)
+
+% analysecplx: wrapper function that automatically decides which statistical test to run for a given data set
+%
+% Inputs--
+% data:  this can either be a vector of complex numbers, or a matrix
+%        if it is a matrix, the first two columns are the x and y (real and imaginary) values of the DV
+%        if there are further columns, these are treated as the group labels and participant IDs
+% group: condition labels indicating the level of the independent variable that each data point corresponds to
+%        this is an optional input, but if it is supplied it supercedes values from the matrix
+% participant: variable storing participant/subject IDs (i.e. the random factor) for repeated measures analysis
+%        this is an optional input, but if it is supplied it supercedes values from the matrix
+%
+% the function first runs the condition index test for each level of the IV
+% if the condition index test is non-significant for all levels, a T-squared-circ test is run for one- and two-sample designs, or an ANOVA-squared-circ test otherwise
+% if the condition index test is significant for any level, a T-squared test or MANOVA is run instead
+% if participant labels are included, a repeated measures test is conducted
+% the Mahalanobis distance effect size statistic (D) is also calculated
+
+grouplabels = [];
+participantlabels = [];
+
+s = size(data);
+
+if (~isreal(data(1)))
+    datavals = data;
+else
+    datavals = complex(data(:,1),data(:,2));
+    
+    if (s(2)>2)
+        grouplabels = data(:,3);
+    end
+    if (s(2)>3)
+        participantlabels = data(:,4);
+    end
+end
+
+if (~isempty(group))
+    grouplabels = group;
+end
+if (~isempty(participant))
+    participantlabels = participant;
+end
+
+factorlist = unique(grouplabels);
+
+if (isempty(grouplabels))
+    grouplabels = ones(size(complexdata));
+end
+isRM = 1;
+if (isempty(participantlabels))
+    isRM = 0;
+end
+
+factorlist = unique(grouplabels);
+ngroups = length(factorlist);
+nobservations = length(complexdata);
+
+disp(strcat('Design has ',num2str(ngroups),' levels, with ',numstr(nobservations/ngroups),' observations per level'));
+if (isRM==1)
+    disp('Design is repeated measures');
+end
+
+
[…]

### 2021-01-07 FourierStatsMatlab/clustercorrect.m

@@ -77,7 +77,7 @@ for n = 1:m
                 allp(n) = p;
                 allt(n) = output.tstat;
             case 2  % one-sample Hotelling's t-squared
-                output = tsq1_test(datax(:,n));
+                output = tsqh_test(datax(:,n));
                 allp(n) = output.pval;
                 allt(n) = output.tsq;
             case 3  % one-sample T-squared-circ
@@ -95,7 +95,9 @@ for n = 1:m
                 allp(n) = p;
                 allt(n) = output.tstat;
             case 2  % two sample Hotelling's t-squared
-                
+                output = tsqh_test(datax(:,n),datay(:,n));
+                allp(n) = output.pval;
+                allt(n) = output.tsq;                
             case 3  % two-sample T-squared-circ
                 output = tsqc_test(datax(:,n),datay(:,n));
                 allp(n) = output.pval;
@@ -185,7 +187,7 @@ if (nclusters>0)
                             [h,p,ci,output] = ttest(tempdataA);
                             tsum = tsum + output.tstat;
                         case 2  % one-sample Hotelling's t-squared
-                            output = tsq1_test(tempdataA);
+                            output = tsqh_test(tempdataA);
                             tsum = tsum + output.tsq;
                         case 3  % one-sample T-squared-circ
                             output = tsqc_test(tempdataA);
@@ -202,7 +204,8 @@ if (nclusters>0)
                             [h,p,ci,output] = ttest2(tempdataA,tempdataB);
                             tsum = tsum + output.tstat;
                         case 2  % two sample Hotelling's t-squared
-                            
+                            output = tsqh_test(tempdataA,tempdataB);
+                            tsum = tsum + output.tsq;
                         case 3  % two-sample T-squared-circ
                             output = tsqc_test(tempdataA,tempdataB);
                             tsum = tsum + output.tsqc;

### 2021-01-07 FourierStatsMatlab/tsq1_test.m

@@ -1,46 +1,0 @@
-function output = tsq1_test(x,mu)
-
-% tsq1_test: function to calculate a one-sample Hotelling's T-squared statistic 
-% the input (x) must be Nx2 or 2xN matrices of numbers
-% or can be stored as a single vector of complex numbers
-% mu is an optional vector to which the data are compared
-% the output is a structure containing the test results
-
-d = size(x);
-if (d(1)<d(2))
-    x = x';
-end
-
-if (~isempty(mu))
-    for n = 1:length(mu)
-        x(:,n) = x(:,n) - mu(n);
-    end
-end
-
-if ~isreal(x)
-    x = [real(x); imag(x)]';
-end
-
-    method = 'One-sample Hotellings t-squared test';
-
-    xbar = mean(x);
-    s = size(x);
-    m = s(2);
-    N = s(1);
-    C = cov(x);
-    Cinv = inv(C);
-    tsq = N * xbar * Cinv * xbar';
-    Fratio = tsq .* (N - m)./(m.*(N-1));
-    df1 = m;
-    df2 = N-m;
-    pval = fcdf(Fratio,df1,df2,'upper');
-    pval = min(pval,1);
-    
-output.tsq = tsq;
-output.Fratio = Fratio;
-output.df1 = df1;
-output.df2 = df2;
-output.pval = pval;
-output.method = method;
-
-end

### 2021-01-07 R/clustercorrect.R

@@ -55,7 +55,7 @@ clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,pair
         allp[n] <- output$p.value
         allt[n] <- output$statistic}
       if (testtype==2){
-        output <- tsq1.test(datax[,n])  # one-sample Hotelling's t-squared
+        output <- tsqh.test(datax[,n])  # one-sample Hotelling's t-squared
         allp[n] <- output$p.value
         allt[n] <- output$tsq}
       if (testtype==3){
@@ -71,9 +71,9 @@ clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,pair
         allp[n] <- output$p.value
         allt[n] <- output$statistic}
       if (testtype==2){
-        output <- Hotelling::hotelling.test(data.frame(Re(datax[,n]),Im(datax[,n])),data.frame(Re(datay[,n]),Im(datay[,n])))  #independent Hotelling's T-squared
-        allp[n] <- output$pval
-        allt[n] <- output$stats$statistic}
+        output <- tsqh.test(datax[,n],datay[,n],paired=FALSE)  # independent T-squared
+        allp[n] <- output$p.value
+        allt[n] <- output$tsq}
       if (testtype==3){
         output <- tsqc.test(datax[,n],datay[,n],paired=FALSE)  # independent T-squared-circ
         allp[n] <- output$p.value
@@ -151,7 +151,7 @@ clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,pair
             output <- t.test(tempdataA)
             tsum <- tsum + output$statistic}
           if (testtype==2){
-            output <- tsq1.test(tempdataA)
+            output <- tsqh.test(tempdataA)
             tsum <- tsum + output$tsq}
           if (testtype==3){
             output <- tsqc.test(tempdataA)
@@ -166,8 +166,8 @@ clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,pair
             output <- t.test(tempdataA,tempdataB)
             tsum <- tsum + output$statistic}
           if (testtype==2){
-            output <- Hotelling::hotelling.test(data.frame(Re(tempdataA),Im(tempdataA)),data.frame(Re(tempdataB),Im(tempdataB)))
-            tsum <- tsum + output$stats$statistic}
+            output <- tsqh.test(tempdataA,tempdataB,paired=FALSE)
[…]

### 2021-01-07 R/tsqc.test.R

@@ -18,9 +18,9 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
   if (!is.complex(x)){x <- complex(real=x[,1],imaginary=x[,2])}
   if (!is.null(y)){if (!is.complex(y)){y <- complex(real=y[,1],imaginary=y[,2])}}
 
-  if (is.null(y)){method <- 'One-sample t-squared circ test'}
-  if (!is.null(y) & paired==TRUE){method <- 'Paired t-squared circ test'}
-  if (!is.null(y) & paired==FALSE){method <- 'Independent samples t-squared circ test'}
+  if (is.null(y)){method <- 'One-sample T-squared circ test'}
+  if (!is.null(y) & paired==TRUE){method <- 'Paired T-squared circ test'}
+  if (!is.null(y) & paired==FALSE){method <- 'Independent samples T-squared circ test'}
 
   if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}
 

### 2021-01-07 manuscript/misc/testanalysecplx.R

@@ -61,3 +61,17 @@ input <- data.frame(xy,group,participant)
 analysecplx(input)
 
 
+
+
+
+N <- 10
+
+x <- rnorm(N)
+y <- rnorm(N)
+xy1 <- complex(real=x,imaginary=y)
+x <- rnorm(N)
+y <- rnorm(N)
+xy2 <- complex(real=x,imaginary=y)
+
+tsqh.test(xy1,xy2)
+h <- Hotelling::hotelling.test(x=data.frame(Re(xy1),Im(xy1)),y=data.frame(Re(xy2),Im(xy2)))

### 2021-01-08 manuscript/manuscript.Rmd
Minor bug fixes

@@
     output <- tsqc.test(simdata,paired=FALSE)
-    if (output$p.value < 0.05){csig[cond] <- csig[cond] + 1}
+    if (output$pval < 0.05){csig[cond] <- csig[cond] + 1}
     output <- tsqh.test(simdata)
-    if (output$p.value < 0.05){hsig[cond] <- hsig[cond] + 1}
+    if (output$pval < 0.05){hsig[cond] <- hsig[cond] + 1}
   }}
@@
   output <- anovacirc.test(data.frame(simdata,grouplabels))
-  if (output$p.value<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
+  if (output$pval<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
 
@@
 There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
-Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = `r round(tresult$statistic,digits=2)`, _df_ = `r tresult$parameter`, _p_ = `r round(tresult$p.value,digits=2)`).
+Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = `r round(tresult$statistic,digits=2)`, _df_ = `r tresult$parameter`, _p_ = `r round(tresult$pval,digits=2)`).
 However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b).
 The condition index test was non-significant for both conditions (sound: CI = `r round(cisound$CI,digits=2)`, _p_ = `r round(cisound$pval,digits=2)`; light: CI = `r round(cilight$CI,digits=2)`, _p_ = `r round(cilight$pval,digits=2)`), so a paired-samples $T^2_{circ}$ test was conducted.
-This revealed a significant difference between conditions ($T^2_{circ}$ = `r round(tsqcresult$tsqc,digits=2)`, $F_{(`r tsqcresult$df1`,`r tsqcresult$df2`)}$ = `r round(tsqcresult$Fratio,digits=2)`, _p_ = `r round(tsqcresult$p.value,digits=3)`) with an effect size of _D_ = `r round(D,digits=2)`.
+This revealed a significant difference between conditions ($T^2_{circ}$ = `r round(tsqcresult$tsqc,digits=2)`, $F_{(`r tsqcresult$df1`,`r tsqcresult$df2`)}$ = `r round(tsqcresult$Fratio,digits=2)`, _p_ = `r round(tsqcresult$pval,digits=3)`) with an effect size of _D_ = `r round(D,digits=2)`.
 This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
@@
 With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = `r format(round(max(allCI),digits=2),nsmall=2)`, all _p_ > `r round(min(allCIp),digits=2)`).
-A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(`r withincirc$dfM`,`r withincirc$dfR`)}$ = `r round(withincirc$Fratio,digits=1)`, _p_ < `r max(round(withincirc$p.value,digits=3),0.001)`).
+A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(`r withincirc$dfM`,`r withincirc$dfR`)}$ = `r round(withincirc$Fratio,digits=1)`, _p_ < `r max(round(withincirc$pval,digits=3),0.001)`).
 Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[3],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[3],digits=2)`, _D_ = `r round(allD[3],digits=2)`, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[4],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[4],digits=2)`, _D_ = `r round(allD[4],digits=2)`, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[5],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[5],digits=2)`, _D_ = `r round(allD[5],digits=2)`, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[6],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[6],digits=2)`, _D_ = `r round(allD[6],digits=2)`, _p_ < 0.001).

### 2021-01-08 R/analysecplx.R

@@ -87,7 +87,7 @@ if (assumptionsmet==1 && ngroups<3){
   output$Fratio <- results$Fratio
   output$df1 <- results$df1
   output$df2 <- results$df2
-  output$p.value <- results$p.value
+  output$pval <- results$pval
 }
 
 # run T-squared
@@ -102,7 +102,7 @@ if (assumptionsmet==0 && ngroups<3){
     output$Fratio <- results$Fratio
     output$df1 <- results$df1
     output$df2 <- results$df2
-    output$p.value <- results$p.value
+    output$pval <- results$pval
     }
   if (ngroups>1){
     dataA <- complexdata[which(grouplabels==factorlist[1])]
@@ -115,7 +115,7 @@ if (assumptionsmet==0 && ngroups<3){
       output$Fratio <- results$Fratio
       output$df1 <- results$df1
       output$df2 <- results$df2
-      output$p.value <- results$p.value
+      output$pval <- results$pval
       message(paste('The test statistic T^2 =',round(results$tsq,digits=2)))
       message(paste('The equivalent F-ratio with',results$df1,'and',results$df2,'degrees of freedom is F =',round(results$Fratio,digits=2)))
     }
@@ -129,7 +129,7 @@ if (assumptionsmet==0 && ngroups<3){
       output$Fratio <- results$Fratio
       output$df1 <- results$df1
       output$df2 <- results$df2
-      output$p.value <- results$p.value
+      output$pval <- results$pval
       }
   }
 }
@@ -145,7 +145,7 @@ if (assumptionsmet==1 && ngroups>2){
     output$Fratio <- results$Fratio
     output$df1 <- results$dfM
     output$df2 <- results$dfR
-    output$p.value <- results$p.value
+    output$pval <- results$pval
 }
 
 # run MANOVA
@@ -165,13 +165,13 @@ if (assumptionsmet==0 && ngroups>2){
     results <- MANOVA.RM::multRM(cbind(dataforManova$Real, dataforManova$Imaginary) ~ Group, data=dataforManova, subject="Participant", within="Group")
     output$testtype <- 'Repeated measures MANOVA'
     }
-  output$teststat <- results$MATS
-  output$p.value <- results$resampling[2]
+  output$teststat <- as.numeric(results$MATS)
+  output$pval <- results$resampling[2]
   message(paste('The Modified ANOVA-like Test Statistic is',round(results$MATS,digits=2)))
 }
 
[…]

### 2021-01-08 R/anovacirc.test.R

@@ -60,9 +60,9 @@ anovacirc.test <- function(data, group=NULL, participant=NULL){
   MSM <- SSM/dfM
   MSR <- SSR/dfR
   Fratio <- MSM/MSR
-  p.value <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+  pval <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
 
-  output <- data.frame(Fratio,p.value,SSM,SSR,dfM,dfR,MSM,MSR)
+  output <- data.frame(Fratio,pval,SSM,SSR,dfM,dfR,MSM,MSR)
 }
 
   # if participant labels have been supplied, run a repeated measures ANOVA-circ
@@ -86,9 +86,9 @@ anovacirc.test <- function(data, group=NULL, participant=NULL){
   MSM <- SSM/dfM
   MSR <- SSR/dfR
   Fratio <- MSM/MSR
-  p.value <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
+  pval <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
 
-  output <- data.frame(Fratio,p.value,SSW,SSM,SSR,dfW,dfM,dfR,MSM,MSR)
+  output <- data.frame(Fratio,pval,SSW,SSM,SSR,dfW,dfM,dfR,MSM,MSR)
   }
 
   return(output)

### 2021-01-08 R/tsqh.test.R

@@ -67,9 +67,9 @@ tsqh.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
 
   }
 
-  p.value <- min(pval,1)
+  pval <- min(pval,1)
 
   # store outputs in a data structure
-  output <- data.frame(tsq,Fratio,df1,df2,p.value,method)
+  output <- data.frame(tsq,Fratio,df1,df2,pval,method)
 
   return(output)}

### 2021-01-08 R/clustercorrect.R
Minor bug fixes

@@ -56,11 +56,11 @@ clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,pair
         allt[n] <- output$statistic}
       if (testtype==2){
         output <- tsqh.test(datax[,n])  # one-sample Hotelling's t-squared
-        allp[n] <- output$p.value
+        allp[n] <- output$pval
         allt[n] <- output$tsq}
       if (testtype==3){
         output <- tsqc.test(datax[,n])  # T-squared-circ
-        allp[n] <- output$p.value
+        allp[n] <- output$pval
         allt[n] <- output$tsqc}
     }
 
@@ -72,11 +72,11 @@ clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,pair
         allt[n] <- output$statistic}
       if (testtype==2){
         output <- tsqh.test(datax[,n],datay[,n],paired=FALSE)  # independent T-squared
-        allp[n] <- output$p.value
+        allp[n] <- output$pval
         allt[n] <- output$tsq}
       if (testtype==3){
         output <- tsqc.test(datax[,n],datay[,n],paired=FALSE)  # independent T-squared-circ
-        allp[n] <- output$p.value
+        allp[n] <- output$pval
         allt[n] <- output$tsqc}
     }
   }
@@ -178,7 +178,7 @@ clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,pair
     }
 
     # compare each cluster to the null distribution, retain the significant ones
-    clusterps <- 0*(1:ccount) + 1
+    clusterps <- 0*(1:ccount)
     if (length(nulldist)==nresamples){
     for (cc in 1:ccount){
       i <- which(abs(nulldist)>abs(sumtvals[cc]))

### 2021-01-08 R/tsqc.test.R

@@ -62,7 +62,7 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
     pval <- pf(Fratio,df1=df1,df2=df2,lower.tail=FALSE)
   }
 
-  p.value <- min(pval,1)
+  pval <- min(pval,1)
 
-  output <- data.frame(tsqc,Fratio,df1,df2,p.value,method)
+  output <- data.frame(tsqc,Fratio,df1,df2,pval,method)
   return(output)}

### 2021-01-09 manuscript/manuscript.Rmd
Added pairwisemahal function. Other minor improvements and fixes.

@@
 
-packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','HDMD','devtools')
+packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','devtools')
 missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
@@
 
-nsims <- 100000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
+nsims <- 100  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
 
@@
 If the test is non-significant, one can proceed with $T^2_{circ}$; if it is significant, $T^2$ should be used instead.
-An _R_ function implementing this test is included in the _FourierStats_ package (the _CI.test_ function).
+A function implementing this test is included in the _FourierStats_ package (the _CI.test_ function in _R_, and _CI_test_ function in _Matlab_).
 
@@
 In all cases, the functions depart from the Type I error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the outlier's Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core _mahalanobis_ function in _R_, return $D^2$, which can be converted to _D_ by taking the square root).
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core _mahalanobis_ function in _R_ (or _mahal_ function in _Matlab_), return $D^2$, which can be converted to _D_ by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the $T^2_{circ}$ test.
@@
 @Giudice2009].
-This is a valuable statistic to include when reporting the results of multivariate tests, and an _R_ function (_pairwise.mahalanobis_) to calculate it is available as part of the [_HDMD_](https://CRAN.R-project.org/package=HDMD) package.
+This is a valuable statistic to include when reporting the results of multivariate tests, and a function (_pairwisemahal_) to calculate it is available as part of the _FourierStats_ package.
 
@@
 A repeated measures version of ${ANOVA}^2_{circ}$ can also be implemented following the same logic.
-The project repository contains an _R_ function (_anovacirc.test_) to run both of these tests for one-way designs.
+The project repository contains a function (_anovacirc.test_ in _R_, _anovacirc_test_ in _Matlab_) to run both of these tests for one-way designs.
 In principle factorial versions might also be derived.
@@
 fores <- rbind(totest,totest2)
-mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=6))
-D <- sqrt(mahala_sq$distance[1,2])
+mahalaD <- pairwisemahal(fores, rep(1:2,each=6))
+D <- mahalaD[1,2]
 
@@
 There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality.
-Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = `r round(tresult$statistic,digits=2)`, _df_ = `r tresult$parameter`, _p_ = `r round(tresult$pval,digits=2)`).
+Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = `r round(tresult$statistic,digits=2)`, _df_ = `r tresult$parameter`, _p_ = `r round(tresult$p.value,digits=2)`).
 However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b).
@@
 fores <- rbind(set1,set2)
-mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=89))
-allD[cond-1] <- sqrt(mahala_sq$distance[1,2])
+mahalaD <- pairwisemahal(fores, rep(1:2,each=89))
+allD[cond-1] <- mahalaD[1,2]
 
@@
 [-@Pei2017] suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat demanding, though an _R_ function is provided to calculate error bars using this method (available through the _amperrors_ function).
+This approach is somewhat demanding, though a function is provided to calculate error bars using this method (available through the _amperrors_ function).
 However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes.

### 2021-01-09 manuscript/manuscript.tex
Added pairwisemahal function. Other minor improvements and fixes.

@@
 \author{Daniel H. Baker}
-\date{2021-01-03}
+\date{2021-01-09}
 
@@
 All scripts used to generate this manuscript are available at: \url{https://github.com/bakerdh/FourierStats}.
-This includes an \emph{R} package called \emph{FourierStats}, featuring functions to implement a one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.
+This includes a \emph{Matlab} toolbox and an \emph{R} package called \emph{FourierStats}, featuring functions to implement Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.
 
@@
 If the test is non-significant, one can proceed with \(T^2_{circ}\); if it is significant, \(T^2\) should be used instead.
-An \emph{R} function implementing this test is included in the \emph{FourierStats} package (the \emph{CI.test} function).
+A function implementing this test is included in the \emph{FourierStats} package (the \emph{CI.test} function in \emph{R}, and \emph{CI\_test} function in \emph{Matlab}).
 
@@
 In all cases, the functions depart from the Type I error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the outlier's Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R}, return \(D^2\), which can be converted to \emph{D} by taking the square root).
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R} (or \emph{mahal} function in \emph{Matlab}), return \(D^2\), which can be converted to \emph{D} by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the \(T^2_{circ}\) test.
@@
 Del Giudice, 2009).
-This is a valuable statistic to include when reporting the results of multivariate tests, and an \emph{R} function (\emph{pairwise.mahalanobis}) to calculate it is available as part of the \href{https://CRAN.R-project.org/package=HDMD}{\emph{HDMD}} package.
+This is a valuable statistic to include when reporting the results of multivariate tests, and a function (\emph{pairwisemahal}) to calculate it is available as part of the \emph{FourierStats} package.
 
@@
 A repeated measures version of \({ANOVA}^2_{circ}\) can also be implemented following the same logic.
-The project repository contains an \emph{R} function (\emph{anovacirc.test}) to run both of these tests for one-way designs.
+The project repository contains a function (\emph{anovacirc.test} in \emph{R}, \emph{anovacirc\_test} in \emph{Matlab}) to run both of these tests for one-way designs.
 In principle factorial versions might also be derived.
@@
 The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted.
-This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of \emph{D} = 1.02.
+This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of \emph{D} = 2.14.
 This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
@@
 
-With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.20, all \emph{p} \textgreater{} 0.19).
+With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.20, all \emph{p} \textgreater{} 0.23).
 A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, \emph{p} \textless{} 0.001).
-Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, \emph{D} = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, \emph{D} = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, \emph{D} = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, \emph{D} = 0.33, \emph{p} \textless{} 0.001).
+Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, \emph{D} = 1.1, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, \emph{D} = 1.04, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, \emph{D} = 0.63, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, \emph{D} = 1.17, \emph{p} \textless{} 0.001).
 The study by Vilidaite et al.
@@
 (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat demanding, though an \emph{R} function is provided to calculate error bars using this method (available through the \emph{amperrors} function).
+This approach is somewhat demanding, though a function is provided to calculate error bars using this method (available through the \emph{amperrors} function).
 However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes.

### 2021-01-09 FourierStatsMatlab/CI_test.m

@@ -10,13 +10,19 @@ function output = CI_test(data,alpha)
 % the index is then compared for significance with an expected distribution function
 % significant tests (p < alpha) violate the assumptions of the T-squared-circ and ANOVA-squared-circ tests
 % see Baker (2021) for further details
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 if (isempty(alpha))
     alpha = 0.05;
 end
 
+d = size(data);
+if (d(1)<d(2))
+    data = data';
+end
+
 if (~isreal(data))
-    data = [real(data); imag(data)]';
+    data = [real(data) imag(data)];
 end
 
 C = cov(data);

### 2021-01-09 FourierStatsMatlab/amperrors.m

@@ -11,6 +11,7 @@ function output = amperrors(input,method,quantiles,nresamples)
 %  the optional 'nresamples' variable allows the number of iterations to be determined for the bootstrapping option (ignored for other methods)
 %  typical values are 95 (for 95% confidence intervals), and 68 (for standard errors)
 %  the function returns the mean amplitude and the upper and lower error bars
+%  this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 if isempty(quantiles)
     quantiles = 95;

### 2021-01-09 FourierStatsMatlab/analysecplx.m

@@ -16,6 +16,7 @@ function output = analysecplx(data, group, participant)
 % if the condition index test is significant for any level, a T-squared test or MANOVA is run instead
 % if participant labels are included, a repeated measures test is conducted
 % the Mahalanobis distance effect size statistic (D) is also calculated
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 grouplabels = [];
 participantlabels = [];
@@ -23,9 +24,9 @@ participantlabels = [];
 s = size(data);
 
 if (~isreal(data(1)))
-    datavals = data;
+    complexdata = data;
 else
-    datavals = complex(data(:,1),data(:,2));
+    complexdata = complex(data(:,1),data(:,2));
     
     if (s(2)>2)
         grouplabels = data(:,3);
@@ -56,7 +57,7 @@ factorlist = unique(grouplabels);
 ngroups = length(factorlist);
 nobservations = length(complexdata);
 
-disp(strcat('Design has ',num2str(ngroups),' levels, with ',numstr(nobservations/ngroups),' observations per level'));
+disp(strcat('Design has ',num2str(ngroups),' levels, with ',num2str(nobservations/ngroups),' observations per level'));
 if (isRM==1)
     disp('Design is repeated measures');
 end
@@ -66,8 +67,8 @@ end
 CIresults = zeros(ngroups,4);
 for n = 1:ngroups
     i = find(grouplabels==factorlist(n));
-    temp = CI_test(complexdata(i));
-    CIresults = [temp.CI temp.N temp.criticalCI temp.pval];
+    temp = CI_test(complexdata(i),[]);
+    CIresults(n,:) = [temp.CI temp.N temp.criticalCI temp.pval];
 end
 nsigCI = length(find(CIresults(:,4)<(0.05/ngroups)));
 disp(strcat('The condition index was significant for ',num2str(nsigCI),' out of ',num2str(ngroups),' levels'));
@@ -112,7 +113,7 @@ if (assumptionsmet==0 && ngroups<3)
     if (ngroups==1)
         disp('Running a one-sample T-squared test');
         output.testtype = 'One-sample T-squared test';
-        results = tsqh.test(complexdata);
+        results = tsqh_test(complexdata,[],[],[]);
         disp(strcat('The test statistic T^2 = ',num2str(results.tsq,2)));
[…]

### 2021-01-09 FourierStatsMatlab/anovacirc_test.m

@@ -14,11 +14,16 @@ function output = anovacirc_test(data,group,participant)
 %        this is an optional input, but if it is supplied it supercedes values from the matrix
 %
 % see Baker (2021) for further details
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 grouplabels = [];
 participantlabels = [];
 
 s = size(data);
+
+if (s(1)<s(2))
+    data = data';
+end
 
 if (~isreal(data(1)))
     datavals = data;
@@ -72,7 +77,7 @@ if (isempty(participantlabels))
     output.dfR = dfR;
     output.MSM = MSM;
     output.MSR = MSR;
-    output.method = 'Between-subjects ANOVA2-circ';
+    output.method = 'One-way between-subjects ANOVA^2-circ';
     
 else
     % participant labels have been supplied, run a repeated measures ANOVA-circ test
@@ -110,7 +115,7 @@ else
     output.dfR = dfR;
     output.MSM = MSM;
     output.MSR = MSR;
-    output.method = 'Repeated measures ANOVA2-circ';
+    output.method = 'One-way repeated measures ANOVA^2-circ';
    
 end
 

### 2021-01-09 FourierStatsMatlab/clustercorrect.m

@@ -14,6 +14,7 @@ function clustout = clustercorrect(datax,datay,adjacencymatrix,testtype,paired,c
 % this should be an m x m matrix (where m is the number of sensors), where values are zero for non-adjacent sensor pairs, and 1 for adjacent pairs
 %
 % outputs an object containing the indices of all significant clusters
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 if (isempty(paired))
     paired = 0;

### 2021-01-09 FourierStatsMatlab/getel.m

@@ -2,6 +2,7 @@ function output = getel(input)
 % getel: helper function that calculates an ellipse bounding some data points
 % the outline of the ellipse is returned, sampled at 200 points
 % the input should be an Nx2 matrix of observations or an N-length vector of complex values
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 compdata = input;
 if (~isreal(input))

### 2021-01-09 FourierStatsMatlab/pairwisemahal.m

@@ -1,0 +1,64 @@
+function output = pairwisemahal(data,grouping)
+
+% calculates the pairwise (group to group) Mahalanobis distance between group means
+%
+% inputs:
+% data - can be either a vector of N complex numbers, a matrix of Nxm complex numbers, or an Nx2 matrix of real numbers
+%        if a vector of complex numbers or matrix of real numbers, the 'grouping' variable is also required
+%        if a matrix of complex numbers is supplied, the columns are interpreted as the groups
+% grouping - a variable containing group membership information for the data
+%
+% output structure (each variable is a subfield):
+% D - an mxm matrix of Mahalanobis distances between the group means
+%     note that the Mahalanobis distance (D) is returned, not D^2
+% groups - a list of group names in the order corresponding to the rows and
+%          columns of the matrix D
+%
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
+
+d = size(data);
+if (d(1)<d(2))
+    data = data';
+end
+d = size(data);
+
+% convert data to an Nxm matrix of real numbers
+if (~isreal(data))
+    if (min(d)==1)  % it is a vector
+        realdata = [real(data) imag(data)];         % convert to real
+    else            % it is a matrix
+        ngroups = d(2);
+        realdata = zeros(d(1),2);
+        for n = 1:ngroups
+            realdata(((n-1)*d(1)+1):(n*d(1)),:) = [real(data(:,n)) imag(data(:,n))];
+            grouping(((n-1)*d(1)+1):(n*d(1))) = n;
+        end
+        
+    end
+else
+    realdata = data;
+end
+
+grouplabels = unique(grouping);
+m = length(grouplabels);
+
+for n = 1:m
+    centroids(n,:) = mean(realdata(find(grouping==grouplabels(n)),:));
+end
+
+for n1 = 1:m
+    for n2 = 1:m
+        CA = cov(realdata(find(grouping==grouplabels(n1)),:)); % covariance matrix for group A
+        CB = cov(realdata(find(grouping==grouplabels(n2)),:)); % covariance matrix for group B
+        N1 = length(find(grouping==grouplabels(n1)));          % sample size for group A
+        N2 = length(find(grouping==grouplabels(n2)));          % sample size for group B
+        pwcov = ((N1-1)*CA + (N2-1)*CB)/(N1+N2-2);            % pooled covariance matrix for the two groups
[…]

### 2021-01-09 FourierStatsMatlab/testpairwisemahal.m

@@ -1,0 +1,12 @@
+% test the pairwisemahal function using simulated data with known effect sizes
+
+
+N = 10000;
+
+dataA = 1*randn(N,2);
+dataB = 1*randn(N,2);
+dataB(:,1) = dataB(:,1) + 1;
+grouplabels = ceil((1:(N*2))./N);
+data = [dataA; dataB];
+output = pairwisemahal(data,grouplabels);
+output.D

### 2021-01-09 FourierStatsMatlab/testscript.m

@@ -1,0 +1,113 @@
+
+load humanSSVEPdata.mat;
+
+subjmeans = squeeze(alltarget(:,31,:));
+
+% for n = 1:7
+%     temp = tsqc_test(subjmeans(:,n),[],[],[]);
+%     alloutput(n) = temp.tsqc;
+% end
+% temp
+% alloutput
+% 
+% 
+% for n = 1:7
+%     temp = tsqc_test(subjmeans(:,1),subjmeans(:,n),0,[]);
+%     alloutput(n) = temp.tsqc;
+% end
+% temp
+% alloutput
+% 
+% 
+% for n = 1:7
+%     temp = tsqc_test(subjmeans(:,1),subjmeans(:,n),1,[]);
+%     alloutput(n) = temp.tsqc;
+% end
+% temp
+% alloutput
+% 
+% 
+% 
+% for n = 1:7
+%     temp = tsqh_test(subjmeans(:,n),[],[],[]);
+%     alloutput(n) = temp.tsq;
+% end
+% temp
+% alloutput
+% 
+% 
+% for n = 1:7
+%     temp = tsqh_test(subjmeans(:,1),subjmeans(:,n),0,[]);
+%     alloutput(n) = temp.tsq;
+% end
+% temp
+% alloutput
+% 
+% 
+% for n = 2:7
+%     temp = tsqh_test(subjmeans(:,1),subjmeans(:,n),1,[]);
+%     alloutput(n) = temp.tsq;
+% end
+% temp
+% alloutput
+% 
+% 
+% 
+% 
+% for n = 1:7
+%     xydata = [real(subjmeans(:,n)) imag(subjmeans(:,n))]';
+%     temp = tsqc_test(xydata,[],[],[]);
+%     alloutput(n) = temp.tsqc;
+% end
+% temp
+% alloutput
+% 
+% 
+% for n = 1:7
+%     xydata = [real(subjmeans(:,n)) imag(subjmeans(:,n))]';
+%     temp = tsqh_test(xydata,[],[],[]);
+%     alloutput(n) = temp.tsq;
+% end
+% temp
+% alloutput
+
+% for n = 1:7
+%     temp = CI_test(subjmeans(:,n),0.05);
+%     alloutput(n) = temp.CI;
+% end
+% temp
+% alloutput
+
+
+% 
+% for n = 1:7
+%     xydata = [real(subjmeans(:,n)) imag(subjmeans(:,n))]';
[…]

### 2021-01-09 FourierStatsMatlab/tsqc_test.m

@@ -6,20 +6,27 @@ function output = tsqc_test(x,y,paired,mu)
 % for a two-sample test, x and y are the data from the two conditions
 % if paired=1, the matrices (x and y) must be the same size
 % the output is a structure containing the test results
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 d = size(x);
 if (d(1)<d(2))
     x = x';
 end
+if (~isempty(y))
 d = size(y);
 if (d(1)<d(2))
     y = y';
+end
 end
 
 if (~isempty(mu))
     for n = 1:length(mu)
         x(:,n) = x(:,n) - mu(n);
     end
+end
+
+if (isempty(paired))
+    paired = 0;
 end
 
 if isreal(x)
@@ -32,13 +39,13 @@ if (~isempty(y))
 end
 
 if (isempty(y))
-    method = 'One-sample t-squared circ test';
+    method = 'One-sample T-squared circ test';
 end
 if (~isempty(y) && paired==1)
-    method = 'Paired t-squared circ test';
+    method = 'Paired samples T-squared circ test';
 end
 if (~isempty(y) && paired==0)
-    method = 'Independent samples t-squared circ test';
+    method = 'Independent samples T-squared circ test';
 end
 
 if (paired==1)

### 2021-01-09 FourierStatsMatlab/tsqh_test.m

@@ -5,6 +5,7 @@ function output = tsqh_test(x,y,paired,mu)
 % for a one-sample test, mu is an optional vector to which the data are compared
 % for a two-sample test, x and y are the data from the two conditions
 % if paired=TRUE, the matrices (x and y) must be the same size
+% this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
 d = size(x);
 if (d(1)<d(2))
@@ -21,12 +22,16 @@ if (~isempty(mu))
     end
 end
 
+if (isempty(paired))
+    paired = 0;
+end
+
 if ~isreal(x)
-    x = [real(x); imag(x)]';
+    x = [real(x) imag(x)];
 end
 if (~isempty(y))
     if ~isreal(y)
-        y = [real(y); imag(y)]';
+        y = [real(y) imag(y)];
     end
 end
 
@@ -49,7 +54,7 @@ end
 
 if (paired==1 || isempty(y))
     % paired or one-sample version of the test
-
+    
     xbar = mean(x);
     s = size(x);
     m = s(2);
@@ -93,7 +98,7 @@ end
 
 pval = min(pval,1);
 
-output.tsqc = tsq;
+output.tsq = tsq;
 output.Fratio = Fratio;
 output.df1 = df1;
 output.df2 = df2;

### 2021-01-09 R/CI.test.R

@@ -9,6 +9,9 @@
 #' @export
 CI.test <- function(data, alpha=0.05){
 
+  if (!is.null(ncol(data))){d <- dim(data)
+  if (d[1]<d[2]){data <- t(data)}}
+
   if (is.complex(data)){data <- data.frame(Re(data),Im(data))}
 
   C <- cov(data)  # covariance matrix

### 2021-01-09 R/analysecplx.R

@@ -153,7 +153,7 @@ if (assumptionsmet==0 && ngroups>2){
 
   if (isRM==0){
     message('Running an independent MANOVA')
-    output$testtype <- 'Independent MANOVA'
+    output$testtype <- 'Independent MANOVA (test statistic is modified ANOVA-like statistic)'
     dataforManova <- data.frame(grouplabels,Re(complexdata),Im(complexdata))
     colnames(dataforManova) <- c("Group","Real","Imaginary")
     results <- MANOVA.RM::MANOVA.wide(cbind(dataforManova$Real, dataforManova$Imaginary) ~ Group, data = dataforManova)
@@ -163,7 +163,7 @@ if (assumptionsmet==0 && ngroups>2){
     dataforManova <- data.frame(grouplabels,participantlabels,Re(complexdata),Im(complexdata))
     colnames(dataforManova) <- c("Group","Participant","Real","Imaginary")
     results <- MANOVA.RM::multRM(cbind(dataforManova$Real, dataforManova$Imaginary) ~ Group, data=dataforManova, subject="Participant", within="Group")
-    output$testtype <- 'Repeated measures MANOVA'
+    output$testtype <- 'Repeated measures MANOVA (test statistic is modified ANOVA-like statistic)'
     }
   output$teststat <- as.numeric(results$MATS)
   output$pval <- results$resampling[2]
@@ -180,8 +180,8 @@ if (ngroups==1){
 }
 if (ngroups > 1){
   # calculate pairwise Mahalanobis distance between each pair of conditions
-  mahala_sq <- HDMD::pairwise.mahalanobis(data.frame(Re(complexdata),Im(complexdata)),grouplabels)
-  D <- max(sqrt(mahala_sq$distance))  # choose the largest effect size
+  pmahal <- pairwisemahal(data.frame(Re(complexdata),Im(complexdata)),grouplabels)
+  D <- max(pmahal)  # choose the largest effect size
 }
 message(paste('The effect size (Mahalanobis distance) is D =',round(D,digits=2)))
 

### 2021-01-09 R/anovacirc.test.R

@@ -21,6 +21,9 @@ anovacirc.test <- function(data, group=NULL, participant=NULL){
   # if the input data is a vector, it must be complex Fourier components
   if (is.null(ncol(data))){datavals <- data}
 
+  if (!is.null(ncol(data))){d <- dim(data)
+  if (d[1]<d[2]){data <- t(data)}}
+
   # if the input is not a vector, then work out what it contains based on its dimensions
   if (!is.null(ncol(data))){d <- dim(data)
 
@@ -62,7 +65,8 @@ anovacirc.test <- function(data, group=NULL, participant=NULL){
   Fratio <- MSM/MSR
   pval <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
 
-  output <- data.frame(Fratio,pval,SSM,SSR,dfM,dfR,MSM,MSR)
+  method <- 'One-way between-subjects ANOVA^2-circ'
+  output <- data.frame(Fratio,pval,SSM,SSR,dfM,dfR,MSM,MSR,method)
 }
 
   # if participant labels have been supplied, run a repeated measures ANOVA-circ
@@ -88,7 +92,8 @@ anovacirc.test <- function(data, group=NULL, participant=NULL){
   Fratio <- MSM/MSR
   pval <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)
 
-  output <- data.frame(Fratio,pval,SSW,SSM,SSR,dfW,dfM,dfR,MSM,MSR)
+  method <- 'One-way repeated measures ANOVA^2-circ'
+  output <- data.frame(Fratio,pval,SSW,SSM,SSR,dfW,dfM,dfR,MSM,MSR,method)
   }
 
   return(output)

### 2021-01-09 R/pairwisemahal.R

@@ -1,0 +1,62 @@
+#' pairwisemahal: calculates the pairwise (group to group) Mahalanobis distance between group means
+#'
+#' inputs:
+#' data - can be either a vector of N complex numbers, a matrix of Nxm complex numbers, or an Nx2 matrix of real numbers
+#'        if a vector of complex numbers or matrix of real numbers, the 'grouping' variable is also required
+#'        if a matrix of complex numbers is supplied, the columns are interpreted as the groups
+#' grouping - a variable containing group membership information for the data
+#'
+#' output:
+#' an mxm data frame of Mahalanobis distances between the group means
+#'     note that the Mahalanobis distance (D) is returned, not D^2
+#' the row and column names of the data frame are the condition names from the grouping input variable (or are otherwise numeric)
+#'
+#' this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
+
+#' @export
+pairwisemahal <- function(data, grouping=NULL){
+
+  d <- dim(data)
+  if (!is.null(ncol(data))){if (d[1]<d[2]){data <- t(data)}}
+  d <- dim(data)
+  if (is.complex(data)){
+    if (min(d)==1){       # it is a vector
+      realdata <- data.frame(Re(data),Im(data))
+    }
+    if (min(d)>1){        # it is a matrix
+      ngroups <- d(2)
+      realdata <- matrix(0,nrow=d(1),ncol=2)
+      grouping <- NULL
+      for (n in 1:ngroups){
+        realdata[((n-1)*d(1)+1):(n*d(1)),1] <- Re(data)
+        realdata[((n-1)*d(1)+1):(n*d(1)),2] <- Im(data)
+        grouping[((n-1)*d(1)+1):(n*d(1))] <- n
+      }
+    }
+  }
+  if (!is.complex(data)){realdata = data}
+
+  grouplabels <- as.factor(grouping)
+  factorlist <- levels(grouplabels)
+  m = length(factorlist)
+
+  centroids <- matrix(0,nrow=m,ncol=2)
+  for (n in 1:m){centroids[n,] <- colMeans(realdata[which(grouplabels==factorlist[n]),])}
+
+  distances <- matrix(0,nrow=m,ncol=m)
+  for (n1 in 1:m){
+    for (n2 in 1:m){
+      CA <- cov(realdata[which(grouplabels==factorlist[n1]),])
[…]

### 2021-01-09 R/tsqc.test.R

@@ -19,7 +19,7 @@ tsqc.test <- function(x,y=NULL,paired=FALSE,mu=NULL){
   if (!is.null(y)){if (!is.complex(y)){y <- complex(real=y[,1],imaginary=y[,2])}}
 
   if (is.null(y)){method <- 'One-sample T-squared circ test'}
-  if (!is.null(y) & paired==TRUE){method <- 'Paired T-squared circ test'}
+  if (!is.null(y) & paired==TRUE){method <- 'Paired samples T-squared circ test'}
   if (!is.null(y) & paired==FALSE){method <- 'Independent samples T-squared circ test'}
 
   if (paired==TRUE){if (is.null(y)==FALSE){x <- x - y}}

### 2021-01-10 FourierStatsMatlab/amperrors.m
Applied 'varargin' to Matlab functions. Improved efficiency of clustercorrect function.

@@ -1,4 +1,4 @@
-function output = amperrors(input,method,quantiles,nresamples)
+function output = amperrors(input,varargin)
 
 % amperrors: calculate error bars on the amplitude component of coherently averaged data
 % the input is a vector of complex numbers, or an Nx2 matrix of real and imaginary components
@@ -13,14 +13,17 @@ function output = amperrors(input,method,quantiles,nresamples)
 %  the function returns the mean amplitude and the upper and lower error bars
 %  this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
-if isempty(quantiles)
-    quantiles = 95;
-end
-if isempty(nresamples)
-    nresamples = 10000;
-end
-if isempty(method)
-    method = 'b';
+method = 'b';
+quantiles = 95;
+nresamples = 10000;
+if nargin>1
+    method = varargin{1};
+    if nargin>2
+        quantiles = varargin{2};
+        if nargin>3
+            nresamples = varargin{3};
+        end
+    end
 end
 
 s = size(input);

### 2021-01-10 FourierStatsMatlab/analysecplx.m

@@ -1,4 +1,4 @@
-function output = analysecplx(data, group, participant)
+function output = analysecplx(data, varargin)
 
 % analysecplx: wrapper function that automatically decides which statistical test to run for a given data set
 %
@@ -17,6 +17,16 @@ function output = analysecplx(data, group, participant)
 % if participant labels are included, a repeated measures test is conducted
 % the Mahalanobis distance effect size statistic (D) is also calculated
 % this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
+
+group = [];
+participant = [];
+if nargin>1
+    group = varargin{1};
+    if nargin>2
+        participant = varargin{2};
+    end
+end
+
 
 grouplabels = [];
 participantlabels = [];

### 2021-01-10 FourierStatsMatlab/anovacirc_test.m

@@ -1,4 +1,4 @@
-function output = anovacirc_test(data,group,participant)
+function output = anovacirc_test(data,varargin)
 
 % anovacirc.test: two-dimensional analysis of variance using complex data
 % an extension of the logic of the T-squared circ statistic of Victor & Mast (1991)
@@ -15,6 +15,15 @@ function output = anovacirc_test(data,group,participant)
 %
 % see Baker (2021) for further details
 % this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
+
+group = [];
+participant = [];
+if nargin>1
+    group = varargin{1};
+    if nargin>2
+        participant = varargin{2};
+    end
+end
 
 grouplabels = [];
 participantlabels = [];

### 2021-01-10 FourierStatsMatlab/testpairwisemahal.m

@@ -8,5 +8,12 @@ dataB = 1*randn(N,2);
 dataB(:,1) = dataB(:,1) + 1;
 grouplabels = ceil((1:(N*2))./N);
 data = [dataA; dataB];
+
 output = pairwisemahal(data,grouplabels);
 output.D
+
+compdata(:,1) = complex(dataA(:,1),dataA(:,2));
+compdata(:,2) = complex(dataB(:,1),dataB(:,2));
+output = pairwisemahal(compdata);
+output.D
+

### 2021-01-10 FourierStatsMatlab/testscript.m

@@ -1,7 +1,7 @@
 
 load humanSSVEPdata.mat;
 
-subjmeans = squeeze(alltarget(:,31,:));
+% subjmeans = squeeze(alltarget(:,31,:));
 
 % for n = 1:7
 %     temp = tsqc_test(subjmeans(:,n),[],[],[]);
@@ -89,25 +89,61 @@ subjmeans = squeeze(alltarget(:,31,:));
 % alloutput
 
 
-data = subjmeans(:);
-participantlabels = repmat(1:100,[7 1])';
-participantlabels = participantlabels(:);
-grouplabels = repmat(1:7,[100 1]);
-grouplabels = grouplabels(:);
+% data = subjmeans(:);
+% participantlabels = repmat(1:100,[7 1])';
+% participantlabels = participantlabels(:);
+% grouplabels = repmat(1:7,[100 1]);
+% grouplabels = grouplabels(:);
 
 % anovacirc_test(data,grouplabels,participantlabels)
 % anovacirc_test(data,grouplabels,[])
 
-mergeddata = [real(data),imag(data),grouplabels,participantlabels];
+% mergeddata = [real(data),imag(data),grouplabels,participantlabels];
 % anovacirc_test(mergeddata,[],[])
-mergeddata = [real(data),imag(data),grouplabels];
+% mergeddata = [real(data),imag(data),grouplabels];
 % anovacirc_test(mergeddata,[],[])
 
 % analysecplx(subjmeans(:,1),[],[])
 
-analysecplx(mergeddata,[],[])
+% analysecplx(mergeddata,[],[])
 
 % output = pairwisemahal(data,grouplabels)
 % output.groups
 % output.D
+
+        threshdist = 0.18;          % arbitrary distance between electrodes that seems to work well in pairing up neighbouring electrodes
+        
+        for m = 1:64                % loop through all electrodes
+            for n = 1:64            % calculate absolute distance from target electrode to each other electrode
+                xy1 = montage.electrodelocs(montage.channelmappings(m),:);
+                xy2 = montage.electrodelocs(montage.channelmappings(n),:);
+                electrodedistances(m,n) = sqrt((xy1(1)-xy2(1)).^2 + (xy1(2)-xy2(2)).^2);
+            end
+        end
+        
+        adjacencymatrix = zeros(size(electrodedistances));      
[…]

### 2021-01-10 FourierStatsMatlab/tsqc_test.m

@@ -1,4 +1,4 @@
-function output = tsqc_test(x,y,paired,mu)
+function output = tsqc_test(x,varargin)
 
 % tsqc_test: function to calculate the T-squared circ statistic from Victor & Mast (1991)
 % the inputs must be Nx2 or 2xN matrices of numbers
@@ -8,10 +8,24 @@ function output = tsqc_test(x,y,paired,mu)
 % the output is a structure containing the test results
 % this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
 
+y = [];
+paired = 0;
+mu = [];
+if nargin>1
+    y = varargin{1};
+    if nargin>2
+        paired = varargin{2};
+        if nargin>3
+            mu = varargin{3};
+        end
+    end
+end
+
 d = size(x);
 if (d(1)<d(2))
     x = x';
 end
+
 if (~isempty(y))
 d = size(y);
 if (d(1)<d(2))

### 2021-01-10 FourierStatsMatlab/tsqh_test.m

@@ -1,4 +1,4 @@
-function output = tsqh_test(x,y,paired,mu)
+function output = tsqh_test(x,varargin)
 
 % tsqh.test: function to calculate a different variants of the T-squared test of Hotelling (1931)
 % the inputs must be Nx2 or 2xN matrices of numbers
@@ -6,6 +6,19 @@ function output = tsqh_test(x,y,paired,mu)
 % for a two-sample test, x and y are the data from the two conditions
 % if paired=TRUE, the matrices (x and y) must be the same size
 % this function is part of the FourierStats package: https://github.com/bakerdh/FourierStats
+
+y = [];
+paired = 0;
+mu = [];
+if nargin>1
+    y = varargin{1};
+    if nargin>2
+        paired = varargin{2};
+        if nargin>3
+            mu = varargin{3};
+        end
+    end
+end
 
 d = size(x);
 if (d(1)<d(2))

### 2021-01-12 manuscript/manuscript.Rmd
Preparing for first public release

@@
 
-nsims <- 100  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
+nsims <- 100000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
 
@@
 
-  polygon(0.4*ellRot2[1, ]+xlocs[m], 0.1*ellRot2[2, ]+1.5, col=addalpha(nbdtpal[2],0.3),border=NA)
+  polygon(0.4*ellRot2[1, ]+xlocs[m], 0.1*ellRot2[2, ]+1.5, col=addalpha(nbdtpal[2],0.3), border=NA)
   lines(0.4*ellRot[1, ]+xlocs[m], 0.1*ellRot[2, ]+1.5, lwd=1)
@@
 
+# this version avoids calling the condition index test to speed things up
+# it means that the probability density function is not calculated on each iteration
+cilist <- seq(1,100,0.001)    # list of possible CIs
+alpha <- 0.05
 nsig <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
 for (s in 1:length(samplesizes)){
+
+  N <- samplesizes[s]
+  # calculate probability density function using modified Edelman equation
+  pdffunction <- ((N-2)*(2^(N-2))) *
+    ((cilist^2 - 1)/((cilist^2+1)^(N-1))) * (cilist^(N-3))
+  cdfinverse <- 1-(cumsum(pdffunction)/sum(pdffunction)) # inverse of cdf
+  criticalCI <- cilist[min(which(cdfinverse<=alpha))]  # find the threshold CI
+
   for (d in 1:length(dlist)){
@@
       data[1,] <- c(dlist[d],0)
-    output <- CI.test(data)
-    if(output$pval<0.05){nsig[s,d] <- nsig[s,d] + 1}
+
+      C <- cov(data)  # covariance matrix
+      N <- nrow(data) # sample size
+      eigVal  <- eigen(C)$values  # calculate eigenvalues
+      CI <- sqrt(eigVal[1]/eigVal[2]) # calculate condition index
+
+      if(CI>criticalCI){nsig[s,d] <- nsig[s,d] + 1}
     }
@@
 
-for (s in 1:length(samplesizes)){
-  lines(dlist,propsig[s,],lwd=3,col=nbdtpal[s])
-}
+for (s in 1:length(samplesizes)){lines(dlist,propsig[s,],lwd=3,col=nbdtpal[s])}
 

### 2021-01-12 FourierStatsMatlab/testpairwisemahal.m

@@ -1,19 +1,0 @@
-% test the pairwisemahal function using simulated data with known effect sizes
-
-
-N = 10000;
-
-dataA = 1*randn(N,2);
-dataB = 1*randn(N,2);
-dataB(:,1) = dataB(:,1) + 1;
-grouplabels = ceil((1:(N*2))./N);
-data = [dataA; dataB];
-
-output = pairwisemahal(data,grouplabels);
-output.D
-
-compdata(:,1) = complex(dataA(:,1),dataA(:,2));
-compdata(:,2) = complex(dataB(:,1),dataB(:,2));
-output = pairwisemahal(compdata);
-output.D
-

### 2021-01-12 FourierStatsMatlab/testscript.m

@@ -1,149 +1,0 @@
-
-load humanSSVEPdata.mat;
-
-% subjmeans = squeeze(alltarget(:,31,:));
-
-% for n = 1:7
-%     temp = tsqc_test(subjmeans(:,n),[],[],[]);
-%     alloutput(n) = temp.tsqc;
-% end
-% temp
-% alloutput
-% 
-% 
-% for n = 1:7
-%     temp = tsqc_test(subjmeans(:,1),subjmeans(:,n),0,[]);
-%     alloutput(n) = temp.tsqc;
-% end
-% temp
-% alloutput
-% 
-% 
-% for n = 1:7
-%     temp = tsqc_test(subjmeans(:,1),subjmeans(:,n),1,[]);
-%     alloutput(n) = temp.tsqc;
-% end
-% temp
-% alloutput
-% 
-% 
-% 
-% for n = 1:7
-%     temp = tsqh_test(subjmeans(:,n),[],[],[]);
-%     alloutput(n) = temp.tsq;
-% end
-% temp
-% alloutput
-% 
-% 
-% for n = 1:7
-%     temp = tsqh_test(subjmeans(:,1),subjmeans(:,n),0,[]);
-%     alloutput(n) = temp.tsq;
-% end
-% temp
-% alloutput
-% 
-% 
-% for n = 2:7
-%     temp = tsqh_test(subjmeans(:,1),subjmeans(:,n),1,[]);
-%     alloutput(n) = temp.tsq;
-% end
-% temp
-% alloutput
-% 
-% 
-% 
-% 
-% for n = 1:7
-%     xydata = [real(subjmeans(:,n)) imag(subjmeans(:,n))]';
-%     temp = tsqc_test(xydata,[],[],[]);
-%     alloutput(n) = temp.tsqc;
-% end
-% temp
-% alloutput
-% 
-% 
-% for n = 1:7
-%     xydata = [real(subjmeans(:,n)) imag(subjmeans(:,n))]';
-%     temp = tsqh_test(xydata,[],[],[]);
-%     alloutput(n) = temp.tsq;
-% end
-% temp
-% alloutput
-
-% for n = 1:7
-%     temp = CI_test(subjmeans(:,n),0.05);
-%     alloutput(n) = temp.CI;
-% end
-% temp
-% alloutput
-
-
-% 
-% for n = 1:7
-%     xydata = [real(subjmeans(:,n)) imag(subjmeans(:,n))]';
-%     temp = CI_test(xydata,0.05);
[…]

### 2021-01-12 R/FourierStats-package.r

@@ -1,5 +1,5 @@
 #' Documentation for FourierStats R package
-#' The FourierStats package implements several statistical tests for use with bivariate data such as the complex (real and imaginary) components of the Fourier spectrum. Included functions implement the T-squared-circ test of Victor & Mast (1991), and the condition index and ANOVA-squared-circ described by Baker (2021).
+#' The FourierStats package implements several statistical tests for use with bivariate data such as the complex (real and imaginary) components of the Fourier spectrum. Included functions implement the T-squared-circ test of Victor & Mast (1991), and the condition index and ANOVA-squared-circ tests described by Baker (2021).
 #' Functions include:
 #'  - analysecplx: a function to analyse complex Fourier data, following the guidelines from the Baker (2021) paper
 #'  - tsqc.test: implementation of the T-squared-circ test of Victor & Mast (1991)
@@ -7,7 +7,11 @@
 #'  - CI.test: condition index test, to test the assumptions of the T-squared-circ test
 #'  - anovacirc.test: implementation of one-way between subjects and repeated measures ANOVA-squared-circ test
 #'  - amperrors: function to calculate error bars for amplitudes incorporating coherent averaging
+#'  - pairwisemahal: calculates the pairwise Mahalanobis distance between two group means
+#'  - clustercorrect: implements cluster correction described by Maris & Oostenveld (2007) for multivariate T-tests
 #'  - getel: helper function that calculates the bounding ellipse for a cloud of points
 #'  - fftshift: helper function that performs the quadrant shift of a 2D Fourier spectrum
 #'
-#'  DHB 26/12/20
+#'  package available from: https://github.com/bakerdh/FourierStats
+#'  for further details see: ArXiV
+#'  DHB 12/01/21

### 2021-01-13 R/CI.test.R
Added ArXiv URL

@@ -5,7 +5,7 @@
 #' the condition index is square root of the the ratio of eigenvalues (eigenvector lengths), calculated longest/shortest
 #' the index is then compared for significance with an expected distribution function
 #' significant tests (p < alpha) violate the assumptions of the T-squared-circ and ANOVA-squared-circ tests
-#' see Baker (2021) for further details
+#' see Baker (2021, http://arxiv.org/abs/2101.04408) for further details
 #' @export
 CI.test <- function(data, alpha=0.05){
 

### 2021-01-13 R/FourierStats-package.r

@@ -13,5 +13,5 @@
 #'  - fftshift: helper function that performs the quadrant shift of a 2D Fourier spectrum
 #'
 #'  package available from: https://github.com/bakerdh/FourierStats
-#'  for further details see: ArXiV
+#'  for further details see: http://arxiv.org/abs/2101.04408
 #'  DHB 12/01/21

### 2021-01-13 R/anovacirc.test.R

@@ -11,7 +11,7 @@
 #' participant: variable storing participant/subject IDs (i.e. the random factor) for repeated measures analysis
 #'        this is an optional input, but if it is supplied it supercedes values from the matrix
 #'
-#' see Baker (2021) for further details
+#' see Baker (2021, http://arxiv.org/abs/2101.04408) for further details
 #' @export
 anovacirc.test <- function(data, group=NULL, participant=NULL){
 

### 2021-03-28 manuscript/manuscript.Rmd
Revisions in response to reviews

@@
 
-nsims <- 100000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
+nsims <- 1000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
 
@@
 Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency.
-The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase).
+The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase), either to test for the presence of a signal, or to compare signals across different conditions.
 However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded.
@@
 The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run.
-Implementations of these novel tools are provided as an _R_ package, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
+Implementations of these novel tools are provided as an _R_ package and a _Matlab_ toolbox, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
 
@@
 A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal.
-The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-defined index of the brain's response (see Figure \@ref(fig:fourierexplain)a,b).
+The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response (see Figure \@ref(fig:fourierexplain)a,b).
 Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \@ref(fig:fourierexplain)c), or equivalently as complex numbers with real and imaginary components (Figure \@ref(fig:fourierexplain)d).
-In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data only.
-However an alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary components).
-Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods.
+In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data (or equivalently power or signal-to-noise ratios) only [e.g.
+@Liu-Shuang2014; @Afsari2014; @Itthipuripat2014; @McFadden2014; @Vanegas2015; @Smith2017; @Hou2020] (note this is a non-exhaustive selection of recent studies).
+Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if signals differ between two (or more) experimental conditions.
+However, there are several objections to using univariate statistics to answer these questions, as will be demonstrated below.
+An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components).
+Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods for typical applications (detecting the presence of a signal, or comparing multiple signals).
 
@@
 This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as will be demonstrated below).
-Conceptually, this statistic takes the vector difference between the centroid ($\bar{x}$) and a comparison point ($\mu$), and scales by the mean length of the residual vector lines joining each data point to the centroid.
+Conceptually, this statistic takes the vector difference between the bivariate sample mean ($\bar{x}$) and a comparison point ($\mu$), and scales by the mean length of the residual vector lines joining each data point to the sample mean.
 Two-sample and repeated measures versions of the $T^2_{circ}$ statistic are also possible.
@@
 ```{r amphists, fig.cap="Demonstration of skew in absolute Fourier amplitudes for signals of different strengths.
-Signal strength is quantified as Cohen's d, defined as the ratio of the mean to the standard deviation of the sample.
+Signal strength is quantified as Cohen's d, defined as the ratio of the sample mean to the sample standard deviation.
 The upper row shows samples of 50 grey points, and the population mean (coloured points).
@@
 The lower row in Figure \@ref(fig:amphists) shows distributions of amplitudes for the same set of signal strengths.
-These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's _d_ > 2; Cohen's _d_ is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]).
+These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's _d_ > 2; Cohen's _d_ is the difference of sample means scaled by the standard deviation, see Cohen [-@Cohen1988]).
 One consequence of this is that T-tests will potentially have an inflated Type I error (false positive) rate for many signals encountered empirically, especially if used to make pointwise comparisons to an amplitude of 0.
 
-Typical solutions, such as log-transforming the data, are unlikely to be equally applicable to all conditions.
+Typical solutions for dealing with skew, such as log-transforming the data, are unlikely to be equally applicable to all conditions.
 For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal.
@@
 Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents.
-Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.
+Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.
 
@@
 Using the $T^2$ or $T^2_{circ}$ statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
-Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$.
-This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant.
-Their simulations show the largest advantage for $T^2_{circ}$ for effect sizes around _d_ = 1 (where the mean signal strength is equal to the standard deviation of the data).
+Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$, which is expected on theoretical grounds because the additional degrees of freedom in the $T^2_{circ}$ test make it more efficient.
+Their simulations involved generating random data sets of different sample sizes, and different signal strengths (sample means), and comparing the number of such tests where each statistic was significant.
+The simulations showed the largest advantage for $T^2_{circ}$ for effect sizes around _d_ = 1 (where the sample mean is equal to the sample standard deviation).
 The advantage appeared to be stronger for smaller sample sizes.
@@
 This advantage is lost for large sample sizes (N > 32) and large effect sizes (when _d_ > 2 and N > 8).
-These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use should be restricted to small sample studies.
+These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use can be restricted to small sample studies.
 
@@
 samplesizes <- ceiling(2^slist)
+actualslog <- log2(samplesizes)
 
@@
 title(paste('d =',dlist[vinds[n]]),cex.main=2)
-lines(slist,tsqpropM[,vinds[n]],lwd=3,col=nbdtpal[1])
-lines(slist,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])
-
-points(slist,tsqpropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
-points(slist,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])
+lines(actualslog,tsqpropM[,vinds[n]],lwd=3,col=nbdtpal[1])
+lines(actualslog,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])
+
+points(actualslog,tsqpropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
+points(actualslog,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])
 
@@
 title(ttext,cex.main=2)
-image(slist,dlist,circpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+image(actualslog,dlist,circpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
 text(1.1,3.8,'(e)',cex=2)
@@
 title(ttext,cex.main=2)
-image(slist,dlist,tsqvscirc,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
-cl <- contourLines(slist,dlist,tsqvscirc,levels=c(0.02,0.05,0.1,0.2,0.4))
+image(actualslog,dlist,tsqvscirc,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+cl <- contourLines(actualslog,dlist,tsqvscirc,levels=c(0.02,0.05,0.1,0.2,0.4))
 for (n in 1:length(cl)){
@@
 When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (horizontal dashed lines).
-However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the variances of the two dependent variables more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2.
+However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the sample variances of the two dependent variables grow more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2.
 In contrast, the $T^2$ statistic, which explicitly takes account of the covariance matrix (see equation \@ref(eq:t2eq)) shows no increase (black curves).
@@
 plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
-axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     
 axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
-mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklabelsx, side = 1, at=ticklocsx)     
 mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
@@
 plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
-axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
+axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)   
 axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
-mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
+mtext(text = ticklabelsx, side = 1, at=ticklocsx)    
 mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
@@
 
-One possible remedy to control the Type I error rate would be to adjust either the alpha level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated).
+One possible remedy to control the Type I error rate would be to adjust either the $\alpha$ level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated).
 However, this will reduce the statistical power of the $T^2_{circ}$ test, and its advantage over $T^2$ is relatively marginal in most situations to begin with (see Figure \@ref(fig:powerfig)).
@@
 \label{eq:edelman1}
-pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)},
+pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)}.
 \end{equation}
 
-Attempts to validate this by simulation suggest that for small sample sizes (_N_ < 10) a closer approximation is given by:
+Attempts to validate this by simulation (100,000 random data sets) found that for small sample sizes (_N_ < 10), there is a mismatch between the equation's predictions and the simulations (the black curve does not match the blue shading in Figure \@ref(fig:distcomparison)a).
+Instead, a closer approximation to the simulations is given by:
 
@@
 
-Figure \@ref(fig:distcomparison)a shows example distributions derived from both expressions (black and red curves), and simulations from 100,000 random data sets with _N_ = 4 (blue shading).
+The red curve in Figure \@ref(fig:distcomparison)a shows the predictions of this modified equation, which are much closer to the simulation results (blue shading).
 The vertical lines show the critical (95%) threshold for the analytic and simulated results.
@@
 The Mahalanobis distance [@Mahalanobis1936] is a useful metric for identifying such multivariate outliers so that they can be excluded.
-It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points.
+It calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points.
 This means that any correlations in the data set are taken into account when calculating the distance metric, _D_.
@@
 In all cases, the functions depart from the Type I error rate ($\alpha$ = 0.05; horizontal dashed line in Figure \@ref(fig:outlierplot)) when the outlier's Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core _mahalanobis_ function in _R_ (or _mahal_ function in _Matlab_), return $D^2$, which can be converted to _D_ by taking the square root).
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the sample mean (note that many implementations of the Mahalanobis distance statistic, such as the core _mahalanobis_ function in _R_, or _mahal_ function in _Matlab_, return $D^2$, which can be converted to _D_ by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the $T^2_{circ}$ test.
@@
 One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's $T^2$, but permits independent variables with more than two levels, as well as factorial designs.
-However, if the assumptions of $T^2_{circ}$ hold for a data set, it might alternatively be possible to extend the logic of the $T^2_{circ}$ test [@Victor1991] to the more general case, and obtain a sensitivity benefit similar to that shown in Figure \@ref(fig:powerfig).
-
-The F-statistic for a one-way ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance.
-For multivariate data, the change in group mean would be calculated using the vector distances between the complex means, and the residuals are the vector distances between each data point and its corresponding group mean.
-For an independent one-way design with $k$ groups (or conditions) and $N$ observations per group, the F-distribution will have $2(k-1)$ and $2((Nk)-k)$ degrees of freedom.
-The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number.
-A _p_-value can then be derived from the F-distribution in the usual way.
+However, if the assumptions of $T^2_{circ}$ hold for a data set, it should alternatively be possible to extend the logic of the $T^2_{circ}$ test [@Victor1991] to the more general multiple group case, and obtain a sensitivity benefit similar to that shown in Figure \@ref(fig:powerfig).
+
+The F-statistic for a one-way independent ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance:
+
+\begin{equation}
+\label{eq:Fratio}
+F = \frac{MS_M}{MS_R},
+\end{equation}
+
+where $MS_M$ is the mean squares of the linear model, and $MS_R$ is the mean squares of the residuals.
+For bivariate data, the linear model is defined as the change in group (condition) means relative to the grand mean, calculated using the vector distances between points:
+
+\begin{equation}
+\label{eq:MSM}
+MS_M = \frac{\Sigma N_k (\bar{x}_k - \bar{x}_{grand})^2}{df_M},
+\end{equation}
+
+where $N_k$ is the number of observations in group _k_, $\bar{x}_k$ is the bivariate sample mean of group _k_, $\bar{x}_{grand}$ is the bivariate mean of the entire sample (including all groups), $\Sigma$ denotes summation across groups, and $df_M$ is the degrees of freedom for the model.
+The model degrees of freedom for _k_ groups is $df_M = 2(k-1)$, with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.
+
+The residuals are the vector distances between each data point and its corresponding group mean. The denominator component for the F-ratio equation is therefore defined as:
+
+\begin{equation}
+\label{eq:MSR}
+MS_R = \frac{\Sigma (x_{i,k} - \bar{x}_k)^2}{df_R},
+\end{equation}
+
+where $x_{i,k}$ is the $i$th data point from group _k_, and other terms are as described previously.
+The degrees of freedom for a balanced design are calculated as $df_R = 2((Nk)-k)$, where _N_ is the number of observations per group, and _k_ is the number of groups.
+A _p_-value can then be determined by comparing the F-ratio calculated from equation \@ref(eq:Fratio) to an F-distribution with $df_M$ and $df_R$ degrees of freedom.
 
@@
 title(paste('d =',dlist[vinds[n]]),cex.main=2)
-lines(slist,manovapropM[,vinds[n]],lwd=3,col=nbdtpal[1])
-lines(slist,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])
-
-points(slist,manovapropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
-points(slist,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])
+lines(actualslog,manovapropM[,vinds[n]],lwd=3,col=nbdtpal[1])
+lines(actualslog,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])
+
+points(actualslog,manovapropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
+points(actualslog,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])
 
@@
 title(expression('MANOVA'),cex.main=2)
-image(slist,dlist,manovapropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+image(actualslog,dlist,manovapropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
 text(1.1,3.8,'(d)',cex=2)
@@
 title(ttext,cex.main=2)
-image(slist,dlist,manvscirc,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
-cl <- contourLines(slist,dlist,manvscirc,levels=c(0.02,0.05,0.1,0.2,0.4))
+image(actualslog,dlist,manvscirc,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+cl <- contourLines(actualslog,dlist,manvscirc,levels=c(0.02,0.05,0.1,0.2,0.4))
 for (n in 1:length(cl)){
@@
 Bonferroni correction).
-Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary; see Figure \@ref(fig:fourierexplain)d).
+Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes of the two measures depend on stimulus phase (which is arbitrary; see Figure \@ref(fig:fourierexplain)d).
 A repeated measures version of ${ANOVA}^2_{circ}$ can also be implemented following the same logic.
@@
 
-The flowchart in Figure \@ref(fig:flowchart) contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
+The flowchart in Figure \@ref(fig:flowchart) illustrates a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
 Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test.
-Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, and the $T^2$ statistic if they are not.
-Comparisons with more than two conditions should be tested with the $ANOVA^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not.
+Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the condition index test is non-significant, and the $T^2$ statistic otherwise.
+Comparisons with more than two conditions should be tested with the $ANOVA^2_{circ}$ statistic if the condition index test is non-significant, or a MANOVA otherwise.
 Many MANOVA implementations cannot deal correctly with random factors (repeated measures), particularly in complex factorial designs.
@@
 This revealed a significant difference between conditions ($T^2_{circ}$ = `r round(tsqcresult$tsqc,digits=2)`, $F_{(`r tsqcresult$df1`,`r tsqcresult$df2`)}$ = `r round(tsqcresult$Fratio,digits=2)`, _p_ = `r round(tsqcresult$pval,digits=3)`) with an effect size of _D_ = `r round(D,digits=2)`.
-This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
+This demonstrates that sound and (optogenetic) light entrain neural responses with different latencies (lags).
+The original study by Hwang et al.
 [-@Hwang2019] went on to explore interactions between these two signals.
@@
 Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \@ref(fig:humanSSVEP)), were well-isolated in the Fourier domain, and increased with stimulus contrast.
-Significant activity was evident at 4% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \@ref(fig:humanSSVEP).
+Significant activity was evident at 4% contrast and above following cluster correction (with very stringent $\alpha$ levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \@ref(fig:humanSSVEP).
 For the main analysis, responses were taken from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), and averaged across repetition for each participant.
@@
 Electrodes marked in red indicate cluster-corrected significance.
-Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point).
-Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.', fig.width=14, fig.height=5.4, fig.align="center", echo=FALSE}
+Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (marked  by the black points in the scalp plots).
+Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the bivariate sample means.', fig.width=14, fig.height=5.4, fig.align="center", echo=FALSE}
 
@@
 It is worth stating explicitly that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations.
-This is the case for most paradigms in which the nervous system is driven by a periodic stimulus.
-However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms [@Berger1929; @Buzsaki2004], which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals [e.g.
+This is the case for most paradigms in which the nervous system is driven by a periodic stimulus, and sometimes also when phase-locked oscillatory responses are induced by a brief stimulus presentation [e.g.
+in time-frequency analysis of event-related potentials, see @Delorme2004; @Pfurtscheller1979].
+However, the methods are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms [@Berger1929; @Buzsaki2004], which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals [e.g.
 @Canolty2010].
@@
 [-@Pei2017] suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat demanding, though a function is provided to calculate error bars using this method (available through the _amperrors_ function).
+This approach is somewhat computationally demanding, though a function is provided to calculate error bars using this method (available through the _amperrors_ function).
 However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes.
@@
 The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled mean amplitudes.
-Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68% and 95% are typical, with the 68% interval corresponding to the standard error).
+Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68% or 95% are typical, with the 68% interval corresponding to the standard error).
 
@@
 The greater power afforded by these tests should in general lead to more accurate statistical inferences when analysing periodic data.
+Additionally, more sensitive methods require fewer observations to reach a given level of power, which has important ethical implications for animal research if fewer experimental subjects are needed.
 

### 2021-03-28 manuscript/manuscript.tex
Preparing for first public release
Revisions in response to reviews

@@
 \author{Daniel H. Baker}
-\date{2021-01-09}
+\date{2021-03-28}
 
@@
 The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run.
-Implementations of these novel tools are provided as an \emph{R} package, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
+Implementations of these novel tools are provided as an \emph{R} package and a \emph{Matlab} toolbox, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.
 
@@
 A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal.
-The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-defined index of the brain's response (see Figure \ref{fig:fourierexplain}a,b).
+The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response (see Figure \ref{fig:fourierexplain}a,b).
 Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \ref{fig:fourierexplain}c), or equivalently as complex numbers with real and imaginary components (Figure \ref{fig:fourierexplain}d).
-In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data only.
-However an alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary components).
-Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods.
+In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data (or equivalently power or signal-to-noise ratios) only (e.g.
+Liu-Shuang et al., 2014; Afsari et al., 2014; Hou et al., 2020; Itthipuripat et al., 2014; McFadden et al., 2014; Smith et al., 2017; Vanegas et al., 2015) (note this is a non-exhaustive selection of recent studies).
+Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if signals differ between two (or more) experimental conditions.
+However, there are several objections to using univariate statistics to answer these questions, as will be demonstrated below.
+An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components).
+Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods for typical applications (detecting the presence of a signal, or comparing multiple signals).
 
@@
 This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as will be demonstrated below).
-Conceptually, this statistic takes the vector difference between the centroid (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the mean length of the residual vector lines joining each data point to the centroid.
+Conceptually, this statistic takes the vector difference between the bivariate sample mean (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the mean length of the residual vector lines joining each data point to the sample mean.
 Two-sample and repeated measures versions of the \(T^2_{circ}\) statistic are also possible.
@@
 \caption{Demonstration of skew in absolute Fourier amplitudes for signals of different strengths.
-Signal strength is quantified as Cohen's d, defined as the ratio of the mean to the standard deviation of the sample.
+Signal strength is quantified as Cohen's d, defined as the ratio of the sample mean to the sample standard deviation.
 The upper row shows samples of 50 grey points, and the population mean (coloured points).
@@
 The lower row in Figure \ref{fig:amphists} shows distributions of amplitudes for the same set of signal strengths.
-These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's \emph{d} \textgreater{} 2; Cohen's \emph{d} is the mean difference scaled by the standard deviation, see Cohen (1988)).
+These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's \emph{d} \textgreater{} 2; Cohen's \emph{d} is the difference of sample means scaled by the standard deviation, see Cohen (1988)).
 One consequence of this is that T-tests will potentially have an inflated Type I error (false positive) rate for many signals encountered empirically, especially if used to make pointwise comparisons to an amplitude of 0.
 
-Typical solutions, such as log-transforming the data, are unlikely to be equally applicable to all conditions.
+Typical solutions for dealing with skew, such as log-transforming the data, are unlikely to be equally applicable to all conditions.
 For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal.
@@
 Using the \(T^2\) or \(T^2_{circ}\) statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
-Victor \& Mast (1991) report simulations showing situations where \(T^2_{circ}\) has greater power than \(T^2\).
-This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant.
-Their simulations show the largest advantage for \(T^2_{circ}\) for effect sizes around \emph{d} = 1 (where the mean signal strength is equal to the standard deviation of the data).
+Victor \& Mast (1991) report simulations showing situations where \(T^2_{circ}\) has greater power than \(T^2\), which is expected on theoretical grounds because the additional degrees of freedom in the \(T^2_{circ}\) test make it more efficient.
+Their simulations involved generating random data sets of different sample sizes, and different signal strengths (sample means), and comparing the number of such tests where each statistic was significant.
+The simulations showed the largest advantage for \(T^2_{circ}\) for effect sizes around \emph{d} = 1 (where the sample mean is equal to the sample standard deviation).
 The advantage appeared to be stronger for smaller sample sizes.
@@
 This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (when \emph{d} \textgreater{} 2 and N \textgreater{} 8).
-These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use should be restricted to small sample studies.
+These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use can be restricted to small sample studies.
 
@@
 When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (horizontal dashed lines).
-However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the variances of the two dependent variables more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2.
+However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the sample variances of the two dependent variables grow more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2.
 In contrast, the \(T^2\) statistic, which explicitly takes account of the covariance matrix (see equation \eqref{eq:t2eq}) shows no increase (black curves).
@@
 
-One possible remedy to control the Type I error rate would be to adjust either the alpha level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated).
+One possible remedy to control the Type I error rate would be to adjust either the \(\alpha\) level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated).
 However, this will reduce the statistical power of the \(T^2_{circ}\) test, and its advantage over \(T^2\) is relatively marginal in most situations to begin with (see Figure \ref{fig:powerfig}).
@@
 \label{eq:edelman1}
-pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)},
+pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)}.
 \end{equation}
 
-Attempts to validate this by simulation suggest that for small sample sizes (\emph{N} \textless{} 10) a closer approximation is given by:
+Attempts to validate this by simulation (100,000 random data sets) found that for small sample sizes (\emph{N} \textless{} 10), there is a mismatch between the equation's predictions and the simulations (the black curve does not match the blue shading in Figure \ref{fig:distcomparison}a).
+Instead, a closer approximation to the simulations is given by:
 
@@
 
-Figure \ref{fig:distcomparison}a shows example distributions derived from both expressions (black and red curves), and simulations from 100,000 random data sets with \emph{N} = 4 (blue shading).
+The red curve in Figure \ref{fig:distcomparison}a shows the predictions of this modified equation, which are much closer to the simulation results (blue shading).
 The vertical lines show the critical (95\%) threshold for the analytic and simulated results.
@@
 The Mahalanobis distance (Mahalanobis, 1936) is a useful metric for identifying such multivariate outliers so that they can be excluded.
-It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points.
+It calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points.
 This means that any correlations in the data set are taken into account when calculating the distance metric, \emph{D}.
@@
 In all cases, the functions depart from the Type I error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the outlier's Mahalanobis distance exceeds a value around 3.
-This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R} (or \emph{mahal} function in \emph{Matlab}), return \(D^2\), which can be converted to \emph{D} by taking the square root).
+This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the sample mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R}, or \emph{mahal} function in \emph{Matlab}, return \(D^2\), which can be converted to \emph{D} by taking the square root).
 Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the \(T^2_{circ}\) test.
@@
 One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's \(T^2\), but permits independent variables with more than two levels, as well as factorial designs.
-However, if the assumptions of \(T^2_{circ}\) hold for a data set, it might alternatively be possible to extend the logic of the \(T^2_{circ}\) test (Victor and Mast, 1991) to the more general case, and obtain a sensitivity benefit similar to that shown in Figure \ref{fig:powerfig}.
+However, if the assumptions of \(T^2_{circ}\) hold for a data set, it should alternatively be possible to extend the logic of the \(T^2_{circ}\) test (Victor and Mast, 1991) to the more general multiple group case, and obtain a sensitivity benefit similar to that shown in Figure \ref{fig:powerfig}.
 
-The F-statistic for a one-way ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance.
-For multivariate data, the change in group mean would be calculated using the vector distances between the complex means, and the residuals are the vector distances between each data point and its corresponding group mean.
-For an independent one-way design with \(k\) groups (or conditions) and \(N\) observations per group, the F-distribution will have \(2(k-1)\) and \(2((Nk)-k)\) degrees of freedom.
-The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number.
-A \emph{p}-value can then be derived from the F-distribution in the usual way.
+The F-statistic for a one-way independent ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance:
+
+\begin{equation}
+\label{eq:Fratio}
+F = \frac{MS_M}{MS_R},
+\end{equation}
+
+where \(MS_M\) is the mean squares of the linear model, and \(MS_R\) is the mean squares of the residuals.
+For bivariate data, the linear model is defined as the change in group (condition) means relative to the grand mean, calculated using the vector distances between points:
+
+\begin{equation}
+\label{eq:MSM}
+MS_M = \frac{\Sigma N_k (\bar{x}_k - \bar{x}_{grand})^2}{df_M},
+\end{equation}
+
+where \(N_k\) is the number of observations in group \emph{k}, \(\bar{x}_k\) is the bivariate sample mean of group \emph{k}, \(\bar{x}_{grand}\) is the bivariate mean of the entire sample (including all groups), \(\Sigma\) denotes summation across groups, and \(df_M\) is the degrees of freedom for the model.
+The model degrees of freedom for \emph{k} groups is \(df_M = 2(k-1)\), with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.
+
+The residuals are the vector distances between each data point and its corresponding group mean. The denominator component for the F-ratio equation is therefore defined as:
+
+\begin{equation}
+\label{eq:MSR}
+MS_R = \frac{\Sigma (x_{i,k} - \bar{x}_k)^2}{df_R},
+\end{equation}
+
+where \(x_{i,k}\) is the \(i\)th data point from group \emph{k}, and other terms are as described previously.
+The degrees of freedom for a balanced design are calculated as \(df_R = 2((Nk)-k)\), where \emph{N} is the number of observations per group, and \emph{k} is the number of groups.
+A \emph{p}-value can then be determined by comparing the F-ratio calculated from equation \eqref{eq:Fratio} to an F-distribution with \(df_M\) and \(df_R\) degrees of freedom.
 
@@
 Following a significant \({ANOVA}^2_{circ}\) test, one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction).
-Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary; see Figure \ref{fig:fourierexplain}d).
+Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes of the two measures depend on stimulus phase (which is arbitrary; see Figure \ref{fig:fourierexplain}d).
 A repeated measures version of \({ANOVA}^2_{circ}\) can also be implemented following the same logic.
@@
 
-The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
+The flowchart in Figure \ref{fig:flowchart} illustrates a proposed decision structure for the analysis of periodic data, once any outliers have been removed.
 Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test.
-Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if they are not.
-Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not.
+Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the condition index test is non-significant, and the \(T^2\) statistic otherwise.
+Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the condition index test is non-significant, or a MANOVA otherwise.
 Many MANOVA implementations cannot deal correctly with random factors (repeated measures), particularly in complex factorial designs.
@@
 This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of \emph{D} = 2.14.
-This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al.
+This demonstrates that sound and (optogenetic) light entrain neural responses with different latencies (lags).
+The original study by Hwang et al.
 (2019) went on to explore interactions between these two signals.
@@
 Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast.
-Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \ref{fig:humanSSVEP}.
+Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent \(\alpha\) levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \ref{fig:humanSSVEP}.
 For the main analysis, responses were taken from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), and averaged across repetition for each participant.
@@
 Electrodes marked in red indicate cluster-corrected significance.
-Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point).
-Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.}\label{fig:humanSSVEP}
+Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (marked  by the black points in the scalp plots).
+Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the bivariate sample means.}\label{fig:humanSSVEP}
 \end{figure}
@@
 It is worth stating explicitly that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations.
-This is the case for most paradigms in which the nervous system is driven by a periodic stimulus.
-However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger, 1929; Buzsáki and Draguhn, 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g.
+This is the case for most paradigms in which the nervous system is driven by a periodic stimulus, and sometimes also when phase-locked oscillatory responses are induced by a brief stimulus presentation (e.g.~in time-frequency analysis of event-related potentials, see Delorme and Makeig, 2004; Pfurtscheller and Aranibar, 1979).
+However, the methods are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger, 1929; Buzsáki and Draguhn, 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g.
 Canolty and Knight, 2010).
@@
 (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude.
-This approach is somewhat demanding, though a function is provided to calculate error bars using this method (available through the \emph{amperrors} function).
+This approach is somewhat computationally demanding, though a function is provided to calculate error bars using this method (available through the \emph{amperrors} function).
 However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes.
@@
 The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled mean amplitudes.
-Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68\% and 95\% are typical, with the 68\% interval corresponding to the standard error).
+Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68\% or 95\% are typical, with the 68\% interval corresponding to the standard error).
 
@@
 
+\leavevmode\hypertarget{ref-Delorme2004}{}%
+Delorme A, Makeig S.
+2004.
+EEGLAB: An open source toolbox for analysis of single-trial EEG dynamics including independent component analysis.
+\emph{J Neurosci Methods} \textbf{134}:9--21.
+doi:\href{https://doi.org/10.1016/j.jneumeth.2003.10.009}{10.1016/j.jneumeth.2003.10.009}
+
 \leavevmode\hypertarget{ref-Edelman1988}{}%
@@
 
+\leavevmode\hypertarget{ref-Hou2020}{}%
+Hou C, Nicholas SC, Verghese P.
+2020.
+Contrast normalization accounts for binocular interactions in human striate and extra-striate visual cortex.
+\emph{J Neurosci} \textbf{40}:2753--2763.
+doi:\href{https://doi.org/10.1523/JNEUROSCI.2043-19.2020}{10.1523/JNEUROSCI.2043-19.2020}
+
 \leavevmode\hypertarget{ref-Hwang2019}{}%
@@
 doi:\href{https://doi.org/10.1038/s41597-020-00621-z}{10.1038/s41597-020-00621-z}
+
+\leavevmode\hypertarget{ref-Itthipuripat2014}{}%
+Itthipuripat S, Garcia JO, Rungratsameetaweemana N, Sprague TC, Serences JT.
+2014.
+Changing the spatial scope of attention alters patterns of neural gain in human cortex.
+\emph{J Neurosci} \textbf{34}:112--23.
+doi:\href{https://doi.org/10.1523/JNEUROSCI.3943-13.2014}{10.1523/JNEUROSCI.3943-13.2014}
 
@@
 doi:\href{https://doi.org/10.1016/j.jneumeth.2007.03.024}{10.1016/j.jneumeth.2007.03.024}
+
+\leavevmode\hypertarget{ref-McFadden2014}{}%
+McFadden KL, Steinmetz SE, Carroll AM, Simon ST, Wallace A, Rojas DC.
+2014.
+Test-retest reliability of the 40 hz EEG auditory steady-state response.
+\emph{PLoS ONE} \textbf{9}:e85748.
+doi:\href{https://doi.org/10.1371/journal.pone.0085748}{10.1371/journal.pone.0085748}
 
@@
 doi:\href{https://doi.org/10.1016/j.visres.2016.03.010}{10.1016/j.visres.2016.03.010}
+
+\leavevmode\hypertarget{ref-Pfurtscheller1979}{}%
+Pfurtscheller G, Aranibar A.
+1979.
+Evaluation of event-related desynchronization (ERD) preceding and following voluntary self-paced movement.
+\emph{Electroencephalogr Clin Neurophysiol} \textbf{46}:138--46.
+doi:\href{https://doi.org/10.1016/0013-4694(79)90063-4}{10.1016/0013-4694(79)90063-4}
 
@@
 
+\leavevmode\hypertarget{ref-Smith2017}{}%
+Smith AK, Wade AR, Penkman KEH, Baker DH.
+2017.
+Dietary modulation of cortical excitation and inhibition.
+\emph{J Psychopharmacol} \textbf{31}:632--637.
+doi:\href{https://doi.org/10.1177/0269881117699613}{10.1177/0269881117699613}
+
 \leavevmode\hypertarget{ref-Snyder1992}{}%
@@
 doi:\href{https://doi.org/10.1073/pnas.1400942111}{10.1073/pnas.1400942111}
+
+\leavevmode\hypertarget{ref-Vanegas2015}{}%
+Vanegas MI, Blangero A, Kelly SP.
+2015.
+Electrophysiological indices of surround suppression in humans.
+\emph{J Neurophysiol} \textbf{113}:1100--9.
+doi:\href{https://doi.org/10.1152/jn.00774.2014}{10.1152/jn.00774.2014}
 

### 2021-06-15 manuscript/manuscript.Rmd
Revisions in response to reviewer comments

--- 2021-03-28 manuscript/manuscript.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+
+nsims <- 100000  # global value for the whole script - increase to 100000 for final run, but note this takes around 24 hours to complete (depending on processor speed)
+buildfigs <- 1
 
@@
 # install the missing ones, and activate all
-
-packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','devtools')
+packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','devtools','knitr')
 missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
@@
 
-nsims <- 1000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)
+knitr::opts_chunk$set(echo = TRUE)
 
@@
 Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli.
-Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency.
+Neural signals recorded using a variety of techniques will then include phase-locked oscillations at the stimulation frequency.
 The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase), either to test for the presence of a signal, or to compare signals across different conditions.
-However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded.
+However, the assumptions of these tests will sometimes be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded.
 An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components.
@@
 
-A widely used paradigm in many branches of neuroscience is to drive the nervous system using periodic stimuli.
+A widely used approach in many branches of neuroscience is to drive the nervous system using periodic stimuli.
 This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations.
@@
 @Liu-Shuang2014; @Afsari2014; @Itthipuripat2014; @McFadden2014; @Vanegas2015; @Smith2017; @Hou2020] (note this is a non-exhaustive selection of recent studies).
-Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if signals differ between two (or more) experimental conditions.
+Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if responses differ in amplitude and/or phase between two (or more) experimental conditions.
 However, there are several objections to using univariate statistics to answer these questions, as will be demonstrated below.
-An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components).
+An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components in a Cartesian space).
 Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods for typical applications (detecting the presence of a signal, or comparing multiple signals).
 
-```{r fourierexplain, fig.cap="Illustration of the principles of Fourier analysis for time-varying signals.
-Panel (a) shows a sinusoidal signal waveform, and simulated neural responses for successive observations (repetitions in the same individual, or recorded from multiple individuals).
-Panel (b) shows the Fourier amplitude spectra of the waveforms in (a), with clear peaks at the signal frequency (5Hz) in each example.
-The Fourier spectrum also includes a phase term, which can be represented in polar coordinates (amplitude and phase); panel (c) shows this for example responses at the signal frequency (grey points) and their average (blue point), which is computed separately for amplitude and phase terms.
-Alternatively, the same information is contained in a Cartesian representation of the real and imaginary parts of the complex spectrum (panel (d)).
-Notice that the individual observations (grey points) are the same in panels (c,d), but the different representations have implications for how the average (blue points) and measures of spread (shaded regions) are calculated.", fig.align="center", echo=FALSE, fig.width=9, fig.height=8}
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/fourierexplain.pdf', bg="transparent", height = 8, width = 9)
 
@@
 
+dev.off()}
+
+```
+
+```{r fourierexplain, fig.cap="Illustration of the principles of Fourier analysis for time-varying signals.
+Panel (a) shows a sinusoidal signal waveform, and simulated neural responses for successive observations (repetitions in the same individual, or recorded from multiple individuals).
+Panel (b) shows the Fourier amplitude spectra of the waveforms in (a), with clear peaks at the signal frequency (5Hz) in each example.
+The Fourier spectrum also includes a phase term, which can be represented in polar coordinates (amplitude and phase); panel (c) shows this for example responses at the signal frequency (grey points) and their average (blue point), which is computed separately for amplitude and phase terms.
+Alternatively, the same information is contained in a Cartesian representation of the real and imaginary parts of the complex spectrum (panel (d)).
+Notice that the individual observations (grey points) are the same in panels (c,d), but the different representations have implications for how the average (blue points) and measures of spread (shaded regions) are calculated.", fig.align="center", echo=FALSE, fig.width=9, fig.height=8}
+
+knitr::include_graphics('figures/fourierexplain.pdf')
+
 ```
@@
 \label{eq:t2c}
-T^2_{circ} = (N-1)\frac{|\bar{x}-\mu|^2}{\Sigma|x_j - \bar{x}|^2}
+T^2_{circ} = (N-1)\frac{|\bar{x}-\mu|^2}{\Sigma|x_j - \bar{x}|^2},
 \end{equation}
@@
 In the present paper, best practice guidelines are developed for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms.
-It is first demonstrated why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed.
+It is first demonstrated why parametric univariate statistics are inappropriate for such data, primarily because amplitudes for weak signals are not normally distributed.
 Then conditions are investigated under which either the $T^2$ or $T^2_{circ}$ statistic should be used.
@@
 
-```{r amphists, fig.cap="Demonstration of skew in absolute Fourier amplitudes for signals of different strengths.
-Signal strength is quantified as Cohen's d, defined as the ratio of the sample mean to the sample standard deviation.
-The upper row shows samples of 50 grey points, and the population mean (coloured points).
-The lower row shows kernel density functions generated from 100,000 amplitude values.
-Note that the mean phase of the signal is irrelevant for these simulations, and is shown in the positive x-direction for consistency.", fig.align="center", echo=FALSE, fig.width=12, fig.height=6}
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/amphists.pdf', bg="transparent", height = 6, width = 12)
 
@@
 
+dev.off()}
+
+```
+
+```{r amphists, fig.cap="Demonstration of skew in absolute Fourier amplitudes for signals of different strengths.
+Signal strength is quantified as Cohen's d, defined as the ratio of the sample mean to the sample standard deviation.
+The upper row shows samples of 50 grey points, and the population mean (coloured points).
+The lower row shows kernel density functions generated from 100,000 amplitude values.
+Note that the mean phase of the signal is irrelevant for these simulations, and is shown in the positive x-direction for consistency.", fig.align="center", echo=FALSE, fig.width=12, fig.height=6}
+
+knitr::include_graphics('figures/amphists.pdf')
+
 ```
@@
 Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents.
-Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.
-
-# Conditions under which $T^2_{circ}$ is more sensitive than $T^2$
-
-Using the $T^2$ or $T^2_{circ}$ statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
+Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, as will be demonstrated in the following section.
+
+# Comparison of statistical power between univariate and multivariate tests
+
+Using the $T^2$ or $T^2_{circ}$ statistic allows the phase information to be retained, and therefore provides greater power than running univariate T-tests on amplitudes, as well as avoiding problems caused by using absolute amplitude values.
 Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$, which is expected on theoretical grounds because the additional degrees of freedom in the $T^2_{circ}$ test make it more efficient.
@@
 
-Here these simulations are replicated and extended (see Figure \@ref(fig:powerfig)), and it is shown that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f).
-However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations.
-This advantage is lost for large sample sizes (N > 32) and large effect sizes (when _d_ > 2 and N > 8).
-These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use can be restricted to small sample studies.
-
-```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes (100,000 simulated data sets per condition).
-Panels a-c replicate conditions reported by Victor and Mast (1991).
-Panels (d) and (e) show a wider range of conditions for each statistic.
-Panel (f) shows the difference between the two statistics, with contour lines indicating differences of 0.02, 0.05, 0.1, 0.2 and 0.4.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+Here these simulations are replicated and extended (see Figure \@ref(fig:powerfig)).
+The real and imaginary Fourier components of a periodic signal could in principle be analysed using two separate univariate t-tests.
+To detect the presence of a signal, each t-test would compare to a value of zero (i.e.
+the origin in Figure \@ref(fig:fourierexplain)d), to test against the null hypothesis that there is no signal ($H_{0x}: x = 0; H_{0y}: y = 0$).
+Without correction for familywise error, this approach has high statistical power for detecting when at least one test is significant (see grey triangles in Figure \@ref(fig:powerfig)a-c).
+However the use of two separate tests inflates the familywise error rate such that 10% of tests are significant (assuming $\alpha = 0.05$), even in the absence of a signal.
+Using Bonferroni correction to ameliorate this problem substantially reduces the power (white triangles in Figure \@ref(fig:powerfig)a-c).
+
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/powerfig.pdf', bg="transparent", height = 8, width = 12)
 
 layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))
-abc <- c('a','b','c')
 
@@
 
+tmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
+tmatrix2 <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
+tmatrix3 <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
 tsqmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
@@
     for (s2 in 1:length(dlist)){
+      baseline <- matrix(rnorm(2*samplesizes[s]),nrow=samplesizes[s],ncol=2)
       data <- matrix(rnorm(2*samplesizes[s]),nrow=samplesizes[s],ncol=2)
@@
       if (a[5]<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
+
+      ampdata <- abs(complex(real = data[,1], imaginary = data[,2]))
+      ampbase <- abs(complex(real = baseline[,1], imaginary = baseline[,2]))
+      a <- t.test(ampbase,ampdata)
+      if (a$p.value<0.05){tmatrix[s,s2] <- tmatrix[s,s2] + 1}
+
+      a <- t.test(data[,1])
+      b <- t.test(data[,2])
+      if (a$p.value<0.05 | b$p.value<0.05){tmatrix2[s,s2] <- tmatrix2[s,s2] + 1}
+      if (a$p.value<0.025 | b$p.value<0.025){tmatrix3[s,s2] <- tmatrix3[s,s2] + 1}
+
     }}
@@
 
+tpropM <- tmatrix/nsims
+tpropM2 <- tmatrix2/nsims
+tpropM3 <- tmatrix3/nsims
 tsqpropM <- tsqmatrix/nsims
@@
   title(paste('d =',dlist[vinds[n]]),cex.main=2)
+  lines(actualslog,tpropM3[,vinds[n]],lwd=3,col='black',lty=2)
+  lines(actualslog,tpropM2[,vinds[n]],lwd=3,col='grey')
+  lines(actualslog,tpropM[,vinds[n]],lwd=3,col=nbdtpal[3])
   lines(actualslog,tsqpropM[,vinds[n]],lwd=3,col=nbdtpal[1])
@@
 
-points(actualslog,tsqpropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
-points(actualslog,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])
-
-text(1.1,0.95,paste('(',abc[n],')',sep=''),cex=2)
-
-if (n==1){legend(3.8,1,c(expression(T^2),expression('T'[circ]^2)),pch=15:16,cex=1.5,col=nbdtpal[1:2],box.lwd=2)}
+  points(actualslog,tpropM3[,vinds[n]],lwd=3,pch=25,cex=1.5,col='black',bg='white')
+  points(actualslog,tpropM2[,vinds[n]],lwd=3,pch=24,cex=1.5,col='grey',bg='grey')
+  points(actualslog,tpropM[,vinds[n]],lwd=3,pch=23,cex=1.5,col=nbdtpal[3],bg=nbdtpal[3])
+  points(actualslog,tsqpropM[,vinds[n]],lwd=3,pch=21,cex=1.5,col=nbdtpal[1],bg=nbdtpal[1])
+  points(actualslog,circpropM[,vinds[n]],lwd=3,pch=22,cex=1.5,col=nbdtpal[2],bg=nbdtpal[2])
+
+  text(1.1,0.95,paste('(',letters[n],')',sep=''),cex=2)
+
+  if (n==1){legend(1.4,1,c('Two univariate, uncorrected','Two univariate, corrected',expression(T^2),expression('T'[circ]^2), 'Univariate amplitudes'),pch=c(24:25,21:23),cex=1.5,pt.bg=c('grey','white',nbdtpal[1:3]),col=c('grey','black',nbdtpal[1:3]),box.lwd=2)}
 }
@@
 title(expression(T^2),cex.main=2)
-image(slist,dlist,tsqpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
+image(actualslog,dlist,tsqpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
 text(1.1,3.8,'(d)',cex=2)
@@
 
-```
+dev.off()}
+
+```
+
+```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes (100,000 simulated data sets per condition).
+Panels a-c replicate conditions reported by Victor and Mast (1991), and extend them to include univariate tests for comparison.
+Panels (d) and (e) show a wider range of conditions for each multivariate statistic.
+Panel (f) shows the difference between the two statistics, with contour lines indicating power differences of 0.02, 0.05, 0.1, 0.2 and 0.4.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+
+knitr::include_graphics('figures/powerfig.pdf')
+
+```
+
+Using multivariate tests allows high power to be maintained without inflating the Type I error rate.
+Here, the multivariate mean (or equivalently, the difference in means between two conditions) is compared to zero, again testing against the null hypothesis that there is no signal ($H_0: x = y = 0$).
+The $T^2_{circ}$ statistic (red squares in Figure \@ref(fig:powerfig)a-c) has greater power than either the $T^2$ statistic (black circles) or the corrected univariate tests (white triangles).
+The power advantage over $T^2$ occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f).
+However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations.
+This advantage is lost for large sample sizes (N > 32) and large effect sizes (when _d_ > 2 and N > 8).
+These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use can be restricted to small-sample studies.
+
+As mentioned above, an alternative approach that is widely used in the literature is to discard the phase information and calculate the amplitude values instead (lengths of the lines in the top row of Figure \@ref(fig:amphists)).
+Because the amplitudes can never be negative, the presence of a signal must be detected by comparison to a baseline condition in which no stimulus was presented, typically using a paired t-test.
+Here, the null hypothesis is that the stimulus-present and stimulus-absent conditions produce equal amplitudes at the stimulus frequency ($H_0: A_{present} = A_{absent}$).
+This approach has markedly lower power than the multivariate tests (see blue diamonds in Figure \@ref(fig:powerfig)a-c).
+In principle, one could also conduct circular statistical tests on the phase component of the signal [@Berens2009].
+However in the limit, testing both the amplitude and phase separately can be no more sensitive than testing the real and imaginary components (because these are just the polar and Cartesian representations of the same information), and suffers the same issues with familywise error from multiple tests.
+Multivariate approaches avoid these problems and provide high statistical power.
 
@@
 Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on satisfying the $T^2_{circ}$ test's more stringent assumptions.
-The two variables must be independent (i.e.
+The two dependent variables must be independent (i.e.
 uncorrelated), and of equal variance.
@@
 
-```{r falsealarms, fig.width=6, fig.height=7, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
-Estimates are for 100,000 simulated data sets per condition, with N = 10 observations.
-The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/falsealarms.pdf', bg="transparent", height = 7, width = 6)
 
@@
 
+dev.off()}
+
+```
+
+```{r falsealarms, fig.width=6, fig.height=7, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
+Estimates are for 100,000 simulated data sets per condition, with N = 10 observations.
+The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}
+
+knitr::include_graphics('figures/falsealarms.pdf')
+
 ```
@@
 
-```{r distcomparison, fig.cap='Logic of the condition index test.
-Panel (a) shows the distribution of condition indices derived from Equations 3 (black curve) and 4 (red curve), and by stochastic simulation (blue shading), for a sample size of N = 4 observations.
-The vertical lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line).
-Ellipse icons above panel (a) illustrate different condition indices between 1 and 19 (note that rotation does not affect the condition index).
-Panel (b) shows how 95 percent thresholds change as a function of the number of observations.', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/distcomparison.pdf', bg="transparent", height = 5, width = 9)
 
@@
 
+dev.off()}
+
+```
+
+```{r distcomparison, fig.cap='Logic of the condition index test.
+Panel (a) shows the distribution of condition indices derived from Equations 3 (black curve) and 4 (red curve), and by stochastic simulation (blue shading), for a sample size of N = 4 observations.
+The vertical lines show the 95 percent thresholds on the distributions (where 95 percent of values lie to the left of the line).
+Ellipse icons above panel (a) illustrate different condition indices between 1 and 19 (note that rotation does not affect the condition index).
+Panel (b) shows how 95 percent thresholds change as a function of the number of observations.', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}
+
+knitr::include_graphics('figures/distcomparison.pdf')
+
 ```
@@
 The Mahalanobis distance [@Mahalanobis1936] is a useful metric for identifying such multivariate outliers so that they can be excluded.
-It calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points.
+Note that outlier exclusion is contentious, but it is beyond the scope of this paper to contribute to this debate beyond noting that preregistration of any outlier exclusion protocols is adviseable.
+The Mahalanobis distance calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points.
 This means that any correlations in the data set are taken into account when calculating the distance metric, _D_.
@@
 
-```{r outlierplot, fig.cap='Simulations illustrating the Mahalanobis distance metric, and showing how a single outlier affects the condition index.
-The upper row shows three example data sets, each with a single outlier shown in red.
-The outliers have approximate Mahalanobis distances of 1, 3 and 5.
-The ellipses are calculated with the outlier included (red) and excluded (black), illustrating how the outlier distorts the aspect ratio of the ellipse.
-The main plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.', fig.align="center", fig.width=7, fig.height=6, echo=FALSE}
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/outlierplot.pdf', bg="transparent", height = 6, width = 7)
 
@@
 
+dev.off()}
+
+```
+
+```{r outlierplot, fig.cap='Simulations illustrating the Mahalanobis distance metric, and showing how a single outlier affects the condition index.
+The upper row shows three example data sets, each with a single outlier shown in red.
+The outliers have approximate Mahalanobis distances of 1, 3 and 5.
+The ellipses are calculated with the outlier included (red) and excluded (black), illustrating how the outlier distorts the aspect ratio of the ellipse.
+The main plot shows how the proportion of significant condition index tests depends on the outlier distance and the sample size.', fig.align="center", fig.width=7, fig.height=6, echo=FALSE}
+
+knitr::include_graphics('figures/outlierplot.pdf')
+
 ```
@@
 A _p_-value can then be determined by comparing the F-ratio calculated from equation \@ref(eq:Fratio) to an F-distribution with $df_M$ and $df_R$ degrees of freedom.
+Appendix 1 summarises the results of stochastic simulations that confirm the F-distribution in the absence of an effect.
 
@@
 
-```{r powerfig2, fig.cap="Simulations comparing the sensitivity of MANOVA and ANOVA2circ.
-The format mirrors that of Figure 3.
-In these simulations, there were three conditions, with the signal being added to one condition only.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/powerfig2.pdf', bg="transparent", height = 8, width = 12)
 
 layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))
-abc <- c('a','b','c')
 
@@
 
-text(1.1,0.95,paste('(',abc[n],')',sep=''),cex=2)
+text(1.1,0.95,paste('(',letters[n],')',sep=''),cex=2)
 
@@
 
+dev.off()}
+
+```
+
+```{r powerfig2, fig.cap="Simulations comparing the sensitivity of MANOVA and ANOVA2circ.
+The format mirrors that of Figure 3.
+In these simulations, there were three conditions, with the signal being added to one condition only.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}
+
+knitr::include_graphics('figures/powerfig2.pdf')
+
 ```
@@
 
-```{r flowchart, fig.cap = 'Flowchart illustrating how one might decide which test to conduct for a given data set, based on the study design and the outcome of the condition index test.', fig.align="center", echo=FALSE}
-
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+pdf('figures/flowchart.pdf', bg="transparent", height = 4, width = 6)
+
+par(mar=c(0.1,0.1,0.1,0.1))  
 pos <- coordinates(c(1,2,4))
@@
 
+dev.off()}
+
+```
+
+```{r flowchart, fig.cap = 'Flowchart illustrating how one might decide which test to conduct for a given data set, based on the study design and the outcome of the condition index test.', fig.align="center", echo=FALSE}
+
+knitr::include_graphics('figures/flowchart.pdf')
+
 ```
@@
 
-```{r mousedata, fig.cap = 'Summary of mouse steady-state responses to 40 Hz stimulation.
-Panel (a) shows the Fourier amplitude spectrum with inset scalp plots for sound (red) and light (blue) stimulation, averaged across repetitions and individuals.
-Grey and black points in the insets indicate electrode locations.
-Panel (b) shows complex (x = real, y = imaginary) Fourier components for 6 individual mice (small points) and their average (large points) for both conditions.', fig.width=10, fig.height=6, fig.align="center", echo=FALSE}
+```{r include=FALSE}
 
@@
 ydata2 <- Im(temp2)
+
+if (buildfigs==1){
+
+pdf('figures/mousedata.pdf', bg="transparent", height = 6, width = 10)
 
@@
 
+dev.off()}
+
 totest <- data.frame(xdata,ydata)
@@
 D <- mahalaD[1,2]
+
+```
+
+```{r mousedata, fig.cap = 'Summary of mouse steady-state responses to 40 Hz stimulation.
+Panel (a) shows the Fourier amplitude spectrum with inset scalp plots for sound (red) and light (blue) stimulation, averaged across repetitions and individuals.
+Grey and black points in the insets indicate electrode locations.
+Panel (b) shows complex (x = real, y = imaginary) Fourier components for 6 individual mice (small points) and their average (large points) for both conditions.', fig.width=10, fig.height=6, fig.align="center", echo=FALSE}
+
+knitr::include_graphics('figures/mousedata.pdf')
 
@@
 
-```{r humanSSVEP, fig.cap = 'Summary of human SSVEP data.
-Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes).
-Electrodes marked in red indicate cluster-corrected significance.
-Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (marked  by the black points in the scalp plots).
-Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the bivariate sample means.', fig.width=14, fig.height=5.4, fig.align="center", echo=FALSE}
+```{r include=FALSE}
 
@@
 
+cleandata <- subjmeans[which(isincluded==1),]
+
+if (buildfigs==1){
+
+pdf('figures/humanSSVEP.pdf', bg="transparent", height = 4, width = 14)
+
+  par(mar=c(0.1,0.1,0.1,0.1))  
 distmatrix <- matrix(0,nrow=64,ncol=64)
@@
 
-cleandata <- subjmeans[which(isincluded==1),]
-
 contrastlevels <- c(0,2,4,8,16,32,64)
+
+for (cond in 1:7){
+  xvals <- Re(cleandata[,cond])/5
+  yvals <- Im(cleandata[,cond])/5
+  ellRot <- getel(data.frame(xvals,yvals))
+  lines(ellRot[1, ]+cond, ellRot[2, ]+0.5)
+  text(cond,1.02,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
+}
+
+dev.off()}
 
@@
   yvals <- Im(cleandata[,cond])/5
-  ellRot <- getel(data.frame(xvals,yvals))
-  lines(ellRot[1, ]+cond, ellRot[2, ]+0.5)
-
-  text(cond,1.02,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
 totest <- data.frame(xvals,yvals)
@@
 
+```{r humanSSVEP, fig.cap = 'Summary of human SSVEP data.
+Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes).
+Electrodes marked in red indicate cluster-corrected significance.
+Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (marked  by the black points in the scalp plots).
+Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the bivariate sample means.', fig.width=14, fig.height=5.4, fig.align="center", echo=FALSE}
+
+knitr::include_graphics('figures/humanSSVEP.pdf')
+
+```
+
 With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = `r format(round(max(allCI),digits=2),nsmall=2)`, all _p_ > `r round(min(allCIp),digits=2)`).
@@
 However, the methods are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms [@Berger1929; @Buzsaki2004], which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals [e.g.
-@Canolty2010].
+@Quinn2021].
 When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average [@Baker2021].
@@
 However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses [@Jeffreys1961].
-Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests [@Rouder2009] and ANOVA [@Rouder2017].
+Subject to determining appropriate priors, Bayes factor scores might be calculated for Bayesian versions of all of the statistics considered here, much as has been done previously for univariate T-tests [@Rouder2009] and ANOVA [@Rouder2017].
 However this is a non-trivial undertaking, and is beyond the scope of the current paper.
@@
 This avoids problems from non-normal distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information.
-Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahalanobis distance reported as a measure of effect size.
+Outliers can be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahalanobis distance should be reported as a measure of effect size.
 For sample sizes of N < 32, the $T^2_{circ}$ and $ANOVA^2_{circ}$ statistics can be used if the condition index test is non-significant for all conditions.
@@
 
+# Appendix 1 - proof by stochastic simulation of asymptotic F-distributions
+
+To confirm that distributions of F-statistics generated by the $ANOVA^2_{circ}$ test have the expected form, simulations were conducted for a range of study designs.
+A null bivariate data set was generated for _N_ participants and _k_ groups, using normally distributed random numbers with underlying condition means of 0 for both the real and imaginary components.
+The F-statistic was then calculated using equations \@ref(eq:Fratio), \@ref(eq:MSM) and \@ref(eq:MSR).
+This procedure was repeated one million times for twenty combinations of _N_ and _k_, and the distributions summarised using histograms.
+Figure \@ref(fig:fdistributions) shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of $df_M = 2(k-1)$ and $df_R = 2((Nk)-k)$ degrees of freedom (black lines, generated using the _df_ function in _R_).
+The correspondence between the stochastic and analytic distributions is excellent, comfirming that the test produces the expected null distribution of test statistics.
+
+```{r include=FALSE}
+
+if (buildfigs==1){
+
+  pdf('figures/Fdistributions.pdf', bg="transparent", height = 5, width = 9)
+
+  nsims <- nsims*10  # increase to 1 million simulations for these
+
+  samplesizes <- 2^(2:6)
+  nconds <- c(3,5,7,9)
+  finex <- seq(0,4,length=1000)
+
+  plotlims <- c(0,30,0,4)
+  ticklocsx <- seq(0,4,1)    # locations of tick marks on x axis
+  ticklocsy <- 0:4    # locations of tick marks on y axis
+  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])
+  axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+  title(xlab="F ratio       ", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+  title(ylab="Probability density", line=1, cex.lab=1.5)
+
+  Fmatrix <- array(0,dim=c(length(samplesizes),length(nconds),nsims))
+
+  for (s in 1:length(samplesizes)){
+
+    axis(1, at=ticklocsx+(6*(s-1)), tck=0.01, lab=F, lwd=2)
+    mtext(text = ticklocsx, side = 1, line=0.2, at=ticklocsx+(6*(s-1)))
+    mtext(text = paste('N =',samplesizes[s]), side=3, at=2+(6*(s-1)))
+
+    for (s2 in 1:length(nconds)){
+
+      grouplabels <- as.factor(rep(1:nconds[s2],each=samplesizes[s]))
+
+      for (n in 1:nsims){
+
+        data <- matrix(rnorm(nconds[s2]*2*samplesizes[s]),nrow=nconds[s2]*samplesizes[s],ncol=2)
+
+        simdata <- complex(real=data[,1],imaginary=data[,2])
+        output <- anovacirc.test(data.frame(simdata,grouplabels))
+        Fmatrix[s,s2,n] <- output$Fratio
+      }
+
+      a <- hist(Fmatrix[s,s2,],breaks = 200, plot=FALSE)
+      axvals <- a$mids
+      ayvals <- 0.95*a$counts/max(a$counts)
+      polygon(c(0,axvals,max(axvals))+(6*(s-1)), c(0,ayvals,0)+(4-s2), col=addalpha(nbdtpal[3],alpha=0.5),border=NA)
+      dfM <- 2*(nconds[s2]-1)
+      dfR <- 2*((samplesizes[s]*nconds[s2]) - nconds[s2])
+      fdist <- df(finex,dfM,dfR)
+      fdist <- 0.95*fdist/max(fdist)
+      lines(finex+(6*(s-1)),(4-s2)+fdist,lwd=3)
+
+    }}
+
+  text(29,3.5,'k = 3')
+  text(29,2.5,'k = 5')
+  text(29,1.5,'k = 7')
+  text(29,0.5,'k = 9')
+
+  dev.off()
+
+}
+
+```
+
+```{r fdistributions, fig.cap = 'Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).', fig.width=9, fig.height=5, fig.align="center", echo=FALSE}
+
+knitr::include_graphics('figures/Fdistributions.pdf')
+
+```
+
 # References

### 2021-06-15 manuscript/manuscript.tex
Updated figures for version 2
Revisions in response to reviewer comments

@@
 \author{Daniel H. Baker}
-\date{2021-03-28}
+\date{2021-06-14}
 
@@
 Liu-Shuang et al., 2014; Afsari et al., 2014; Hou et al., 2020; Itthipuripat et al., 2014; McFadden et al., 2014; Smith et al., 2017; Vanegas et al., 2015) (note this is a non-exhaustive selection of recent studies).
-Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if signals differ between two (or more) experimental conditions.
+Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if responses differ in amplitude and/or phase between two (or more) experimental conditions.
 However, there are several objections to using univariate statistics to answer these questions, as will be demonstrated below.
-An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components).
+An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components in a Cartesian space).
 Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods for typical applications (detecting the presence of a signal, or comparing multiple signals).
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/fourierexplain-1}
+{\centering \includegraphics{figures/fourierexplain}
 
@@
 In the present paper, best practice guidelines are developed for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms.
-It is first demonstrated why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed.
+It is first demonstrated why parametric univariate statistics are inappropriate for such data, primarily because amplitudes for weak signals are not normally distributed.
 Then conditions are investigated under which either the \(T^2\) or \(T^2_{circ}\) statistic should be used.
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/amphists-1}
+{\centering \includegraphics{figures/amphists}
 
@@
 Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents.
-Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.
+Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, as will be demonstrated in the following section.
 
-\hypertarget{conditions-under-which-t2_circ-is-more-sensitive-than-t2}{%
-\section{\texorpdfstring{Conditions under which \(T^2_{circ}\) is more sensitive than \(T^2\)}{Conditions under which T\^{}2\_\{circ\} is more sensitive than T\^{}2}}\label{conditions-under-which-t2_circ-is-more-sensitive-than-t2}}
+\hypertarget{comparison-of-statistical-power-between-univariate-and-multivariate-tests}{%
+\section{Comparison of statistical power between univariate and multivariate tests}\label{comparison-of-statistical-power-between-univariate-and-multivariate-tests}}
 
-Using the \(T^2\) or \(T^2_{circ}\) statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values.
+Using the \(T^2\) or \(T^2_{circ}\) statistic allows the phase information to be retained, and therefore provides greater power than running univariate T-tests on amplitudes, as well as avoiding problems caused by using absolute amplitude values.
 Victor \& Mast (1991) report simulations showing situations where \(T^2_{circ}\) has greater power than \(T^2\), which is expected on theoretical grounds because the additional degrees of freedom in the \(T^2_{circ}\) test make it more efficient.
@@
 
-Here these simulations are replicated and extended (see Figure \ref{fig:powerfig}), and it is shown that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f).
-However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations.
-This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (when \emph{d} \textgreater{} 2 and N \textgreater{} 8).
-These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use can be restricted to small sample studies.
+Here these simulations are replicated and extended (see Figure \ref{fig:powerfig}).
+The real and imaginary Fourier components of a periodic signal could in principle be analysed using two separate univariate t-tests.
+To detect the presence of a signal, each t-test would compare to a value of zero (i.e.~the origin in Figure \ref{fig:fourierexplain}d), to test against the null hypothesis that there is no signal (\(H_{0x}: x = 0; H_{0y}: y = 0\)).
+Without correction for familywise error, this approach has high statistical power for detecting when at least one test is significant (see grey triangles in Figure \ref{fig:powerfig}a-c).
+However the use of two separate tests inflates the familywise error rate such that 10\% of tests are significant (assuming \(\alpha = 0.05\)), even in the absence of a signal.
+Using Bonferroni correction to ameliorate this problem substantially reduces the power (white triangles in Figure \ref{fig:powerfig}a-c).
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/powerfig-1}
+{\centering \includegraphics{figures/powerfig}
 
@@
 \caption{Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes (100,000 simulated data sets per condition).
-Panels a-c replicate conditions reported by Victor and Mast (1991).
-Panels (d) and (e) show a wider range of conditions for each statistic.
-Panel (f) shows the difference between the two statistics, with contour lines indicating differences of 0.02, 0.05, 0.1, 0.2 and 0.4.}\label{fig:powerfig}
+Panels a-c replicate conditions reported by Victor and Mast (1991), and extend them to include univariate tests for comparison.
+Panels (d) and (e) show a wider range of conditions for each multivariate statistic.
+Panel (f) shows the difference between the two statistics, with contour lines indicating power differences of 0.02, 0.05, 0.1, 0.2 and 0.4.}\label{fig:powerfig}
 \end{figure}
+
+Using multivariate tests allows high power to be maintained without inflating the Type I error rate.
+Here, the multivariate mean (or equivalently, the difference in means between two conditions) is compared to zero, again testing against the null hypothesis that there is no signal (\(H_0: x = y = 0\)).
+The \(T^2_{circ}\) statistic (red squares in Figure \ref{fig:powerfig}a-c) has greater power than either the \(T^2\) statistic (black circles) or the corrected univariate tests (white triangles).
+The power advantage over \(T^2\) occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f).
+However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations.
+This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (when \emph{d} \textgreater{} 2 and N \textgreater{} 8).
+These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use can be restricted to small-sample studies.
+
+As mentioned above, an alternative approach that is widely used in the literature is to discard the phase information and calculate the amplitude values instead (lengths of the lines in the top row of Figure \ref{fig:amphists}).
+Because the amplitudes can never be negative, the presence of a signal must be detected by comparison to a baseline condition in which no stimulus was presented, typically using a paired t-test.
+Here, the null hypothesis is that the stimulus-present and stimulus-absent conditions produce equal amplitudes at the stimulus frequency (\(H_0: A_{present} = A_{absent}\)).
+This approach has markedly lower power than the multivariate tests (see blue diamonds in Figure \ref{fig:powerfig}a-c).
+In principle, one could also conduct circular statistical tests on the phase component of the signal (Berens, 2009).
+However in the limit, testing both the amplitude and phase separately can be no more sensitive than testing the real and imaginary components (because these are just the polar and Cartesian representations of the same information), and suffers the same issues with familywise error from multiple tests.
+Multivariate approaches avoid these problems and provide high statistical power.
 
@@
 Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on satisfying the \(T^2_{circ}\) test's more stringent assumptions.
-The two variables must be independent (i.e.~uncorrelated), and of equal variance.
+The two dependent variables must be independent (i.e.~uncorrelated), and of equal variance.
 These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not.
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/falsealarms-1.pdf}
+\includegraphics{figures/falsealarms.pdf}
 \caption{\label{fig:falsealarms}Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b).
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/distcomparison-1}
+{\centering \includegraphics{figures/distcomparison}
 
@@
 The Mahalanobis distance (Mahalanobis, 1936) is a useful metric for identifying such multivariate outliers so that they can be excluded.
-It calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points.
+Note that outlier exclusion is contentious, but it is beyond the scope of this paper to contribute to this debate beyond noting that preregistration of any outlier exclusion protocols is adviseable.
+The Mahalanobis distance calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points.
 This means that any correlations in the data set are taken into account when calculating the distance metric, \emph{D}.
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/outlierplot-1}
+{\centering \includegraphics{figures/outlierplot}
 
@@
 A \emph{p}-value can then be determined by comparing the F-ratio calculated from equation \eqref{eq:Fratio} to an F-distribution with \(df_M\) and \(df_R\) degrees of freedom.
+Appendix 1 summarises the results of stochastic simulations that confirm the F-distribution in the absence of an effect.
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/powerfig2-1}
+{\centering \includegraphics{figures/powerfig2}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/flowchart-1}
+{\centering \includegraphics{figures/flowchart}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/mousedata-1}
+{\centering \includegraphics{figures/mousedata}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/humanSSVEP-1}
+{\centering \includegraphics{figures/humanSSVEP}
 
@@
 However, the methods are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger, 1929; Buzsáki and Draguhn, 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g.
-Canolty and Knight, 2010).
+Quinn et al., 2021).
 When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al., 2021).
@@
 However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses (Jeffreys, 1961).
-Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al., 2009) and ANOVA (Rouder et al., 2017).
+Subject to determining appropriate priors, Bayes factor scores might be calculated for Bayesian versions of all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al., 2009) and ANOVA (Rouder et al., 2017).
 However this is a non-trivial undertaking, and is beyond the scope of the current paper.
@@
 This avoids problems from non-normal distributions of amplitudes violating the test assumptions, and also provides a sensitivity benefit from the inclusion of phase information.
-Outliers should be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahalanobis distance reported as a measure of effect size.
+Outliers can be removed when they have a Mahalanobis distance exceeding 3, and the pairwise Mahalanobis distance should be reported as a measure of effect size.
 For sample sizes of N \textless{} 32, the \(T^2_{circ}\) and \(ANOVA^2_{circ}\) statistics can be used if the condition index test is non-significant for all conditions.
@@
 The greater power afforded by these tests should in general lead to more accurate statistical inferences when analysing periodic data.
+Additionally, more sensitive methods require fewer observations to reach a given level of power, which has important ethical implications for animal research if fewer experimental subjects are needed.
+
+\hypertarget{appendix-1---proof-by-stochastic-simulation-of-asymptotic-f-distributions}{%
+\section{Appendix 1 - proof by stochastic simulation of asymptotic F-distributions}\label{appendix-1---proof-by-stochastic-simulation-of-asymptotic-f-distributions}}
+
+To confirm that distributions of F-statistics generated by the \(ANOVA^2_{circ}\) test have the expected form, simulations were conducted for a range of study designs.
+A null bivariate data set was generated for \emph{N} participants and \emph{k} groups, using normally distributed random numbers with underlying condition means of 0 for both the real and imaginary components.
+The F-statistic was then calculated using equations \eqref{eq:Fratio}, \eqref{eq:MSM} and \eqref{eq:MSR}.
+This procedure was repeated one million times for twenty combinations of \emph{N} and \emph{k}, and the distributions summarised using histograms.
+Figure \ref{fig:fdistributions} shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of \(df_M = 2(k-1)\) and \(df_R = 2((Nk)-k)\) degrees of freedom (black lines, generated using the \emph{df} function in \emph{R}).
+The correspondence between the stochastic and analytic distributions is excellent, comfirming that the test produces the expected null distribution of test statistics.
+
+\begin{figure}
+
+{\centering \includegraphics{figures/Fdistributions}
+
+}
+
+\caption{Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).}\label{fig:fdistributions}
+\end{figure}
 
@@
 
+\leavevmode\hypertarget{ref-Berens2009}{}%
+Berens P. 2009. CircStat: A MATLAB Toolbox for circular statistics. \emph{Journal of Statistical Software} \textbf{31}:1--21. doi:\href{https://doi.org/10.18637/jss.v031.i10}{10.18637/jss.v031.i10}
+
 \leavevmode\hypertarget{ref-Berger1929}{}%
@@
 Buzsáki G, Draguhn A. 2004. Neuronal oscillations in cortical networks. \emph{Science} \textbf{304}:1926--9. doi:\href{https://doi.org/10.1126/science.1099745}{10.1126/science.1099745}
-
-\leavevmode\hypertarget{ref-Canolty2010}{}%
-Canolty RT, Knight RT.
-2010.
-The functional role of cross-frequency coupling.
-\emph{Trends Cogn Sci} \textbf{14}:506--15.
-doi:\href{https://doi.org/10.1016/j.tics.2010.09.001}{10.1016/j.tics.2010.09.001}
 
@@
 
+\leavevmode\hypertarget{ref-Quinn2021}{}%
+Quinn A, Lopes-dos-Santos V, Dupret D, Nobre A.
+2021.
+EMD: Empirical mode decomposition and hilbert-huang spectral analyses in python.
+\emph{Journal of Open Source Software} \textbf{6}:2977.
+doi:\href{https://doi.org/10.21105/joss.02977}{10.21105/joss.02977}
+
 \leavevmode\hypertarget{ref-Rees1986}{}%

### 2021-06-15 manuscript/misc/appendixsims.R

@@ -1,0 +1,77 @@
+
+# install the FourierStats package from github using the devtools package
+packagelist <- c('FourierStats')
+missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
+if (length(missingpackages)>0){devtools::install_github("bakerdh/FourierStats")}
+toinstall <- packagelist[which(!packagelist %in% (.packages()))]
+invisible(lapply(toinstall,library,character.only=TRUE))
+addalpha <- function(col, alpha=1){apply(sapply(col, col2rgb)/255, 2, function(x) rgb(x[1], x[2], x[3], alpha=alpha))}
+
+library(tictoc)
+
+tic()
+
+nsims <- 1000000
+
+samplesizes <- 2^(2:6)
+nconds <- c(3,5,7,9)
+finex <- seq(0,4,length=1000)
+
+
+pdf('Fdistributions.pdf', bg="transparent", height = 5, width = 9)
+
+plotlims <- c(0,30,0,4)
+ticklocsx <- seq(0,4,1)    # locations of tick marks on x axis
+ticklocsy <- 0:4    # locations of tick marks on y axis
+plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])
+axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
+# mtext(text = ticklocsy[1:2], side = 2, at=ticklocsy[1:2], line=0.2, las=1)
+title(xlab="F ratio       ", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
+title(ylab="Probability density", line=1, cex.lab=1.5)
+
+Fmatrix <- array(0,dim=c(length(samplesizes),length(nconds),nsims))
+
+for (s in 1:length(samplesizes)){
+
+axis(1, at=ticklocsx+(6*(s-1)), tck=0.01, lab=F, lwd=2)
+mtext(text = ticklocsx, side = 1, line=0.2, at=ticklocsx+(6*(s-1)))
+mtext(text = paste('N =',samplesizes[s]), side=3, at=2+(6*(s-1)))
+
+       for (s2 in 1:length(nconds)){
+
+      grouplabels <- as.factor(rep(1:nconds[s2],each=samplesizes[s]))
+
[…]

### 2021-06-26 manuscript/manuscript.Rmd
Changed aspect ratio of figure 4
Fixed a minor error

@@
 
-pdf('figures/falsealarms.pdf', bg="transparent", height = 7, width = 6)
-
-par(mfrow=c(2,1), mar=c(3,4,1,2))
+pdf('figures/falsealarms.pdf', bg="transparent", height = 4, width = 15)
+
+par(mfrow=c(1,2), mar=c(3,4,1,2))
 
@@
 A repeated measures $ANOVA^2_{circ}$ test was conducted, revealing a significant effect of stimulus contrast ($F_{(`r withincirc$dfM`,`r withincirc$dfR`)}$ = `r round(withincirc$Fratio,digits=1)`, _p_ < `r max(round(withincirc$pval,digits=3),0.001)`).
-Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.08$) revealed significant differences at 8% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[3],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[3],digits=2)`, _D_ = `r round(allD[3],digits=2)`, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[4],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[4],digits=2)`, _D_ = `r round(allD[4],digits=2)`, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[5],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[5],digits=2)`, _D_ = `r round(allD[5],digits=2)`, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[6],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[6],digits=2)`, _D_ = `r round(allD[6],digits=2)`, _p_ < 0.001).
+Pairwise $T^2_{circ}$ statistics comparing the baseline (0% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to $\alpha = 0.008$) revealed significant differences at 8% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[3],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[3],digits=2)`, _D_ = `r round(allD[3],digits=2)`, _p_ < 0.001), 16% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[4],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[4],digits=2)`, _D_ = `r round(allD[4],digits=2)`, _p_ < 0.001), 32% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[5],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[5],digits=2)`, _D_ = `r round(allD[5],digits=2)`, _p_ < 0.001) and 64% contrast ($T^2_{circ}$ = `r format(round(alltsq$tsqc[6],digits=2),nsmall=2)`, $F_{(2,176)}$ = `r round(alltsq$Fratio[6],digits=2)`, _D_ = `r round(allD[6],digits=2)`, _p_ < 0.001).
 The study by Vilidaite et al.

### 2021-06-26 manuscript/manuscript.tex
Fixed a minor error

@@
 \author{Daniel H. Baker}
-\date{2021-06-14}
+\date{2021-06-26}
 
@@
 A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, \emph{p} \textless{} 0.001).
-Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, \emph{D} = 1.1, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, \emph{D} = 1.04, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, \emph{D} = 0.63, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, \emph{D} = 1.17, \emph{p} \textless{} 0.001).
+Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.008\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, \emph{D} = 1.1, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, \emph{D} = 1.04, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, \emph{D} = 0.63, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, \emph{D} = 1.17, \emph{p} \textless{} 0.001).
 The study by Vilidaite et al.

### 2021-06-26 manuscript/manuscript.Rmd
Fixed a typo

@@
 Figure \@ref(fig:fdistributions) shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of $df_M = 2(k-1)$ and $df_R = 2((Nk)-k)$ degrees of freedom (black lines, generated using the _df_ function in _R_).
-The correspondence between the stochastic and analytic distributions is excellent, comfirming that the test produces the expected null distribution of test statistics.
+The correspondence between the stochastic and analytic distributions is excellent, confirming that the test produces the expected null distribution of test statistics.
 

### 2021-06-26 manuscript/manuscript.tex
Fixed a typo

@@
 Figure \ref{fig:fdistributions} shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of \(df_M = 2(k-1)\) and \(df_R = 2((Nk)-k)\) degrees of freedom (black lines, generated using the \emph{df} function in \emph{R}).
-The correspondence between the stochastic and analytic distributions is excellent, comfirming that the test produces the expected null distribution of test statistics.
+The correspondence between the stochastic and analytic distributions is excellent, confirming that the test produces the expected null distribution of test statistics.
 

### 2021-07-06 manuscript/manuscript.Rmd
Minor edits following additional reviewer comments (Appendix removed)

@@
 where $N_k$ is the number of observations in group _k_, $\bar{x}_k$ is the bivariate sample mean of group _k_, $\bar{x}_{grand}$ is the bivariate mean of the entire sample (including all groups), $\Sigma$ denotes summation across groups, and $df_M$ is the degrees of freedom for the model.
+Note that by using the vector distances between points, this equation reduces multivariate data to a single scalar value, in a similar way to equations \@ref(eq:t2eq) and \@ref(eq:t2c).
 The model degrees of freedom for _k_ groups is $df_M = 2(k-1)$, with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.
@@
 A _p_-value can then be determined by comparing the F-ratio calculated from equation \@ref(eq:Fratio) to an F-distribution with $df_M$ and $df_R$ degrees of freedom.
-Appendix 1 summarises the results of stochastic simulations that confirm the F-distribution in the absence of an effect.
 
@@
 
-# Appendix 1 - proof by stochastic simulation of asymptotic F-distributions
-
-To confirm that distributions of F-statistics generated by the $ANOVA^2_{circ}$ test have the expected form, simulations were conducted for a range of study designs.
+<!-- Note that the appendix simulations below were added at the request of a reviewer, and then subsequently removed following the request of the same reviewer (and the editor). -->
+<!-- # Appendix 1 - proof by stochastic simulation of asymptotic F-distributions -->
+
+<!-- To confirm that distributions of F-statistics generated by the $ANOVA^2_{circ}$ test have the expected form, simulations were conducted for a range of study designs.
 A null bivariate data set was generated for _N_ participants and _k_ groups, using normally distributed random numbers with underlying condition means of 0 for both the real and imaginary components.
@@
 The correspondence between the stochastic and analytic distributions is excellent, confirming that the test produces the expected null distribution of test statistics.
-
-```{r include=FALSE}
-
-if (buildfigs==1){
-
-  pdf('figures/Fdistributions.pdf', bg="transparent", height = 5, width = 9)
-
-  nsims <- nsims*10  # increase to 1 million simulations for these
-
-  samplesizes <- 2^(2:6)
-  nconds <- c(3,5,7,9)
-  finex <- seq(0,4,length=1000)
-
-  plotlims <- c(0,30,0,4)
-  ticklocsx <- seq(0,4,1)    # locations of tick marks on x axis
-  ticklocsy <- 0:4    # locations of tick marks on y axis
-  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])
-  axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
-  title(xlab="F ratio       ", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
-  title(ylab="Probability density", line=1, cex.lab=1.5)
-
-  Fmatrix <- array(0,dim=c(length(samplesizes),length(nconds),nsims))
-
-  for (s in 1:length(samplesizes)){
-
-    axis(1, at=ticklocsx+(6*(s-1)), tck=0.01, lab=F, lwd=2)
-    mtext(text = ticklocsx, side = 1, line=0.2, at=ticklocsx+(6*(s-1)))
-    mtext(text = paste('N =',samplesizes[s]), side=3, at=2+(6*(s-1)))
-
-    for (s2 in 1:length(nconds)){
-
-      grouplabels <- as.factor(rep(1:nconds[s2],each=samplesizes[s]))
-
-      for (n in 1:nsims){
-
-        data <- matrix(rnorm(nconds[s2]*2*samplesizes[s]),nrow=nconds[s2]*samplesizes[s],ncol=2)
-
-        simdata <- complex(real=data[,1],imaginary=data[,2])
-        output <- anovacirc.test(data.frame(simdata,grouplabels))
-        Fmatrix[s,s2,n] <- output$Fratio
-      }
-
-      a <- hist(Fmatrix[s,s2,],breaks = 200, plot=FALSE)
-      axvals <- a$mids
-      ayvals <- 0.95*a$counts/max(a$counts)
-      polygon(c(0,axvals,max(axvals))+(6*(s-1)), c(0,ayvals,0)+(4-s2), col=addalpha(nbdtpal[3],alpha=0.5),border=NA)
-      dfM <- 2*(nconds[s2]-1)
-      dfR <- 2*((samplesizes[s]*nconds[s2]) - nconds[s2])
-      fdist <- df(finex,dfM,dfR)
-      fdist <- 0.95*fdist/max(fdist)
-      lines(finex+(6*(s-1)),(4-s2)+fdist,lwd=3)
-
-    }}
-
-  text(29,3.5,'k = 3')
-  text(29,2.5,'k = 5')
-  text(29,1.5,'k = 7')
-  text(29,0.5,'k = 9')
-
-  dev.off()
-
-}
-
-```
-
-```{r fdistributions, fig.cap = 'Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).', fig.width=9, fig.height=5, fig.align="center", echo=FALSE}
-
-knitr::include_graphics('figures/Fdistributions.pdf')
-
-```
+-->
+
+<!-- ```{r include=FALSE} -->
+
+<!-- if (buildfigs==1){ -->
+
+<!--   pdf('figures/Fdistributions.pdf', bg="transparent", height = 5, width = 9) -->
+
+<!--   nsims <- nsims*10  # increase to 1 million simulations for these -->
+
+<!--   samplesizes <- 2^(2:6) -->
+<!--   nconds <- c(3,5,7,9) -->
+<!--   finex <- seq(0,4,length=1000) -->
+
+<!--   plotlims <- c(0,30,0,4) -->
+<!--   ticklocsx <- seq(0,4,1)    # locations of tick marks on x axis -->
+<!--   ticklocsy <- 0:4    # locations of tick marks on y axis -->
+<!--   plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4]) -->
+<!--   axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2) -->
+<!--   title(xlab="F ratio       ", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5) -->
+<!--   title(ylab="Probability density", line=1, cex.lab=1.5) -->
+
+<!--   Fmatrix <- array(0,dim=c(length(samplesizes),length(nconds),nsims)) -->
+
+<!--   for (s in 1:length(samplesizes)){ -->
+
+<!--     axis(1, at=ticklocsx+(6*(s-1)), tck=0.01, lab=F, lwd=2) -->
+<!--     mtext(text = ticklocsx, side = 1, line=0.2, at=ticklocsx+(6*(s-1))) -->
+<!--     mtext(text = paste('N =',samplesizes[s]), side=3, at=2+(6*(s-1))) -->
+
+<!--     for (s2 in 1:length(nconds)){ -->
+
+<!--       grouplabels <- as.factor(rep(1:nconds[s2],each=samplesizes[s])) -->
+
+<!--       for (n in 1:nsims){ -->
+
+<!--         data <- matrix(rnorm(nconds[s2]*2*samplesizes[s]),nrow=nconds[s2]*samplesizes[s],ncol=2) -->
+
+<!--         simdata <- complex(real=data[,1],imaginary=data[,2]) -->
+<!--         output <- anovacirc.test(data.frame(simdata,grouplabels)) -->
+<!--         Fmatrix[s,s2,n] <- output$Fratio -->
+<!--       } -->
+
+<!--       a <- hist(Fmatrix[s,s2,],breaks = 200, plot=FALSE) -->
+<!--       axvals <- a$mids -->
+<!--       ayvals <- 0.95*a$counts/max(a$counts) -->
+<!--       polygon(c(0,axvals,max(axvals))+(6*(s-1)), c(0,ayvals,0)+(4-s2), col=addalpha(nbdtpal[3],alpha=0.5),border=NA) -->
+<!--       dfM <- 2*(nconds[s2]-1) -->
+<!--       dfR <- 2*((samplesizes[s]*nconds[s2]) - nconds[s2]) -->
+<!--       fdist <- df(finex,dfM,dfR) -->
+<!--       fdist <- 0.95*fdist/max(fdist) -->
+<!--       lines(finex+(6*(s-1)),(4-s2)+fdist,lwd=3) -->
+
+<!--     }} -->
+
+<!--   text(29,3.5,'k = 3') -->
+<!--   text(29,2.5,'k = 5') -->
+<!--   text(29,1.5,'k = 7') -->
+<!--   text(29,0.5,'k = 9') -->
+
+<!--   dev.off() -->
+
+<!-- } -->
+
+<!-- ``` -->
+
+<!-- ```{r fdistributions, fig.cap = 'Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).', fig.width=9, fig.height=5, fig.align="center", echo=FALSE} -->
+
+<!-- knitr::include_graphics('figures/Fdistributions.pdf') -->
+
+<!-- ``` -->
 

### 2021-07-06 manuscript/manuscript.tex
Minor edits following additional reviewer comments (Appendix removed)

@@
 \author{Daniel H. Baker}
-\date{2021-06-26}
+\date{2021-07-06}
 
@@
 where \(N_k\) is the number of observations in group \emph{k}, \(\bar{x}_k\) is the bivariate sample mean of group \emph{k}, \(\bar{x}_{grand}\) is the bivariate mean of the entire sample (including all groups), \(\Sigma\) denotes summation across groups, and \(df_M\) is the degrees of freedom for the model.
+Note that by using the vector distances between points, this equation reduces multivariate data to a single scalar value, in a similar way to equations \eqref{eq:t2eq} and \eqref{eq:t2c}.
 The model degrees of freedom for \emph{k} groups is \(df_M = 2(k-1)\), with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.
@@
 A \emph{p}-value can then be determined by comparing the F-ratio calculated from equation \eqref{eq:Fratio} to an F-distribution with \(df_M\) and \(df_R\) degrees of freedom.
-Appendix 1 summarises the results of stochastic simulations that confirm the F-distribution in the absence of an effect.
 
@@
 Additionally, more sensitive methods require fewer observations to reach a given level of power, which has important ethical implications for animal research if fewer experimental subjects are needed.
-
-\hypertarget{appendix-1---proof-by-stochastic-simulation-of-asymptotic-f-distributions}{%
-\section{Appendix 1 - proof by stochastic simulation of asymptotic F-distributions}\label{appendix-1---proof-by-stochastic-simulation-of-asymptotic-f-distributions}}
-
-To confirm that distributions of F-statistics generated by the \(ANOVA^2_{circ}\) test have the expected form, simulations were conducted for a range of study designs.
-A null bivariate data set was generated for \emph{N} participants and \emph{k} groups, using normally distributed random numbers with underlying condition means of 0 for both the real and imaginary components.
-The F-statistic was then calculated using equations \eqref{eq:Fratio}, \eqref{eq:MSM} and \eqref{eq:MSR}.
-This procedure was repeated one million times for twenty combinations of \emph{N} and \emph{k}, and the distributions summarised using histograms.
-Figure \ref{fig:fdistributions} shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of \(df_M = 2(k-1)\) and \(df_R = 2((Nk)-k)\) degrees of freedom (black lines, generated using the \emph{df} function in \emph{R}).
-The correspondence between the stochastic and analytic distributions is excellent, confirming that the test produces the expected null distribution of test statistics.
-
-\begin{figure}
-
-{\centering \includegraphics{figures/Fdistributions}
-
-}
-
-\caption{Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).}\label{fig:fdistributions}
-\end{figure}
 
